---
title: "pathway-overlap - ratio-plot - SNV -downsample"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

## Load libraries
```{r}
library(ggplot2)
library(stringr)
library(dplyr)
library(ggpubr)
library(purrr)
library(rstatix)
library(EnhancedVolcano)
library(clusterProfiler)
library(enrichplot)
library(UpSetR)
library(viridis)

hg19_size <- 3036303846
```

## Functions
```{r}
# combine the permutation pvalues from each sample
# to assess if group of samples is significantly different from expected
combine_pvalues_fisher <- function(p_values) {
    chi_sq_stat = -2 * sum(log(p_values))
    degrees_of_freedom = 2 * length(p_values)
    p_combined = pchisq(chi_sq_stat, degrees_of_freedom, lower.tail = FALSE)
    return(p_combined)
}

# function to calculate significance with results coming from
# observed and downsampled
ratio_significance <- function(combined_regions, group_vars, region_levels, by.region = TRUE) {
  
  if(by.region){
  res <- combined_regions %>%
    group_by_at(vars(one_of(c("region", group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue.adj)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  res$region <- factor(res$region, levels = region_levels)
  } else{
    res <- combined_regions %>%
    group_by_at(vars(one_of(c(group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue.adj)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  }
  return(res)
}

# Using pvalue here instead of pvalue.adj because many pathways were tested
ratio_significance_pathway <- function(combined_regions, group_vars, region_levels, by.region = TRUE) {
  
  if(by.region){
  res <- combined_regions %>%
    group_by_at(vars(one_of(c("region", group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  res$region <- factor(res$region, levels = region_levels)
  } else{
    res <- combined_regions %>%
    group_by_at(vars(one_of(c(group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  }
  return(res)
}

# function to calculate significance between condition groups
group_significance <- function(combined_regions, group_vars, condition, region_levels){
  formula <- as.formula(paste("log2fc ~", condition))
  
  results <- combined_regions %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(
      p_value = wilcox.test(formula = formula, 
                            data = cur_data(),
                            exact = FALSE)$p.value
    ) %>%
    ungroup() %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)
    )

  # Order by region
  results$region <- factor(results$region, levels = region_levels)
  
  return(results)
}


# function to calculate significance between condition groups
# also reports significance if seen in both downsampled and observed groups
# adjusted p-values by bejamini-hochberg
# combined_regions <- go.expressed_regions_combined
# group_vars <- c("group", "region")
# condition <- "condition"
# region_levels <- go.expressed_regions_levels

group_significance_go <- function(combined_regions, group_vars, condition, region_levels){
  formula <- as.formula(paste("log2fc ~", condition))
  
  results <- combined_regions %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(
      p_value = t.test(formula = formula, data = cur_data())$p.value
    ) %>%
    ungroup() %>%
    mutate(
           significant_p_adjusted = case_when(
                            p_value <= 0.0001 ~ '****',
                            p_value <= 0.001 ~ '***',
                            p_value <= 0.01 ~ '**',
                            p_value <= 0.05 ~ '*',
                            TRUE ~ NA_character_)
    )
  
  # Identify regions with p_adjusted < 0.05 in both groups
  regions_significant_in_both_groups <- as.data.frame(results) %>%
    filter(p_value < 0.05) %>%
    dplyr::count(region) %>%
    filter(n == 2) %>%
    pull(region)
  
  # Create new column based on the condition
  results <- results %>%
    mutate(significant_p_adjusted_both_groups = ifelse(region %in% regions_significant_in_both_groups, significant_p_adjusted, NA_character_))

  # Order by region
  results$region <- factor(results$region, levels = region_levels)
  
  return(results)
}

calculate_stat_summary <- function(observed, group_vars, mut_var){
  results <- observed %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(mean = mean(!!sym(mut_var), na.rm = TRUE),
              sd = sd(!!sym(mut_var), na.rm = TRUE))
}

theme_Publication <- function(base_size=16, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family = "")
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               #legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_fill_Publication_2 <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#7fc97f","#ef3b2c","#386cb0","#fdb462","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

wrap_labels <- function(labels, width = 5) {
  wrapped_labels <- str_wrap(labels, width)
  names(wrapped_labels) <- names(labels)
  return(wrapped_labels)
}
```

## Load sample table
```{r}
sample_table <- read.table("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/1-data/sample_table_filtered.txt", header = TRUE, sep = "\t")
```

## Load genomic distribution results
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/2-calculate-ratio/snv")
# expressed_nonexpressed
  # observed
  expressed_nonexpressed <- read.csv("expressed_nonexpressed_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)

# essential
  # observed
  essential_nonessential <- read.csv("essential_genes_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)
  
# go
  go_ids <- read.csv("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/1-load-data/expressed_genes_gene_ontology.csv", header = TRUE)
  
  # observed
  go_expressed <- read.csv("go.expressed_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)

# go shuffle
  # observed
  go_expressed_shuffled<- read.csv("go.expressed.shuffled_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)
  
# kegg
  # observed
  kegg_expressed <- read.csv("kegg.expressed_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)

# kegg shuffle
  # observed
  kegg_expressed_shuffled <- read.csv("kegg.expressed.shuffled_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)
```

# Analysis

## Expressed and Nonexpressed genes

### By condition
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/expressed_nonexpressed")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/expressed_nonexpressed")

expressed_nonexpressed_levels <- unique(expressed_nonexpressed$region)

expressed_nonexpressed_regions <- merge(expressed_nonexpressed, sample_table, by = "sample")
expressed_nonexpressed_regions$region <- factor(expressed_nonexpressed_regions$region, expressed_nonexpressed_levels)


expressed_nonexpressed_regions_ratio_significance <- ratio_significance(expressed_nonexpressed_regions,
                                                   "condition",
                                                   expressed_nonexpressed_levels)

expressed_nonexpressed_regions_group_significance <- group_significance(expressed_nonexpressed_regions,
                                                         c("region"), 
                                                         "condition",
                                                         expressed_nonexpressed_levels)
# observed
pdf("SNV_expressed_nonexpressed_condition_observed_downsampled.pdf",
    width = 5,
    height = 5)
ggplot(expressed_nonexpressed_regions, aes(x = region, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.6) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    geom_text(data = expressed_nonexpressed_regions_ratio_significance, 
              aes(x = region, y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 7, angle = 90) +
    geom_text(data = expressed_nonexpressed_regions_group_significance,
              aes(x = region, y = Inf, fill = NA, label = significant_p_adjusted), # pvalue of control vs enu
              vjust = 15, hjust = 0.5, size = 8, color = "firebrick") +
    labs(x = "Region", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication(name = "Condition", labels = c("Control", "ENU")) +
    theme_Publication() +
    theme(legend.justification = c("center","top"),
          legend.position = "top",
          axis.text.x = element_text(size = 12)) +
    ylim(NA, max(expressed_nonexpressed_regions$log2fc + 0.1))
dev.off()
```

### By cycle
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/expressed_nonexpressed")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/expressed_nonexpressed")

expressed_nonexpressed_levels <- unique(expressed_nonexpressed$region)

expressed_nonexpressed_regions <- merge(expressed_nonexpressed, sample_table, by = "sample")
expressed_nonexpressed_regions$region <- factor(expressed_nonexpressed_regions$region, expressed_nonexpressed_levels)

expressed_nonexpressed_regions_ratio_significance <- ratio_significance(expressed_nonexpressed_regions,
                                                   c("cycle", "condition"),
                                                   expressed_nonexpressed_levels)

# observed
pdf("SNV_expressed_nonexpressed_cycle_observed_downsampled.pdf",
    width = 6,
    height = 5)
ggplot(expressed_nonexpressed_regions, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.7) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Cycle", y = "Log2(observed/expected)\nSNVs Per Cell", fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_x_discrete(labels = c("Control", "ENU")) + 
    theme_Publication() +
    geom_text(data = expressed_nonexpressed_regions_ratio_significance,
              aes(x = condition, y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    facet_grid(.~region) +
    theme(legend.justification = c("center","top"),
          legend.position = "top",
          axis.text.x = element_text(size = 12)) +
    ylim(NA, max(expressed_nonexpressed_regions$log2fc) + 0.1)
dev.off()
```

## Essential and Nonessential genes

### By condition
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/essential_nonessential")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/essential_nonessential")

essential_nonessential_levels <- unique(essential_nonessential$region)

essential_nonessential_regions <- merge(essential_nonessential, sample_table, by = "sample")
essential_nonessential_regions$region <- factor(essential_nonessential_regions$region, essential_nonessential_levels)


essential_nonessential_regions_ratio_significance <- ratio_significance(essential_nonessential_regions,
                                                   "condition",
                                                   essential_nonessential_levels)

essential_nonessential_regions_group_significance <- group_significance(essential_nonessential_regions,
                                                         c("region"), 
                                                         "condition",
                                                         essential_nonessential_levels)
# observed
pdf("SNV_essential_nonessential_condition_observed_downsampled.pdf",
    width = 5,
    height = 5)
ggplot(essential_nonessential_regions, aes(x = region, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.6) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    geom_text(data = essential_nonessential_regions_ratio_significance, 
              aes(x = region, y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 7, angle = 90) +
    geom_text(data = essential_nonessential_regions_group_significance,
              aes(x = region, y = Inf, fill = NA, label = significant_p_adjusted), # pvalue of control vs enu
              vjust = 15, hjust = 0.5, size = 8, color = "firebrick") +
    labs(x = "Region", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication(name = "Condition", labels = c("Control", "ENU")) +
    theme_Publication() +
    theme(legend.justification = c("center","top"),
          legend.position = "top",
          axis.text.x = element_text(size = 12)) +
    ylim(NA, max(essential_nonessential_regions$log2fc + 0.1))
dev.off()
```

### By cycle
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/essential_nonessential")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/essential_nonessential")

essential_nonessential_levels <- unique(essential_nonessential$region)

essential_nonessential_regions <- merge(essential_nonessential, sample_table, by = "sample")
essential_nonessential_regions$region <- factor(essential_nonessential_regions$region, essential_nonessential_levels)

essential_nonessential_regions_ratio_significance <- ratio_significance(essential_nonessential_regions,
                                                   c("cycle", "condition"),
                                                   essential_nonessential_levels)

# observed
pdf("SNV_essential_nonessential_cycle_observed_downsampled.pdf",
    width = 6,
    height = 5)
ggplot(essential_nonessential_regions, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.7) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Cycle", y = "Log2(observed/expected)\nSNVs Per Cell", fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_x_discrete(labels = c("Control", "ENU")) + 
    theme_Publication() +
    geom_text(data = essential_nonessential_regions_ratio_significance,
              aes(x = condition, y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    facet_grid(.~region) +
    theme(legend.justification = c("center","top"),
          legend.position = "top",
          axis.text.x = element_text(size = 12)) +
    ylim(NA, max(essential_nonessential_regions$log2fc) + 0.1)
dev.off()
```



## Gene ontology
- Improvement: Incorporate average expression of each pathway

### Compare ratios wihtin conditions

#### ENU

##### Observed
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")

go_expressed_regions_levels <- unique(go_expressed$region)

go_expressed_regions <- merge(go_expressed, sample_table, by = "sample")

go_expressed_regions$region <- factor(go_expressed_regions$region, go_expressed_regions_levels)

go_expressed_regions <- subset(go_expressed_regions, condition == "enu")

go_expressed_regions_ratio_significance <- ratio_significance_pathway(go_expressed_regions,
                                                   "condition",
                                                   go_expressed_regions_levels)


# summarize by getting average log2fc for each group and then compare groups
go_expressed_regions_summary <- subset(go_expressed_regions) %>%
  group_by(region, condition) %>%
  summarise(median_log2fc = median(log2fc, na.rm = TRUE),
            median.obs.mut = median(obs.mut, na.rm = TRUE))

# add significance values
go_expressed_regions_summary <- merge(go_expressed_regions_summary, 
                                      go_expressed_regions_ratio_significance, 
                                      by = "region")
# add GO IDs
go_expressed_regions_summary$GO.ID <- go_ids$ID[match(go_expressed_regions_summary$region, go_ids$Description)]

# add size of gene set
go_expressed_regions_summary$genes <- go_ids$geneID[match(go_expressed_regions_summary$region, go_ids$Description)]
go_expressed_regions_summary$gene_length <- unlist(lapply(strsplit(go_expressed_regions_summary$genes, "/"), length))

# add p-value correction
go_expressed_regions_summary$pvalue.adjust <- p.adjust(go_expressed_regions_summary$pvalue, method = "BH")

# filter out genes with 0 median obs mutations
go_expressed_regions_summary <- go_expressed_regions_summary %>%
  filter(median.obs.mut >= 1)

# get significant hits
go_expressed_regions_summary_sig <- go_expressed_regions_summary %>%
  filter(pvalue.adjust <= 0.05)

# enu vs control
write.csv(go_expressed_regions_summary, "enu_ratio_results.csv", row.names = FALSE)

# volcano
pdf("enu_ratio_volcano.pdf",
    width = 6,
    height = 6)
EnhancedVolcano(go_expressed_regions_summary,
                lab = go_expressed_regions_summary$region,
                title = "ENU-treated cells",
                subtitle = "Log2(Expected vs Observed SNV ratios)",
                caption = NULL,
                x = 'median_log2fc',
                y = 'pvalue',
                pCutoffCol = "pvalue.adjust",
                pointSize = 4.0,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 4,
                labCol = 'black',
                #labFace = 'bold',
                #boxedLabels = TRUE,
                drawConnectors = TRUE,
                #widthConnectors = 1.0,
                #colConnectors = 'black',
                #maxoverlapsConnectors = Inf,
                #selectLab = KO_vs_WT$volcano.label,
                #ylim = c(0,7),
                # xlim = c(-3, 1.5),
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
dev.off()
```

###### Relationships
```{r}
pdf("enu_ratio_sig_pvalue_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go_expressed_regions_summary$pvalue,
     go_expressed_regions_summary$gene_length)
dev.off()

pdf("enu_ratio_sig_log2fc_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go_expressed_regions_summary$median_log2fc,
     go_expressed_regions_summary$gene_length)
dev.off()

pdf("enu_ratio_sig_pvalue_vs_obs_mut.pdf",
    width = 4,
    height = 4)
plot(go_expressed_regions_summary$pvalue,
     go_expressed_regions_summary$median.obs.mut)
dev.off()

pdf("enu_ratio_gene_length_vs_obs_mut.pdf",
    width = 4,
    height = 4)
plot(go_expressed_regions_summary$gene_length,
     go_expressed_regions_summary$median.obs.mut)
dev.off()

pdf("enu_ratio_obs_mut_vs_log2fc.pdf",
    width = 4,
    height = 4)
plot(go_expressed_regions_summary$median.obs.mut,
     go_expressed_regions_summary$median_log2fc)
dev.off()
```

##### Shuffled
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_shuffled_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_shuffled_regions")

go_expressed_shuffled_regions_levels <- unique(go_expressed_shuffled$region)

go_expressed_shuffled_regions <- merge(go_expressed_shuffled, sample_table, by = "sample")

go_expressed_shuffled_regions$region <- factor(go_expressed_shuffled_regions$region, go_expressed_shuffled_regions_levels)

go_expressed_shuffled_regions <- subset(go_expressed_shuffled_regions, condition == "enu")

go_expressed_shuffled_regions_ratio_significance <- ratio_significance_pathway(go_expressed_shuffled_regions,
                                                   "condition",
                                                   go_expressed_shuffled_regions_levels)


# summarize by getting average log2fc for each group and then compare groups
go_expressed_shuffled_regions_summary <- subset(go_expressed_shuffled_regions) %>%
  group_by(region, condition) %>%
  summarise(median_log2fc = median(log2fc, na.rm = TRUE),
            median.obs.mut = median(obs.mut, na.rm = TRUE))

# add significance values
go_expressed_shuffled_regions_summary <- merge(go_expressed_shuffled_regions_summary, 
                                      go_expressed_shuffled_regions_ratio_significance, 
                                      by = "region")
# add GO IDs
go_expressed_shuffled_regions_summary$GO.ID <- go_ids$ID[match(go_expressed_shuffled_regions_summary$region, go_ids$Description)]

# add size of gene set
go_expressed_shuffled_regions_summary$genes <- go_ids$geneID[match(go_expressed_shuffled_regions_summary$region, go_ids$Description)]
go_expressed_shuffled_regions_summary$gene_length <- unlist(lapply(strsplit(go_expressed_shuffled_regions_summary$genes, "/"), length))

# add p-value correction
go_expressed_shuffled_regions_summary$pvalue.adjust <- p.adjust(go_expressed_shuffled_regions_summary$pvalue, method = "BH")

# filter out genes with 0 median obs mutations
go_expressed_shuffled_regions_summary <- go_expressed_shuffled_regions_summary %>%
  filter(median.obs.mut >= 1)

# get significant hits
go_expressed_shuffled_regions_summary_sig <- go_expressed_shuffled_regions_summary %>%
  filter(pvalue.adjust <= 0.05)

# enu vs control
write.csv(go_expressed_shuffled_regions_summary, "enu_ratio_shuffled_results.csv", row.names = FALSE)

# volcano
pdf("enu_ratio_shuffled_volcano.pdf",
    width = 6,
    height = 6)
EnhancedVolcano(go_expressed_shuffled_regions_summary,
                lab = go_expressed_shuffled_regions_summary$region,
                title = "ENU-treated cells",
                subtitle = "Log2(Expected vs Observed SNV ratios)",
                caption = NULL,
                x = 'median_log2fc',
                y = 'pvalue',
                pCutoffCol = "pvalue.adjust",
                pointSize = 4.0,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 4,
                labCol = 'black',
                #labFace = 'bold',
                #boxedLabels = TRUE,
                drawConnectors = TRUE,
                #widthConnectors = 1.0,
                #colConnectors = 'black',
                #maxoverlapsConnectors = Inf,
                #selectLab = KO_vs_WT$volcano.label,
                #ylim = c(0,7),
                # xlim = c(-3, 1.5),
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
dev.off()
```

###### Relationships
```{r}
pdf("enu_ratio_sig_pvalue_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go_expressed_shuffled_regions_summary$pvalue,
     go_expressed_shuffled_regions_summary$gene_length)
dev.off()

pdf("enu_ratio_sig_log2fc_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go_expressed_shuffled_regions_summary$median_log2fc,
     go_expressed_shuffled_regions_summary$gene_length)
dev.off()

pdf("enu_ratio_sig_pvalue_vs_obs_mut.pdf",
    width = 4,
    height = 4)
plot(go_expressed_shuffled_regions_summary$pvalue,
     go_expressed_shuffled_regions_summary$median.obs.mut)
dev.off()

pdf("enu_ratio_gene_length_vs_obs_mut.pdf",
    width = 4,
    height = 4)
plot(go_expressed_shuffled_regions_summary$gene_length,
     go_expressed_shuffled_regions_summary$median.obs.mut)
dev.off()

pdf("enu_ratio_obs_mut_vs_log2fc.pdf",
    width = 4,
    height = 4)
plot(go_expressed_shuffled_regions_summary$median.obs.mut,
     go_expressed_shuffled_regions_summary$median_log2fc)
dev.off()
```

##### Overlap of genes in significant pathways
- No overlaps
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/1-load-data")

# load granges and results of enrichments
go_expressed_granges <- readRDS("go.expressed.RDS")
enrich.go_expressed <- readRDS("enrich.go.expressed.RDS")

# filter significant hits in real based on shuffle
go_expressed_regions_summary_sig <- go_expressed_regions_summary_sig %>%
  filter(!GO.ID %in% go_expressed_shuffled_regions_summary_sig$GO.ID)

# filter for sig hits
go_expressed_granges_sig <- go_expressed_granges[names(go_expressed_granges) %in% go_expressed_regions_summary_sig$region]
enrich.go_expressed.sig <- filter(enrich.go_expressed, ID %in% go_expressed_regions_summary_sig$GO.ID)

# get sequence length distribution and add to results
lengths <- unlist(lapply(go_expressed_granges_sig, function(x) sum(width(x))))
go_expressed_regions_summary_sig$sequence_length <- lengths[match(go_expressed_regions_summary_sig$region, names(lengths))]

setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")

# edox <- pairwise_termsim(enrich.go_expressed.sig)
# pdf("enu_ratio_sig_filtered_tree.pdf",
#     width = 10,
#     height = 10)
# treeplot(edox)
# dev.off()

# using all go terms too complicated
# pdf("enu_ratio_sig_filtered_cnet.pdf",
#     width = 10,
#     height = 10)
# cnetplot(edox, 
#          showCategory = 5)
# dev.off()

# get shared genes amongst significant go terms
# genelist <- strsplit(go_expressed_regions_summary_sig$genes, "/")
# names(genelist) <- go_expressed_regions_summary_sig$region
# genelist <- fromList(genelist)
# genelist.reduced <- genelist[rowSums(genelist[, names(genelist)]) > 1,] # filter out sets with no overlap between others
#                        
# pdf("enu_ratio_sig_filtered_upset.pdf",
#     width = 8,
#     height = 5)
# upset(genelist.reduced,
#       nsets	= 10,
#       nintersects = NA,
#       order.by = "freq",
#       show.numbers = FALSE)
# dev.off()

# save
# write.csv(go_expressed_regions_summary_sig, "enu_ratio_results_filtered.csv", row.names = FALSE)
```

##### Final plots
- No significant genes
```{r}
# plot bar plot of significant hits
go_expressed_regions_summary_sig <- go_expressed_regions_summary_sig[order(go_expressed_regions_summary_sig$median_log2fc, decreasing = TRUE),]
go_expressed_regions_summary_sig$region <- factor(go_expressed_regions_summary_sig$region, levels = go_expressed_regions_summary_sig$region)

pdf("enu_ratio_sig_filtered_bar.pdf",
    width = 8,
    height = 5)
ggplot(go_expressed_regions_summary_sig[1:16,], aes(x = region, y = median_log2fc, fill = pvalue.adjust)) +
  geom_bar(stat = "identity") +
  scale_y_reverse() +
  coord_flip() +
  scale_fill_viridis() +
  theme_Publication() +
  theme(legend.justification = c("right","top"),
        legend.position = "top",
        legend.title = element_text(size=15),
        legend.text = element_text(size=10),
        axis.title.x = element_text(size = 12)) +
  labs(x = "GO Biological Process", 
       y = "Mean Log2(observed/expected)\nSNVs Per Cell",
       fill = "Adjusted p-value")
dev.off()

# bar plot of number of increased estimated mutations
bar_plot <- subset(go_expressed_regions, group == "observed") %>%
  mutate(est.mut = obs.mut / Cov / Sen_SNV * hg19_size) %>%
  group_by(region, condition) %>%
  summarise(mean.est.mut = mean(est.mut, na.rm = TRUE)) %>%
  group_by(region) %>%
  spread(key = condition, value = mean.est.mut) %>%
  mutate(group_diff = abs(enu - control)) %>%
  filter(region %in% go_expressed_regions_summary_sig[1:16,]$region)

bar_plot <- bar_plot[match(go_expressed_regions_summary_sig[1:16,]$region, bar_plot$region),]
bar_plot$region <- factor(bar_plot$region, levels = bar_plot$region)
bar_plot$group_diff

pdf("enu_number_sig_filtered_bar.pdf",
    width = 8,
    height = 4.5)
ggplot(bar_plot, aes(x = region, y = group_diff)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_Publication() +
  theme( 
        axis.title.x = element_text(size = 12)) +
  labs(x = "GO Biological Process", 
       y = "Mean Estimated SNV increase\nin ENU-treated cells")
dev.off()
```


#### Control

##### Observed
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")

go_expressed_regions_levels <- unique(go_expressed$region)

go_expressed_regions <- merge(go_expressed, sample_table, by = "sample")

go_expressed_regions$region <- factor(go_expressed_regions$region, go_expressed_regions_levels)

go_expressed_regions <- subset(go_expressed_regions, condition == "control")

go_expressed_regions_ratio_significance <- ratio_significance_pathway(go_expressed_regions,
                                                   "condition",
                                                   go_expressed_regions_levels)


# summarize by getting average log2fc for each group and then compare groups
go_expressed_regions_summary <- subset(go_expressed_regions) %>%
  group_by(region, condition) %>%
  summarise(median_log2fc = median(log2fc, na.rm = TRUE),
            median.obs.mut = median(obs.mut, na.rm = TRUE))

# add significance values
go_expressed_regions_summary <- merge(go_expressed_regions_summary, 
                                      go_expressed_regions_ratio_significance, 
                                      by = "region")
# add GO IDs
go_expressed_regions_summary$GO.ID <- go_ids$ID[match(go_expressed_regions_summary$region, go_ids$Description)]

# add size of gene set
go_expressed_regions_summary$genes <- go_ids$geneID[match(go_expressed_regions_summary$region, go_ids$Description)]
go_expressed_regions_summary$gene_length <- unlist(lapply(strsplit(go_expressed_regions_summary$genes, "/"), length))

# add p-value correction
go_expressed_regions_summary$pvalue.adjust <- p.adjust(go_expressed_regions_summary$pvalue, method = "BH")

# filter out genes with 0 median obs mutations
go_expressed_regions_summary <- go_expressed_regions_summary %>%
  filter(median.obs.mut >= 1)

# get significant hits
go_expressed_regions_summary_sig <- go_expressed_regions_summary %>%
  filter(pvalue.adjust <= 0.05)

# control vs control
write.csv(go_expressed_regions_summary, "control_ratio_results.csv", row.names = FALSE)

# volcano
pdf("control_ratio_volcano.pdf",
    width = 6,
    height = 6)
EnhancedVolcano(go_expressed_regions_summary,
                lab = go_expressed_regions_summary$region,
                title = "control-treated cells",
                subtitle = "Log2(Expected vs Observed SNV ratios)",
                caption = NULL,
                x = 'median_log2fc',
                y = 'pvalue',
                pCutoffCol = "pvalue.adjust",
                pointSize = 4.0,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 4,
                labCol = 'black',
                #labFace = 'bold',
                #boxedLabels = TRUE,
                drawConnectors = TRUE,
                #widthConnectors = 1.0,
                #colConnectors = 'black',
                #maxoverlapsConnectors = Inf,
                #selectLab = KO_vs_WT$volcano.label,
                #ylim = c(0,7),
                # xlim = c(-3, 1.5),
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
dev.off()
```

###### Relationships
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")

pdf("control_ratio_sig_pvalue_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go_expressed_regions_summary$pvalue,
     go_expressed_regions_summary$gene_length)
dev.off()

pdf("control_ratio_sig_log2fc_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go_expressed_regions_summary$median_log2fc,
     go_expressed_regions_summary$gene_length)
dev.off()

pdf("control_ratio_sig_pvalue_vs_obs_mut.pdf",
    width = 4,
    height = 4)
plot(go_expressed_regions_summary$pvalue,
     go_expressed_regions_summary$median.obs.mut)
dev.off()

pdf("control_ratio_gene_length_vs_obs_mut.pdf",
    width = 4,
    height = 4)
plot(go_expressed_regions_summary$gene_length,
     go_expressed_regions_summary$median.obs.mut)
dev.off()

pdf("control_ratio_obs_mut_vs_log2fc.pdf",
    width = 4,
    height = 4)
plot(go_expressed_regions_summary$median.obs.mut,
     go_expressed_regions_summary$median_log2fc)
dev.off()
```

##### Shuffled
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")

go_expressed_shuffled_regions_levels <- unique(go_expressed_shuffled$region)

go_expressed_shuffled_regions <- merge(go_expressed_shuffled, sample_table, by = "sample")

go_expressed_shuffled_regions$region <- factor(go_expressed_shuffled_regions$region, go_expressed_shuffled_regions_levels)

go_expressed_shuffled_regions <- subset(go_expressed_shuffled_regions, condition == "control")

go_expressed_shuffled_regions_ratio_significance <- ratio_significance_pathway(go_expressed_shuffled_regions,
                                                   "condition",
                                                   go_expressed_shuffled_regions_levels)


# summarize by getting average log2fc for each group and then compare groups
go_expressed_shuffled_regions_summary <- subset(go_expressed_shuffled_regions) %>%
  group_by(region, condition) %>%
  summarise(median_log2fc = median(log2fc, na.rm = TRUE),
            median.obs.mut = median(obs.mut, na.rm = TRUE))

# add significance values
go_expressed_shuffled_regions_summary <- merge(go_expressed_shuffled_regions_summary, 
                                      go_expressed_shuffled_regions_ratio_significance, 
                                      by = "region")
# add GO IDs
go_expressed_shuffled_regions_summary$GO.ID <- go_ids$ID[match(go_expressed_shuffled_regions_summary$region, go_ids$Description)]

# add size of gene set
go_expressed_shuffled_regions_summary$genes <- go_ids$geneID[match(go_expressed_shuffled_regions_summary$region, go_ids$Description)]
go_expressed_shuffled_regions_summary$gene_length <- unlist(lapply(strsplit(go_expressed_shuffled_regions_summary$genes, "/"), length))

# add p-value correction
go_expressed_shuffled_regions_summary$pvalue.adjust <- p.adjust(go_expressed_shuffled_regions_summary$pvalue, method = "BH")

# filter out genes with 0 median obs mutations
go_expressed_shuffled_regions_summary <- go_expressed_shuffled_regions_summary %>%
  filter(median.obs.mut >= 1)

# get significant hits
go_expressed_shuffled_regions_summary_sig <- go_expressed_shuffled_regions_summary %>%
  filter(pvalue.adjust <= 0.05)

# control vs control
write.csv(go_expressed_shuffled_regions_summary, "control_ratio_shuffled_results.csv", row.names = FALSE)

# volcano
pdf("control_ratio_shuffled_volcano.pdf",
    width = 6,
    height = 6)
EnhancedVolcano(go_expressed_shuffled_regions_summary,
                lab = go_expressed_shuffled_regions_summary$region,
                title = "control-treated cells",
                subtitle = "Log2(Expected vs Observed SNV ratios)",
                caption = NULL,
                x = 'median_log2fc',
                y = 'pvalue',
                pCutoffCol = "pvalue.adjust",
                pointSize = 4.0,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 4,
                labCol = 'black',
                #labFace = 'bold',
                #boxedLabels = TRUE,
                drawConnectors = TRUE,
                #widthConnectors = 1.0,
                #colConnectors = 'black',
                #maxoverlapsConnectors = Inf,
                #selectLab = KO_vs_WT$volcano.label,
                #ylim = c(0,7),
                # xlim = c(-3, 1.5),
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
dev.off()
```

###### Relationships
```{r}
pdf("control_ratio_sig_pvalue_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go_expressed_shuffled_regions_summary$pvalue,
     go_expressed_shuffled_regions_summary$gene_length)
dev.off()

pdf("control_ratio_sig_log2fc_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go_expressed_shuffled_regions_summary$median_log2fc,
     go_expressed_shuffled_regions_summary$gene_length)
dev.off()

pdf("control_ratio_sig_pvalue_vs_obs_mut.pdf",
    width = 4,
    height = 4)
plot(go_expressed_shuffled_regions_summary$pvalue,
     go_expressed_shuffled_regions_summary$median.obs.mut)
dev.off()

pdf("control_ratio_gene_length_vs_obs_mut.pdf",
    width = 4,
    height = 4)
plot(go_expressed_shuffled_regions_summary$gene_length,
     go_expressed_shuffled_regions_summary$median.obs.mut)
dev.off()

pdf("control_ratio_obs_mut_vs_log2fc.pdf",
    width = 4,
    height = 4)
plot(go_expressed_shuffled_regions_summary$median.obs.mut,
     go_expressed_shuffled_regions_summary$median_log2fc)
dev.off()
```

##### Overlap of genes in significant pathways
- No overlaps
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/1-load-data")

# load granges and results of enrichments
go_expressed_granges <- readRDS("go.expressed.RDS")
enrich.go_expressed <- readRDS("enrich.go.expressed.RDS")

# filter significant hits in real based on shuffle
go_expressed_regions_summary_sig <- go_expressed_regions_summary_sig %>%
  filter(!GO.ID %in% go_expressed_shuffled_regions_summary_sig$GO.ID)

# filter for sig hits
go_expressed_granges_sig <- go_expressed_granges[names(go_expressed_granges) %in% go_expressed_regions_summary_sig$region]
enrich.go_expressed.sig <- filter(enrich.go_expressed, ID %in% go_expressed_regions_summary_sig$GO.ID)

# get sequence length distribution and add to results
lengths <- unlist(lapply(go_expressed_granges_sig, function(x) sum(width(x))))
go_expressed_regions_summary_sig$sequence_length <- lengths[match(go_expressed_regions_summary_sig$region, names(lengths))]

setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")

# edox <- pairwise_termsim(enrich.go_expressed.sig)
# pdf("control_ratio_sig_filtered_tree.pdf",
#     width = 10,
#     height = 10)
# treeplot(edox)
# dev.off()

# using all go terms too complicated
# pdf("control_ratio_sig_filtered_cnet.pdf",
#     width = 10,
#     height = 10)
# cnetplot(edox, 
#          showCategory = 5)
# dev.off()

# get shared genes amongst significant go terms
# genelist <- strsplit(go_expressed_regions_summary_sig$genes, "/")
# names(genelist) <- go_expressed_regions_summary_sig$region
# genelist <- fromList(genelist)
# genelist.reduced <- genelist[rowSums(genelist[, names(genelist)]) > 1,] # filter out sets with no overlap between others
#                        
# pdf("control_ratio_sig_filtered_upset.pdf",
#     width = 8,
#     height = 5)
# upset(genelist.reduced,
#       nsets	= 10,
#       nintersects = NA,
#       order.by = "freq",
#       show.numbers = FALSE)
# dev.off()

# save
# write.csv(go_expressed_regions_summary_sig, "control_ratio_results_filtered.csv", row.names = FALSE)
```

##### Final plots
- No significant genes
```{r, eval = FALSE}
# plot bar plot of significant hits
go_expressed_regions_summary_sig <- go_expressed_regions_summary_sig[order(go_expressed_regions_summary_sig$median_log2fc, decreasing = TRUE),]
go_expressed_regions_summary_sig$region <- factor(go_expressed_regions_summary_sig$region, levels = go_expressed_regions_summary_sig$region)

pdf("control_ratio_sig_filtered_bar.pdf",
    width = 8,
    height = 5)
ggplot(go_expressed_regions_summary_sig[1:16,], aes(x = region, y = median_log2fc, fill = pvalue.adjust)) +
  geom_bar(stat = "identity") +
  scale_y_reverse() +
  coord_flip() +
  scale_fill_viridis() +
  theme_Publication() +
  theme(legend.justification = c("right","top"),
        legend.position = "top",
        legend.title = element_text(size=15),
        legend.text = element_text(size=10),
        axis.title.x = element_text(size = 12)) +
  labs(x = "GO Biological Process", 
       y = "Mean Log2(observed/expected)\nSNVs Per Cell",
       fill = "Adjusted p-value")
dev.off()

# bar plot of number of increased estimated mutations
bar_plot <- subset(go_expressed_regions) %>%
  mutate(est.mut = obs.mut / Cov / Sen_SNV * hg19_size) %>%
  group_by(region, condition) %>%
  summarise(mean.est.mut = mean(est.mut, na.rm = TRUE)) %>%
  group_by(region) %>%
  spread(key = condition, value = mean.est.mut) %>%
  mutate(group_diff = abs(control - control)) %>%
  filter(region %in% go_expressed_regions_summary_sig[1:16,]$region)

bar_plot <- bar_plot[match(go_expressed_regions_summary_sig[1:16,]$region, bar_plot$region),]
bar_plot$region <- factor(bar_plot$region, levels = bar_plot$region)
bar_plot$group_diff

pdf("control_number_sig_filtered_bar.pdf",
    width = 8,
    height = 4.5)
ggplot(bar_plot, aes(x = region, y = group_diff)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_Publication() +
  theme( 
        axis.title.x = element_text(size = 12)) +
  labs(x = "GO Biological Process", 
       y = "Mean Estimated SNV increase\nin control-treated cells")
dev.off()
```

### ENU vs Control

#### Observed
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/downsampled/go_expressed_regions")

go_expressed_regions_levels <- unique(go_expressed$region)

go_expressed_regions <- merge(go_expressed, sample_table, by = "sample")

go_expressed_regions$region <- factor(go_expressed_regions$region, go_expressed_regions_levels)

go_expressed_regions_ratio_significance <- ratio_significance_pathway(go_expressed_regions,
                                                   "condition",
                                                   go_expressed_regions_levels)

go_expressed_regions_group_significance <- group_significance(go_expressed_regions,
                                                         c("group", "region"),
                                                         "condition",
                                                         go_expressed_regions_levels)

# summarize by getting average log2fc for each group and then compare groups
go_expressed_regions_summary <- subset(go_expressed_regions) %>%
  group_by(region, condition) %>%
  summarise(median_log2fc = mean(log2fc, na.rm = TRUE)) %>%
  group_by(region) %>%
  spread(key = condition, value = median_log2fc) %>%
  mutate(group_diff = enu - control)

# add group significance values
go_expressed_regions_summary <- merge(go_expressed_regions_summary, 
                                      go_expressed_regions_group_significance, 
                                      by = "region")
# add GO IDs
go_expressed_regions_summary$GO.ID <- go_ids$ID[match(go_expressed_regions_summary$region, go_ids$Description)]

# add p-value corerection
go_expressed_regions_summary$p_value.adjust <- p.adjust(go_expressed_regions_summary$p_value, method = "BH")

# enu vs control
write.csv(go_expressed_regions_summary, "enu_vs_control_ratio_results.csv", row.names = FALSE)

pdf("enu_vs_control_ratio_volcano.pdf",
    width = 6,
    height = 6)
EnhancedVolcano(go_expressed_regions_summary,
                lab = go_expressed_regions_summary$region,
                title = "ENU vs Control expected vs observed ratios",
                subtitle = NULL,
                caption = NULL,
                x = 'group_diff',
                y = 'p_value',
                pCutoffCol = "p_value.adjust",
                pointSize = 4.0,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 4,
                labCol = 'black',
                labFace = 'bold',
                #boxedLabels = TRUE,
                #drawConnectors = TRUE,
                #widthConnectors = 1.0,
                #colConnectors = 'black',
                #maxoverlapsConnectors = Inf,
                #selectLab = KO_vs_WT$volcano.label,
                ylim = c(0,3),
                xlim = c(-2,2),
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
dev.off()
```