---
title: "VEP analysis"
output: html_document
date: "2023-10-12"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

## Load libraries
```{r}
library(ggplot2)
library(stringr)
library(dplyr)
library(ggpubr)
library(purrr)
library(rstatix)
library(viridis)
library(tidyr)
library(emmeans)
```

## Functions
```{r}
# combine the permutation pvalues from each sample
# using fishers combined test statistic
combine_pvalues_fisher <- function(p_values) {
  
    chi_sq_stat = -2 * sum(log(p_values), na.rm = TRUE)
    degrees_of_freedom = 2 * length(p_values)
    p_combined = pchisq(chi_sq_stat, degrees_of_freedom, lower.tail = FALSE)
    return(p_combined)
    
}

consequences_combined_significance <- function(ratios, group_vars, group_levels) {
  
  res <- ratios %>%
    subset(variant_type %in% group_levels) %>%
    mutate(pvalue = if_else(pvalue == 0, 0.0001, pvalue)) %>% # give values to very low pvalues that are 0 
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(p_value = combine_pvalues_fisher(pvalue)) %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)) %>%
    ungroup()
  
  return(res)
}

consequences_combined_significance_CADD <- function(ratios, group_vars, group_levels) {
  
  res <- ratios %>%
    mutate(pvalue = if_else(pvalue == 0, 0.001, pvalue)) %>% # give values to very low pvalues that are 0 
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(p_value = combine_pvalues_fisher(pvalue)) %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)) %>%
    ungroup()
  
  return(res)
}

consequences_group_significance <- function(ratios, group_vars, group_levels) {
  
  res <- ratios %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(p_value = suppressWarnings(wilcox.test(log2fc, mu = 0)$p.value)) %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)) %>%
    ungroup()

  
  return(res)
}

score_combined_significance <- function(ratios, group_vars, group_levels) {
  
  res <- ratios %>%
    mutate(pvalue = if_else(pvalue == 0, 0.001, pvalue)) %>% # give values to very low pvalues that are 0 
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(p_value = combine_pvalues_fisher(pvalue)) %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)) %>%
    ungroup()
  
  return(res)
}

# function to calculate significance between condition groups
group_significance <- function(combined_regions, group_vars, condition, group_levels){
  formula <- as.formula(paste("log2fc ~", condition))
  
  results <- combined_regions %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(
      p_value = wilcox.test(formula = formula, 
                            data = cur_data(),
                            exact = FALSE)$p.value
    ) %>%
    ungroup() %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)
    )

  # Order by region
  results$variant_type <- factor(results$variant_type, levels = group_levels)
  
  return(results)
}

ratio_significance_summarize <- function(ratios, group_vars, group_levels) {
  
  res <- ratios %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(p_value = combine_pvalues_fisher(pvalue)) %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)) %>%
    ungroup()
  
  # Order by region
  res$type <- factor(res$type, levels = group_levels)
  
  return(res)
}

group_significance_summarize <- function(combined_regions, group_vars, condition, group_levels){
  formula <- as.formula(paste("log2fc ~", condition))
  
  results <- combined_regions %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(
      p_value = wilcox.test(formula = formula, 
                            data = cur_data(),
                            exact = FALSE)$p.value
    ) %>%
    ungroup() %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)
    )

  # Order by region
  results$type <- factor(results$type, levels = group_levels)
  
  return(results)
}

calculate_dnds <- function(df, variant_types, sample_table) {
  variant_type2 <- "synonymous variant" # Fixed variant type2
  
  # Initialize an empty data frame to store results
  all_variant_ratios <- data.frame()
  
  for (variant_type1 in variant_types) {
    # Filter and calculate dnds for observed mutations
    variant.obs <- df %>%
      filter(variant_type %in% c(variant_type1, variant_type2)) %>%
      select(sample, variant_type, obs.mut) %>%
      mutate(obs.mut = obs.mut + 1) %>%
      spread(key = variant_type, value = obs.mut, fill = 1) %>%  # fill = 1 to handle cases with 0 count
      mutate(dnds = !!sym(variant_type1) / !!sym(variant_type2),
             variant_type = variant_type1) %>%
      select(sample, dnds, variant_type)
    
    # Combine results of this variant type with the results of previous ones
    all_variant_ratios <- rbind(all_variant_ratios, variant.obs)
  }
      
  # Combine with meta data
  all_variant_ratios <- merge(all_variant_ratios, sample_table, by = "sample")
  
  return(all_variant_ratios)
}

calculate_dnds_ratio <- function(df, variant_types, sample_table) {
  variant_type2 <- "synonymous variant" # Fixed variant type2
  
  # Initialize an empty data frame to store results
  all_variant_ratios <- data.frame()
  
  for (variant_type1 in variant_types) {
    # Filter and calculate dnds for observed mutations
    variant.obs <- df %>%
      filter(variant_type %in% c(variant_type1, variant_type2)) %>%
      select(sample, variant_type, obs.mut) %>%
      mutate(obs.mut = obs.mut + 1) %>%
      spread(key = variant_type, value = obs.mut, fill = 1) %>%  # fill = 1 to handle cases with 0 count
      mutate(dnds.obs = log2(!!sym(variant_type1)/ !!sym(variant_type2)))
    
    # Filter and calculate dnds for simulated mutations
    variant.sim <- df %>%
      filter(variant_type %in% c(variant_type1, variant_type2)) %>%
      select(sample, variant_type, sim.mut) %>%
      mutate(sim.mut = sim.mut + 1) %>%
      spread(key = variant_type, value = sim.mut, fill = 1) %>%
      mutate(dnds.sim = log2(!!sym(variant_type1)/ !!sym(variant_type2)))
    
    # Calculate the observed/simulated ratio
    variant.ratio <- variant.obs
    colnames(variant.ratio)[2] <- "nonsynonymous variant"
    variant.ratio$dnds.sim <- variant.sim$dnds.sim
    variant.ratio$dnds.ratio <- variant.ratio$dnds.obs - variant.ratio$dnds.sim
    variant.ratio$variant_type <- variant_type1  # Add column to indicate the current variant type
    
    
    # Combine results of this variant type with the results of previous ones
    all_variant_ratios <- rbind(all_variant_ratios, variant.ratio)
  }
      
  # Combine with meta data
  all_variant_ratios <- merge(all_variant_ratios, sample_table, by = "sample")
  
  return(all_variant_ratios)
}

ratio_significance_dnds <- function(combined_regions, group_vars, group_levels, variable){
  
  results <- combined_regions %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(
      p_value = wilcox.test(x = !!sym(variable), mu = 0, exact = FALSE)$p.value
    ) %>%
    ungroup() %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
             p_adjusted <= 0.0001 ~ '****',
             p_adjusted <= 0.001 ~ '***',
             p_adjusted <= 0.01 ~ '**',
             p_adjusted <= 0.05 ~ '*',
             TRUE ~ NA_character_)
    )

  # Order by region
  results$variant_type <- factor(results$variant_type, levels = group_levels)
  
  return(results)
}

group_significance_dnds <- function(combined_regions, group_vars, condition, group_levels, variable){
  formula <- as.formula(paste(variable, " ~", condition))
  
  results <- combined_regions %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(
      p_value = wilcox.test(formula = formula, 
                            data = cur_data(),
                            exact = FALSE)$p.value
    ) %>%
    ungroup() %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)
    )

  # Order by region
  results$variant_type <- factor(results$variant_type, levels = group_levels)
  
  return(results)
}

calculate_stat_summary <- function(observed, group_vars, mut_var){
  results <- observed %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(mean = mean(!!sym(mut_var), na.rm = TRUE),
              sd = sd(!!sym(mut_var), na.rm = TRUE))
}

theme_Publication <- function(base_size=16, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family = "")
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               #legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_fill_Publication_2 <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#7fc97f","#ef3b2c","#386cb0","#fdb462","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

wrap_labels <- function(labels, width = 5) {
  wrapped_labels <- str_wrap(labels, width)
  names(wrapped_labels) <- names(labels)
  return(wrapped_labels)
}
```

## Load sample table
```{r}
sample_table <- read.table("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/1-data/sample_table_filtered.txt", header = TRUE, sep = "\t")
```

## Load ratio results
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/231009 multiple ENU analysis/5-variant-effect-prediction/1-calculate-ratio/cells-combined")

# consequences
consequences.main.mean <- read.csv("consequence_main_oe_mean.csv", header = TRUE, row.names = NULL)
consequences.syn_non.mean <- read.csv("consequence_synonymous_nonsynonymous_oe_mean.csv", header = TRUE, row.names = NULL)
```

# Consequences

## Number of different types of variants
```{r}
print(consequences.main.mean %>%
  group_by(variant_type) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

print(consequences.syn_non.mean %>%
  group_by(variant_type) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)
```

## Get number of SNV vs INDEL variants
```{r}
consequences.main.mean <- consequences.main.mean %>%
  mutate(mutation_type = case_when(
    variant_type == "inframe insertion" ~ "INDEL",
    variant_type == "inframe deletion" ~ "INDEL",
    variant_type == "frameshift variant" ~ "INDEL",
    .default = "SNV"
    )
  )

consequences.main.mean %>%
  group_by(mutation_type) %>%
  summarize(total = sum(obs.mut))
```

## Get number of coding vs non-coding
```{r}
consequences.main.mean <- consequences.main.mean %>%
  mutate(group = case_when(
    variant_type == "synonymous variant" ~ "coding",
    variant_type == "missense variant" ~ "coding",
    variant_type == "start lost" ~ "coding",
    variant_type == "stop gained" ~ "coding",
    variant_type == "stop lost" ~ "coding",
    variant_type == "splice acceptor variant" ~ "coding",
    variant_type == "splice donor variant" ~ "coding",
    variant_type == "inframe insertion" ~ "coding",
    variant_type == "inframe deletion" ~ "coding",
    variant_type == "frameshift variant" ~ "coding",
    variant_type == "coding sequence variant / NMD transcript variant" ~ "coding",
    variant_type == "synonymous variant / NMD transcript variant" ~ "coding",
    .default = "non-coding"
    )
  )

consequences.main.mean %>%
  group_by(group) %>%
  summarize(total = sum(obs.mut))
```

### All Variants
- Using mean as this allows the estimates to be more accurate
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences")

# plot mean
consequences.main.mean$cycle <- factor(consequences.main.mean$cycle, 
                                       levels = c(1, 3, 6, 9))
consequences.main.mean$condition <- factor(consequences.main.mean$condition, 
                                           levels = c("control", "enu"))

pdf("consequences_cycle_main_mean.pdf",
    width = 30,
    height = 30)
ggplot(consequences.main.mean, aes(x = condition, y = log2fc, color = cycle)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=log2fc - log2fc.sd, ymax=log2fc + log2fc.sd),
                    position=position_dodge(width = 0.7),
                    fatten = 12) +
    guides(fill = guide_legend(override.aes = list(shape = 16)), 
           color = guide_legend(override.aes = list(shape = 16))) +
    labs(x = "Consequence", y = "Log2(observed/expected)\nVariants Per Cell") +
    facet_wrap(.~variant_type, scales = "free", labeller = labeller(variant_type = label_wrap_gen(30))) + 
    scale_color_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.justification = c("center","top"),
          legend.position = "top",
          axis.text.x = element_text(size = 12))
dev.off()

# plot median
# consequences.main.median$cycle <- factor(consequences.main.median$cycle, 
#                                        levels = c(1, 3, 6, 9))
# consequences.main.median$condition <- factor(consequences.main.median$condition, 
#                                            levels = c("control", "enu"))
# 
# pdf("consequences_cycle_main_median.pdf",
#     width = 30,
#     height = 30)
# ggplot(consequences.main.median, aes(x = condition, y = log2fc, color = cycle)) + 
#     geom_hline(yintercept = 0, linetype = "dashed") +
#     geom_pointrange(aes(ymin=log2fc - log2fc.sd, ymax=log2fc + log2fc.sd),
#                     position=position_dodge(width = 0.7),
#                     size = 1.5) +
#     guides(fill = guide_legend(override.aes = list(shape = 16)), 
#            color = guide_legend(override.aes = list(shape = 16))) +
#     labs(x = "Consequence", y = "Log2(observed/expected)\nVariants Per Cell") +
#     facet_wrap(.~variant_type, scales = "free", labeller = labeller(variant_type = label_wrap_gen(30))) + 
#     scale_color_viridis(discrete = TRUE, end = 0.8) +
#     theme_Publication() +
#     theme(legend.justification = c("center","top"),
#           legend.position = "top",
#           axis.text.x = element_text(size = 12))
# dev.off()
```

### Introns
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/introns")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/introns")

# plot
pdf("consequences.main.mean.intron.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(consequences.main.mean, variant_type == "intron"), aes(x = condition, y = log2fc, color = cycle)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=log2fc - log2fc.sd, ymax=log2fc + log2fc.sd),
                    position=position_dodge(width = 0.7),
                    fatten = 12) +
    guides(fill = guide_legend(override.aes = list(shape = 16)), 
           color = guide_legend(override.aes = list(shape = 16))) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU"))
    #scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1))
dev.off()
```

### Synonymous and Non-Synonymous

#### Unnormalized
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/synonymous_nonsynonymous")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/synonymous_nonsynonymous")

# set levels
consequences.syn_non.mean$variant_type <- factor(consequences.syn_non.mean$variant_type,
                                                 levels = c("synonymous", "nonsynonymous"))

consequences.syn_non.mean$cycle <- factor(consequences.syn_non.mean$cycle, 
                                       levels = c(1, 3, 6, 9))
consequences.syn_non.mean$condition <- factor(consequences.syn_non.mean$condition, 
                                           levels = c("control", "enu"))

# new facet labels
labels <- c("Synonymous",
             "Non-Synonymous")
names(labels) <- c("synonymous", "nonsynonymous")

# plot
pdf("consequences.syn_non.mean.pdf",
    width = 5,
    height = 4)
ggplot(consequences.syn_non.mean, aes(x = condition, y = log2fc, color = cycle)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=log2fc - log2fc.sd, ymax=log2fc + log2fc.sd),
                    position=position_dodge(width = 0.7),
                    fatten = 8) +
    guides(fill = guide_legend(override.aes = list(shape = 16)), 
           color = guide_legend(override.aes = list(shape = 16))) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~variant_type,
               scale = "fixed",
               nrow = 1,
               labeller = labeller(variant_type = labels))
    #scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1))
dev.off()
```

#### Intron Normalized
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/synonymous_nonsynonymous")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/synonymous_nonsynonymous")

# normalize by intron
consequences.syn_non.mean.norm <- consequences.syn_non.mean %>%
  merge(consequences.main.mean[consequences.main.mean$variant_type == "intron", c("condition", "cycle", "log2fc")], 
        by = c("condition", "cycle")) %>%
  mutate(`log2fc.norm` = `log2fc.x` - `log2fc.y`)

# set levels
consequences.syn_non.mean.norm$variant_type <- factor(consequences.syn_non.mean.norm$variant_type,
                                                 levels = c("synonymous", "nonsynonymous"))

consequences.syn_non.mean.norm$cycle <- factor(consequences.syn_non.mean.norm$cycle, 
                                       levels = c(1, 3, 6, 9))

consequences.syn_non.mean.norm$condition <- factor(consequences.syn_non.mean.norm$condition, 
                                           levels = c("control", "enu"))

# facet labels
labels <- c("Synonymous",
             "Non-Synonymous")
names(labels) <- c("synonymous", "nonsynonymous")

# plot
pdf("consequences.syn_non.mean.norm.pdf",
    width = 5,
    height = 4)
ggplot(consequences.syn_non.mean.norm, aes(x = condition, y = log2fc.norm, color = cycle)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=log2fc.norm - log2fc.sd, ymax=log2fc.norm + log2fc.sd),
                    position=position_dodge(width = 0.7),
                    fatten = 5) +
    guides(fill = guide_legend(override.aes = list(shape = 16)), 
           color = guide_legend(override.aes = list(shape = 16))) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~variant_type,
               scale = "fixed",
               nrow = 1,
               labeller = labeller(variant_type = labels))
    #scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1))
dev.off()
```

#### N/S ratio
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/synonymous_nonsynonymous")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/synonymous_nonsynonymous")

# calculate the ratio and standard deviation
syn_non.mean.ratio <- consequences.syn_non.mean.norm %>%
  group_by(cycle, condition) %>%
  summarise(
    nonsyn_log2fc = log2fc.norm[variant_type == "nonsynonymous"],
    syn_log2fc = log2fc.norm[variant_type == "synonymous"],
    nonsyn_sd = log2fc.sd[variant_type == "nonsynonymous"],
    syn_sd = log2fc.sd[variant_type == "synonymous"],
    log2fc.norm.ns.ratio = nonsyn_log2fc - syn_log2fc,
    log2fc.norm.ns.ratio.sd = sqrt(nonsyn_sd^2 + syn_sd^2)
  )

# plot
pdf("consequences.syn_non.mean.norm.ratio.pdf",
    width = 3,
    height = 4)
ggplot(syn_non.mean.ratio, aes(x = condition, y = log2fc.norm.ns.ratio, color = cycle)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=log2fc.norm.ns.ratio - log2fc.norm.ns.ratio.sd, ymax=log2fc.norm.ns.ratio + log2fc.norm.ns.ratio.sd),
                    position=position_dodge(width = 0.7),
                    fatten = 5) +
    guides(fill = guide_legend(override.aes = list(shape = 16)), 
           color = guide_legend(override.aes = list(shape = 16))) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Non-Syn. / Syn."))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU"))
    #scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1))
dev.off()
```

#### Specific SNV coding variants
- Normalized by introns
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/snv_coding")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/snv_coding")

# intron normalization
consequences.main.mean.coding.norm <- consequences.main.mean %>%
  filter(variant_type %in% c("synonymous variant", 
                             "stop retained variant", 
                             "missense variant", 
                             "start lost", 
                             "stop gained",
                             "stop lost")) %>%
  merge(consequences.main.mean[consequences.main.mean$variant_type == "intron", c("condition", "cycle", "log2fc")], 
        by = c("condition", "cycle")) %>%
  mutate(`log2fc.norm` = `log2fc.x` - `log2fc.y`)

# set levels
consequences.main.mean.coding.norm$variant_type <- factor(consequences.main.mean.coding.norm$variant_type,
                                                        levels = c("synonymous variant", 
                                                                   "stop retained variant", 
                                                                   "missense variant", 
                                                                   "start lost", 
                                                                   "stop gained",
                                                                   "stop lost"))

consequences.main.mean.coding.norm$cycle <- factor(consequences.main.mean.coding.norm$cycle, 
                                       levels = c(1, 3, 6, 9))

consequences.main.mean.coding.norm$condition <- factor(consequences.main.mean.coding.norm$condition, 
                                           levels = c("control", "enu"))

# new facet labels
labels <- c("Synonymous",
            "Stop retained",
            "Missense",
            "Start lost",
            "Stop gained",
            "Stop lost")

names(labels) <- levels(consequences.main.mean.coding.norm$variant_type)

# plot
pdf("consequences.main.mean.coding.norm.pdf",
    width = 10,
    height = 4)
ggplot(consequences.main.mean.coding.norm, aes(x = condition, y = log2fc.norm, color = cycle)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=log2fc.norm - log2fc.sd, ymax=log2fc.norm + log2fc.sd),
                    position=position_dodge(width = 0.7),
                    fatten = 5) +
    guides(fill = guide_legend(override.aes = list(shape = 16)), 
           color = guide_legend(override.aes = list(shape = 16))) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~variant_type,
               scale = "fixed",
               nrow = 1,
               labeller = labeller(variant_type = labels))
    #scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1))
dev.off()
```

#### Non-coding variants
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/snv_noncoding")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/snv_noncoding")

# subset 
consequences_levels <- c("intron",
                         "intergenic variant",
                         "upstream gene variant",
                         "downstream gene variant",
                         "3 prime UTR variant",
                         "non coding transcript exon variant",
                         "5 prime UTR variant",
                         "splice",
                         "TF binding site variant",
                         "mature miRNA variant")
consequences.main.mean.snv.noncode <- subset(consequences.main.mean, variant_type %in% consequences_levels)

# set levels
consequences.main.mean.snv.noncode$variant_type <- factor(consequences.main.mean.snv.noncode$variant_type, 
                                                          levels = consequences_levels)

consequences.main.mean.snv.noncode$cycle <- factor(consequences.main.mean.snv.noncode$cycle, 
                                       levels = c(1, 3, 6, 9))

consequences.main.mean.snv.noncode$condition <- factor(consequences.main.mean.snv.noncode$condition, 
                                           levels = c("control", "enu"))

# facet labels
labels <- c("Intron",
            "Intergenic",
           "5KB Upstream",
           "5KB Downstream",
           "3' UTR",
           "lncRNA",
           "5' UTR",
           "Splice Region",
           "TF Binding Site",
           "miRNA")

names(labels) <- consequences_levels

# plot
pdf("consequences_cycle_main_mean_snv_noncode.pdf",
    width = 10,
    height = 5.5)
ggplot(consequences.main.mean.snv.noncode, aes(x = condition, y = log2fc, color = cycle)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=log2fc - log2fc.sd, ymax=log2fc + log2fc.sd),
                    position=position_dodge(width = 0.7),
                    fatten = 5) +
    guides(fill = guide_legend(override.aes = list(shape = 16)), 
           color = guide_legend(override.aes = list(shape = 16))) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~variant_type,
               scale = "free",
               nrow = 2,
               labeller = labeller(variant_type = wrap_labels(labels, width = 10)))
    #scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1))
dev.off()
```

#### INDEL Variants
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/indel_coding")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cells-combined/consequences/indel_coding")

# intron normalization
consequences.main.mean.indel.norm <- consequences.main.mean %>%
  filter(variant_type %in% c("inframe insertion",
                             "inframe deletion",
                             "frameshift variant")) %>%
  merge(consequences.main.mean[consequences.main.mean$variant_type == "intron", c("condition", "cycle", "log2fc")], 
        by = c("condition", "cycle")) %>%
  mutate(`log2fc.norm` = `log2fc.x` - `log2fc.y`)

# set levels
consequences.main.mean.indel.norm$variant_type <- factor(consequences.main.mean.indel.norm$variant_type,
                                                        levels = c("inframe insertion",
                                                                   "inframe deletion",
                                                                   "frameshift variant"))

consequences.main.mean.indel.norm$cycle <- factor(consequences.main.mean.indel.norm$cycle, 
                                       levels = c(1, 3, 6, 9))

consequences.main.mean.indel.norm$condition <- factor(consequences.main.mean.indel.norm$condition, 
                                           levels = c("control", "enu"))

labels <- c("Inframe\ninsertion",
            "Inframe\ndeletion",
            "Frameshift")
names(labels) <- levels(consequences.main.mean.indel.norm$variant_type)

# plot
pdf("consequences_cycle_main_indel_code.pdf",
    width = 5.5,
    height = 4)
ggplot(consequences.main.mean.indel.norm, aes(x = condition, y = log2fc.norm, color = cycle)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=log2fc.norm - log2fc.sd, ymax=log2fc.norm + log2fc.sd),
                    position=position_dodge(width = 0.7),
                    fatten = 5) +
    guides(fill = guide_legend(override.aes = list(shape = 16)), 
           color = guide_legend(override.aes = list(shape = 16))) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~variant_type,
               scale = "fixed",
               nrow = 1,
               labeller = labeller(variant_type = labels))
    #scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1))
dev.off()
```


# CADD
- Only SNVs

## Raw score

### Number of different types of variants
```{r}
print(cadd.raw.mean %>%
  group_by(type) %>%
  summarize(total = sum(obs_mut)),
  n = Inf)

print(cadd.raw.mean %>%
  group_by(bin) %>%
  summarize(total = sum(obs_mut)),
  n = Inf)
```

### By cycle

#### Synonymous
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/cycle")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/cycle")

# subset
cadd.raw.mean.coding.synonymous <- subset(cadd.raw.mean, grepl("snv coding synonymous", type))

# remove NA
cadd.raw.mean.coding.synonymous <- cadd.raw.mean.coding.synonymous[!is.na(cadd.raw.mean.coding.synonymous$bin),]

# set levels
bin <- c("[-4,-3.5]", "(-3.5,-3)", "(-3,-2.5]", "(-2.5,-2]", "(-2,-1.5]", "(-1.5,-1]",
  "(-1,-0.5]", "(-0.5,0]", "(0,0.5]", "(0.5,1]", "(1,1.5]", "(1.5,2]", "(2,2.5]",
  "(2.5,3]", "(3,3.5]", "(3.5,4]")
cadd.raw.mean.coding.synonymous$bin <- factor(cadd.raw.mean.coding.synonymous$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.raw.mean.coding.synonymous.combined.pvalue <- consequences_combined_significance_CADD(cadd.raw.mean.coding.synonymous,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.raw.mean.coding.synonymous.combined.pvalue, "cadd.raw.mean.coding.synonymous.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.raw.mean.coding.synonymous.cellreps.pvalue <- consequences_group_significance(cadd.raw.mean.coding.synonymous,
#                                                    c("condition", "cycle", "variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.raw.mean.coding.synonymous.cellreps.pvalue, "cadd.raw.mean.coding.synonymous.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.raw.mean.coding.synonymous %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.raw.mean.coding.synonymous.summary.csv")


# plot
pdf("cadd.raw.mean.coding.synonymous.synonymous.pdf",
    width = 8,
    height = 8)
ggplot(cadd.raw.mean.coding.synonymous, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

#### Non-Synonymous
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/cycle")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/cycle")

# subset
cadd.raw.mean.coding.nonsynonymous <- subset(cadd.raw.mean, grepl("snv coding nonsynonymous", type))

# remove NA
cadd.raw.mean.coding.nonsynonymous <- cadd.raw.mean.coding.nonsynonymous[!is.na(cadd.raw.mean.coding.nonsynonymous$bin),]

# set levels
bin <- c("[-4,-3.5]", "(-3.5,-3)", "(-3,-2.5]", "(-2.5,-2]", "(-2,-1.5]", "(-1.5,-1]",
  "(-1,-0.5]", "(-0.5,0]", "(0,0.5]", "(0.5,1]", "(1,1.5]", "(1.5,2]", "(2,2.5]",
  "(2.5,3]", "(3,3.5]", "(3.5,4]")
cadd.raw.mean.coding.nonsynonymous$bin <- factor(cadd.raw.mean.coding.nonsynonymous$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.raw.mean.coding.nonsynonymous.combined.pvalue <- consequences_combined_significance_CADD(cadd.raw.mean.coding.nonsynonymous,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.raw.mean.coding.nonsynonymous.combined.pvalue, "cadd.raw.mean.coding.nonsynonymous.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.raw.mean.coding.nonsynonymous.cellreps.pvalue <- consequences_group_significance(cadd.raw.mean.coding.nonsynonymous,
#                                                    c("condition", "cycle", "variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.raw.mean.coding.nonsynonymous.cellreps.pvalue, "cadd.raw.mean.coding.nonsynonymous.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.raw.mean.coding.nonsynonymous %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.raw.mean.coding.nonsynonymous.summary.csv")


# plot
pdf("cadd.raw.mean.coding.nonsynonymous.synonymous.pdf",
    width = 8,
    height = 8)
ggplot(cadd.raw.mean.coding.nonsynonymous, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

## PHRED score
- binned mutation frequency has been noramlized for total amount of mutations falling in all bins 
-  a scaled C-score of greater of equal 10 indicates that these are predicted to be the 10% most deleterious substitutions that you can do to the human genome, a score of greater or equal 20 indicates the 1% most deleterious and so on.
- Identify which bins have a sufficient number of SNVs to be analyzed and then look at those
- Is there a score at which we start to see negative and posiive selection? From non synonymous this looks like 12-21

### Number of different types of variants - modes
- nonsynonymous = 21-35
- splice = 21-35
- 5 prime UTR = 12-21
- synonymous = 7-12
- 3prim UTR = 7-12
- nmd = 4-7
- upstream = 4-7
- downstream = 4-7
- noncoding transcript exon = 4-7
- intron = 2-4
- regulatory = 2-4
- noncoding transcript intron = 2-4
- intergenic = 0-1

- knowing the median cadd, compare O/E of all variants of each type where ones with higher means are expected to have more negative selection
- for each type, get bin (or top 2) with most variants, and then compare O/E for that amongst types
- within each, divide into 10 log10 bins
- within each, divide into 3 bins, high medium and low

Note that this does not sum up to the total amount of observed variants, meaning that some were left un annotated
```{r}
print(cadd.phred.mean %>%
  group_by(variant_type) %>%
  summarize(total = sum(obs.mut, na.rm = TRUE)),
  n = Inf)

temp <- cadd.phred.mean %>%
  group_by(variant_type) %>%
  summarize(total = sum(obs.mut, na.rm = TRUE))
print(cadd.phred.bin.mean %>%
  group_by(variant_type, bin) %>%
  summarize(total = sum(obs.mut, na.rm = TRUE)),
  n = Inf)
```

### By cycle

#### Without binning

##### O/E ratio of frequency
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/")
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/")
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")

# add sample info
cadd.phred.mean <- merge(cadd.phred.mean,
                                    sample_table,
                                    by = "sample")

# remove undefined
# remove intergenic because this is probably an underestimate
cadd.phred.mean <- subset(cadd.phred.mean, !variant_type %in% c("undefined", "snv noncoding intergenic"))

# set levels
levels <- c("snv coding synonymous",
            "snv coding nonsynonymous",
            "snv noncoding intron",
            "snv noncoding transcript intron",
            "snv noncoding upstream gene",
            "snv noncoding downstream gene",
            "snv noncoding regulatory",
            "snv noncoding nmd",
            "snv noncoding 3prime UTR",
            "snv noncoding transcript exon",
            "snv noncoding splice",
            "snv noncoding 5prime UTR")

cadd.phred.mean$variant_type <- factor(cadd.phred.mean$variant_type,
                                       levels = levels)

# combine p-values from each cell into group pvalue
cadd.phred.mean.pvalue <- consequences_combined_significance(cadd.phred.mean,
                                                   c("condition", "cycle", "variant_type"),
                                                   levels)
write.csv(cadd.phred.mean.pvalue, "cadd.phred.mean.pvalue.csv")

stats_summary <- cadd.phred.mean %>%
  group_by(condition, cycle, variant_type) %>%
  summarise(log2fc.mean = mean(log2fc, na.rm = TRUE),
            log2fc.sd = sd(log2fc, na.rm = TRUE),
            obs.mut.mean = mean(obs.mut, na.rm = TRUE),
            obs.mut.sd = sd(obs.mut, na.rm = TRUE),
            obs.sim.diff.mean = mean(obs.mut - sim.mut, na.rm = TRUE),
            obs.sim.diff.sd = sd(obs.mut - sim.mut, na.rm = TRUE),
            obs.prop.total.mean = mean(obs.mut / SNV_obs, na.rm = TRUE),
            obs.prop.total.mean = sd(obs.mut / SNV_obs, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.mean.summary.csv")

labels = c("Synonymous",
           "Non-Synonymous",
           "Intron",
           "Non-coding RNA Intron",
           "5KB Upstream Gene",
           "5KB Downstream Gene",
           "Non-coding Regulatory",
           "Nonsense Mediated Decay",
           "3' UTR",
           "Non-coding RNA Exon",
           "Splice Site",
           "5' UTR")

names(labels) <- levels

# plot all
pdf("cadd.phred.mean.pdf",
    width = 10,
    height = 5)
ggplot(cadd.phred.mean, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Median CADD Score Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 10.5)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~variant_type,
                 scale = "fixed",
                 nrow = 2,
                 labeller = labeller(variant_type = wrap_labels(labels, width = 15)))
      #scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1))
dev.off()

# coding
pdf("cadd.phred.mean.coding.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(cadd.phred.mean, grepl("synonymous", variant_type)), aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Median CADD Score Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 10.5)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~variant_type,
                 scale = "fixed",
                 nrow = 1,
                 labeller = labeller(variant_type = wrap_labels(labels, width = 15))) +
      scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1))
dev.off()

# noncoding
pdf("cadd.phred.mean.noncoding.pdf",
    width = 8.5,
    height = 5)
ggplot(subset(cadd.phred.mean, !grepl("synonymous", variant_type)), aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Median CADD Score Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 10.5)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~variant_type,
                 scale = "fixed",
                 nrow = 2,
                 labeller = labeller(variant_type = wrap_labels(labels, width = 15))) +
      scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1))
dev.off()
```

##### Observed median CADD score
- using median because low amount of observed variants in some cells
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")

# add sample info
cadd.phred.mean.score <- merge(cadd.phred.mean.score,
                                    sample_table,
                                    by = "sample")


cadd.phred.median.score <- cadd.phred.mean.score %>%
  select(c("sample", "type", "median")) %>%
  mutate(group = case_when(grepl("snv coding", type) ~ "coding",
         .default = "noncoding")
  ) %>%
  subset(type != "undefined")

# set levels
cadd.phred.median.score$type <- factor(cadd.phred.median.score$type,
                                                 levels = c("snv coding nonsynonymous",
                                                            "snv coding synonymous",
                                                            "snv noncoding 5prime UTR",
                                                            "snv noncoding 3prime UTR",
                                                            "snv noncoding intron",
                                                            "snv noncoding nmd",
                                                            "snv noncoding splice",
                                                            "snv noncoding upstream gene",
                                                            "snv noncoding downstream gene",
                                                            "snv noncoding transcript exon",
                                                            "snv noncoding transcript intron",
                                                            "snv noncoding regulatory",
                                                            "snv noncoding intergenic"
                                                            ))

# Perform ANOVA
# anova_result <- summary(aov(median ~ type, data = cadd.phred.median.score))
# 
# # Perform linear model
# lm_result <- lm(median ~ type, data = cadd.phred.median.score)
# 
# # Compute emmeans and pairwise comparisons
# emm_interaction <- emmeans(lm_result, ~ median)
# pairwise_results <- pairs(emm_interaction)
# 
# # Adjust for multiple comparisons and add significance
# adjusted_results <- na.omit(summary(pairwise_results, adjust = "tukey")) %>%
#                     add_significance("p.value")
# 
# # Write the results to a CSV file
# write.csv(adjusted_results, "cadd.phred.median.score.pvalue.csv")
# anova_summary_text <- write(capture.output(anova_result), 
#                             file = output_filename, 
#                             append = TRUE, 
#                             sep = "\n")

# get mean and sd for each group
stats_summary <- cadd.phred.median.score %>%
  group_by(type) %>%
  summarise(mean = mean(median, na.rm = TRUE),
            sd = sd(median, na.rm = TRUE)) %>%
  arrange(desc(mean))

# sort from least to greatest
cadd.phred.median.score$type <- factor(cadd.phred.median.score$type,
                                       levels = stats_summary$type)

labels = c("Non-Synonymous",
           "Splice Site",
           "5' UTR",
           "Synonymous",
           "3' UTR",
           "Non-coding RNA Exon",
           "Non-coding Regulatory",
           "5KB Upstream Gene",
           "Intron",
           "Nonsesnse Mediated Decay",
           "5KB Downstream Gene",
           "Non-coding RNA Intron",
           "Intergenic")
# bar plot
pdf("cadd.phred.mean.score.pdf",
    width = 7,
    height = 4)
ggplot(cadd.phred.median.score, aes(x = type, y = median, fill = type)) + 
    geom_hline(yintercept = 15, linetype = "dashed") +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +  # Average bar plot
    geom_errorbar(data = stats_summary, aes(x = type, y = mean, ymin = mean - sd, ymax = mean + sd),
                    width = 0.5, position = position_dodge(0.9)) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", y = "Median CADD Score Per Cell", fill = "Variant Type/Location") +
    scale_fill_viridis(option = "H",
                       discrete = TRUE,
                       labels = labels,
                       begin = 0.1,
                       end = 0.9) +
    theme_Publication() +
    theme(legend.position = "right",  # Place legend on the right side
        legend.direction = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
dev.off()
```

#### With binning
- This is now using the frequency of variants per bin normalized by the total amount of variants per cell

##### All variant types
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/")
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")

# add sample info
cadd.phred.bin.mean <- merge(cadd.phred.bin.mean,
                                    sample_table,
                                    by = "sample")

# remove NA
cadd.phred.bin.mean <- cadd.phred.bin.mean[!is.na(cadd.phred.bin.mean$bin),]

# check which bin has most
print(cadd.phred.bin.mean %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut, na.rm = TRUE)),
  n = Inf)


# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean$bin <- factor(cadd.phred.bin.mean$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean,
                                                   c("condition", "cycle", "variant_type", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.combined.pvalue, "cadd.phred.bin.mean.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean,
#                                                    c("condition", "cycle", "variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.cellreps.pvalue, "cadd.phred.bin.mean.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.summary.csv")


# plot all bins
pdf("cadd.phred.bin.mean.pdf",
    width = 10,
    height = 10)
ggplot(cadd.phred.bin.mean, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Binned CADD Score Frequency"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
  facet_wrap(.~bin,
                 scale = "fixed")
dev.off()
```

##### Coding Combined
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/")
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")

# add sample info
cadd.phred.coding.combined.bin.mean <- merge(cadd.phred.coding.combined.bin.mean,
                                    sample_table,
                                    by = "sample")

# subset
cadd.phred.mean.coding <- subset(cadd.phred.coding.combined.bin.mean, grepl("snv coding", variant_type))

# remove NA
cadd.phred.mean.coding <- cadd.phred.mean.coding[!is.na(cadd.phred.mean.coding$bin),]

# check which bin has most
print(cadd.phred.mean.coding %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)
 

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.mean.coding$bin <- factor(cadd.phred.mean.coding$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.mean.coding.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.mean.coding,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.mean.coding.combined.pvalue, "cadd.phred.mean.coding.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.mean.coding.cellreps.pvalue <- consequences_group_significance(cadd.phred.mean.coding,
#                                                    c("condition", "cycle", "variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.mean.coding.cellreps.pvalue, "cadd.phred.mean.coding.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.mean.coding %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.mean.coding.summary.csv")


# plot all bins
pdf("cadd.phred.mean.coding.pdf",
    width = 9,
    height = 6)
ggplot(cadd.phred.mean.coding, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Binned CADD Score Frequency"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 2)
dev.off()

# plot subset bins
# nonsynonymous ~25
# synonymous ~10
cadd.phred.mean.coding.sub <- subset(cadd.phred.mean.coding, bin %in% c("(7.71,12.8]", "(21.4,35.7]"))

# combine p-values from each cell into group pvalue
cadd.phred.mean.coding.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.mean.coding.sub,
                                                   c("condition", "cycle", "bin"))

write.csv(cadd.phred.mean.coding.combined.pvalue, "cadd.phred.mean.coding.subset.bins.pvalue.csv")

# new facet labels
labels <- c("Non-synonymous\n(21.4 - 35.7)",
            "Synonymous\n(7.7 - 12.8)")

names(labels) <- unique(cadd.phred.mean.coding.sub$bin)

pdf("cadd.phred.mean.coding.subset.bins.pdf",
    width = 4.5,
    height = 4)
ggplot(cadd.phred.mean.coding.sub, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    scale_y_continuous(breaks = c(-0.25, 0, 0.25), lim = c(-0.25, 0.25)) +
    facet_wrap(.~bin,
                 scale = "fixed",
               nrow = 1,
               labeller = labeller(bin = labels))
dev.off()
```

##### Synonymous
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/")
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")

# subset
cadd.phred.bin.mean.coding.synonymous <- subset(cadd.phred.bin.mean, grepl("snv coding synonymous", variant_type))

# remove NA
cadd.phred.bin.mean.coding.synonymous <- cadd.phred.bin.mean.coding.synonymous[!is.na(cadd.phred.bin.mean.coding.synonymous$bin),]

# check which bin has most
print(cadd.phred.bin.mean.coding.synonymous %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)


# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.coding.synonymous$bin <- factor(cadd.phred.bin.mean.coding.synonymous$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.coding.synonymous.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.coding.synonymous,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.coding.synonymous.combined.pvalue, "cadd.phred.bin.mean.coding.synonymous.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.coding.synonymous.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.coding.synonymous,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.coding.synonymous.cellreps.pvalue, "cadd.phred.bin.mean.coding.synonymous.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.coding.synonymous %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.coding.synonymous.summary.csv")


# plot
pdf("cadd.phred.bin.mean.coding.synonymous.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.coding.synonymous, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-Synonymous
- need to normalize by intron
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.coding.nonsynonymous <- subset(cadd.phred.bin.mean, grepl("snv coding nonsynonymous", variant_type))

# remove NA
cadd.phred.bin.mean.coding.nonsynonymous <- cadd.phred.bin.mean.coding.nonsynonymous[!is.na(cadd.phred.bin.mean.coding.nonsynonymous$bin),]

# check which bin has most
print(cadd.phred.bin.mean.coding.nonsynonymous %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.coding.nonsynonymous$bin <- factor(cadd.phred.bin.mean.coding.nonsynonymous$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.coding.nonsynonymous.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.coding.nonsynonymous,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.coding.nonsynonymous.combined.pvalue, "cadd.phred.bin.mean.coding.nonsynonymous.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.coding.nonsynonymous.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.coding.nonsynonymous,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.coding.nonsynonymous.cellreps.pvalue, "cadd.phred.bin.mean.coding.nonsynonymous.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.coding.nonsynonymous %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.coding.nonsynonymous.summary.csv")


# plot
pdf("cadd.phred.bin.mean.coding.nonsynonymous.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.coding.nonsynonymous, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-coding 3' UTR
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.noncoding.3primeUTR <- subset(cadd.phred.bin.mean, grepl("snv noncoding 3prime UTR", variant_type))

# remove NA
cadd.phred.bin.mean.noncoding.3primeUTR <- cadd.phred.bin.mean.noncoding.3primeUTR[!is.na(cadd.phred.bin.mean.noncoding.3primeUTR$bin),]

# check which bin has most
print(cadd.phred.bin.mean.noncoding.3primeUTR %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.noncoding.3primeUTR$bin <- factor(cadd.phred.bin.mean.noncoding.3primeUTR$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.noncoding.3primeUTR.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.noncoding.3primeUTR,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.noncoding.3primeUTR.combined.pvalue, "cadd.phred.bin.mean.noncoding.3primeUTR.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.noncoding.3primeUTR.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.noncoding.3primeUTR,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.noncoding.3primeUTR.cellreps.pvalue, "cadd.phred.bin.mean.noncoding.3primeUTR.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.noncoding.3primeUTR %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.noncoding.3primeUTR.summary.csv")


# plot
pdf("cadd.phred.bin.mean.noncoding.3primeUTR.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.noncoding.3primeUTR, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-coding 5' UTR
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.noncoding.5primeUTR <- subset(cadd.phred.bin.mean, grepl("snv noncoding 5prime UTR", variant_type))

# remove NA
cadd.phred.bin.mean.noncoding.5primeUTR <- cadd.phred.bin.mean.noncoding.5primeUTR[!is.na(cadd.phred.bin.mean.noncoding.5primeUTR$bin),]

# check which bin has most
print(cadd.phred.bin.mean.noncoding.5primeUTR %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.noncoding.5primeUTR$bin <- factor(cadd.phred.bin.mean.noncoding.5primeUTR$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.noncoding.5primeUTR.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.noncoding.5primeUTR,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.noncoding.5primeUTR.combined.pvalue, "cadd.phred.bin.mean.noncoding.5primeUTR.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.noncoding.5primeUTR.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.noncoding.5primeUTR,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.noncoding.5primeUTR.cellreps.pvalue, "cadd.phred.bin.mean.noncoding.5primeUTR.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.noncoding.5primeUTR %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.noncoding.5primeUTR.summary.csv")


# plot
pdf("cadd.phred.bin.mean.noncoding.5primeUTR.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.noncoding.5primeUTR, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-coding upstream
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.noncoding.upstream <- subset(cadd.phred.bin.mean, grepl("snv noncoding upstream", variant_type))

# remove NA
cadd.phred.bin.mean.noncoding.upstream <- cadd.phred.bin.mean.noncoding.upstream[!is.na(cadd.phred.bin.mean.noncoding.upstream$bin),]

# check which bin has most
print(cadd.phred.bin.mean.noncoding.upstream %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.noncoding.upstream$bin <- factor(cadd.phred.bin.mean.noncoding.upstream$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.noncoding.upstream.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.noncoding.upstream,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.noncoding.upstream.combined.pvalue, "cadd.phred.bin.mean.noncoding.upstream.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.noncoding.upstream.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.noncoding.upstream,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.noncoding.upstream.cellreps.pvalue, "cadd.phred.bin.mean.noncoding.upstream.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.noncoding.upstream %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.noncoding.upstream.summary.csv")


# plot
pdf("cadd.phred.bin.mean.noncoding.upstream.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.noncoding.upstream, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-coding downstream
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.noncoding.downstream <- subset(cadd.phred.bin.mean, grepl("snv noncoding downstream", variant_type))

# remove NA
cadd.phred.bin.mean.noncoding.downstream <- cadd.phred.bin.mean.noncoding.downstream[!is.na(cadd.phred.bin.mean.noncoding.downstream$bin),]

# check which bin has most
print(cadd.phred.bin.mean.noncoding.downstream %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.noncoding.downstream$bin <- factor(cadd.phred.bin.mean.noncoding.downstream$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.noncoding.downstream.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.noncoding.downstream,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.noncoding.downstream.combined.pvalue, "cadd.phred.bin.mean.noncoding.downstream.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.noncoding.downstream.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.noncoding.downstream,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.noncoding.downstream.cellreps.pvalue, "cadd.phred.bin.mean.noncoding.downstream.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.noncoding.downstream %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.noncoding.downstream.summary.csv")


# plot
pdf("cadd.phred.bin.mean.noncoding.downstream.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.noncoding.downstream, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-coding intron
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.noncoding.intron <- subset(cadd.phred.bin.mean, grepl("snv noncoding intron", variant_type))

# remove NA
cadd.phred.bin.mean.noncoding.intron <- cadd.phred.bin.mean.noncoding.intron[!is.na(cadd.phred.bin.mean.noncoding.intron$bin),]

# check which bin has most
print(cadd.phred.bin.mean.noncoding.intron %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.noncoding.intron$bin <- factor(cadd.phred.bin.mean.noncoding.intron$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.noncoding.intron.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.noncoding.intron,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.noncoding.intron.combined.pvalue, "cadd.phred.bin.mean.noncoding.intron.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.noncoding.intron.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.noncoding.intron,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.noncoding.intron.cellreps.pvalue, "cadd.phred.bin.mean.noncoding.intron.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.noncoding.intron %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.noncoding.intron.summary.csv")


# plot
pdf("cadd.phred.bin.mean.noncoding.intron.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.noncoding.intron, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-coding NMD
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.noncoding.nmd <- subset(cadd.phred.bin.mean, grepl("snv noncoding nmd", variant_type))

# remove NA
cadd.phred.bin.mean.noncoding.nmd <- cadd.phred.bin.mean.noncoding.nmd[!is.na(cadd.phred.bin.mean.noncoding.nmd$bin),]

# check which bin has most
print(cadd.phred.bin.mean.noncoding.nmd %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.noncoding.nmd$bin <- factor(cadd.phred.bin.mean.noncoding.nmd$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.noncoding.nmd.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.noncoding.nmd,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.noncoding.nmd.combined.pvalue, "cadd.phred.bin.mean.noncoding.nmd.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.noncoding.nmd.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.noncoding.nmd,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.noncoding.nmd.cellreps.pvalue, "cadd.phred.bin.mean.noncoding.nmd.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.noncoding.nmd %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.noncoding.nmd.summary.csv")


# plot
pdf("cadd.phred.bin.mean.noncoding.nmd.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.noncoding.nmd, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-coding intergenic
- all negative values, this doesn't make sense
- Are these variants really in intergenic regions?
- Could these be used to normalize non-coding variants to account for?
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.noncoding.intergenic <- subset(cadd.phred.bin.mean, grepl("snv noncoding intergenic", variant_type))

# remove NA
cadd.phred.bin.mean.noncoding.intergenic <- cadd.phred.bin.mean.noncoding.intergenic[!is.na(cadd.phred.bin.mean.noncoding.intergenic$bin),]

# check which bin has most
print(cadd.phred.bin.mean.noncoding.intergenic %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.noncoding.intergenic$bin <- factor(cadd.phred.bin.mean.noncoding.intergenic$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.noncoding.intergenic.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.noncoding.intergenic,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.noncoding.intergenic.combined.pvalue, "cadd.phred.bin.mean.noncoding.intergenic.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.noncoding.intergenic.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.noncoding.intergenic,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.noncoding.intergenic.cellreps.pvalue, "cadd.phred.bin.mean.noncoding.intergenic.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.noncoding.intergenic %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.noncoding.intergenic.summary.csv")


# plot
pdf("cadd.phred.bin.mean.noncoding.intergenic.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.noncoding.intergenic, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-coding regulatory
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.noncoding.regulatory <- subset(cadd.phred.bin.mean, grepl("snv noncoding regulatory", variant_type))

# remove NA
cadd.phred.bin.mean.noncoding.regulatory <- cadd.phred.bin.mean.noncoding.regulatory[!is.na(cadd.phred.bin.mean.noncoding.regulatory$bin),]

# check which bin has most
print(cadd.phred.bin.mean.noncoding.regulatory %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.noncoding.regulatory$bin <- factor(cadd.phred.bin.mean.noncoding.regulatory$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.noncoding.regulatory.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.noncoding.regulatory,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.noncoding.regulatory.combined.pvalue, "cadd.phred.bin.mean.noncoding.regulatory.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.noncoding.regulatory.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.noncoding.regulatory,
#                                                    c("condition", "cycle", "variant_variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.noncoding.regulatory.cellreps.pvalue, "cadd.phred.bin.mean.noncoding.regulatory.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.noncoding.regulatory %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.noncoding.regulatory.summary.csv")


# plot
pdf("cadd.phred.bin.mean.noncoding.regulatory.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.noncoding.regulatory, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_variant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-coding splice
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.noncoding.splice <- subset(cadd.phred.bin.mean, grepl("snv noncoding splice", variant_type))

# remove NA
cadd.phred.bin.mean.noncoding.splice <- cadd.phred.bin.mean.noncoding.splice[!is.na(cadd.phred.bin.mean.noncoding.splice$bin),]

# check which bin has most
print(cadd.phred.bin.mean.noncoding.splice %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.noncoding.splice$bin <- factor(cadd.phred.bin.mean.noncoding.splice$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.noncoding.splice.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.noncoding.splice,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.noncoding.splice.combined.pvalue, "cadd.phred.bin.mean.noncoding.splice.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.noncoding.splice.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.noncoding.splice,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.noncoding.splice.cellreps.pvalue, "cadd.phred.bin.mean.noncoding.splice.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.noncoding.splice %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.noncoding.splice.summary.csv")


# plot
pdf("cadd.phred.bin.mean.noncoding.splice.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.noncoding.splice, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-coding transcript exon
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.noncoding.exon <- subset(cadd.phred.bin.mean, grepl("snv noncoding transcript exon", variant_type))

# remove NA
cadd.phred.bin.mean.noncoding.exon <- cadd.phred.bin.mean.noncoding.exon[!is.na(cadd.phred.bin.mean.noncoding.exon$bin),]

# check which bin has most
print(cadd.phred.bin.mean.noncoding.exon %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.noncoding.exon$bin <- factor(cadd.phred.bin.mean.noncoding.exon$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.noncoding.exon.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.noncoding.exon,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.noncoding.exon.combined.pvalue, "cadd.phred.bin.mean.noncoding.exon.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.noncoding.exon.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.noncoding.exon,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.noncoding.exon.cellreps.pvalue, "cadd.phred.bin.mean.noncoding.exon.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.noncoding.exon %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.noncoding.exon.summary.csv")


# plot
pdf("cadd.phred.bin.mean.noncoding.exon.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.noncoding.exon, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

##### Non-coding transcript intron
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/cadd/phred/cycle/")
# subset
cadd.phred.bin.mean.noncoding.transcript.intron <- subset(cadd.phred.bin.mean, grepl("snv noncoding transcript intron", variant_type))

# remove NA
cadd.phred.bin.mean.noncoding.transcript.intron <- cadd.phred.bin.mean.noncoding.transcript.intron[!is.na(cadd.phred.bin.mean.noncoding.transcript.intron$bin),]

# check which bin has most
print(cadd.phred.bin.mean.noncoding.transcript.intron %>%
  group_by(bin) %>%
  summarize(total = sum(obs.mut)),
  n = Inf)

# set levels
bin <- c("[0,1]", "(1,1.67]", "(1.67,2.78]", "(2.78,4.63]", "(4.63,7.71]", "(7.71,12.8]", "(12.8,21.4]", "(21.4,35.7]", "(35.7,59.4]", "(59.4,99]")

cadd.phred.bin.mean.noncoding.transcript.intron$bin <- factor(cadd.phred.bin.mean.noncoding.transcript.intron$bin,
                            levels = bin)

# combine p-values from each cell into group pvalue
cadd.phred.bin.mean.noncoding.transcript.intron.combined.pvalue <- consequences_combined_significance_CADD(cadd.phred.bin.mean.noncoding.transcript.intron,
                                                   c("condition", "cycle", "bin"),
                                                   bin)
write.csv(cadd.phred.bin.mean.noncoding.transcript.intron.combined.pvalue, "cadd.phred.bin.mean.noncoding.transcript.intron.combined.pvalue.csv")

# calculate group pvalue with all cells as replicates
# cadd.phred.bin.mean.noncoding.transcript.intron.cellreps.pvalue <- consequences_group_significance(cadd.phred.bin.mean.noncoding.transcript.intron,
#                                                    c("condition", "cycle", "variant_variant_type"),
#                                                    consequences_levels)
# write.csv(cadd.phred.bin.mean.noncoding.transcript.intron.cellreps.pvalue, "cadd.phred.bin.mean.noncoding.transcript.intron.cellreps.pvalue.csv")

# get mean and sd for each group
stats_summary <- cadd.phred.bin.mean.noncoding.transcript.intron %>%
  group_by(condition, cycle, bin) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE))

write.csv(stats_summary, "cadd.phred.bin.mean.noncoding.transcript.intron.summary.csv")


# plot
pdf("cadd.phred.bin.mean.noncoding.transcript.intron.pdf",
    width = 8,
    height = 8)
ggplot(cadd.phred.bin.mean.noncoding.transcript.intron, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Frequency Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~bin,
                 scale = "fixed",
                 nrow = 4)
dev.off()
```

## Alpha missense

### By cycle
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/alpha-missense")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/alpha-missense")

# mean
am.mean$cycle <- factor(am.mean$cycle)
pdf("am_cycle_mean.pdf",
    width = 5,
    height = 4)
ggplot(am.mean, aes(x = condition, y = log2fc, fill = cycle)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1)) +
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

# median
am.median$cycle <- factor(am.median$cycle)
pdf("am_cycle_median.pdf",
    width = 5,
    height = 4)
ggplot(am.median, aes(x = condition, y = log2fc, fill = cycle)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1)) +
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

am.sum$cycle <- factor(am.sum$cycle)
pdf("am_cycle_sum.pdf",
    width = 5,
    height = 4)
ggplot(am.sum, aes(x = condition, y = log2fc, fill = cycle)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1)) +
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()
```


## GERP

### By condition

#### Mean
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/gerp")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/gerp")

# gerp_levels <- unique(gerp.mean$variant_type)
# 
# # combine p-values from each cell into group pvalue
# gerp.mean.combined_pvalue <- consequences_combined_significance(gerp.mean,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.mean.combined_pvalue, "gerp_mean_combined_pvalue.csv")
# 
# # calculate group pvalue with all cells as replicates
# gerp.mean.cellreps_pvalue <- consequences_group_significance(gerp.mean,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.mean.cellreps_pvalue, "gerp_mean_cellreps_pvalue.csv")
# 
# # get mean and sd for each group
# stats_summary <- consequences.main.mean.snv.code %>%
#   group_by(condition, variant_variant_type) %>%
#   summarise(mean = mean(log2fc, na.rm = TRUE),
#             sd = sd(log2fc, na.rm = TRUE))
# 
# write.csv(stats_summary, "consequences_condition_main_mean_snv_code_stat_summary.csv")

# control
pdf("gerp_condition_mean_control.pdf",
    width = 5,
    height = 4)
ggplot(subset(gerp.mean, condition == "control"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n Mean GERP Score Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    # geom_text(data = subset(consequences.main.mean.snv.code.cellreps_pvalue, condition == "control"),
    #           aes(x = variant_type, y = Inf, label = significant_p_adjusted), # pvalue of observed/expected
    #           vjust = 4, hjust = 0.75) +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()

# enu
pdf("gerp_condition_mean_enu.pdf",
    width = 5,
    height = 4)
ggplot(subset(gerp.mean, condition == "enu"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n Mean GERP Score Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    # geom_text(data = subset(consequences.main.mean.snv.code.cellreps_pvalue, condition == "control"),
    #           aes(x = variant_type, y = Inf, label = significant_p_adjusted), # pvalue of observed/expected
    #           vjust = 4, hjust = 0.75) +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()
```

## Transcription factor

### By cycle
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/transcription-factor")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/transcription-factor")

# mean
tf.mean$cycle <- factor(tf.mean$cycle)
pdf("tf_cycle_mean.pdf",
    width = 5,
    height = 4)
ggplot(tf.mean, aes(x = condition, y = log2fc, fill = cycle)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1)) +
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

# median
tf.median$cycle <- factor(tf.median$cycle)
pdf("tf_cycle_median.pdf",
    width = 5,
    height = 4)
ggplot(tf.median, aes(x = condition, y = log2fc, fill = cycle)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1)) +
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()
```

#### Median
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/transcription-factor")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/transcription-factor")

# gerp_levels <- unique(gerp.mean$variant_type)
# 
# # combine p-values from each cell into group pvalue
# gerp.mean.combined_pvalue <- consequences_combined_significance(gerp.mean,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.mean.combined_pvalue, "gerp_mean_combined_pvalue.csv")
# 
# # calculate group pvalue with all cells as replicates
# gerp.mean.cellreps_pvalue <- consequences_group_significance(gerp.mean,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.mean.cellreps_pvalue, "gerp_mean_cellreps_pvalue.csv")
# 
# # get mean and sd for each group
# stats_summary <- consequences.main.mean.snv.code %>%
#   group_by(condition, variant_variant_type) %>%
#   summarise(mean = mean(log2fc, na.rm = TRUE),
#             sd = sd(log2fc, na.rm = TRUE))
# 
# write.csv(stats_summary, "consequences_condition_main_mean_snv_code_stat_summary.csv")

# control
pdf("tf_condition_median_control.pdf",
    width = 5,
    height = 4)
ggplot(subset(tf.median, condition == "control"), aes(x = condition, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n median Motif Score Change Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    # geom_text(data = subset(consequences.main.median.snv.code.cellreps_pvalue, condition == "control"),
    #           aes(x = variant_type, y = Inf, label = significant_p_adjusted), # pvalue of observed/expected
    #           vjust = 4, hjust = 0.75) +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()

# enu
pdf("tf_condition_median_enu.pdf",
    width = 5,
    height = 4)
ggplot(subset(tf.median, condition == "enu"), aes(x = condition, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n median Motif Score Change Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    # geom_text(data = subset(consequences.main.median.snv.code.cellreps_pvalue, condition == "control"),
    #           aes(x = variant_type, y = Inf, label = significant_p_adjusted), # pvalue of observed/expected
    #           vjust = 4, hjust = 0.75) +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()
```

#### Sum
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/transcription-factor")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/transcription-factor")

# gerp_levels <- unique(gerp.mean$variant_type)
# 
# # combine p-values from each cell into group pvalue
# gerp.mean.combined_pvalue <- consequences_combined_significance(gerp.mean,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.mean.combined_pvalue, "gerp_mean_combined_pvalue.csv")
# 
# # calculate group pvalue with all cells as replicates
# gerp.mean.cellreps_pvalue <- consequences_group_significance(gerp.mean,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.mean.cellreps_pvalue, "gerp_mean_cellreps_pvalue.csv")
# 
# # get mean and sd for each group
# stats_summary <- consequences.main.mean.snv.code %>%
#   group_by(condition, variant_variant_type) %>%
#   summarise(mean = mean(log2fc, na.rm = TRUE),
#             sd = sd(log2fc, na.rm = TRUE))
# 
# write.csv(stats_summary, "consequences_condition_main_mean_snv_code_stat_summary.csv")

# control
pdf("tf_condition_sum_control.pdf",
    width = 5,
    height = 4)
ggplot(subset(tf.sum, condition == "control"), aes(x = condition, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n Summed Motif Score Change Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    # geom_text(data = subset(consequences.main.mean.snv.code.cellreps_pvalue, condition == "control"),
    #           aes(x = variant_type, y = Inf, label = significant_p_adjusted), # pvalue of observed/expected
    #           vjust = 4, hjust = 0.75) +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()

# enu
pdf("tf_condition_sum_enu.pdf",
    width = 5,
    height = 4)
ggplot(subset(tf.sum, condition == "enu"), aes(x = condition, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n Summed Motif Score Change Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    # geom_text(data = subset(consequences.main.mean.snv.code.cellreps_pvalue, condition == "control"),
    #           aes(x = variant_type, y = Inf, label = significant_p_adjusted), # pvalue of observed/expected
    #           vjust = 4, hjust = 0.75) +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()
```

## PhyloP

### By condition

#### Mean
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/phylop")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/phylop")

# gerp_levels <- unique(gerp.mean$variant_type)
# 
# # combine p-values from each cell into group pvalue
# gerp.mean.combined_pvalue <- consequences_combined_significance(gerp.mean,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.mean.combined_pvalue, "gerp_mean_combined_pvalue.csv")
# 
# # calculate group pvalue with all cells as replicates
# gerp.mean.cellreps_pvalue <- consequences_group_significance(gerp.mean,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.mean.cellreps_pvalue, "gerp_mean_cellreps_pvalue.csv")
# 
# # get mean and sd for each group
# stats_summary <- consequences.main.mean.snv.code %>%
#   group_by(condition, variant_variant_type) %>%
#   summarise(mean = mean(log2fc, na.rm = TRUE),
#             sd = sd(log2fc, na.rm = TRUE))
# 
# write.csv(stats_summary, "consequences_condition_main_mean_snv_code_stat_summary.csv")

# control
pdf("phylop_condition_mean_control.pdf",
    width = 5,
    height = 4)
ggplot(subset(phylop.mean, condition == "control"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n Mean phylop Score Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()

# enu
pdf("phylop_condition_mean_enu.pdf",
    width = 5,
    height = 4)
ggplot(subset(phylop.mean, condition == "enu"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n Mean phylop Score Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()
```

#### Median
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/phylop")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/phylop")

# gerp_levels <- unique(gerp.median$variant_type)
# 
# # combine p-values from each cell into group pvalue
# gerp.median.combined_pvalue <- consequences_combined_significance(gerp.median,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.median.combined_pvalue, "gerp_median_combined_pvalue.csv")
# 
# # calculate group pvalue with all cells as replicates
# gerp.median.cellreps_pvalue <- consequences_group_significance(gerp.median,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.median.cellreps_pvalue, "gerp_median_cellreps_pvalue.csv")
# 
# # get median and sd for each group
# stats_summary <- consequences.main.median.snv.code %>%
#   group_by(condition, variant_variant_type) %>%
#   summarise(median = median(log2fc, na.rm = TRUE),
#             sd = sd(log2fc, na.rm = TRUE))
# 
# write.csv(stats_summary, "consequences_condition_main_median_snv_code_stat_summary.csv")

# control
pdf("phylop_condition_median_control.pdf",
    width = 5,
    height = 4)
ggplot(subset(phylop.median, condition == "control"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n median phylop Score Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()

# enu
pdf("phylop_condition_median_enu.pdf",
    width = 5,
    height = 4)
ggplot(subset(phylop.median, condition == "enu"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n median phylop Score Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()
```

## PhastCons

### By condition

#### Mean
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/phastcons")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/phastcons")

# gerp_levels <- unique(gerp.mean$variant_type)
# 
# # combine p-values from each cell into group pvalue
# gerp.mean.combined_pvalue <- consequences_combined_significance(gerp.mean,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.mean.combined_pvalue, "gerp_mean_combined_pvalue.csv")
# 
# # calculate group pvalue with all cells as replicates
# gerp.mean.cellreps_pvalue <- consequences_group_significance(gerp.mean,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.mean.cellreps_pvalue, "gerp_mean_cellreps_pvalue.csv")
# 
# # get mean and sd for each group
# stats_summary <- consequences.main.mean.snv.code %>%
#   group_by(condition, variant_variant_type) %>%
#   summarise(mean = mean(log2fc, na.rm = TRUE),
#             sd = sd(log2fc, na.rm = TRUE))
# 
# write.csv(stats_summary, "consequences_condition_main_mean_snv_code_stat_summary.csv")

# control
pdf("phastcons_condition_mean_control.pdf",
    width = 5,
    height = 4)
ggplot(subset(phastcons.mean, condition == "control"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n Mean phastcons Score Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()

# enu
pdf("phastcons_condition_mean_enu.pdf",
    width = 5,
    height = 4)
ggplot(subset(phastcons.mean, condition == "enu"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n Mean phastcons Score Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()
```

#### Median
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/phastcons")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/phastcons")

# gerp_levels <- unique(gerp.median$variant_type)
# 
# # combine p-values from each cell into group pvalue
# gerp.median.combined_pvalue <- consequences_combined_significance(gerp.median,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.median.combined_pvalue, "gerp_median_combined_pvalue.csv")
# 
# # calculate group pvalue with all cells as replicates
# gerp.median.cellreps_pvalue <- consequences_group_significance(gerp.median,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.median.cellreps_pvalue, "gerp_median_cellreps_pvalue.csv")
# 
# # get median and sd for each group
# stats_summary <- consequences.main.median.snv.code %>%
#   group_by(condition, variant_variant_type) %>%
#   summarise(median = median(log2fc, na.rm = TRUE),
#             sd = sd(log2fc, na.rm = TRUE))
# 
# write.csv(stats_summary, "consequences_condition_main_median_snv_code_stat_summary.csv")

# control
pdf("phastcons_condition_median_control.pdf",
    width = 5,
    height = 4)
ggplot(subset(phastcons.median, condition == "control"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n median phastcons Score Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()

# enu
pdf("phastcons_condition_median_enu.pdf",
    width = 5,
    height = 4)
ggplot(subset(phastcons.median, condition == "enu"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = "Log2(Observed / Expected)\n median phastcons Score Per Cell") +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()
```

#### sum
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/phastcons")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/5-variant-effect-prediction/2-ratio-plot/phastcons")

# gerp_levels <- unique(gerp.sum$variant_type)
# 
# # combine p-values from each cell into group pvalue
# gerp.sum.combined_pvalue <- consequences_combined_significance(gerp.sum,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.sum.combined_pvalue, "gerp_sum_combined_pvalue.csv")
# 
# # calculate group pvalue with all cells as replicates
# gerp.sum.cellreps_pvalue <- consequences_group_significance(gerp.sum,
#                                                    c("condition", "variant_type"),
#                                                    gerp_levels)
# write.csv(gerp.sum.cellreps_pvalue, "gerp_sum_cellreps_pvalue.csv")
# 
# # get sum and sd for each group
# stats_summary <- consequences.main.sum.snv.code %>%
#   group_by(condition, variant_variant_type) %>%
#   summarise(sum = sum(log2fc, na.rm = TRUE),
#             sd = sd(log2fc, na.rm = TRUE))
# 
# write.csv(stats_summary, "consequences_condition_main_sum_snv_code_stat_summary.csv")

# control
pdf("phastcons_condition_sum_control.pdf",
    width = 5,
    height = 4)
ggplot(subset(phastcons.sum, condition == "control"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell")))) +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()

# enu
pdf("phastcons_condition_sum_enu.pdf",
    width = 5,
    height = 4)
ggplot(subset(phastcons.sum, condition == "enu"), aes(x = variant_type, y = log2fc, fill = variant_type)) + 
    geom_hline(yintercept = 0, linevariant_type = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(fill = "Variant variant_type") +
    labs(x = "", y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell")))) +
    scale_fill_brewer(palette = "Dark2") +
    theme_Publication() +
    theme(legend.position = "right",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.direction = "vertical",  # Ensure legend items are vertically aligned
        legend.justification = "left",  # Left justify the legend text
          legend.text = element_text(size = 10),  # Adjust legend text size
          legend.title = element_text(size = 12),  # Adjust legend title size
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
dev.off()
```