---
title: "compare dndscv and simulation"
output: html_document
date: "2023-09-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

## Load libraries
```{r}
library(dndscv)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(stringr)
library(tidyr)
library(viridis)
library(reshape2)
```

## Functions
```{r}
theme_Publication <- function(base_size=16, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family = "")
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               #legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}
```


## Load results
```{r}
# sample table
sample_table <- read.table("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/1-data/sample_table_filtered.txt", header = TRUE, sep = "\t")

# dnds
obs_all <- readRDS("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed/obs_all.RDS")

# simulation
vep_sim <- read.csv("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/5-variant-effect-prediction/1-calculate-ratio/cells/consequence_main_oe_mean.csv", header = TRUE)
```

# Compare calculation of expected to results from SigProfilerSimulation

## Run dndscv per cell
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/4-compare-vep")

# split into cells
cell_list <- split(obs_all, obs_all$sampleID)

# subset just for ENU
cell_list_subset <- cell_list[sample_table$condition == "enu"]

cell_list_subset_dnds <- lapply(cell_list_subset, function(x) 
              dndscv(mutations = x,
                     gene_list = NULL,
                     max_muts_per_gene_per_sample = Inf,
                     max_coding_muts_per_sample = Inf,
                     refdb = "hg19",
                     kc = ""))

saveRDS(cell_list_subset_dnds, "global_cell_dndscv.RDS")
```

### Synonymous

### Get number of observed and expected variants per gene and sum per cell from dnds results
```{r}
cell_list_subset_dnds_expected <- lapply(cell_list_subset_dnds, function(x) x$genemuts)

# sum for each cell
cell_list_subset_dnds_expected_sum <- unlist(lapply(cell_list_subset_dnds_expected, function(x) sum(x$exp_syn, na.rm = TRUE)))

cell_list_subset_dnds_observed_sum <- unlist(lapply(cell_list_subset_dnds_expected, function(x) sum(x$n_syn, na.rm = TRUE)))

# make df
dnds_synonymous <- data.frame(dnds_expected = cell_list_subset_dnds_expected_sum,
                         dnds_observed = cell_list_subset_dnds_observed_sum,
                         expected_vs_observed = round(log2(cell_list_subset_dnds_expected_sum / cell_list_subset_dnds_observed_sum)),
                         row.names = names(cell_list_subset_dnds_expected_sum))
```

### Get number of observed and expected variants from vep O/E results
```{r}
vep_sim_synonymous <- subset(vep_sim, variant_type == "synonymous variant")
rownames(vep_sim_synonymous) <- vep_sim_synonymous$sample
```

### Compare dndscv and variant effect prediction
```{r}
# merge dndscv, vep, and sample tabler
df_compare <- merge(dnds_synonymous, vep_sim_synonymous, by = "row.names")
df_compare <- merge(df_compare, sample_table, by = "sample")
rownames(df_compare) <- df_compare$Row.names
df_compare$Row.names <- NULL

# correlation plots
pdf("synonymous_expected_cor.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = sim.mut, y = dnds_expected)) +
    geom_smooth(method = lm) +  # Fitted line without fill
    stat_cor() + # Correlation coefficient
    geom_point(aes(color = as.factor(cycle)), size = 2) +  # Apply fill only to points
    labs(y = "dndscv method",
         x = "Our method", 
         color = "Cycle",
         subtitle = "Expected Synonymous Variants Per Cell") + 
    theme_Publication(base_size = 16) +
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()

pdf("synonymous_observed_cor.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = obs.mut, y = dnds_observed)) +
    geom_smooth(method = lm) +  # Fitted line without fill
    stat_cor() + # Correlation coefficient
    geom_point(aes(color = as.factor(cycle)), size = 2) +  # Apply fill only to points
    labs(y = "dndscv method",
         x = "Our method", 
         color = "Cycle",
         subtitle = "Observed Synonymous Variants Per Cell") + 
    theme_Publication(base_size = 16) +
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()

pdf("synonymous_expected_vs_observed_cor.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = log2fc, y = expected_vs_observed)) +
  geom_point(aes(color = as.factor(cycle)), size = 2) +
  #geom_hline(yintercept = 1, linetype = "dashed") +
  #scale_y_continuous(limits = c(0.99, 1.01)) +
  labs(y = "dndscv method",
       x = "Our method", 
       color = "Cycle",
       subtitle = expression(Log[2] * "(Observed/Expected)\nSynonymous Variants Per Cell")) + 
  theme_Publication(base_size = 16) +
  theme(legend.justification = c("center","top"),
        legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()

# box plots 
df_box <- df_compare[,c("sample", "dnds_expected", "sim.mut")]
df_box <- melt(df_box, id_vars='sample', var_name='variable', value_name='value')

pdf("synonymous_expected_box.pdf",
    width = 5,
    height = 5)
ggpaired(df_box, x = "variable", y = "value",
         point.size = 2) +
  labs(x = NULL,
       y = "Expected Synonymous Variants Per Cell") + 
    theme_Publication(base_size = 16) +
    scale_x_discrete(labels = c("dndscv", "Our Method")) + 
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  stat_compare_means(paired = TRUE)
dev.off()

df_box <- df_compare[,c("sample", "dnds_observed", "obs.mut")]
df_box <- melt(df_box, id_vars='sample', var_name='variable', value_name='value')

pdf("synonymous_observed_box.pdf",
    width = 5,
    height = 5)
ggpaired(df_box, x = "variable", y = "value",
         point.size = 2) +
  labs(x = NULL,
       y = "Observed Synonymous Variants Per Cell") + 
    theme_Publication(base_size = 16) +
    scale_x_discrete(labels = c("dndscv", "Our Method")) + 
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  stat_compare_means(paired = TRUE)
dev.off()
```

## Missense

### Get expected number per gene and sum per cell
```{r}
cell_list_subset_dnds_expected <- lapply(cell_list_subset_dnds, function(x) x$genemuts)

# sum for each cell
cell_list_subset_dnds_expected_sum <- unlist(lapply(cell_list_subset_dnds_expected, function(x) sum(x$exp_mis, na.rm = TRUE)))

cell_list_subset_dnds_observed_sum <- unlist(lapply(cell_list_subset_dnds_expected, function(x) sum(x$n_mis, na.rm = TRUE)))

# make df
dnds_missense <- data.frame(dnds_expected = cell_list_subset_dnds_expected_sum,
                         dnds_observed = cell_list_subset_dnds_observed_sum,
                         expected_vs_observed = log2(cell_list_subset_dnds_expected_sum / cell_list_subset_dnds_observed_sum),
                         row.names = names(cell_list_subset_dnds_expected_sum))
```

### Get number of observed and expected variants from vep O/E results
```{r}
vep_sim_missense <- subset(vep_sim, variant_type == "missense variant")
rownames(vep_sim_missense) <- vep_sim_missense$sample
```

### Compare dndscv and variant effect prediction
```{r}
# merge dndscv, vep, and sample tabler
df_compare <- merge(dnds_missense, vep_sim_missense, by = "row.names")
df_compare <- merge(df_compare, sample_table, by = "sample")
rownames(df_compare) <- df_compare$Row.names
df_compare$Row.names <- NULL

# correlation plots
pdf("missense_expected_cor.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = sim.mut, y = dnds_expected)) +
    geom_smooth(method = lm) +  # Fitted line without fill
    stat_cor() + # Correlation coefficient
    geom_point(aes(color = as.factor(cycle)), size = 2) +  # Apply fill only to points
    labs(y = "dndscv method",
         x = "Our method", 
         color = "Cycle",
         subtitle = "Expected Missense Variants Per Cell") + 
    theme_Publication(base_size = 16) +
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()

pdf("missense_observed_cor.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = obs.mut, y = dnds_observed)) +
    geom_smooth(method = lm) +  # Fitted line without fill
    stat_cor() + # Correlation coefficient
    geom_point(aes(color = as.factor(cycle)), size = 2) +  # Apply fill only to points
    labs(y = "dndscv method",
         x = "Our method", 
         color = "Cycle",
         subtitle = "Observed Missense Variants Per Cell") + 
    theme_Publication(base_size = 16) +
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()

pdf("missense_expected_vs_observed_cor.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = log2fc, y = expected_vs_observed)) +
  geom_smooth(method = lm) +  # Fitted line without fill
  stat_cor() + # Correlation coefficient
  geom_point(aes(color = as.factor(cycle)), size = 2) +
  labs(y = "dndscv method",
       x = "Our method", 
       color = "Cycle",
       subtitle = expression(Log[2] * "(Observed/Expected)\nMissense Variants Per Cell")) + 
  theme_Publication(base_size = 16) +
  theme(legend.justification = c("center","top"),
        legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()

# box plots 
df_box <- df_compare[,c("sample", "dnds_expected", "sim.mut")]
df_box <- melt(df_box, id_vars='sample', var_name='variable', value_name='value')

pdf("missense_expected_box.pdf",
    width = 5,
    height = 5)
ggpaired(df_box, x = "variable", y = "value",
         point.size = 2) +
  labs(x = NULL,
       y = "Expected Missense Variants Per Cell") + 
    theme_Publication(base_size = 16) +
    scale_x_discrete(labels = c("dndscv", "Our Method")) + 
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  stat_compare_means(paired = TRUE)
dev.off()

df_box <- df_compare[,c("sample", "dnds_observed", "obs.mut")]
df_box <- melt(df_box, id_vars='sample', var_name='variable', value_name='value')

pdf("missense_observed_box.pdf",
    width = 5,
    height = 5)
ggpaired(df_box, x = "variable", y = "value",
         point.size = 2) +
  labs(x = NULL,
       y = "Observed Missense Variants Per Cell") + 
    theme_Publication(base_size = 16) +
    scale_x_discrete(labels = c("dndscv", "Our Method")) + 
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  stat_compare_means(paired = TRUE)
dev.off()
```