---
title: "Pathway O/E Analysis"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

## Load libraries
```{r}
library(ggplot2)
library(stringr)
library(dplyr)
library(ggpubr)
library(purrr)
library(rstatix)
library(emmeans)
library(viridis)
library(GO.db)
library(metap)
library(ggvenn)

hg19_size <- 3036303846
```

## Functions
```{r}
# intron freq can be negative due to shuffling, so set these to 0
set_negative_to_zero <- function(x) {
  
  pmax(x, 0)
  
}


# Function to combine p-values using Fisher's method, returning NA if fewer than two valid p-values
combine_pvalues <- function(pvalues) {
  valid_pvalues <- pvalues[!is.na(pvalues)]
  if (length(valid_pvalues) >= 2) {
    tryCatch(
      sumlog(valid_pvalues)$p,
      error = function(e) NA_real_
    )
  } else {
    NA_real_
  }
}

combine_pvalues_fisher <- function(p_values) {
  
    chi_sq_stat = -2 * sum(log(p_values), na.rm = TRUE)
    degrees_of_freedom = 2 * length(p_values)
    p_combined = pchisq(chi_sq_stat, degrees_of_freedom, lower.tail = FALSE)
    return(p_combined)
    
}

combined_significance <- function(ratios, group_vars) {
  
  res <- ratios %>%
    mutate(pvalue = if_else(pvalue == 0, 0.0001, pvalue)) %>% # give values to very low pvalues that are 0 
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(p_value = combine_pvalues_fisher(pvalue)) %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)) %>%
    ungroup()
  
  return(res)
  
}

# function to calculate significance with results coming from
# observed and downsampled
ratio_significance <- function(combined_regions, group_vars, region_levels, by.region = TRUE) {
  
  if(by.region){
  res <- combined_regions %>%
    group_by_at(vars(one_of(c("region", group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue.adj)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  res$region <- factor(res$region, levels = region_levels)
  } else{
    res <- combined_regions %>%
    group_by_at(vars(one_of(c(group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue.adj)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  }
  return(res)
  
}

# Using pvalue here instead of pvalue.adj because many pathways were tested
ratio_significance_pathway <- function(combined_regions, group_vars, region_levels, by.region = TRUE) {
  
  if(by.region){
  res <- combined_regions %>%
    group_by_at(vars(one_of(c("region", group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  res$region <- factor(res$region, levels = region_levels)
  } else{
    res <- combined_regions %>%
    group_by_at(vars(one_of(c(group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  }
  return(res)
  
}

# function to calculate significance between condition groups
group_significance <- function(combined_regions, group_vars, condition, region_levels){
  formula <- as.formula(paste("log2fc ~", condition))
  
  results <- combined_regions %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(
      p_value = wilcox.test(formula = formula, 
                            data = cur_data(),
                            exact = FALSE)$p.value
    ) %>%
    ungroup() %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)
    )

  # Order by region
  results$region <- factor(results$region, levels = region_levels)
  
  return(results)
}

score_combined_significance <- function(ratios, group_vars, group_levels) {
  
  res <- ratios %>%
    mutate(pvalue = if_else(pvalue == 0, 0.001, pvalue)) %>% # give values to very low pvalues that are 0 
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(p_value = combine_pvalues_fisher(pvalue)) %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)) %>%
    ungroup()
  
  # Order by region
  #res$type <- factor(res$type, levels = group_levels)
  
  return(res)
}
# function to calculate significance between condition groups
# also reports significance if seen in both downsampled and observed groups
# adjusted p-values by bejamini-hochberg
# combined_regions <- go.expressed_regions_combined
# group_vars <- c("group", "region")
# condition <- "condition"
# region_levels <- go.expressed_regions_levels

group_significance_go <- function(combined_regions, group_vars, condition, region_levels){
  formula <- as.formula(paste("log2fc ~", condition))
  
  results <- combined_regions %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(
      p_value = t.test(formula = formula, data = cur_data())$p.value
    ) %>%
    ungroup() %>%
    mutate(
           significant_p_adjusted = case_when(
                            p_value <= 0.0001 ~ '****',
                            p_value <= 0.001 ~ '***',
                            p_value <= 0.01 ~ '**',
                            p_value <= 0.05 ~ '*',
                            TRUE ~ NA_character_)
    )
  
  # Identify regions with p_adjusted < 0.05 in both groups
  regions_significant_in_both_groups <- as.data.frame(results) %>%
    filter(p_value < 0.05) %>%
    dplyr::count(region) %>%
    filter(n == 2) %>%
    pull(region)
  
  # Create new column based on the condition
  results <- results %>%
    mutate(significant_p_adjusted_both_groups = ifelse(region %in% regions_significant_in_both_groups, significant_p_adjusted, NA_character_))

  # Order by region
  results$region <- factor(results$region, levels = region_levels)
  
  return(results)
}

calculate_stat_summary <- function(observed, group_vars, mut_var){
  results <- observed %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(mean = mean(!!sym(mut_var), na.rm = TRUE),
              sd = sd(!!sym(mut_var), na.rm = TRUE))
}

theme_Publication <- function(base_size=16, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family = "")
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               #legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_fill_Publication_2 <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#7fc97f","#ef3b2c","#386cb0","#fdb462","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

wrap_labels <- function(labels, width = 5) {
  wrapped_labels <- str_wrap(labels, width)
  names(wrapped_labels) <- names(labels)
  return(wrapped_labels)
}
```

## Load sample table
```{r}
sample_table <- read.table("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/1-data/sample_table_filtered.txt", header = TRUE, sep = "\t")
```

## Load oe results
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/2-calculate-ratio/")

# expressed
expressed_regions_exons_introns_promoters_enhancers <- read.csv("expressed_regions_exons_introns_promoters_enhancers_res.csv", header = TRUE, row.names = NULL)
expressed_regions_exons_promoters_enhancers <- read.csv("expressed_regions_exons_promoters_enhancers_res_observed_variant_required.csv", header = TRUE, row.names = NULL)

# nonexpressed
nonexpressed_regions_exons_introns_promoters_enhancers <- read.csv("nonexpressed_regions_exons_introns_promoters_enhancers_res.csv", header = TRUE, row.names = NULL)
nonexpressed_regions_exons_promoters_enhancers <- read.csv("nonexpressed_regions_exons_promoters_enhancers_res_observed_variant_required.csv", header = TRUE, row.names = NULL)

# combine expressed and non expressed
expressed_nonexpressed_regions_exons_introns_promoters_enhancers <- rbind(expressed_regions_exons_introns_promoters_enhancers,
                                                                          nonexpressed_regions_exons_introns_promoters_enhancers)
expressed_nonexpressed_regions_exons_promoters_enhancers <- rbind(expressed_regions_exons_promoters_enhancers,
                                                                          nonexpressed_regions_exons_promoters_enhancers)
# gene ontology observed
go_expressed_regions_exons_introns_promoters_enhancers <- read.csv("go_expressed_regions_exons_introns_promoters_enhancers_res.csv", header = TRUE, row.names = NULL)
go_expressed_regions_exons_promoters_enhancers <- read.csv("go_expressed_regions_exons_promoters_enhancers_res_observed_variant_required.csv", header = TRUE, row.names = NULL)

# gene ontology shuffled
go_expressed_regions_exons_introns_promoters_enhancers_shuffled <- lapply(list.files(pattern = "go_expressed_shuffled_regions_exons_introns_promoters_enhancers_res"), function(file) {
  read.csv(file, header = TRUE, row.names = NULL)
})

go_expressed_regions_exons_promoters_enhancers_shuffled <- lapply(list.files(pattern = "go_expressed_shuffled_regions_exons_promoters_enhancers_res"), function(file) {
  read.csv(file, header = TRUE, row.names = NULL)
})

# essential and nonessential
AchillesCommonEssentialNonessentialControls_all_regions_res <- read.csv("AchillesCommonEssentialNonessentialControls_all_regions_res.csv", header = TRUE, row.names = NULL)
AchillesCommonEssentialNonessentialControls_no_introns_res <- read.csv("AchillesCommonEssentialNonessentialControls_no_introns_res_observed_variant_required.csv", header = TRUE, row.names = NULL)

# haploinsufficient
haplo_genes_all_regions_res <- read.csv("haplo_genes_all_regions_res.csv", header = TRUE, row.names = NULL)
haplo_genes_no_introns_res <- read.csv("haplo_genes_no_introns_res_observed_variant_required.csv", header = TRUE, row.names = NULL)

observed.variant.required <- read.csv("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/5-variant-effect-prediction/5-count-statistics/results_combined.csv", header = TRUE)
```

## Add meta data info
```{r}
# add sample info
expressed_nonexpressed_regions_exons_promoters_enhancers <- merge(expressed_nonexpressed_regions_exons_promoters_enhancers,
                                                                  sample_table,
                                                                  by = "sample")

expressed_regions_exons_introns_promoters_enhancers <- merge(expressed_regions_exons_introns_promoters_enhancers,
                                                                  sample_table,
                                                                  by = "sample")

go_expressed_regions_exons_promoters_enhancers <- merge(go_expressed_regions_exons_promoters_enhancers,
                                                                  sample_table,
                                                                  by = "sample")

go_expressed_regions_exons_introns_promoters_enhancers <- merge(go_expressed_regions_exons_introns_promoters_enhancers,
                                                                  sample_table,
                                                                  by = "sample")

AchillesCommonEssentialNonessentialControls_all_regions_res <- merge(AchillesCommonEssentialNonessentialControls_all_regions_res,
                                                                  sample_table,
                                                                  by = "sample")

AchillesCommonEssentialNonessentialControls_no_introns_res <- merge(AchillesCommonEssentialNonessentialControls_no_introns_res,
                                                                  sample_table,
                                                                  by = "sample")

haplo_genes_all_regions_res <- merge(haplo_genes_all_regions_res,
                                                                  sample_table,
                                                                  by = "sample")

haplo_genes_no_introns_res <- merge(haplo_genes_no_introns_res,
                                                                  sample_table,
                                                                  by = "sample")

```


# Analysis

## Expressed and Nonexpressed genes
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/expressed_nonexpressed")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/expressed_nonexpressed")

# combine p-values from each cell into group pvalue
expressed_nonexpressed_regions_exons_promoters_enhancers.pvalue <- combined_significance(expressed_nonexpressed_regions_exons_promoters_enhancers,
                                                                   c("condition", "cycle", "region"))
write.csv(expressed_nonexpressed_regions_exons_promoters_enhancers.pvalue, "expressed_nonexpressed_regions_exons_promoters_enhancers.pvalue.csv")

# plot
expressed_nonexpressed_regions_exons_promoters_enhancers$region <- factor(expressed_nonexpressed_regions_exons_promoters_enhancers$region,
                                                                          levels = c("nonexpressed_regions_exons_promoters_enhancers",
                                                                                     "expressed_regions_exons_promoters_enhancers"))
labels <- c("Non-expressed",
            "Expressed")
names(labels) <- levels(expressed_nonexpressed_regions_exons_promoters_enhancers$region)

# frequency
for (region in unique(expressed_nonexpressed_regions_exons_promoters_enhancers$region)) {
  
  filtered_data <- expressed_nonexpressed_regions_exons_promoters_enhancers %>%
                         filter(region == region)
  
  # Perform ANOVA
  anova_result <- summary(aov(obs.mut ~ condition * cycle, data = filtered_data))
  
  # Perform linear model
  lm_result <- lm(obs.mut ~ condition * cycle, data = filtered_data)

  # Compute emmeans and pairwise comparisons
  emm_interaction <- emmeans(lm_result, ~ condition * cycle)
  pairwise_results <- pairs(emm_interaction)

  # Adjust for multiple comparisons and add significance
  adjusted_results <- na.omit(summary(pairwise_results, adjust = "tukey")) %>%
                      add_significance("p.value")

  # Construct the output filename
  output_filename <- paste("expressed_nonexpressed_regions_exons_promoters_enhancers.pvalue", 
                            gsub(" ", "_", region), ".csv", sep = "")

  # Write the results to a CSV file
  write.csv(adjusted_results, output_filename)
  anova_summary_text <- write(capture.output(anova_result), 
                              file = output_filename, 
                              append = TRUE, 
                              sep = "\n")
  
}

stats_summary <- expressed_nonexpressed_regions_exons_promoters_enhancers %>%
  group_by(condition, cycle, region) %>%
  summarise(mean = mean(obs.mut, na.rm = TRUE),
            sd = sd(obs.mut, na.rm = TRUE))

write.csv(stats_summary, "expressed_nonexpressed_regions_exons_promoters_enhancers_frequency.csv")

pdf("expressed_nonexpressed_regions_exons_promoters_enhancers_frequency.pdf",
    width = 4.5,
    height = 4)
ggplot(expressed_nonexpressed_regions_exons_promoters_enhancers, aes(x = condition, y = obs.mut, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +  # Average bar plot
    geom_errorbar(data = stats_summary, aes(x = condition, fill = as.factor(cycle), y = mean, ymin = mean - sd, ymax = mean + sd),
                    width = 0.5, position = position_dodge(0.9)) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", y = "Observed Variants Per Cell", fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

# ratio by cell
pdf("expressed_nonexpressed_regions_exons_promoters_enhancers_cells.pdf",
    width = 4.5,
    height = 4)
ggplot(expressed_nonexpressed_regions_exons_promoters_enhancers, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1)) +
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

# ratio by group
stats_summary <- expressed_nonexpressed_regions_exons_promoters_enhancers %>%
  group_by(condition, cycle, region) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE),
            observed.variant.requirement.met = sum(observed.variant.requirement.met, na.rm = TRUE) / n() > 0.5)
stats_summary$observed.variant.requirement.met <- factor(stats_summary$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

write.csv(stats_summary, "expressed_nonexpressed_regions_exons_promoters_enhancers_group.csv")

pdf("expressed_nonexpressed_regions_exons_promoters_enhancers_control.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "control"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "Control Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

pdf("expressed_nonexpressed_regions_exons_promoters_enhancers_enu.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "enu"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, begin = 0.2, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "ENU-Treated Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()
```

### Intron normalization
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/expressed_nonexpressed")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/expressed_nonexpressed")

# normalize using intron log2fc
expressed_nonexpressed_regions_exons_promoters_enhancers_norm <- expressed_nonexpressed_regions_exons_promoters_enhancers %>%
  merge(expressed_regions_exons_introns_promoters_enhancers[, c("sample", "log2fc")], 
        by = "sample") %>%
  mutate(`log2fc.norm` = `log2fc.x` - `log2fc.y`)

# plot
expressed_nonexpressed_regions_exons_promoters_enhancers_norm$region <- factor(expressed_nonexpressed_regions_exons_promoters_enhancers_norm$region,
                                                                          levels = c("nonexpressed_regions_exons_promoters_enhancers",
                                                                                     "expressed_regions_exons_promoters_enhancers"))
labels <- c("Non-expressed",
            "Expressed")
names(labels) <- levels(expressed_nonexpressed_regions_exons_promoters_enhancers_norm$region)

# ratio by cell
pdf("expressed_nonexpressed_regions_exons_promoters_enhancers_norm_cells.pdf",
    width = 4.5,
    height = 4)
ggplot(expressed_nonexpressed_regions_exons_promoters_enhancers_norm, aes(x = condition, y = log2fc.norm, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1)) +
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

# ratio by group
stats_summary <- expressed_nonexpressed_regions_exons_promoters_enhancers_norm %>%
  group_by(condition, cycle, region) %>%
  summarise(mean = mean(log2fc.norm, na.rm = TRUE),
            sd = sd(log2fc.norm, na.rm = TRUE),
            observed.variant.requirement.met = sum(observed.variant.requirement.met, na.rm = TRUE) / n() > 0.5)
stats_summary$observed.variant.requirement.met <- factor(stats_summary$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

write.csv(stats_summary, "expressed_nonexpressed_regions_exons_promoters_enhancers_norm_group.csv")

pdf("expressed_nonexpressed_regions_exons_promoters_enhancers_norm_group.pdf",
    width = 4.5,
    height = 4)
ggplot(stats_summary, aes(x = condition, y = mean, color = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    fatten = 5) +
    guides(fill = guide_legend(override.aes = list(shape = 16)), 
           color = guide_legend(override.aes = list(shape = 16))) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

pdf("expressed_nonexpressed_regions_exons_promoters_enhancers_norm_control.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "control"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "Control Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

pdf("expressed_nonexpressed_regions_exons_promoters_enhancers_norm_enu.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "enu"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, begin = 0.2, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "ENU-Treated Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()
```

## Essential and Nonessential genes
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/essential_nonessential")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/essential_nonessential")

# combine p-values from each cell into group pvalue
AchillesCommonEssentialNonessentialControls_no_introns_res.pvalue <- combined_significance(AchillesCommonEssentialNonessentialControls_no_introns_res,
                                                                   c("condition", "cycle", "region"))
write.csv(AchillesCommonEssentialNonessentialControls_no_introns_res.pvalue, "AchillesCommonEssentialNonessentialControls_no_introns_res.pvalue.csv")

# plot
AchillesCommonEssentialNonessentialControls_no_introns_res$region <- factor(AchillesCommonEssentialNonessentialControls_no_introns_res$region,
                                                                          levels = c("nonessential",
                                                                                     "essential"))
labels <- c("Nonessential",
            "Essential")
names(labels) <- levels(AchillesCommonEssentialNonessentialControls_no_introns_res$region)

# frequency
for (region in unique(AchillesCommonEssentialNonessentialControls_no_introns_res$region)) {
  
  filtered_data <- AchillesCommonEssentialNonessentialControls_no_introns_res %>%
                         filter(region == region)
  
  # Perform ANOVA
  anova_result <- summary(aov(obs.mut ~ condition * cycle, data = filtered_data))
  
  # Perform linear model
  lm_result <- lm(obs.mut ~ condition * cycle, data = filtered_data)

  # Compute emmeans and pairwise comparisons
  emm_interaction <- emmeans(lm_result, ~ condition * cycle)
  pairwise_results <- pairs(emm_interaction)

  # Adjust for multiple comparisons and add significance
  adjusted_results <- na.omit(summary(pairwise_results, adjust = "tukey")) %>%
                      add_significance("p.value")

  # Construct the output filename
  output_filename <- paste("AchillesCommonEssentialNonessentialControls_no_introns_res.pvalue", 
                            gsub(" ", "_", region), ".csv", sep = "")

  # Write the results to a CSV file
  write.csv(adjusted_results, output_filename)
  anova_summary_text <- write(capture.output(anova_result), 
                              file = output_filename, 
                              append = TRUE, 
                              sep = "\n")
  
}

stats_summary <- AchillesCommonEssentialNonessentialControls_no_introns_res %>%
  group_by(condition, cycle, region) %>%
  summarise(mean = mean(obs.mut, na.rm = TRUE),
            sd = sd(obs.mut, na.rm = TRUE),
            observed.variant.requirement.met = sum(observed.variant.requirement.met, na.rm = TRUE) / n() > 0.5)
stats_summary$observed.variant.requirement.met <- factor(stats_summary$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

write.csv(stats_summary, "AchillesCommonEssentialNonessentialControls_no_introns_res_frequency.csv")

pdf("AchillesCommonEssentialNonessentialControls_no_introns_res_frequency.pdf",
    width = 4.5,
    height = 4)
ggplot(AchillesCommonEssentialNonessentialControls_no_introns_res, aes(x = condition, y = obs.mut, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +  # Average bar plot
    geom_errorbar(data = stats_summary, aes(x = condition, fill = as.factor(cycle), y = mean, ymin = mean - sd, ymax = mean + sd),
                    width = 0.5, position = position_dodge(0.9)) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", y = "Observed Variants Per Cell", fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

# ratio by cell
pdf("AchillesCommonEssentialNonessentialControls_no_introns_res_cells.pdf",
    width = 4.5,
    height = 4)
ggplot(AchillesCommonEssentialNonessentialControls_no_introns_res, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1)) +
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

# ratio by group
stats_summary <- AchillesCommonEssentialNonessentialControls_no_introns_res %>%
  group_by(condition, cycle, region) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE),
            observed.variant.requirement.met = sum(observed.variant.requirement.met, na.rm = TRUE) / n() > 0.5)
stats_summary$observed.variant.requirement.met <- factor(stats_summary$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

write.csv(stats_summary, "AchillesCommonEssentialNonessentialControls_no_introns_res_group.csv")

pdf("AchillesCommonEssentialNonessentialControls_no_introns_res_control.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "control"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "Control Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

pdf("AchillesCommonEssentialNonessentialControls_no_introns_res_enu.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "enu"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, begin = 0.2, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "ENU-Treated Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()
```

### Intron normalization
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/essential_nonessential")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/essential_nonessential")

# normalize using intron log2fc
AchillesCommonEssentialNonessentialControls_no_introns_res_norm <- AchillesCommonEssentialNonessentialControls_no_introns_res %>%
  merge(AchillesCommonEssentialNonessentialControls_all_regions_res[, c("sample", "log2fc")], 
        by = "sample") %>%
  mutate(`log2fc.norm` = `log2fc.x` - `log2fc.y`)

# plot
AchillesCommonEssentialNonessentialControls_no_introns_res_norm$region <- factor(AchillesCommonEssentialNonessentialControls_no_introns_res_norm$region,
                                                                          levels = c("nonessential",
                                                                                     "essential"))

labels <- c("Nonessential",
            "Essential")
names(labels) <- levels(AchillesCommonEssentialNonessentialControls_no_introns_res_norm$region)

# ratio by cell
pdf("AchillesCommonEssentialNonessentialControls_no_introns_res_norm_cells.pdf",
    width = 4.5,
    height = 4)
ggplot(AchillesCommonEssentialNonessentialControls_no_introns_res_norm, aes(x = condition, y = log2fc.norm, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1)) +
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

# ratio by group
stats_summary <- AchillesCommonEssentialNonessentialControls_no_introns_res_norm %>%
  group_by(condition, cycle, region) %>%
  summarise(mean = mean(log2fc.norm, na.rm = TRUE),
            sd = sd(log2fc.norm, na.rm = TRUE),
            observed.variant.requirement.met = sum(observed.variant.requirement.met, na.rm = TRUE) / n() > 0.5)
stats_summary$observed.variant.requirement.met <- factor(stats_summary$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

write.csv(stats_summary, "AchillesCommonEssentialNonessentialControls_no_introns_res_norm_group.csv")

pdf("AchillesCommonEssentialNonessentialControls_no_introns_res_norm_control.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "control"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "Control Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

pdf("AchillesCommonEssentialNonessentialControls_no_introns_res_norm_enu.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "enu"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, begin = 0.2, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "ENU-Treated Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()
```

## Haplosufficient and Haploinsufficient genes
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/haploinsufficient")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/haploinsufficient")

# combine p-values from each cell into group pvalue
haplo_genes_no_introns_res.pvalue <- combined_significance(haplo_genes_no_introns_res,
                                                                   c("condition", "cycle", "region"))
write.csv(haplo_genes_no_introns_res.pvalue, "haplo_genes_no_introns_res.pvalue.csv")

# plot
haplo_genes_no_introns_res$region <- factor(haplo_genes_no_introns_res$region,
                                                                          levels = c("haplosufficient",
                                                                                     "haploinsufficient"))

labels <- c("Haplosufficient",
            "Haploinsufficient")
names(labels) <- levels(haplo_genes_no_introns_res$region)

# frequency
for (region in unique(haplo_genes_no_introns_res$region)) {
  
  filtered_data <- haplo_genes_no_introns_res %>%
                         filter(region == region)
  
  # Perform ANOVA
  anova_result <- summary(aov(obs.mut ~ condition * cycle, data = filtered_data))
  
  # Perform linear model
  lm_result <- lm(obs.mut ~ condition * cycle, data = filtered_data)

  # Compute emmeans and pairwise comparisons
  emm_interaction <- emmeans(lm_result, ~ condition * cycle)
  pairwise_results <- pairs(emm_interaction)

  # Adjust for multiple comparisons and add significance
  adjusted_results <- na.omit(summary(pairwise_results, adjust = "tukey")) %>%
                      add_significance("p.value")

  # Construct the output filename
  output_filename <- paste("haplo_genes_no_introns_res.pvalue", 
                            gsub(" ", "_", region), ".csv", sep = "")

  # Write the results to a CSV file
  write.csv(adjusted_results, output_filename)
  anova_summary_text <- write(capture.output(anova_result), 
                              file = output_filename, 
                              append = TRUE, 
                              sep = "\n")
  
}

stats_summary <- haplo_genes_no_introns_res %>%
  group_by(condition, cycle, region) %>%
  summarise(mean = mean(obs.mut, na.rm = TRUE),
            sd = sd(obs.mut, na.rm = TRUE))

write.csv(stats_summary, "haplo_genes_no_introns_res_frequency.csv")

pdf("haplo_genes_no_introns_res_frequency.pdf",
    width = 4.5,
    height = 4)
ggplot(haplo_genes_no_introns_res, aes(x = condition, y = obs.mut, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +  # Average bar plot
    geom_errorbar(data = stats_summary, aes(x = condition, fill = as.factor(cycle), y = mean, ymin = mean - sd, ymax = mean + sd),
                    width = 0.5, position = position_dodge(0.9)) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", y = "Observed Variants Per Cell", fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

# ratio by cell
pdf("haplo_genes_no_introns_res_cells.pdf",
    width = 4.5,
    height = 4)
ggplot(haplo_genes_no_introns_res, aes(x = condition, y = log2fc, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1)) +
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

# ratio by group
stats_summary <- haplo_genes_no_introns_res %>%
  group_by(condition, cycle, region) %>%
  summarise(mean = mean(log2fc, na.rm = TRUE),
            sd = sd(log2fc, na.rm = TRUE),
            observed.variant.requirement.met = sum(observed.variant.requirement.met, na.rm = TRUE) / n() > 0.5)
stats_summary$observed.variant.requirement.met <- factor(stats_summary$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

write.csv(stats_summary, "haplo_genes_no_introns_res_group.csv")

pdf("haplo_genes_no_introns_res_control.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "control"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "Control Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

pdf("haplo_genes_no_introns_res_enu.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "enu"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, begin = 0.2, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "ENU-Treated Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()
```

### Intron normalization
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/haploinsufficient")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/haploinsufficient")

# normalize using intron log2fc
haplo_genes_no_introns_res_norm <- haplo_genes_no_introns_res %>%
  merge(haplo_genes_all_regions_res[, c("sample", "log2fc")], 
        by = "sample") %>%
  mutate(`log2fc.norm` = `log2fc.x` - `log2fc.y`)

# plot
haplo_genes_no_introns_res_norm$region <- factor(haplo_genes_no_introns_res_norm$region,
                                                                          levels = c("haplosufficient",
                                                                                     "haploinsufficient"))

labels <- c("Haplosufficient",
            "Haploinsufficient")
names(labels) <- levels(haplo_genes_no_introns_res_norm$region)

# ratio by cell
pdf("haplo_genes_no_introns_res_norm_cells.pdf",
    width = 4.5,
    height = 4)
ggplot(haplo_genes_no_introns_res_norm, aes(x = condition, y = log2fc.norm, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(size = 2, alpha = 0.5, position = position_dodge(width = 0.75)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Cell"))), 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +  # Adjust legend title size
    scale_y_continuous(breaks = c(-1, 0, 1), lim = c(-1, 1)) +
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

# ratio by group
stats_summary <- haplo_genes_no_introns_res_norm %>%
  group_by(condition, cycle, region) %>%
  summarise(mean = mean(log2fc.norm, na.rm = TRUE),
            sd = sd(log2fc.norm, na.rm = TRUE),
            observed.variant.requirement.met = sum(observed.variant.requirement.met, na.rm = TRUE) / n() > 0.5)
stats_summary$observed.variant.requirement.met <- factor(stats_summary$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

write.csv(stats_summary, "haplo_genes_no_introns_res_norm_group.csv")

pdf("haplo_genes_no_introns_res_norm_control.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "control"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "Control Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()

pdf("haplo_genes_no_introns_res_norm_enu.pdf",
    width = 4.5,
    height = 4)
ggplot(subset(stats_summary, condition == "enu"), aes(x = condition, y = mean, color = as.factor(cycle), shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=mean - sd, ymax=mean + sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, begin = 0.2, end = 0.8) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(x = "ENU-Treated Cells", 
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "Variants Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +  # Adjust legend title size
    facet_grid(.~region, labeller = labeller(region = labels))
dev.off()
```

## Gene ontology
- Improvement: Incorporate average expression of each pathway
- Calculate pvalue using normalized OE ratio

### Observed 

#### Calculate Summmary

##### Without introns
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/observed")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/observed")

# add GO ID
goterms <- Term(GOTERM)
go.map <- goterms[goterms %in% unique(go_expressed_regions_exons_promoters_enhancers$region)]
go_expressed_regions_exons_promoters_enhancers$go.id <- names(go.map)[match(go_expressed_regions_exons_promoters_enhancers$region, go.map)]

# combine samples into groups and calculate summary statistics
go_expressed_regions_exons_promoters_enhancers_summary <- go_expressed_regions_exons_promoters_enhancers %>%
  group_by(condition, cycle, region) %>%
  summarise(log2fc.mean = mean(log2fc, na.rm = TRUE),
            log2fc.sd = sd(log2fc, na.rm = TRUE),
            obs.mut.sum = sum(obs.mut, na.rm = TRUE),
            obs.mut.mean = mean(obs.mut, na.rm = TRUE),
            obs.mut.sd = sd(obs.mut, na.rm = TRUE),
            obs.mut.cv = sd(obs.mut, na.rm = TRUE)/mean(obs.mut, na.rm = TRUE),
            sim.mut.sum = sum(sim.mut, na.rm = TRUE),
            sim.mut.mean = mean(sim.mut, na.rm = TRUE)) %>%
  left_join(combined_significance(go_expressed_regions_exons_promoters_enhancers, c("condition", "cycle", "region")),
            by = c("condition", "cycle", "region")) %>%
  mutate(go.id = names(go.map)[match(region, go.map)])

go_expressed_regions_exons_promoters_enhancers_summary$cycle <- factor(go_expressed_regions_exons_promoters_enhancers_summary$cycle,
                                                                       levels = c(1,3,6,9))

# add in observed variant required data
results_pval05 <- observed.variant.required %>%
  filter(pvalue < 0.05)

get_required_count <- function(lfc_val, data, tol = 1e-4) {
  out <- data %>%
    mutate(dist = abs(log2_oe - lfc_val)) %>%
    filter(dist < tol)
  if (nrow(out) == 0) return(NA_real_)
  min(out$observed.count)
}

go_expressed_regions_exons_promoters_enhancers_summary <- go_expressed_regions_exons_promoters_enhancers_summary %>%
  rowwise() %>%
  mutate(
    observed.variant.required = get_required_count(abs(log2fc.mean), results_pval05, tol = 1e-1),
    observed.variant.requirement.met = obs.mut.sum >= observed.variant.required
  ) %>%
  ungroup()
  
go_expressed_regions_exons_promoters_enhancers_summary$observed.variant.requirement.met <- factor(go_expressed_regions_exons_promoters_enhancers_summary$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

# output
write.csv(go_expressed_regions_exons_promoters_enhancers_summary, "go_expressed_regions_exons_promoters_enhancers_summary.csv")
```

##### With introns
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/observed")

# add GO ID
goterms <- Term(GOTERM)
go.map <- goterms[goterms %in% unique(go_expressed_regions_exons_introns_promoters_enhancers$region)]
go_expressed_regions_exons_introns_promoters_enhancers$go.id <- names(go.map)[match(go_expressed_regions_exons_introns_promoters_enhancers$region, go.map)]

# combine samples into groups and calculate summary statistics
go_expressed_regions_exons_introns_promoters_enhancers_summary <- go_expressed_regions_exons_introns_promoters_enhancers %>%
  group_by(condition, cycle, region) %>%
  summarise(log2fc.mean = mean(log2fc, na.rm = TRUE),
            obs.mut.sum = sum(obs.mut, na.rm = TRUE),
            obs.mut.mean = mean(obs.mut, na.rm = TRUE),
            obs.mut.sd = sd(obs.mut, na.rm = TRUE),
            obs.mut.cv = sd(obs.mut, na.rm = TRUE)/mean(obs.mut, na.rm = TRUE),
            sim.mut.mean = mean(sim.mut, na.rm = TRUE)) %>%
  left_join(combined_significance(go_expressed_regions_exons_introns_promoters_enhancers, c("condition", "cycle", "region")),
            by = c("condition", "cycle", "region")) %>%
  mutate(go.id = names(go.map)[match(region, go.map)])

go_expressed_regions_exons_introns_promoters_enhancers_summary$cycle <- factor(go_expressed_regions_exons_introns_promoters_enhancers_summary$cycle,
                                                                       levels = c(1,3,6,9))

# add in observed variant required data
results_pval05 <- observed.variant.required %>%
  filter(pvalue < 0.05)

get_required_count <- function(lfc_val, data, tol = 1e-4) {
  out <- data %>%
    mutate(dist = abs(log2_oe - lfc_val)) %>%
    filter(dist < tol)
  if (nrow(out) == 0) return(NA_real_)
  min(out$observed.count)
}

go_expressed_regions_exons_introns_promoters_enhancers_summary <- go_expressed_regions_exons_introns_promoters_enhancers_summary %>%
  rowwise() %>%
  mutate(
    observed.variant.required = get_required_count(abs(log2fc.mean), results_pval05, tol = 1e-1),
    observed.variant.requirement.met = obs.mut.sum >= observed.variant.required
  ) %>%
  ungroup()
  
go_expressed_regions_exons_introns_promoters_enhancers_summary$observed.variant.requirement.met <- factor(go_expressed_regions_exons_introns_promoters_enhancers_summary$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

# output
write.csv(go_expressed_regions_exons_introns_promoters_enhancers_summary, "go_expressed_regions_exons_introns_promoters_enhancers_summary.csv")
```

#### SNV frequency
```{r}
pdf("control+enu_violin_frequency.pdf",
    width = 4.5,
    height = 4)
ggplot(go_expressed_regions_exons_promoters_enhancers_summary, aes(x = condition, y = obs.mut.mean, fill = as.factor(cycle))) + 
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(data = subset(go_expressed_regions_exons_promoters_enhancers_summary,
                            condition == "enu" & observed.variant.requirement.met == FALSE),
              aes(shape = observed.variant.requirement.met,
                  group = interaction(condition, cycle)),
              size = 1.5, alpha = 0.25,
              position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +
  geom_jitter(data = subset(go_expressed_regions_exons_promoters_enhancers_summary,
                            condition == "enu" & observed.variant.requirement.met == TRUE),
              aes(shape = observed.variant.requirement.met,
                  group = interaction(condition, cycle)),
              size = 1.5, alpha = 0.8,
              position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
      scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(y = "Mean Observed\nVariants Per Pathway", 
         fill = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "vertical",
        axis.title.x = element_blank(),
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control+enu_violin_frequency_observed_requirement.pdf",
    width = 4.5,
    height = 4)
ggplot(go_expressed_regions_exons_promoters_enhancers_summary, 
       aes(x = condition, y = obs.mut.mean, fill = as.factor(cycle))) + 
  geom_violin(position = position_dodge(width = 0.85), 
              scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
  geom_jitter(aes(color = observed.variant.requirement.met,
                  group = interaction(condition, cycle)), 
              size = 1, alpha = 0.5, 
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.85)) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")) +
  guides(fill = guide_legend(order = 1, override.aes = list(shape = NA)),
  color = guide_legend(order = 2)) +
  labs(x = "", y = "Mean Observed\nVariants Per Pathway", fill = "Cycle", 
       color = "Observed Variant\nCount Sufficient?") +
  scale_fill_viridis(discrete = TRUE, end = 0.8) +
  theme_Publication() +
  theme(legend.position = "top", legend.box = "vertical") +
  scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control+enu_violin_frequency_pvalue.pdf",
    width = 4.5,
    height = 4)
ggplot(go_expressed_regions_exons_promoters_enhancers_summary, aes(x = condition, y = obs.mut.mean, fill = as.factor(cycle))) + 
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted), size = 1, alpha = 0.5, position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", y = "Mean Observed Variants Per Pathway", fill = "Cycle", color = "Adjusted p-value") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_summary$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_summary$p_value))) +
    theme_Publication() +
    theme(legend.position = "none",  # Place legend on the top
          legend.box = "vertical") +  # Stack legend items horizontally
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()
```

#### O/E ratio
```{r}
pdf("control+enu_violin_all.pdf",
    width = 4.5,
    height = 4.5)
ggplot(go_expressed_regions_exons_promoters_enhancers_summary, aes(x = condition, y = log2fc.mean, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted), size = 1, alpha = 0.5, position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", color = "Adjusted p-value") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_summary$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_summary$p_value))) +
    theme_Publication() +
    theme(legend.position = "none",  # Place legend on the top
          legend.box = "vertical") +  # Stack legend items horizontally
    scale_y_continuous(breaks = seq(-2, 2, 1), lim = c(-2, 1.5)) +
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control+enu_violin_legend.pdf",
    width = 7,
    height = 5)
ggplot(go_expressed_regions_exons_promoters_enhancers_summary, aes(x = condition, y = log2fc.mean, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted), size = 1, alpha = 0.25, position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", color = "Adjusted p-value") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_summary$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_summary$p_value))) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "horizontal") +  # Stack legend items horizontally
    scale_y_continuous(breaks = seq(-2, 2, 1), lim = c(-2, 1.5)) +
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control_violin_oe.pdf",
    width = 4,
    height = 5)
ggplot(subset(go_expressed_regions_exons_promoters_enhancers_summary,
              condition == "control"), aes(x = condition, y = log2fc.mean, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted,
                    shape = observed.variant.requirement.met,
                    group = interaction(condition, cycle)), size = 2, alpha = 0.5, 
                position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Control Cells",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", 
         color = "Adjusted p-value",
         shape = "Observed Variant\nCount Sufficient?") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_summary$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_summary$p_value))) +
    scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "vertical",
          axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12))  # Adjust legend title size
dev.off()

pdf("enu_violin_oe.pdf",
    width = 4,
    height = 5)
ggplot(subset(go_expressed_regions_exons_promoters_enhancers_summary,
              condition == "enu"), aes(x = condition, y = log2fc.mean, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
   geom_jitter(data = subset(go_expressed_regions_exons_promoters_enhancers_summary,
                            condition == "enu" & observed.variant.requirement.met == FALSE),
              aes(color = p_adjusted, shape = observed.variant.requirement.met,
                  group = interaction(condition, cycle)),
              size = 2, alpha = 0.25,
              position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +

  geom_jitter(data = subset(go_expressed_regions_exons_promoters_enhancers_summary,
                            condition == "enu" & observed.variant.requirement.met == TRUE),
              aes(color = p_adjusted, shape = observed.variant.requirement.met,
                  group = interaction(condition, cycle)),
              size = 2, alpha = 0.9,
              position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "ENU-Treated Cells",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", 
         color = "Adjusted p-value",
         shape = "Observed Variant\nCount Sufficient?") +
    scale_fill_viridis(discrete = TRUE, end = 0.8, begin = 0.2) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_summary$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_summary$p_value))) +
    scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "vertical",
          axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12))  # Adjust legend title size
dev.off()
```

##### Intron normalization
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/observed")

# normalize at the summary level
go_expressed_regions_exons_promoters_enhancers_summary_norm <- go_expressed_regions_exons_promoters_enhancers_summary %>%
  merge(go_expressed_regions_exons_introns_promoters_enhancers_summary[, c("condition", "cycle", "region", "log2fc.mean")], 
        by = c("condition", "cycle", "region")) %>%
  mutate(`log2fc.norm` = `log2fc.mean.x` - `log2fc.mean.y`) %>%
  mutate(`log2fc.norm.sd` = sd(log2fc.norm, na.rm = TRUE))

go_expressed_regions_exons_promoters_enhancers_summary_norm$observed.variant.requirement.met <- factor(go_expressed_regions_exons_promoters_enhancers_summary_norm$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

write.csv(go_expressed_regions_exons_promoters_enhancers_summary_norm, "go_expressed_regions_exons_promoters_enhancers_summary_norm.csv")

pdf("control+enu_violin_oe_norm.pdf",
    width = 4.5,
    height = 4.5)
ggplot(go_expressed_regions_exons_promoters_enhancers_summary_norm, aes(x = condition, y = log2fc.norm, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted), size = 1, alpha = 0.5, position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", color = "Adjusted p-value") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_summary_norm$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_summary_norm$p_value))) +
    theme_Publication() +
    theme(legend.position = "none",  # Place legend on the top
          legend.box = "vertical") +  # Stack legend items horizontally
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control+enu_violin_legend_oe_norm.pdf",
    width = 7,
    height = 5)
ggplot(go_expressed_regions_exons_promoters_enhancers_summary_norm, aes(x = condition, y = log2fc.norm, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted), size = 1, alpha = 0.25, position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", color = "Adjusted p-value") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_summary_norm$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_summary_norm$p_value))) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "horizontal") +  # Stack legend items horizontally
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control_violin_oe_norm.pdf",
    width = 4,
    height = 5)
ggplot(subset(go_expressed_regions_exons_promoters_enhancers_summary_norm,
              condition == "control"), aes(x = condition, y = log2fc.norm, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted,
                    shape = observed.variant.requirement.met,
                    group = interaction(condition, cycle)), size = 2, alpha = 0.5, 
                position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Control Cells",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", 
         color = "Adjusted p-value",
         shape = "Observed Variant\nCount Sufficient?") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_summary_norm$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_summary_norm$p_value))) +
    scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "vertical",
          axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12))  # Adjust legend title size
dev.off()

pdf("enu_violin_oe_norm.pdf",
    width = 4,
    height = 5)
ggplot(subset(go_expressed_regions_exons_promoters_enhancers_summary_norm,
              condition == "enu"), aes(x = condition, y = log2fc.norm, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
   geom_jitter(data = subset(go_expressed_regions_exons_promoters_enhancers_summary_norm,
                            condition == "enu" & observed.variant.requirement.met == FALSE),
              aes(color = p_adjusted, shape = observed.variant.requirement.met,
                  group = interaction(condition, cycle)),
              size = 2, alpha = 0.25,
              position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +

  geom_jitter(data = subset(go_expressed_regions_exons_promoters_enhancers_summary_norm,
                            condition == "enu" & observed.variant.requirement.met == TRUE),
              aes(color = p_adjusted, shape = observed.variant.requirement.met,
                  group = interaction(condition, cycle)),
              size = 2, alpha = 0.9,
              position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "ENU-Treated Cells",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", 
         color = "Adjusted p-value",
         shape = "Observed Variant\nCount Sufficient?") +
    scale_fill_viridis(discrete = TRUE, end = 0.8, begin = 0.2) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_summary_norm$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_summary_norm$p_value))) +
    scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "vertical",
          axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12))  # Adjust legend title size
dev.off()
```

#### Significant pathways
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/observed")

go_expressed_regions_exons_promoters_enhancers_summary[go_expressed_regions_exons_promoters_enhancers_summary$p_adjusted < 0.05,]

sig.pathways <- c("response to oxidative stress",
                  "respiratory system development",
                  "cellular response to chemical stress",
                  "respiratory tube development",
                  "proteasome-mediated ubiquitin-dependent protein catabolic process",
                  "regulation of response to DNA damage stimulus")

go_expressed_regions_exons_promoters_enhancers_summary_norm$cycle <- factor(go_expressed_regions_exons_promoters_enhancers_summary_norm$cycle,
                                                                            levels = c(9, 6, 3))

# plot oe
pdf("enu_cycle_9_sig_pathways_oe_norm.pdf",
    width = 6,  
    height = 7)
ggplot(subset(go_expressed_regions_exons_promoters_enhancers_summary_norm, 
              region %in% sig.pathways & condition == "enu"), aes(x = region, y = log2fc.norm, color = as.factor(cycle),
                                             shape = observed.variant.requirement.met)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=log2fc.norm - log2fc.norm.sd, 
                        ymax=log2fc.norm + log2fc.norm.sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    labs(x = "ENU-Treated Cells",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Cell"))), 
         color = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
      guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    scale_color_viridis(discrete = TRUE, end = 0.8, begin = 0.2, direction = -1) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "vertical",  # Stack legend items vertically
        legend.text = element_text(size = 16),  # Adjust legend text size
        legend.title = element_text(size = 16),
        strip.text = element_text(size = 12),
        axis.title.y = element_blank()) +  # Adjust legend title size
  coord_flip() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 25))
dev.off()
```

##### Reduction in SNVs
```{r}
# reduction in SNV frequency
go_expressed_regions_exons_promoters_enhancers$obs.sim.mut.diff <- go_expressed_regions_exons_promoters_enhancers$obs.mut -
  go_expressed_regions_exons_promoters_enhancers$sim.mut 

sig.pathways.protection <- go_expressed_regions_exons_promoters_enhancers %>%
  subset(region %in% sig.pathways) %>%
  subset(condition == "enu") %>%
  group_by(condition, cycle, region) %>%
  summarize(obs.sim.mut.diff.mean = mean(obs.sim.mut.diff, na.rm = TRUE),
            obs.sim.mut.diff.sd = sd(obs.sim.mut.diff, na.rm = TRUE)) %>%
  ungroup()

sig.pathways.protection$cycle <- factor(sig.pathways.protection$cycle,
                                        levels = c(9, 6, 3))

pdf("enu_cycle_9_sig_pathways_obs-sim_diff.pdf",
    width = 6,  
    height = 7)
ggplot(sig.pathways.protection, aes(x = region, y = obs.sim.mut.diff.mean, color = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=obs.sim.mut.diff.mean - obs.sim.mut.diff.sd, 
                        ymax=obs.sim.mut.diff.mean + obs.sim.mut.diff.sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    labs(x = "",
         y = "Reduction in SNV Frequency",
         color = "Cycle") +
      guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, end = 0.8, begin = 0.2, direction = -1) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 16),  # Adjust legend text size
        legend.title = element_text(size = 16),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
  coord_flip() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 25)) +
  scale_y_continuous(breaks = c(-15, -10, -5, 0, 5, 10))
dev.off()

# percent reduction
go_expressed_regions_exons_promoters_enhancers$obs.sim.mut.diff.percent <- ((go_expressed_regions_exons_promoters_enhancers$obs.mut + 1) -
  (go_expressed_regions_exons_promoters_enhancers$sim.mut + 1))/(go_expressed_regions_exons_promoters_enhancers$sim.mut + 1)

sig.pathways.protection <- go_expressed_regions_exons_promoters_enhancers %>%
  subset(region %in% sig.pathways) %>%
  subset(condition == "enu") %>%
  group_by(condition, cycle, region) %>%
  summarize(obs.sim.mut.diff.percent.mean = mean(obs.sim.mut.diff.percent, na.rm = TRUE),
            obs.sim.mut.diff.percent.sd = sd(obs.sim.mut.diff.percent, na.rm = TRUE)) %>%
  ungroup()

sig.pathways.protection$cycle <- factor(sig.pathways.protection$cycle,
                                        levels = c(9, 6, 3))

pdf("enu_cycle_9_sig_pathways_obs-sim_diff_percent.pdf",
    width = 6,
    height = 7)
ggplot(sig.pathways.protection, aes(x = region, y = obs.sim.mut.diff.percent.mean, color = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_pointrange(aes(ymin=obs.sim.mut.diff.percent.mean - obs.sim.mut.diff.percent.sd, 
                        ymax=obs.sim.mut.diff.percent.mean + obs.sim.mut.diff.percent.sd),
                    position=position_dodge(width = 0.7),
                    size = 1.1) +
    labs(x = "",
         y = "Percent Reduction in SNVs",
         color = "Cycle") +
      guides(color = guide_legend(override.aes = list(shape = 16, size = 0.75), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
    scale_color_viridis(discrete = TRUE, end = 0.8, begin = 0.2, direction = -1) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 16),  # Adjust legend text size
        legend.title = element_text(size = 16),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
  coord_flip() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 25)) +
  scale_y_continuous(breaks = c(-0.5, -0.25, 0, 0.25, 0.5), labels = scales::percent_format(accuracy = 1))
dev.off()
```

### Shuffled 

#### Get average of all instances
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/shuffled")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/shuffled")

# average all shuffled instances
go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg <- bind_rows(go_expressed_regions_exons_introns_promoters_enhancers_shuffled) %>%
  group_by(region, sample) %>%
  summarise(
    obs.mut = mean(obs.mut, na.rm = TRUE),
    sim.mut = mean(sim.mut, na.rm = TRUE),
    log2fc = mean(log2fc, na.rm = TRUE),
    pvalue = combine_pvalues(pvalue), 
    .groups = "drop"
  ) %>%
  ungroup()

go_expressed_regions_exons_promoters_enhancers_shuffled_avg <- bind_rows(go_expressed_regions_exons_promoters_enhancers_shuffled) %>%
  group_by(region, sample) %>%
  summarise(
        obs.mut = mean(obs.mut, na.rm = TRUE),
        sim.mut = mean(sim.mut, na.rm = TRUE),
        log2fc = mean(log2fc, na.rm = TRUE),
        pvalue = combine_pvalues(pvalue), 
    .groups = "drop"
  ) %>%
  ungroup()

# add meta data
go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg <- merge(go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg,
                                                                  sample_table,
                                                                  by = "sample")

go_expressed_regions_exons_promoters_enhancers_shuffled_avg <- merge(go_expressed_regions_exons_promoters_enhancers_shuffled_avg,
                                                                  sample_table,
                                                                  by = "sample")
```

#### Calculate Summmary

##### Without introns
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/shuffled")

# add GO ID
goterms <- Term(GOTERM)
go.map <- goterms[goterms %in% unique(go_expressed_regions_exons_promoters_enhancers_shuffled_avg$region)]
go_expressed_regions_exons_promoters_enhancers_shuffled_avg$go.id <- names(go.map)[match(go_expressed_regions_exons_promoters_enhancers_shuffled_avg$region, go.map)]

# combine samples into groups and calculate summary statistics
go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary <- go_expressed_regions_exons_promoters_enhancers_shuffled_avg %>%
  group_by(condition, cycle, region) %>%
  summarise(log2fc.mean = mean(log2fc, na.rm = TRUE),
            log2fc.sd = sd(log2fc, na.rm = TRUE),
            obs.mut.sum = sum(obs.mut, na.rm = TRUE),
            obs.mut.mean = mean(obs.mut, na.rm = TRUE),
            obs.mut.sd = sd(obs.mut, na.rm = TRUE),
            obs.mut.cv = sd(obs.mut, na.rm = TRUE)/mean(obs.mut, na.rm = TRUE),
            sim.mut.sum = sum(sim.mut, na.rm = TRUE),
            sim.mut.mean = mean(sim.mut, na.rm = TRUE)) %>%
  left_join(combined_significance(go_expressed_regions_exons_promoters_enhancers_shuffled_avg, c("condition", "cycle", "region")),
            by = c("condition", "cycle", "region")) %>%
  mutate(go.id = names(go.map)[match(region, go.map)])

go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$cycle <- factor(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$cycle,
                                                                       levels = c(1,3,6,9))

# add in observed variant required data
results_pval05 <- observed.variant.required %>%
  filter(pvalue < 0.05)

get_required_count <- function(lfc_val, data, tol = 1e-4) {
  out <- data %>%
    mutate(dist = abs(log2_oe - lfc_val)) %>%
    filter(dist < tol)
  if (nrow(out) == 0) return(NA_real_)
  min(out$observed.count)
}

go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary <- go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary %>%
  rowwise() %>%
  mutate(
    observed.variant.required = get_required_count(abs(log2fc.mean), results_pval05, tol = 1e-1),
    observed.variant.requirement.met = obs.mut.sum >= observed.variant.required
  ) %>%
  ungroup()
  
go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$observed.variant.requirement.met <- factor(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

# output
write.csv(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary, "go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary.csv")
```

##### With introns
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/shuffled")

# add GO ID
goterms <- Term(GOTERM)
go.map <- goterms[goterms %in% unique(go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg$region)]
go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg$go.id <- names(go.map)[match(go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg$region, go.map)]

# combine samples into groups and calculate summary statistics
go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg_summary <- go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg %>%
  group_by(condition, cycle, region) %>%
  summarise(log2fc.mean = mean(log2fc, na.rm = TRUE),
            obs.mut.sum = sum(obs.mut, na.rm = TRUE),
            obs.mut.mean = mean(obs.mut, na.rm = TRUE),
            obs.mut.sd = sd(obs.mut, na.rm = TRUE),
            obs.mut.cv = sd(obs.mut, na.rm = TRUE)/mean(obs.mut, na.rm = TRUE),
            sim.mut.mean = mean(sim.mut, na.rm = TRUE)) %>%
  left_join(combined_significance(go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg, c("condition", "cycle", "region")),
            by = c("condition", "cycle", "region")) %>%
  mutate(go.id = names(go.map)[match(region, go.map)])

go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg_summary$cycle <- factor(go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg_summary$cycle,
                                                                       levels = c(1,3,6,9))

# add in observed variant required data
results_pval05 <- observed.variant.required %>%
  filter(pvalue < 0.05)

get_required_count <- function(lfc_val, data, tol = 1e-4) {
  out <- data %>%
    mutate(dist = abs(log2_oe - lfc_val)) %>%
    filter(dist < tol)
  if (nrow(out) == 0) return(NA_real_)
  min(out$observed.count)
}

go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg_summary <- go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg_summary %>%
  rowwise() %>%
  mutate(
    observed.variant.required = get_required_count(abs(log2fc.mean), results_pval05, tol = 1e-1),
    observed.variant.requirement.met = obs.mut.sum >= observed.variant.required
  ) %>%
  ungroup()
  
go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg_summary$observed.variant.requirement.met <- factor(go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg_summary$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

# output
write.csv(go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg_summary, "go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg_summary.csv")
```

#### SNV frequency
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/shuffled")

go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$observed.variant.requirement.met <-
  as.logical(as.character(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$observed.variant.requirement.met))

go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$jitter_alpha <-
  ifelse(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$observed.variant.requirement.met, 0.8, 0.6)

pdf("control+enu_violin_frequency.pdf",
    width = 4.5,
    height = 4)
ggplot(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary, aes(x = condition, y = obs.mut.mean, fill = as.factor(cycle))) + 
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
geom_jitter(
  data = subset(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary,
                condition == "enu"),
  aes(shape = observed.variant.requirement.met,
      group = interaction(condition, cycle),
      alpha = jitter_alpha),
  size = 1.5,
  position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85),
  show.legend = FALSE  # optional if alpha legend not needed
) +
    guides(fill = guide_legend(override.aes = list(shape = NA), order = 1),
           shape = guide_legend(override.aes = list(size = 0.75)), order = 2) +
      scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    labs(y = "Mean Observed\nVariants Per Pathway", 
         fill = "Cycle",
         shape = "Observed Variant\nCount Sufficient?") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "vertical",
        axis.title.x = element_blank(),
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12)) +
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control+enu_violin_frequency_observed_requirement.pdf",
    width = 4.5,
    height = 4)
ggplot(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary, 
       aes(x = condition, y = obs.mut.mean, fill = as.factor(cycle))) + 
  geom_violin(position = position_dodge(width = 0.85), 
              scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
  geom_jitter(aes(color = observed.variant.requirement.met,
                  group = interaction(condition, cycle)), 
              size = 1, alpha = 0.5, 
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.9)) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")) +
  guides(fill = guide_legend(order = 1, override.aes = list(shape = NA)),
  color = guide_legend(order = 2)) +
  labs(x = "", y = "Mean Observed\nVariants Per Pathway", fill = "Cycle", 
       color = "Observed Variant\nCount Sufficient?") +
  scale_fill_viridis(discrete = TRUE, end = 0.8) +
  theme_Publication() +
  theme(legend.position = "top", legend.box = "vertical") +
  scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control+enu_violin_frequency_pvalue.pdf",
    width = 4.5,
    height = 4)
ggplot(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary, aes(x = condition, y = obs.mut.mean, fill = as.factor(cycle))) + 
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted), size = 1, alpha = 0.5, position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "", y = "Mean Observed Variants Per Pathway", fill = "Cycle", color = "Adjusted p-value") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$p_value))) +
    theme_Publication() +
    theme(legend.position = "none",  # Place legend on the top
          legend.box = "vertical") +  # Stack legend items horizontally
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()
```

#### O/E ratio
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/shuffled")

pdf("control+enu_violin_all.pdf",
    width = 4.5,
    height = 4.5)
ggplot(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary, aes(x = condition, y = log2fc.mean, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted), size = 1, alpha = 0.5, position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", color = "Adjusted p-value") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$p_value))) +
    theme_Publication() +
    theme(legend.position = "none",  # Place legend on the top
          legend.box = "vertical") +  # Stack legend items horizontally
    scale_y_continuous(breaks = seq(-2, 2, 1), lim = c(-2, 1.5)) +
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control+enu_violin_legend.pdf",
    width = 7,
    height = 5)
ggplot(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary, aes(x = condition, y = log2fc.mean, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted), size = 1, alpha = 0.25, position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", color = "Adjusted p-value") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$p_value))) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "horizontal") +  # Stack legend items horizontally
    scale_y_continuous(breaks = seq(-2, 2, 1), lim = c(-2, 1.5)) +
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control_violin_oe.pdf",
    width = 4,
    height = 5)
ggplot(subset(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary,
              condition == "control"), aes(x = condition, y = log2fc.mean, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted,
                    shape = observed.variant.requirement.met,
                    group = interaction(condition, cycle)), size = 2, alpha = 0.5, 
                position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Control Cells",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", 
         color = "Adjusted p-value",
         shape = "Observed Variant\nCount Sufficient?") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$p_value))) +
    scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "vertical",
          axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12))  # Adjust legend title size
dev.off()

pdf("enu_violin_oe.pdf",
    width = 4,
    height = 5)
ggplot(subset(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary,
              condition == "enu"), aes(x = condition, y = log2fc.mean, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
   geom_jitter(data = subset(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary,
                            condition == "enu" & observed.variant.requirement.met == FALSE),
              aes(color = p_adjusted, shape = observed.variant.requirement.met,
                  group = interaction(condition, cycle)),
              size = 2, alpha = 0.25,
              position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +

  geom_jitter(data = subset(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary,
                            condition == "enu" & observed.variant.requirement.met == TRUE),
              aes(color = p_adjusted, shape = observed.variant.requirement.met,
                  group = interaction(condition, cycle)),
              size = 2, alpha = 0.9,
              position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "ENU-Treated Cells",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", 
         color = "Adjusted p-value",
         shape = "Observed Variant\nCount Sufficient?") +
    scale_fill_viridis(discrete = TRUE, end = 0.8, begin = 0.2) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary$p_value))) +
    scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "vertical",
          axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12))  # Adjust legend title size
dev.off()
```

##### Intron normalization
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/shuffled")

# normalize at the summary level
go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm <- go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary %>%
  merge(go_expressed_regions_exons_introns_promoters_enhancers_shuffled_avg_summary[, c("condition", "cycle", "region", "log2fc.mean")], 
        by = c("condition", "cycle", "region")) %>%
  mutate(`log2fc.norm` = `log2fc.mean.x` - `log2fc.mean.y`) %>%
  mutate(`log2fc.norm.sd` = sd(log2fc.norm, na.rm = TRUE))

go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm$observed.variant.requirement.met <- factor(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm$observed.variant.requirement.met,
                                                         levels = c(TRUE, FALSE))

write.csv(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm, "go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm.csv")

pdf("control+enu_violin_oe_norm.pdf",
    width = 4.5,
    height = 4.5)
ggplot(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm, aes(x = condition, y = log2fc.norm, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted), size = 1, alpha = 0.5, position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", color = "Adjusted p-value") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm$p_value))) +
    theme_Publication() +
    theme(legend.position = "none",  # Place legend on the top
          legend.box = "vertical") +  # Stack legend items horizontally
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control+enu_violin_legend_oe_norm.pdf",
    width = 7,
    height = 5)
ggplot(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm, aes(x = condition, y = log2fc.norm, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted), size = 1, alpha = 0.25, position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", color = "Adjusted p-value") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm$p_value))) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "horizontal") +  # Stack legend items horizontally
    scale_x_discrete(labels = c("Control", "ENU"))
dev.off()

pdf("control_violin_oe_norm.pdf",
    width = 4,
    height = 5)
ggplot(subset(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm,
              condition == "control"), aes(x = condition, y = log2fc.norm, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
    geom_jitter(aes(color = p_adjusted,
                    shape = observed.variant.requirement.met,
                    group = interaction(condition, cycle)), size = 2, alpha = 0.5, 
                position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Control Cells",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", 
         color = "Adjusted p-value",
         shape = "Observed Variant\nCount Sufficient?") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm$p_value))) +
    scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "vertical",
          axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12))  # Adjust legend title size
dev.off()

pdf("enu_violin_oe_norm.pdf",
    width = 4,
    height = 5)
ggplot(subset(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm,
              condition == "enu"), aes(x = condition, y = log2fc.norm, fill = as.factor(cycle))) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_violin(position = position_dodge(width = 0.85), scale = "width", width = 0.75, draw_quantiles = c(0.5)) +
  geom_jitter(
  data = subset(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm,
                condition == "enu"),
  aes(shape = observed.variant.requirement.met,
      group = interaction(condition, cycle),
      alpha = jitter_alpha,
      color = p_adjusted),
  size = 1.5,
  position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85),
  show.legend = FALSE  # optional if alpha legend not needed
) +
  #  geom_jitter(data = subset(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm,
  #                           condition == "enu" & observed.variant.requirement.met == FALSE),
  #             aes(color = p_adjusted, shape = observed.variant.requirement.met,
  #                 group = interaction(condition, cycle)),
  #             size = 2, alpha = 0.25,
  #             position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +
  # 
  # geom_jitter(data = subset(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm,
  #                           condition == "enu" & observed.variant.requirement.met == TRUE),
  #             aes(color = p_adjusted, shape = observed.variant.requirement.met,
  #                 group = interaction(condition, cycle)),
  #             size = 2, alpha = 0.9,
  #             position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.85)) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "ENU-Treated Cells",
         y = expression(bold(atop(Log[2] * "(Observed / Expected)", "SNVs Per Pathway"))),
         fill = "Cycle", 
         color = "Adjusted p-value",
         shape = "Observed Variant\nCount Sufficient?") +
    scale_fill_viridis(discrete = TRUE, end = 0.8, begin = 0.2) +
    scale_color_gradientn(colors = c("red", "black"), 
                          limits = c(min(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm$p_value), 
                                     max(go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm$p_value))) +
    scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 15), 
                       labels = c(`TRUE` = "True", `FALSE` = "False")) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the top
          legend.box = "vertical",
          axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12))  # Adjust legend title size
dev.off()
```

### Observed vs shuffled

#### Correlation
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/observed_vs_shuffled")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/observed_vs_shuffled")

observed_vs_shuffled <- left_join(go_expressed_regions_exons_promoters_enhancers_summary_norm,
                                  go_expressed_regions_exons_promoters_enhancers_shuffled_avg_summary_norm,
                                  by = c("condition", "cycle", "region"))

pdf("control_vs_shuffled_log2fc.norm.pdf",
    width = 5,
    height = 5)
ggplot(subset(observed_vs_shuffled, condition == "enu" & cycle == "9"), aes(x = log2fc.norm.x, y = log2fc.norm.y)) + 
    geom_point() +
    geom_smooth(method=lm) +
    stat_cor() + 
    labs(x = "Observed",
         y = "Shuffled",
         title = "ENU Cycle 9",
         subtitle = expression(Log[2] * "(Observed / Expected) SNVs Per Pathway")) + 
    theme_Publication()
dev.off()

pdf("control_vs_shuffled_pvalue.pdf",
    width = 5,
    height = 5)
ggplot(subset(observed_vs_shuffled, condition == "enu" & cycle == "9"), aes(x = p_value.x, y = p_value.y)) + 
    geom_point() +
    geom_smooth(method=lm) +
    stat_cor() + 
    labs(x = "Observed",
         y = "Shuffled",
         title = "ENU Cycle 9",
         subtitle = "P-value") + 
    theme_Publication()
dev.off()
```

#### Overlap
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/3-ratio-plot/go_expressed_regions/observed_vs_shuffled")

# get significant pathways
observed_significant <- observed_vs_shuffled %>%
  filter(p_adjusted.x < 0.05) %>%
  pull(region)

shuffled_significant <- observed_vs_shuffled %>%
  filter(p_adjusted.y < 0.05) %>%
  pull(region)

pdf("control_vs_shuffled_overlap_0.05.pdf",
    width = 4,
    height = 4)
ggvenn(list(Observed = observed_significant,
            Shuffled = shuffled_significant),
       auto_scale = TRUE) 
dev.off()

# get significant pathways
observed_significant <- observed_vs_shuffled %>%
  filter(p_adjusted.x < 0.1) %>%
  pull(region)

shuffled_significant <- observed_vs_shuffled %>%
  filter(p_adjusted.y < 0.1) %>%
  pull(region)

pdf("control_vs_shuffled_overlap_0.1.pdf",
    width = 4,
    height = 4)
ggvenn(list(Observed = observed_significant,
            Shuffled = shuffled_significant),
       auto_scale = TRUE) 
dev.off()
```