---
title: "mutation-dsitribution - plot-ratio-snv - multiple ENU"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

## Load libraries
```{r}
library(ggplot2)
library(stringr)
library(dplyr)
library(ggpubr)
library(purrr)
library(rstatix)
library(EnhancedVolcano)
library(clusterProfiler)
library(enrichplot)
library(UpSetR)
library(viridis)

hg19_size <- 3036303846
```

## Functions
```{r}
# function to calculate ratio with results coming from
# shuffle, downsample, and observed
# this is for control vs enu since these are the only groups in shuffled
combine_observed_downsampled <- function(observed, downsample, region_levels, sample_table) {
  combined_df <- rbind(observed, downsample)
  combined_df$group <- c(rep("observed", nrow(observed)), rep("downsampled", nrow(downsample)))
  combined_df$group <- factor(combined_df$group, levels = c("observed", "downsampled"))
  # combined_df <- combined_df %>%
  #   mutate(Condition = 
  #            case_when(
  #              grepl("control", sample) ~ "Control",
  #              grepl("enu", sample) ~ "ENU",
  #              grepl("C", sample) ~ "Control",
  #              grepl("E", sample) ~ "ENU",
  #              .default = "NA"
  #            )
  #          )
  
  combined_df$region <- factor(combined_df$region, levels = region_levels)
  
  combined_df <- merge(combined_df, sample_table, by = "sample")
  
  return(combined_df)
}

combine_pvalues_fisher <- function(p_values) {
    chi_sq_stat = -2 * sum(log(p_values))
    degrees_of_freedom = 2 * length(p_values)
    p_combined = pchisq(chi_sq_stat, degrees_of_freedom, lower.tail = FALSE)
    return(p_combined)
}

# function to calculate significance with results coming from
# observed and downsampled
ratio_significance <- function(combined_regions, group_vars, region_levels, by.region = TRUE) {
  
  if(by.region){
  res <- combined_regions %>%
    group_by_at(vars(one_of(c("group", "region", group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue.adj)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  res$region <- factor(res$region, levels = region_levels)
  } else{
    res <- combined_regions %>%
    group_by_at(vars(one_of(c("group", group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue.adj)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  }
  return(res)
}

ratio_significance_go <- function(combined_regions, group_vars, region_levels, by.region = TRUE) {
  
  if(by.region){
  res <- combined_regions %>%
    group_by_at(vars(one_of(c("group", "region", group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  res$region <- factor(res$region, levels = region_levels)
  } else{
    res <- combined_regions %>%
    group_by_at(vars(one_of(c("group", group_vars)))) %>%
    summarise(pvalue = combine_pvalues_fisher(pvalue.adj)) %>%
    mutate(pvalue_symbol = case_when(
      pvalue <= 0.0001 ~ '****',
      pvalue <= 0.001 ~ '***',
      pvalue <= 0.01 ~ '**',
      pvalue <= 0.05 ~ '*',
      TRUE ~ NA
    ))
  }
  return(res)
}

group_significance <- function(combined_regions, group_vars, condition, region_levels){
  formula <- as.formula(paste("log2fc ~", condition))
  
  results <- combined_regions %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(
      p_value = t.test(formula = formula, data = cur_data())$p.value
    ) %>%
    ungroup() %>%
    mutate(p_adjusted = p.adjust(p_value, method = "BH"),
           significant_p_adjusted = case_when(
                            p_adjusted <= 0.0001 ~ '****',
                            p_adjusted <= 0.001 ~ '***',
                            p_adjusted <= 0.01 ~ '**',
                            p_adjusted <= 0.05 ~ '*',
                            TRUE ~ NA_character_)
    )
  
  # Identify regions with p_adjusted < 0.05 in both groups
  regions_significant_in_both_groups <- results %>%
    filter(p_adjusted < 0.05) %>%
    dplyr::count(region) %>%
    filter(n == 2) %>%
    pull(region)
  
  # Create new column based on the condition
  results <- results %>%
    mutate(significant_p_adjusted_both_groups = ifelse(region %in% regions_significant_in_both_groups, significant_p_adjusted, NA_character_))

  # Order by region
  results$region <- factor(results$region, levels = region_levels)
  
  return(results)
}

# function to calculate significance between condition groups
# also reports significance if seen in both downsampled and observed groups
# adjusted p-values by bejamini-hochberg
# combined_regions <- go.expressed_regions_combined
# group_vars <- c("group", "region")
# condition <- "condition"
# region_levels <- go.expressed_regions_levels

group_significance_go <- function(combined_regions, group_vars, condition, region_levels){
  formula <- as.formula(paste("log2fc ~", condition))
  
  results <- combined_regions %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(
      p_value = t.test(formula = formula, data = cur_data())$p.value
    ) %>%
    ungroup() %>%
    mutate(
           significant_p_adjusted = case_when(
                            p_value <= 0.0001 ~ '****',
                            p_value <= 0.001 ~ '***',
                            p_value <= 0.01 ~ '**',
                            p_value <= 0.05 ~ '*',
                            TRUE ~ NA_character_)
    )
  
  # Identify regions with p_adjusted < 0.05 in both groups
  regions_significant_in_both_groups <- as.data.frame(results) %>%
    filter(p_value < 0.05) %>%
    dplyr::count(region) %>%
    filter(n == 2) %>%
    pull(region)
  
  # Create new column based on the condition
  results <- results %>%
    mutate(significant_p_adjusted_both_groups = ifelse(region %in% regions_significant_in_both_groups, significant_p_adjusted, NA_character_))

  # Order by region
  results$region <- factor(results$region, levels = region_levels)
  
  return(results)
}

calculate_stat_summary <- function(observed, group_vars, mut_var){
  results <- observed %>%
    group_by_at(vars(one_of(group_vars))) %>%
    summarise(mean = mean(!!sym(mut_var), na.rm = TRUE),
              sd = sd(!!sym(mut_var), na.rm = TRUE))
}

theme_Publication <- function(base_size=16, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family = "")
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               #legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_fill_Publication_2 <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#7fc97f","#ef3b2c","#386cb0","#fdb462","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

wrap_labels <- function(labels, width = 5) {
  wrapped_labels <- str_wrap(labels, width)
  names(wrapped_labels) <- names(labels)
  return(wrapped_labels)
}
```

## Load sample table
```{r}
sample_table <- read.table("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/1-data/sample_table_filtered.txt", header = TRUE, sep = "\t")
```

## Load genomic distribution results
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/2-calculate-ratio/snv")
# essential
  # observed
  essential_genes <- read.csv("essential_genes_regions_res_snv.csv", header = TRUE, row.names = NULL)
  # downsample
  essential_genes_downsample <- read.csv("essential_genes_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)

# expressed_nonexpressed
  # observed
  expressed_nonexpressed_regions <- read.csv("expressed_nonexpressed_regions_res_snv.csv", header = TRUE, row.names = NULL)
  # downsample
  expressed_nonexpressed_regions_downsample <- read.csv("expressed_nonexpressed_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)
  
# go
  go_ids <- read.csv("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/1-load-data/expressed_genes_gene_ontology.csv", header = TRUE)
  
  # observed
  go.expressed_regions <- read.csv("go.expressed_regions_res_snv.csv", header = TRUE, row.names = NULL)
  # downsample
  go.expressed_regions_downsample <- read.csv("go.expressed_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)

# go shuffle
  # observed
  go.expressed.shuffled_regions <- read.csv("go.expressed.shuffled_regions_res_snv.csv", header = TRUE, row.names = NULL)
  # downsample
  go.expressed.shuffled_regions_downsample <- read.csv("go.expressed.shuffled_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)
  
# kegg
  # observed
  kegg.expressed_regions <- read.csv("kegg.expressed_regions_res_snv.csv", header = TRUE, row.names = NULL)
  # downsample
  kegg.expressed_regions_downsample <- read.csv("kegg.expressed_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)

# kegg shuffle
  # observed
  kegg.expressed.shuffled_regions <- read.csv("kegg.expressed.shuffled_regions_res_snv.csv", header = TRUE, row.names = NULL)
  # downsample
  kegg.expressed.shuffled_regions_downsample <- read.csv("kegg.expressed.shuffled_regions_res_downsample_snv.csv", header = TRUE, row.names = NULL)
```

# Analysis

## Expressed and Nonexpressed genes

### By condition
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/expressed_nonexpressed")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/expressed_nonexpressed")

expressed_nonexpressed_regions_levels <- c("expressed", "nonexpressed")

expressed_nonexpressed_regions_combined <- combine_observed_downsampled(expressed_nonexpressed_regions, 
                                                expressed_nonexpressed_regions_downsample, 
                                                expressed_nonexpressed_regions_levels, 
                                                sample_table)

expressed_nonexpressed_regions_ratio_significance <- ratio_significance(expressed_nonexpressed_regions_combined,
                                                   "condition",
                                                   expressed_nonexpressed_regions_levels)

# expressed_nonexpressed_regions_group_significance <- group_significance(expressed_nonexpressed_regions_combined,
#                                                          c("group", "region"),
#                                                          "condition",
#                                                          expressed_nonexpressed_regions_levels)

# observed
# downsampled
pdf("SNV_expressed_nonexpressed_regions_condition_observed_downsampled.pdf",
    width = 5,
    height = 5)
ggplot(expressed_nonexpressed_regions_combined, aes(x = group, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Group", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    geom_text(data = expressed_nonexpressed_regions_ratio_significance, aes(x = group, y = Inf, label = pvalue_symbol),
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    facet_grid(.~region) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.y = element_text(size = 12))
dev.off()

# observed
pdf("SNV_expressed_nonexpressed_regions_condition_observed.pdf",
    width = 4,
    height = 4)
ggplot(subset(expressed_nonexpressed_regions_combined, group == "observed"), aes(x = region, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    geom_text(data = subset(expressed_nonexpressed_regions_ratio_significance, group == "observed"), 
              aes(x = region, y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    # geom_text(data = subset(expressed_nonexpressed_regions_group_significance, group == "observed"),
    #           aes(x = Inf, y = Inf, fill = NA, label = significant_p_adjusted_both_groups), # pvalue of control vs enu
    #           vjust = 12, hjust = 2, size = 8, color = "firebrick") +
    labs(x = "Region", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_discrete(name = "Condition", labels = c("Control", "ENU")) +
    scale_fill_Publication() +
    theme_Publication() +
    ylim(NA, max(subset(expressed_nonexpressed_regions_combined, group == "observed")$log2fc + 0.1))
dev.off()
```

### By cycle
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/expressed_nonexpressed")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/expressed_nonexpressed")

expressed_nonexpressed_regions_levels <- c("expressed", "nonexpressed")

expressed_nonexpressed_regions_combined <- combine_observed_downsampled(expressed_nonexpressed_regions, 
                                                expressed_nonexpressed_regions_downsample, 
                                                expressed_nonexpressed_regions_levels, 
                                                sample_table)

expressed_nonexpressed_regions_ratio_significance <- ratio_significance(expressed_nonexpressed_regions_combined,
                                                   c("cycle", "condition"),
                                                   expressed_nonexpressed_regions_levels)

# expressed_nonexpressed_regions_group_significance <- group_significance(expressed_nonexpressed_regions,
#                                                          c("group", "region"), 
#                                                          "cycle",
#                                                          expressed_nonexpressed_regions_levels)

# observed
pdf("SNV_expressed_nonexpressed_regions_cycle_observed.pdf",
    width = 10,
    height = 5)
ggplot(subset(expressed_nonexpressed_regions_combined, group == "observed"), aes(x = as.factor(cycle), y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Cycle", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    geom_text(data = subset(expressed_nonexpressed_regions_ratio_significance, group == "observed"),
              aes(x = as.factor(cycle), y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    facet_grid(.~region)
dev.off()

# downsampled
pdf("SNV_expressed_nonexpressed_cycle_downsampled.pdf",
    width = 10,
    height = 5)
ggplot(subset(expressed_nonexpressed_regions_combined, group == "downsampled"), aes(x = as.factor(cycle), y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Cycle", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    geom_text(data = subset(expressed_nonexpressed_regions_ratio_significance, group == "downsampled"),
              aes(x = as.factor(cycle), y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    facet_grid(.~region)
dev.off()
```



## Essential genes

### By condition
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/essential")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/essential")

essential_genes_levels <- c("essential", "nonessential")

essential_genes_combined <- combine_observed_downsampled(essential_genes, 
                                                essential_genes_downsample, 
                                                essential_genes_levels, 
                                                sample_table)

essential_genes_ratio_significance <- ratio_significance(essential_genes_combined,
                                                   "condition",
                                                   essential_genes_levels)

essential_genes_group_significance <- group_significance(essential_genes_combined,
                                                         c("group", "region"),
                                                         "condition",
                                                         essential_genes_levels)

# observed
# downsampled
pdf("SNV_essential_condition_observed_downsampled.pdf",
    width = 5,
    height = 5)
ggplot(essential_genes_combined, aes(x = group, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Group", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    geom_text(data = essential_genes_ratio_significance, aes(x = group, y = Inf, label = pvalue_symbol),
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    facet_grid(.~region) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.y = element_text(size = 12))
dev.off()

# observed
pdf("SNV_essential_condition_observed.pdf",
    width = 4,
    height = 4)
ggplot(subset(essential_genes_combined, group == "observed"), aes(x = region, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    geom_text(data = subset(essential_genes_ratio_significance, group == "observed"),
              aes(x = region, y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    geom_text(data = subset(essential_genes_group_significance, group == "observed"),
              aes(x = Inf, y = Inf, fill = NA, label = significant_p_adjusted_both_groups), # pvalue of control vs enu
              vjust = 12, hjust = 2, size = 8, color = "firebrick") +
    labs(x = "Condition", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_discrete(name = "Condition", labels = c("Control", "ENU")) +
    scale_fill_Publication() +
    theme_Publication() +
    ylim(NA, max(subset(essential_genes_combined, group == "observed")$log2fc + 0.1))
dev.off()
```

### By cycle
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/essential")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/essential")

essential_genes_levels <- c("essential", "nonessential")

essential_genes_combined <- combine_observed_downsampled(essential_genes, 
                                                essential_genes_downsample, 
                                                essential_genes_levels, 
                                                sample_table)

essential_genes_ratio_significance <- ratio_significance(essential_genes_combined,
                                                   c("cycle", "condition"),
                                                   essential_genes_levels)

# essential_genes_group_significance <- group_significance(essential_genes,
#                                                          c("group", "region"), 
#                                                          "cycle",
#                                                          essential_genes_levels)

# observed
pdf("SNV_essential_cycle_observed.pdf",
    width = 10,
    height = 5)
ggplot(subset(essential_genes_combined, group == "observed"), aes(x = as.factor(cycle), y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Cycle", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    geom_text(data = subset(essential_genes_ratio_significance, group == "observed"),
              aes(x = as.factor(cycle), y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    facet_grid(.~region)
dev.off()

# downsampled
pdf("SNV_essential_cycle_downsampled.pdf",
    width = 10,
    height = 5)
ggplot(subset(essential_genes_combined, group == "downsampled"), aes(x = as.factor(cycle), y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Cycle", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    geom_text(data = subset(essential_genes_ratio_significance, group == "downsampled"),
              aes(x = as.factor(cycle), y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    facet_grid(.~region)
dev.off()
```



## Gene ontology
- Improvement: Incorporate average expression of each pathway

### Compare ratios wihtin groups

#### ENU Observed
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")

go.expressed_regions_levels <- unique(go.expressed_regions$region)

go.expressed_regions_combined <- combine_observed_downsampled(go.expressed_regions, 
                                                go.expressed_regions_downsample, 
                                                go.expressed_regions_levels, 
                                                sample_table)

go.expressed_regions_ratio_significance <- ratio_significance_go(go.expressed_regions_combined,
                                                   "condition",
                                                   go.expressed_regions_levels)

# summarize by getting average log2fc for each group and then compare groups
go.expressed_regions_combined_summary <- subset(go.expressed_regions_combined, group == "observed" & condition == "enu") %>%
  group_by(region, condition) %>%
  summarise(average_log2fc = mean(log2fc, na.rm = TRUE),
            median.obs.mut = median(obs.mut, na.rm = TRUE))

# add significance values
go.expressed_regions_combined_summary <- merge(go.expressed_regions_combined_summary, 
                                               subset(go.expressed_regions_ratio_significance, group == "observed" & condition == "enu"), 
                                               by = "region")
# add GO IDs
go.expressed_regions_combined_summary$GO.ID <- go_ids$ID[match(go.expressed_regions_combined_summary$region, go_ids$Description)]

# add size of gene set
go.expressed_regions_combined_summary$genes <- go_ids$geneID[match(go.expressed_regions_combined_summary$region, go_ids$Description)]

go.expressed_regions_combined_summary$gene_length <- unlist(lapply(strsplit(go.expressed_regions_combined_summary$genes, "/"), length))

# add p-value correction
go.expressed_regions_combined_summary$pvalue.adjust <- p.adjust(go.expressed_regions_combined_summary$pvalue, method = "BH")

# filter out genes with 0 median obs mutations
go.expressed_regions_combined_summary <- go.expressed_regions_combined_summary %>%
  filter(median.obs.mut >= 1)

# get significant hits
go.expressed_regions_combined_summary.sig <- go.expressed_regions_combined_summary %>%
  filter(pvalue.adjust <= 0.05)

# enu vs control
write.csv(go.expressed_regions_combined_summary, "enu_ratio_sig.csv", row.names = FALSE)

# volcano
pdf("enu_ratio_sig.pdf",
    width = 6,
    height = 6)
EnhancedVolcano(go.expressed_regions_combined_summary,
                lab = go.expressed_regions_combined_summary$region,
                title = "ENU-treated cells",
                subtitle = "Log2(Expected vs Observed SNV ratios)",
                caption = NULL,
                x = 'average_log2fc',
                y = 'pvalue',
                pCutoffCol = "pvalue.adjust",
                pointSize = 4.0,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 4,
                labCol = 'black',
                #labFace = 'bold',
                #boxedLabels = TRUE,
                drawConnectors = TRUE,
                #widthConnectors = 1.0,
                #colConnectors = 'black',
                #maxoverlapsConnectors = Inf,
                #selectLab = KO_vs_WT$volcano.label,
                ylim = c(0,7),
                # xlim = c(-3, 1.5),
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
dev.off()


pdf("enu_ratio_sig_pvalue_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go.expressed_regions_combined_summary$pvalue,
     go.expressed_regions_combined_summary$gene_length)
dev.off()

pdf("enu_ratio_sig_log2fc_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go.expressed_regions_combined_summary$average_log2fc,
     go.expressed_regions_combined_summary$gene_length)
dev.off()

pdf("enu_ratio_sig_pvalue_vs_obs_mut.pdf",
    width = 4,
    height = 4)
plot(go.expressed_regions_combined_summary$pvalue,
     go.expressed_regions_combined_summary$median.obs.mut)
dev.off()

pdf("enu_ratio_gene_length_vs_obs_mut.pdf",
    width = 4,
    height = 4)
plot(go.expressed_regions_combined_summary$gene_length,
     go.expressed_regions_combined_summary$median.obs.mut)
dev.off()

pdf("enu_ratio_obs_mut_vs_log2fc.pdf",
    width = 4,
    height = 4)
plot(go.expressed_regions_combined_summary$median.obs.mut,
     go.expressed_regions_combined_summary$average_log2fc)
dev.off()
```

#### ENU Shuffled
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")

go.expressed.shuffled_regions_levels <- unique(go.expressed.shuffled_regions$region)

go.expressed.shuffled_regions_combined <- combine_observed_downsampled(go.expressed.shuffled_regions, 
                                                go.expressed.shuffled_regions_downsample, 
                                                go.expressed.shuffled_regions_levels, 
                                                sample_table)

go.expressed.shuffled_regions_ratio_significance <- ratio_significance_go(go.expressed.shuffled_regions_combined,
                                                   "condition",
                                                   go.expressed.shuffled_regions_levels)

# summarize by getting average log2fc for each group and then compare groups
go.expressed.shuffled_regions_combined_summary <- subset(go.expressed.shuffled_regions_combined, group == "observed" & condition == "enu") %>%
  group_by(region, condition) %>%
  summarise(average_log2fc = mean(log2fc, na.rm = TRUE),
            median.obs.mut = median(obs.mut, na.rm = TRUE))

# add significance values
go.expressed.shuffled_regions_combined_summary <- merge(go.expressed.shuffled_regions_combined_summary, 
                                               subset(go.expressed.shuffled_regions_ratio_significance, group == "observed" & condition == "enu"), 
                                               by = "region")
# add GO IDs
go.expressed.shuffled_regions_combined_summary$GO.ID <- go_ids$ID[match(go.expressed.shuffled_regions_combined_summary$region, go_ids$Description)]

# add size of gene set
go.expressed.shuffled_regions_combined_summary$genes <- go_ids$geneID[match(go.expressed.shuffled_regions_combined_summary$region, go_ids$Description)]

go.expressed.shuffled_regions_combined_summary$gene_length <- unlist(lapply(strsplit(go.expressed.shuffled_regions_combined_summary$genes, "/"), length))

# add p-value correction
go.expressed.shuffled_regions_combined_summary$pvalue.adjust <- p.adjust(go.expressed.shuffled_regions_combined_summary$pvalue, method = "BH")

# filter out genes with 0 median obs mutations
go.expressed.shuffled_regions_combined_summary <- go.expressed.shuffled_regions_combined_summary %>%
  filter(median.obs.mut >= 1)

# get significant hits
go.expressed.shuffled_regions_combined_summary.sig <- go.expressed.shuffled_regions_combined_summary %>%
  filter(pvalue.adjust <= 0.05)

# enu vs control
write.csv(go.expressed.shuffled_regions_combined_summary, "enu_ratio_sig_shuffled.csv", row.names = FALSE)
pdf("enu_ratio_sig_shuffled.pdf",
    width = 6,
    height = 6)
EnhancedVolcano(go.expressed.shuffled_regions_combined_summary,
                lab = go.expressed.shuffled_regions_combined_summary$region,
                title = "ENU-treated cells - shuffled pathways",
                subtitle = "Log2(Expected vs Observed SNV ratios)",
                caption = NULL,
                x = 'average_log2fc',
                y = 'pvalue',
                pCutoffCol = "pvalue.adjust",
                pointSize = 4.0,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 4,
                labCol = 'black',
                #labFace = 'bold',
                #boxedLabels = TRUE,
                drawConnectors = TRUE,
                #widthConnectors = 1.0,
                #colConnectors = 'black',
                #maxoverlapsConnectors = Inf,
                #selectLab = KO_vs_WT$volcano.label,
                ylim = c(0,7),
                # xlim = c(-3, 1.5),
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
dev.off()

pdf("enu_ratio_sig_shuffled_pvalue_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go.expressed.shuffled_regions_combined_summary$pvalue,
     go.expressed.shuffled_regions_combined_summary$gene_length)
dev.off()

pdf("enu_ratio_sig_shuffled_log2fc_vs_gene_length.pdf",
    width = 4,
    height = 4)
plot(go.expressed.shuffled_regions_combined_summary$average_log2fc,
     go.expressed.shuffled_regions_combined_summary$gene_length)
dev.off()

# overlap between two list


```

#### Overlap of genes in significant pathways
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/1-load-data")

# filter significant hits in real based on shuffle
go.expressed_regions_combined_summary.sig <- go.expressed_regions_combined_summary.sig %>%
  filter(!GO.ID %in% go.expressed.shuffled_regions_combined_summary.sig$GO.ID)

# granges 
go.expressed <- readRDS("go.expressed.RDS")

# enrichment results
enrich.go.expressed <- readRDS("enrich.go.expressed.RDS")

# filter for sig hits
go.expressed.sig <- go.expressed[names(go.expressed) %in% go.expressed_regions_combined_summary.sig$region]

enrich.go.expressed.sig <- filter(enrich.go.expressed, ID %in% go.expressed_regions_combined_summary.sig$GO.ID)

# get sequence length distribution and add to results
lengths <- unlist(lapply(go.expressed.sig, function(x) sum(width(x))))
go.expressed_regions_combined_summary.sig$sequence_length <- lengths[match(go.expressed_regions_combined_summary.sig$region, names(lengths))]

setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")

edox <- pairwise_termsim(enrich.go.expressed.sig)
pdf("enu_ratio_sig_filtered_tree.pdf",
    width = 10,
    height = 10)
treeplot(edox)
dev.off()

# using all go terms too complicated
pdf("enu_ratio_sig_filtered_cnet.pdf",
    width = 10,
    height = 10)
cnetplot(edox, 
         showCategory = 5)
dev.off()

# get shared genes amongst significant go terms
genelist <- strsplit(go.expressed_regions_combined_summary.sig$genes, "/")
names(genelist) <- go.expressed_regions_combined_summary.sig$region
genelist <- fromList(genelist)
genelist.reduced <- genelist[rowSums(genelist[, names(genelist)]) > 1,] # filter out sets with no overlap between others
                       
pdf("enu_ratio_sig_filtered_upset.pdf",
    width = 8,
    height = 5)
upset(genelist.reduced,
      nsets	= 10,
      nintersects = NA,
      order.by = "freq",
      show.numbers = FALSE)
dev.off()

# save
write.csv(go.expressed_regions_combined_summary.sig, "enu_ratio_sig_filtered.csv", row.names = FALSE)
```

#### Final plots
```{r}
# plot bar plot of significant hits
go.expressed_regions_combined_summary.sig <- go.expressed_regions_combined_summary.sig[order(go.expressed_regions_combined_summary.sig$average_log2fc, decreasing = TRUE),]
go.expressed_regions_combined_summary.sig$region <- factor(go.expressed_regions_combined_summary.sig$region, levels = go.expressed_regions_combined_summary.sig$region)

pdf("enu_ratio_sig_filtered_bar.pdf",
    width = 8,
    height = 5)
ggplot(go.expressed_regions_combined_summary.sig[1:16,], aes(x = region, y = average_log2fc, fill = pvalue.adjust)) +
  geom_bar(stat = "identity") +
  scale_y_reverse() +
  coord_flip() +
  scale_fill_viridis() +
  theme_Publication() +
  theme(legend.justification = c("right","top"),
        legend.position = "top",
        legend.title = element_text(size=15),
        legend.text = element_text(size=10),
        axis.title.x = element_text(size = 12)) +
  labs(x = "GO Biological Process", 
       y = "Mean Log2(observed/expected)\nSNVs Per Cell",
       fill = "Adjusted p-value")
dev.off()

# bar plot of number of increased estimated mutations
bar_plot <- subset(go.expressed_regions_combined, group == "observed") %>%
  mutate(est.mut = obs.mut / Cov / Sen_SNV * hg19_size) %>%
  group_by(region, condition) %>%
  summarise(mean.est.mut = mean(est.mut, na.rm = TRUE)) %>%
  group_by(region) %>%
  spread(key = condition, value = mean.est.mut) %>%
  mutate(group_diff = abs(enu - control)) %>%
  filter(region %in% go.expressed_regions_combined_summary.sig[1:16,]$region)

bar_plot <- bar_plot[match(go.expressed_regions_combined_summary.sig[1:16,]$region, bar_plot$region),]
bar_plot$region <- factor(bar_plot$region, levels = bar_plot$region)
bar_plot$group_diff

pdf("enu_number_sig_filtered_bar.pdf",
    width = 8,
    height = 4.5)
ggplot(bar_plot, aes(x = region, y = group_diff)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_Publication() +
  theme( 
        axis.title.x = element_text(size = 12)) +
  labs(x = "GO Biological Process", 
       y = "Mean Estimated SNV increase\nin ENU-treated cells")
dev.off()
```

### Compare ratios between groups

#### Observed
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")

go.expressed_regions_levels <- unique(go.expressed_regions$region)

go.expressed_regions_combined <- combine_observed_downsampled(go.expressed_regions, 
                                                go.expressed_regions_downsample, 
                                                go.expressed_regions_levels, 
                                                sample_table)

go.expressed_regions_group_significance <- group_significance_go(go.expressed_regions_combined,
                                                         c("group", "region"),
                                                         "condition",
                                                         go.expressed_regions_levels)

# summarize by getting average log2fc for each group and then compare groups
go.expressed_regions_combined_summary <- subset(go.expressed_regions_combined, group == "observed") %>%
  group_by(region, condition) %>%
  summarise(average_log2fc = mean(log2fc, na.rm = TRUE)) %>%
  group_by(region) %>%
  spread(key = condition, value = average_log2fc) %>%
  mutate(group_diff = enu - control)

# add group significance values
go.expressed_regions_combined_summary <- merge(go.expressed_regions_combined_summary, 
                                               subset(go.expressed_regions_group_significance, group == "observed"), 
                                               by = "region")
# add GO IDs
go.expressed_regions_combined_summary$GO.ID <- go_ids$ID[match(go.expressed_regions_combined_summary$region, go_ids$Description)]

# add p-value corerection
go.expressed_regions_combined_summary$p_value.adjust <- p.adjust(go.expressed_regions_combined_summary$p_value, method = "BH")

# enu vs control
write.csv(go.expressed_regions_combined_summary, "enu_vs_control_ratio_group_sig.csv", row.names = FALSE)
pdf("enu_vs_control_ratio_group_sig.pdf",
    width = 6,
    height = 6)
EnhancedVolcano(go.expressed_regions_combined_summary,
                lab = go.expressed_regions_combined_summary$region,
                title = "ENU vs Control expected vs observed ratios",
                subtitle = NULL,
                caption = NULL,
                x = 'group_diff',
                y = 'p_value',
                pCutoffCol = "p_value",
                pointSize = 4.0,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 4,
                labCol = 'black',
                labFace = 'bold',
                #boxedLabels = TRUE,
                #drawConnectors = TRUE,
                #widthConnectors = 1.0,
                #colConnectors = 'black',
                #maxoverlapsConnectors = Inf,
                #selectLab = KO_vs_WT$volcano.label,
                ylim = c(0,5),
                xlim = c(-3, 1.5),
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
dev.off()
```

#### Downsample
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")

go.expressed_regions_levels <- unique(go.expressed_regions$region)

go.expressed_regions_combined <- combine_observed_downsampled(go.expressed_regions, 
                                                go.expressed_regions_downsample, 
                                                go.expressed_regions_levels, 
                                                sample_table)

go.expressed_regions_group_significance <- group_significance_go(go.expressed_regions_combined,
                                                         c("group", "region"),
                                                         "condition",
                                                         go.expressed_regions_levels)

# summarize by getting average log2fc for each group and then compare groups
go.expressed_regions_combined_summary <- subset(go.expressed_regions_combined, group == "downsampled") %>%
  group_by(region, condition) %>%
  summarise(average_log2fc = mean(log2fc, na.rm = TRUE)) %>%
  group_by(region) %>%
  spread(key = condition, value = average_log2fc) %>%
  mutate(group_diff = enu - control) # relative change when subtracting two log2fc

# add group significance values
go.expressed_regions_combined_summary <- merge(go.expressed_regions_combined_summary, 
                                               subset(go.expressed_regions_group_significance, group == "downsampled"), 
                                               by = "region")
# add GO IDs
go.expressed_regions_combined_summary$GO.ID <- go_ids$ID[match(go.expressed_regions_combined_summary$region, go_ids$Description)]

# add p-value corerection
go.expressed_regions_combined_summary$p_value.adjust <- p.adjust(go.expressed_regions_combined_summary$p_value, method = "BH")

# enu vs control
write.csv(go.expressed_regions_combined_summary, "enu_vs_control_ratio_group_sig_downsample.csv", row.names = FALSE)
pdf("enu_vs_control_ratio_group_sig_downsample.pdf",
    width = 6,
    height = 6)
EnhancedVolcano(go.expressed_regions_combined_summary,
                lab = go.expressed_regions_combined_summary$region,
                title = "ENU vs Control expected vs observed ratios",
                subtitle = NULL,
                caption = NULL,
                x = 'group_diff',
                y = 'p_value',
                pCutoffCol = "p_value",
                pointSize = 4.0,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 4,
                labCol = 'black',
                labFace = 'bold',
                #boxedLabels = TRUE,
                #drawConnectors = TRUE,
                #widthConnectors = 1.0,
                #colConnectors = 'black',
                #maxoverlapsConnectors = Inf,
                #selectLab = KO_vs_WT$volcano.label,
                ylim = c(0,5),
                xlim = c(-3, 1.5),
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
dev.off()
```

### Top pathways
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")

go.expressed_regions_levels <- unique(go.expressed_regions$region)

go.expressed_regions_combined <- combine_observed_downsampled(go.expressed_regions, 
                                                go.expressed_regions_downsample, 
                                                go.expressed_regions_levels, 
                                                sample_table)

go.expressed_regions_ratio_significance <- ratio_significance(go.expressed_regions_combined,
                                                   "condition",
                                                   go.expressed_regions_levels)

go.expressed_regions_group_significance <- group_significance_go(go.expressed_regions_combined,
                                                         c("group", "region"),
                                                         "condition",
                                                         go.expressed_regions_levels)

# observed
# downsampled
pdf("SNV_go.expressed_regions_condition_observed_downsampled.pdf",
    width = 50,
    height = 50)
ggplot(go.expressed_regions_combined, aes(x = group, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Group", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    geom_text(data = go.expressed_regions_ratio_significance, aes(x = group, y = Inf, label = pvalue_symbol),
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    facet_wrap(.~region) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.y = element_text(size = 12))
dev.off()

# observed
pdf("SNV_go.expressed_regions_condition_observed.pdf",
    width = 50,
    height = 50)
ggplot(subset(go.expressed_regions_combined, group == "observed"), aes(x = group, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Group", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    geom_text(data = subset(go.expressed_regions_ratio_significance, group == "observed"), aes(x = group, y = Inf, label = pvalue_symbol),
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
      geom_text(data = subset(chromhmm_regions_group_significance, group == "observed"),
              aes(x = "control", y = Inf, fill = NA, label = significant_p_adjusted_both_groups), # pvalue of control vs enu
              vjust = 8, hjust = -0.5, size = 8, color = "firebrick") +
    facet_wrap(.~region) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.y = element_text(size = 12))
dev.off()
```

#### Top pathways ratio sig
```{r}
# top 10 significant
dist_test_combine_enu_top_10 <- go.expressed_regions_combined %>%
  filter(condition == 'enu' &
          group == "observed") %>%    # Filter for rows where condition is 'enu'
  arrange(pvalue) %>%           # Sort these rows by the pval column in descending order
  slice_head(n = 10)                # Get the first 10 rows after sorting
unique_regions <- distinct(dist_test_combine_enu_top_10, region)

non_enu_data <- go.expressed_regions_combined %>%
  filter(condition != 'enu' &
          group == "observed") %>%  # Keep the rest of the data as is
  semi_join(unique_regions, by = "region")
  
dist_test_combine_top_10 <- bind_rows(dist_test_combine_enu_top_10, non_enu_data)

dist_test_combine_top_10_ratio_significance <- ratio_significance(dist_test_combine_top_10,
                                                   "condition",
                                                   go.expressed_regions_levels)

# dist_test_combine_top_10_group_significance <- group_significance_go(dist_test_combine_top_10,
#                                                          c("group", "region"),
#                                                          "condition",
#                                                          go.expressed_regions_levels)

# observed
pdf("SNV_go.expressed_regions_condition_observed_top_ratio_sig.pdf",
    width = 10,
    height = 10)
ggplot(subset(dist_test_combine_top_10, group == "observed"), aes(x = group, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Group", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    geom_text(data = subset(dist_test_combine_top_10_ratio_significance, group == "observed"), aes(x = group, y = Inf, label = pvalue_symbol),
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
      # geom_text(data = subset(chromhmm_regions_group_significance, group == "observed"),
      #         aes(x = "control", y = Inf, fill = NA, label = significant_p_adjusted_both_groups), # pvalue of control vs enu
      #         vjust = 8, hjust = -0.5, size = 8, color = "firebrick") +
    facet_wrap(.~region) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.y = element_text(size = 12))
dev.off()
```

#### Top pathways group sig
```{r}
top10_group_sig <- go.expressed_regions_group_significance[order(go.expressed_regions_group_significance$p_value),]$region[1:10]

# observed
pdf("SNV_go.expressed_regions_condition_observed_top_group_sig.pdf",
    width = 10,
    height = 10)
ggplot(subset(go.expressed_regions_combined, region %in% top10_group_sig), aes(x = group, y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Group", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    # geom_text(data = subset(go.expressed_regions_ratio_significance, region %in% top10_group_sig), aes(x = group, y = Inf, label = pvalue_symbol),
    #           vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    #   geom_text(data = subset(go.expressed_regions_group_significance, region %in% top10_group_sig),
    #           aes(x = "control", y = Inf, fill = NA, label = significant_p_adjusted_both_groups), # pvalue of control vs enu
    #           vjust = 8, hjust = -0.5, size = 8, color = "firebrick") +
    facet_wrap(.~region) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.y = element_text(size = 12))
dev.off()
```

### By cycle
```{r}
dir.create("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/231009\ multiple\ ENU\ analysis/6-pathway-overlap/3-ratio-plot/go.expressed_regions")

go.expressed_regions_levels <- c("expressed", "nonexpressed")

go.expressed_regions_combined <- combine_observed_downsampled(go.expressed_regions, 
                                                go.expressed_regions_downsample, 
                                                go.expressed_regions_levels, 
                                                sample_table)

go.expressed_regions_ratio_significance <- ratio_significance(go.expressed_regions_combined,
                                                   c("cycle", "condition"),
                                                   go.expressed_regions_levels)

# go.expressed_regions_group_significance <- group_significance(go.expressed_regions,
#                                                          c("group", "region"), 
#                                                          "cycle",
#                                                          go.expressed_regions_levels)

# observed
pdf("SNV_go.expressed_regions_cycle_observed.pdf",
    width = 10,
    height = 5)
ggplot(subset(go.expressed_regions_combined, group == "observed"), aes(x = as.factor(cycle), y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Cycle", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    geom_text(data = subset(go.expressed_regions_ratio_significance, group == "observed"),
              aes(x = as.factor(cycle), y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    facet_grid(.~region)
dev.off()

# downsampled
pdf("SNV_go.expressed_regions_cycle_downsampled.pdf",
    width = 10,
    height = 5)
ggplot(subset(go.expressed_regions_combined, group == "downsampled"), aes(x = as.factor(cycle), y = log2fc, fill = condition)) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.5) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Cycle", y = "Log2(observed/expected)\nSNVs Per Cell") +
    scale_fill_Publication() +
    theme_Publication() +
    geom_text(data = subset(go.expressed_regions_ratio_significance, group == "downsampled"),
              aes(x = as.factor(cycle), y = Inf, label = pvalue_symbol), # pvalue of observed/expected
              vjust = 0.8, hjust = 1, position=position_dodge(width = 0.75), size = 6, angle = 90) +
    facet_grid(.~region)
dev.off()
```


