---
title: "mutation-distribution - calculate-ratio - multiple ENU"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

## Load libraries
```{r}
library(VariantAnnotation)
library(GenomicRanges)
library(GenomicFeatures)
library(GenomicScores)
library(rtracklayer)
library(MutationalPatterns)
library(ggplot2)
library(AnnotationHub)
library(tidyr)
library(tools)
library(foreach)
library(doParallel)
```

## Functions
```{r}
# obs = obs_snv[grep("C1_05_24", names(obs_snv))]
# sim = sim_snv[grep("C1_05_24", names(sim_snv))]
# regions = generic_regions[1:2]

calc_OE_ratio <- function(obs, sim, regions){
  
  # Preallocate vector for observed overlap
  obs_overlaps <- numeric(length(regions))
  
  # get number of observed mutations overlapping genomic region of interest
  obs_overlaps <- sapply(regions, function(region) countOverlaps(obs, region, type = "any", ignore.strand = TRUE))
  names(obs_overlaps) <- names(regions)
  
  # Preallocate matrix for simulated overlaps
  sim_overlaps_df <- matrix(nrow = length(sim), ncol = length(regions))
  colnames(sim_overlaps_df) <- names(regions)

  # get number of simulated mutations overlapping genomic region of interest
  # there should be at least 1000 simulation for 1 observed sample
  sim_overlaps_df <- t(sapply(sim, function(sim_item) {
    sapply(regions, function(region) {
      countOverlaps(GRangesList(sim_item), region, type = "any", ignore.strand = TRUE)
    })
  }))
  
  # get normality of distribution for simulated samples
  sim.norm.pvalues <- apply(sim_overlaps_df, 2, function(column) shapiro.test(column)$p.value)
  
  # adjust normality test pvalue for multiple testing
  sim.norm.pvalues.adj <- p.adjust(sim.norm.pvalues, method = "BH")
  
  # create log2 ratio using observed / (average simulated i.e. expected)
  # using median here instead of mean in case distributions are not normal
  sim.overlaps.median <- colMedians(sim_overlaps_df) 
  log2fc <- log2((obs_overlaps + 1) / (sim.overlaps.median + 1))
  
  # permutation test to calculate significance
  # how many simulation had equal or more/less mutations compared to observed (one-sided)
  perm.test.pvalues <- sapply(seq_along(regions), function(i) {
    region_name <- names(regions)[i]
    sim.median <- median(sim_overlaps_df[, region_name])
    sim.log2fc <-  log2((sim.median + 1) / (sim_overlaps_df[, region_name] + 1))
    obs.log2fc <-  log2((obs_overlaps[region_name] + 1) / (sim.median + 1))
    mean(abs(sim.log2fc)>=abs(obs.log2fc))
  })
  names(perm.test.pvalues) <- names(regions)

  # adjust permutation test pvalue for multiple testing
  perm.test.pvalues.adj <- p.adjust(perm.test.pvalues, method = "BH")
  
  res <- data.frame(region = names(regions),
                    sample = names(obs),
                    obs.mut = obs_overlaps,
                    sim.mut = sim.overlaps.median,
                    log2fc = log2fc,
                    pvalue = perm.test.pvalues,
                    pvalue.adj = perm.test.pvalues.adj,
                    sim.distr.norm.pvalue.adj = sim.norm.pvalues.adj)
  
  return(res)
}

calc_OE_ratio_parallel <- function(samples, obs_snv, sim_snv, regions) {
    # Function to process a single sample
    process_sample <- function(sample, obs_snv, sim_snv, regions) {
        obs_sample <- obs_snv[grep(sample, names(obs_snv))]
        sim_sample <- sim_snv[grep(sample, names(sim_snv))]
        calc_OE_ratio(obs = obs_sample, sim = sim_sample, regions = regions)
    }

    # Detect the number of cores and register parallel backend
    no_cores <- detectCores() - 1  # Leaving one core free
    registerDoParallel(no_cores)

    # Apply calc_OE_ratio in parallel and row-bind the results
    results_df <- foreach(sample = samples, .combine = 'rbind', .packages = c("GenomicRanges")) %dopar% {
        process_sample(sample, obs_snv, sim_snv, regions)
    }
    return(results_df)
}

theme_Publication <- function(base_size=16, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family = "")
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               #legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

wrap_labels <- function(labels, width = 5) {
  wrapped_labels <- str_wrap(labels, width)
  names(wrapped_labels) <- names(labels)
  return(wrapped_labels)
}
```

## Load data
```{r}
setwd("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/1-load-data")

sample_table <- read.table("sample_table_filtered.txt", header = TRUE, sep = "\t")

sample_table_downsample <- sample_table[!(sample_table$condition == "enu" & sample_table$cycle == "3"),]

# observed mutations
obs_snv <- readRDS("obs_snv.RDS")
obs_indel <- readRDS("obs_indel.RDS")

# simulated mutations
sim_snv <- readRDS("sim_snv.RDS")
sim_indel <- readRDS("sim_indel.RDS")

# downsampled observed mutations
obs_snv_downsampled <- readRDS("obs_snv_downsampled.RDS")
obs_indel_downsampled <- readRDS("obs_indel_downsampled.RDS")

# downsampled simulated mutations
sim_snv_downsampled <- readRDS("sim_snv_downsampled.RDS")
sim_indel_downsampled <- readRDS("sim_indel_downsampled.RDS")

# annotations
expressed_nonexpressed_regions <- readRDS("expressed_nonexpressed_regions.RDS")
go.expressed_regions <- readRDS("go.expressed.RDS")
go.expressed.shuffled_regions <- readRDS("go.expressed.shuffled.RDS")
kegg.expressed_regions <- readRDS("kegg.expressed.RDS")
kegg.expressed.shuffled_regions <- readRDS("kegg.expressed.shuffled.RDS")
essential_genes_regions <- readRDS("essential_genes.RDS")
marker_genes_regions <- GRangesList(marker_genes = readRDS("marker_genes.RDS"))
```

# Mutation distribution analysis
- Improvement: also get where in the genomic region this is occuring - i.e. the topography (left, right, center)
- Improvement: Examine mutation density vs epigenetic signals (e.g. RNA-Seq), use simulated to calculate background signal

## INDEL

### expressed_nonexpressed_regions
```{r}
dir.create("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")
setwd("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")

# observed 
expressed_nonexpressed_regions_res_indel <- calc_OE_ratio_parallel(samples = sample_table$sample,
                                                  obs_indel = obs_indel,
                                                  sim_indel = sim_indel,
                                                  regions = expressed_nonexpressed_regions)
write.csv(expressed_nonexpressed_regions_res_indel, "expressed_nonexpressed_regions_res_indel.csv")

# downsample mutations to match in all samples
expressed_nonexpressed_regions_res_downsample_indel <- calc_OE_ratio_parallel(samples = sample_table_downsample$sample,
                                                             obs_indel = obs_indel_downsampled,
                                                             sim_indel = sim_indel_downsampled,
                                                             regions = expressed_nonexpressed_regions)
write.csv(expressed_nonexpressed_regions_res_downsample_indel, "expressed_nonexpressed_regions_res_downsample_indel.csv")
```

### go.expressed_regions 
```{r}
dir.create("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")
setwd("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")

# observed 
go.expressed_regions_res_indel <- calc_OE_ratio_parallel(samples = sample_table$sample,
                                                  obs_indel = obs_indel,
                                                  sim_indel = sim_indel,
                                                  regions = go.expressed_regions )
write.csv(go.expressed_regions_res_indel, "go.expressed_regions_res_indel.csv")

# downsample mutations to match in all samples
go.expressed_regions_res_downsample_indel <- calc_OE_ratio_parallel(samples = sample_table_downsample$sample,
                                                             obs_indel = obs_indel_downsampled,
                                                             sim_indel = sim_indel_downsampled,
                                                             regions = go.expressed_regions )
write.csv(go.expressed_regions_res_downsample_indel, "go.expressed_regions_res_downsample_indel.csv")
```

### go.expressed.shuffled_regions
```{r}
dir.create("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")
setwd("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")

# observed 
go.expressed.shuffled_regions_res_indel <- calc_OE_ratio_parallel(samples = sample_table$sample,
                                                  obs_indel = obs_indel,
                                                  sim_indel = sim_indel,
                                                  regions = go.expressed.shuffled_regions)
write.csv(go.expressed.shuffled_regions_res_indel, "go.expressed.shuffled_regions_res_indel.csv")

# downsample mutations to match in all samples
go.expressed.shuffled_regions_res_downsample_indel <- calc_OE_ratio_parallel(samples = sample_table_downsample$sample,
                                                             obs_indel = obs_indel_downsampled,
                                                             sim_indel = sim_indel_downsampled,
                                                             regions = go.expressed.shuffled_regions)
write.csv(go.expressed.shuffled_regions_res_downsample_indel, "go.expressed.shuffled_regions_res_downsample_indel.csv")

```

### kegg.expressed_regions 
```{r}
dir.create("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")
setwd("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")

# observed 
kegg.expressed_regions_res_indel <- calc_OE_ratio_parallel(samples = sample_table$sample,
                                                  obs_indel = obs_indel,
                                                  sim_indel = sim_indel,
                                                  regions = kegg.expressed_regions )
write.csv(kegg.expressed_regions_res_indel, "kegg.expressed_regions_res_indel.csv")

# downsample mutations to match in all samples
kegg.expressed_regions_res_downsample_indel <- calc_OE_ratio_parallel(samples = sample_table_downsample$sample,
                                                             obs_indel = obs_indel_downsampled,
                                                             sim_indel = sim_indel_downsampled,
                                                             regions = kegg.expressed_regions )
write.csv(kegg.expressed_regions_res_downsample_indel, "kegg.expressed_regions_res_downsample_indel.csv")
```

### kegg.expressed.shuffled_regions
```{r}
dir.create("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")
setwd("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")

# observed 
kegg.expressed.shuffled_regions_res_indel <- calc_OE_ratio_parallel(samples = sample_table$sample,
                                                  obs_indel = obs_indel,
                                                  sim_indel = sim_indel,
                                                  regions = kegg.expressed.shuffled_regions)
write.csv(kegg.expressed.shuffled_regions_res_indel, "kegg.expressed.shuffled_regions_res_indel.csv")

# downsample mutations to match in all samples
kegg.expressed.shuffled_regions_res_downsample_indel <- calc_OE_ratio_parallel(samples = sample_table_downsample$sample,
                                                             obs_indel = obs_indel_downsampled,
                                                             sim_indel = sim_indel_downsampled,
                                                             regions = kegg.expressed.shuffled_regions)
write.csv(kegg.expressed.shuffled_regions_res_downsample_indel, "kegg.expressed.shuffled_regions_res_downsample_indel.csv")

```

### essential_genes_regions
```{r}
dir.create("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")
setwd("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")

# observed 
essential_genes_regions_res_indel <- calc_OE_ratio_parallel(samples = sample_table$sample,
                                                  obs_indel = obs_indel,
                                                  sim_indel = sim_indel,
                                                  regions = essential_genes_regions)
write.csv(essential_genes_regions_res_indel, "essential_genes_regions_res_indel.csv")

# downsample mutations to match in all samples
essential_genes_regions_res_downsample_indel <- calc_OE_ratio_parallel(samples = sample_table_downsample$sample,
                                                             obs_indel = obs_indel_downsampled,
                                                             sim_indel = sim_indel_downsampled,
                                                             regions = essential_genes_regions)
write.csv(essential_genes_regions_res_downsample_indel, "essential_genes_regions_res_downsample_indel.csv")

```

### marker_genes_regions
```{r}
dir.create("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")
setwd("/gs/gsfs0/users/rcutler/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway-overlap/2-calculate-ratio/indel")

# observed 
marker_genes_regions_res_indel <- calc_OE_ratio_parallel(samples = sample_table$sample,
                                                  obs_indel = obs_indel,
                                                  sim_indel = sim_indel,
                                                  regions = marker_genes_regions)
write.csv(marker_genes_regions_res_indel, "marker_genes_regions_res_indel.csv")

# downsample mutations to match in all samples
marker_genes_regions_res_downsample_indel <- calc_OE_ratio_parallel(samples = sample_table_downsample$sample,
                                                             obs_indel = obs_indel_downsampled,
                                                             sim_indel = sim_indel_downsampled,
                                                             regions = marker_genes_regions)
write.csv(marker_genes_regions_res_downsample_indel, "marker_genes_regions_res_downsample_indel.csv")
```