---
title: "dN/dS ratio analysis"
output: html_document
date: "2023-09-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Following this tutorial: http://htmlpreview.github.io/?http://github.com/im3sanger/dndscv/blob/master/vignettes/dNdScv.html

# Preparation

## Load libraries
```{r}
library(dndscv)
library(MutationalPatterns)
library(BSgenome.Hsapiens.UCSC.hg19)
library(ggplot2)
library(zFPKM)
library(biomaRt)
library(dplyr)
library(ggpubr)
library(stringr)
library(tidyr)
library(viridis)
```

## Functions
```{r}
compare_control_enu <- function(data) {
  # Ensure the 'data' argument is a dataframe
  if (!is.data.frame(data)) {
    stop("Input must be a dataframe.")
  }

  # Check if necessary columns exist
  required_columns <- c("name", "mle", "cilow", "cihigh", "list_name")
  if (!all(required_columns %in% names(data))) {
    stop("Dataframe must contain the following columns: ", paste(required_columns, collapse=", "))
  }

  # Function to calculate standard error from confidence intervals
  calc_se <- function(cilow, cihigh) {
    (cihigh - cilow) / (2 * 1.96)
  }

  # Adding standard errors to the data frame
  data$se <- mapply(calc_se, data$cilow, data$cihigh)

  # Function to calculate Z score and raw p-value
  compare_groups <- function(group1, group2) {
    difference_in_means <- group1$mle - group2$mle
    se_combined <- sqrt(group1$se^2 + group2$se^2)
    z_score <- difference_in_means / se_combined
    p_value <- 2 * (1 - pnorm(abs(z_score)))
    return(p_value)
  }

  # Compare control vs enu for each name and store results
  comparison_results <- data.frame(name = character(), p_value = numeric())
  for (name in unique(data$name)) {
    control_group <- data[data$name == name & data$list_name == "control", ]
    enu_group <- data[data$name == name & data$list_name == "enu", ]
    if (nrow(control_group) == 1 && nrow(enu_group) == 1) {
      p_value <- compare_groups(control_group, enu_group)
      comparison_results <- rbind(comparison_results, data.frame(name = name, p_value = p_value))
    }
  }

  # Adjusting p-values for multiple comparisons
  comparison_results$adjusted_p_value <- p.adjust(comparison_results$p_value, method = "bonferroni")

  # Return the results
  return(comparison_results)
}

compare_condition_cycles <- function(data) {
  # Ensure the 'data' argument is a dataframe
  if (!is.data.frame(data)) {
    stop("Input must be a dataframe.")
  }

  # Check if necessary columns exist
  required_columns <- c("name", "mle", "cilow", "cihigh", "condition_cycle")
  if (!all(required_columns %in% names(data))) {
    stop("Dataframe must contain the following columns: ", paste(required_columns, collapse=", "))
  }

  # Function to calculate standard error from confidence intervals
  calc_se <- function(cilow, cihigh) {
    (cihigh - cilow) / (2 * 1.96)
  }

  # Adding standard errors to the data frame
  data$se <- mapply(calc_se, data$cilow, data$cihigh)

  # Function to calculate Z score and raw p-value
  compare_groups <- function(group1, group2) {
    difference_in_means <- group1$mle - group2$mle
    se_combined <- sqrt(group1$se^2 + group2$se^2)
    z_score <- difference_in_means / se_combined
    p_value <- 2 * (1 - pnorm(abs(z_score)))
    return(p_value)
  }

  # Compare all pairs of condition_cycles for each name and store results
  comparison_results <- data.frame(name = character(), 
                                   condition_cycle1 = character(), 
                                   condition_cycle2 = character(), 
                                   p_value = numeric())
  unique_names <- unique(data$name)
  unique_condition_cycles <- unique(data$condition_cycle)

  for (name in unique_names) {
    for (i in 1:(length(unique_condition_cycles) - 1)) {
      for (j in (i + 1):length(unique_condition_cycles)) {
        group1 <- data[data$name == name & data$condition_cycle == unique_condition_cycles[i], ]
        group2 <- data[data$name == name & data$condition_cycle == unique_condition_cycles[j], ]
        if (nrow(group1) == 1 && nrow(group2) == 1) {
          p_value <- compare_groups(group1, group2)
          comparison_results <- rbind(comparison_results, 
                                      data.frame(name = name, 
                                                 condition_cycle1 = unique_condition_cycles[i], 
                                                 condition_cycle2 = unique_condition_cycles[j], 
                                                 p_value = p_value))
        }
      }
    }
  }

  # Adjusting p-values for multiple comparisons
  comparison_results$adjusted_p_value <- p.adjust(comparison_results$p_value, method = "bonferroni")

  # Return the results
  return(comparison_results)
}

theme_Publication <- function(base_size=16, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family = "")
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               #legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

# grlist_to_dataframe_parallel <- function(gr_list) {
#   numCores <- detectCores() - 1
#   registerDoParallel(cores=numCores)
# 
#   df_list <- foreach(i = seq_along(gr_list), .packages = c("GenomicRanges")) %dopar% {
#     gr <- gr_list[[i]]
# 
#     # Use safer methods to extract data from GRanges objects
#       sampleID = names(gr_list)[i],
#       chr = as.character(seqnames(gr)),
#       pos = ranges(gr)@start,
#       ref = sapply(gr$REF, as.character),
#       mut = sapply(gr$ALT, as.character)
# 
#     # Create the data frame
#     data.frame(sampleID, chr, pos, ref, mut, stringsAsFactors = FALSE)
#   }
# 
#   combined_df <- do.call(rbind, df_list)
#   stopImplicitCluster()
#   return(combined_df)
# }

grlist_to_dataframe <- function(gr_list) {
  # Use lapply to iterate over GRangesList and perform operations
  df_list <- lapply(seq_along(gr_list), function(i) {
    gr <- gr_list[[i]]
    data.frame(
      sampleID = names(gr_list)[i],
      chr = as.character(seqnames(gr)),
      pos = start(ranges(gr)),
      ref = as.character(gr$REF),
      mut = as.character(gr$ALT)
    )
  })
  
  # Combine all the data frames into a single data frame using do.call and rbind
  combined_df <- do.call(rbind, df_list)

  return(combined_df)
}

# function to convert to dataframe in correct format
grlist_to_dataframe_old <- function(gr_list) {
  # Initialize an empty list to store the data frames
  df_list <- list()
  
  # Loop over each GRanges object in the GRangesList
  for (i in seq_along(gr_list)) {
    # Extract the ith GRanges object
    gr <- gr_list[[i]]
    
    # Convert the GRanges object to a data frame
    df <- data.frame(
      chr = as.character(seqnames(gr)),
      pos = ranges(gr)@start,
      ref = sapply(gr$REF, as.character),
      mut = sapply(gr$ALT, as.character)
    )
    
    # Add sampleID column
    df$sampleID <- names(gr_list)[i]
    
    # Append to the list
    df_list[[i]] <- df
  }
  
  # Combine all the data frames into a single data frame
  combined_df <- do.call(rbind, df_list)
  combined_df <- combined_df[,c(5,1,2,3,4)] # make sampleID first column
  return(combined_df)
}


test_mean_significance <- function(mean, lower_limit, upper_limit, sample_size) {
  # Calculate the standard error from the confidence interval
  standard_error <- (upper_limit - lower_limit) / (2 * 1.96)
  
  # Calculate the t statistic
  t_statistic <- mean / standard_error
  
  # Calculate degrees of freedom
  df <- sample_size - 1
  
  # Calculate the p-value
  p_value <- 2 * pt(-abs(t_statistic), df)
  
  # Return a list containing the t statistic, the p-value, and whether the mean is significantly different from 0
  significantly_different <- p_value < 0.05
  return(list(t_statistic = t_statistic, p_value = p_value, significantly_different = significantly_different))
}

wrap_labels <- function(labels, width = 5) {
  wrapped_labels <- str_wrap(labels, width)
  names(wrapped_labels) <- names(labels)
  return(wrapped_labels)
}
```

## Load data
```{r}
sample_table <- read.table("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/1-data/sample_table_filtered.txt", header = TRUE, sep = "\t")
```

## Load mutations
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/5-mutation-distribution/1-calculate-ratio")

# observed mutations
obs_snv <- readRDS("obs_snv.RDS")
seqlevelsStyle(obs_snv) <- "NCBI"
obs_snv <- grlist_to_dataframe_old(obs_snv)

obs_indel <- readRDS("obs_indel.RDS")
seqlevelsStyle(obs_indel) <- "NCBI"
obs_indel <- grlist_to_dataframe_old(obs_indel)

obs_all <- rbind(obs_snv, obs_indel)

# save
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed")
saveRDS(obs_all, "obs_all.RDS")
obs_all <- readRDS("obs_all.RDS")
```

## Load expressed genes
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/annotation/rna")

# rna expression table
rna <- read.table("ENCFF245ZOB-IMR90-total-RNA-seq-hg19.tsv", header = TRUE)

# remove version from ensemble gene id
rna$gene_id <- gsub("\\..*", "", rna$gene_id)

# plot TPM distribution
nrow(rna)
hist(log2(rna$TPM))

# zFPKM calculation
zfpkm <- zFPKM(rna[,"TPM", drop = FALSE])
zFPKMPlot(rna[,"TPM", drop = FALSE])

# plot zfpkm  distribution
hist(zfpkm$TPM)

# keep genes with > -3 zfpkm
rna.filter <- rna[zfpkm$TPM > -3,]

# plot filtered TPM distribution
nrow(rna.filter)
hist(log2(rna.filter$TPM))

# get entrz id
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
gene_names <- getBM(attributes = c('ensembl_gene_id', 'external_gene_name', 'entrezgene_id'),
                filters = 'ensembl_gene_id',
                values = rna$gene_id, 
                mart = ensembl)

rna.filter <- merge(rna.filter, gene_names, by.x = "gene_id", by.y = "ensembl_gene_id")
sum(duplicated(rna.filter$entrezgene_id))
rna.filter <- rna.filter[!duplicated(rna.filter$entrezgene_id),] # remove duplicates

# get overlap with dnds ref object
data("refcds_hg19", package = "dndscv")
expressed.genes <- intersect(unlist(lapply(RefCDS, function(x) x$gene_name)), rna.filter$external_gene_name)

# save
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed")
saveRDS(expressed.genes, "expressed.genes.RDS")
expressed.genes <- readRDS("expressed.genes.RDS")
```

## Load essential genes
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/annotation/gene-sets")

essential <- read.csv("AchillesCommonEssentialControls.csv", header = TRUE)
essential$Gene <- sub(" \\(.*", "", essential$Gene)

nonessential <- read.csv("AchillesNonessentialControls.csv", header = TRUE)
nonessential$Gene <- sub(" \\(.*", "", nonessential$Gene)

# subset essential and non essential genes which are expressed
essential.expressed <- subset(essential, Gene %in% expressed.genes)
nonessential.expressed <- subset(nonessential, Gene %in% expressed.genes)

# save
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed")
saveRDS(essential.expressed$Gene, "essential.expressed.genes.RDS")
saveRDS(nonessential.expressed$Gene, "nonessential.expressed.genes.RDS")
```

## Load cancer genes
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/annotation/gene-sets")

# load
df <- read.csv("Census_allMon Mar 24 23_07_21 2025.csv", header = TRUE)
cancer.genes <- df$Gene.Symbol

# filter for expressed
cancer.genes.expressed <- cancer.genes[cancer.genes %in% expressed.genes]

# save
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed")
saveRDS(cancer.genes.expressed, "cancer.genes.expressed.RDS")
```

# All genes

## By Cycle

#### Run dndscv
- Needing to group control 1 and control 9 due to low number of mutations
- not filtering out cancer genes from indel background model
- using all mutations per gene instead of limiting to 3
- using all coding mutations per sample instead of limiting to 3000
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed/all")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed/all")

# split into groups
control <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "control")$sample,]
enu_3 <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "enu" & cycle == 3)$sample,]
enu_6 <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "enu" & cycle == 6)$sample,]
enu_9 <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "enu" & cycle == 9)$sample,]

mutations_list <- list(control = control,
                       enu_3 = enu_3,
                       enu_6 = enu_6,
                       enu_9 = enu_9)

observed.res.list <- lapply(mutations_list, function(x) 
              dndscv(mutations = x,
                     gene_list = NULL,
                     max_muts_per_gene_per_sample = Inf,
                     max_coding_muts_per_sample = Inf,
                     refdb = "hg19",
                     kc = ""))

saveRDS(observed.res.list, "observed_global_dndscv.RDS")
observed.res.list <- readRDS("observed_global_dndscv.RDS")
```

#### Global dN/dS
- wmis = missense
- wnon = nonsense
- wspl = essential splice site mutations
- wtru = truncating substitutions
- wall = all non-synonymous substitutions

- Global dN/dS ratios in somatic evolution of cancer, and seemingly of healthy somatic tissues, appear to show a near-universal pattern of dN/dS~1, with exome-wide dN/dS ratios typically slightly higher than 1
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed/all")

# merge conditions
globaldnds_list <- lapply(observed.res.list, function(x) x$globaldnds)
names(globaldnds_list)[1] <- "control_1"

globaldnds_merged <- do.call(rbind, lapply(names(globaldnds_list), function(df_name) {
  tmp_df <- globaldnds_list[[df_name]]
  tmp_df$list_name <- df_name
  return(tmp_df)}))

# remove Inf
globaldnds_merged <- data.frame(lapply(globaldnds_merged, function(x) replace(x, is.infinite(x), NA)))

# add group info
globaldnds_merged <- separate(globaldnds_merged, col = list_name, into = c("condition", "cycle"), sep = "_", fill = "right")



pdf("global_dnds_condition_bar.pdf",
    width = 6,
    height = 4)
 ggplot(globaldnds_merged, aes(x = condition, y = mle, fill = condition)) +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +  # Average bar plot
    geom_errorbar(aes(ymin = cilow, ymax = cihigh),
                    width = 0.5, position = position_dodge(0.9)) +
       geom_hline(yintercept = 1, linetype = "dashed") +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(y = "Observed\nGlobal dN/dS Ratio", fill = "Condition") + 
    scale_fill_Publication(labels = c("Control", "ENU")) +
    theme_Publication(base_size = 16) +
   scale_y_continuous(trans=scales::pseudo_log_trans(base = 10),
                   breaks = c(0, 1, 10, 100),
                   limits = c(0, 100)) +
    facet_wrap(.~name, ncol = 5, labeller = labeller(name = wrap_labels(new_labels, 12))) +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()

new_labels <- c(
  wmis = "Missense",
  wnon = "Nonsense",
  wspl = "Essential\nsplice site",
  wtru = "Truncating",
  wall = "All mutations"
)

# plot
pdf("global.dnds.pdf",
    width = 8,
    height = 4)
ggplot(globaldnds_merged, aes(x = condition, y = mle, color = cycle)) + 
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_pointrange(aes(ymin=cilow, ymax=cihigh),
                    position=position_dodge(width = 0.7),
                    fatten = 5) +
    guides(fill = guide_legend(override.aes = list(shape = 16)), 
           color = guide_legend(override.aes = list(shape = 16))) +
    scale_color_viridis(discrete = TRUE, end = 0.8) +
    labs(x = "", 
         y = "Global dN/dS", 
         fill = "Cycle") +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    theme_Publication() +
    theme(legend.position = "top",  # Place legend on the right side
        legend.box = "horizontal",  # Stack legend items vertically
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12)) +  # Adjust legend title size
    scale_x_discrete(labels = c("Control", "ENU")) +
    facet_wrap(.~name,
               scale = "fixed",
               nrow = 1,
               labeller = labeller(name = new_labels))
dev.off()
```

#### Gene level dN/dS
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed/all")


# merge conditions
genednds_list <- lapply(observed.res.list, function(x) x$sel_cv)

genednds_merged <- do.call(rbind, lapply(names(genednds_list), function(df_name) {
  tmp_df <- genednds_list[[df_name]]
  tmp_df$list_name <- df_name
  return(tmp_df)}))



# make long for plotting
genednds_merged_long <- pivot_longer(genednds_merged, 
                                     cols = c("wmis_cv", "wnon_cv", "wspl_cv"),
                                     names_to = "variable", 
                                     values_to = "value")

# plot by group
genednds_merged_long$list_name <- factor(genednds_merged_long$list_name, levels = unique(genednds_merged$list_name))

new_labels <- c(
  wmis_cv = "missense",
  wnon_cv = "nonsense",
  wspl_cv = "splice"
)

pdf("gene_dNdScv_condition_violin.pdf",
    width = 6,
    height = 5)
 ggplot(genednds_merged_long, aes(x = list_name, y = value, fill = list_name)) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_violin() +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(y = "Gene dN/dS ratio", fill = "Condition") + 
    scale_fill_Publication() +
    theme_Publication(base_size = 16) +
    facet_wrap(.~variable, labeller = labeller(variable = wrap_labels(new_labels, 12))) +
   scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()

# subset significant
genednds_merged_sig <- genednds_merged %>%
  filter(pallsubs_cv < 0.1)

# make long for plotting
genednds_merged_long <- pivot_longer(genednds_merged_sig, 
                                     cols = c("wmis_cv", "wnon_cv"),
                                     names_to = "variable", 
                                     values_to = "value")

# plot by group
genednds_merged_long$list_name <- factor(genednds_merged_long$list_name, levels = c("control", "enu"))

new_labels <- c(
  wmis_cv = "missense",
  wnon_cv = "nonsense"
)

pdf("gene_dNdScv_sig_condition_bar.pdf",
    width = 4,
    height = 4)
 ggplot(genednds_merged_long, aes(x = list_name, y = value, fill = list_name)) +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +
    stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), geom="errorbar", color="black", width=0.5, position = position_dodge(0.9)) +
    stat_compare_means(comparisons = list(c("control", "enu")), method = "wilcox.test", tip.length = 0, label = "p.signif", vjust = 0.5, hide.ns = TRUE)+
    geom_hline(yintercept = 1, linetype = "dashed") +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(y = "Gene dN/dS ratio", fill = "Condition") + 
    scale_fill_Publication() +
    theme_Publication(base_size = 16) +
    facet_wrap(.~variable, labeller = labeller(variable = wrap_labels(new_labels, 12))) +
    scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()
```

###### Overlap with cancer driver genes
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed/all")

# filter for genes with hits in cancer driver that have significant dnds
genednds_merged_hits_cancer_driver <- genednds_merged %>% 
  filter(n_syn > 0 | n_mis > 0 | n_non > 0 | n_spl > 0) %>%
  filter(gene_name %in% cancer.genes)

genednds_merged_hits_cancer_driver_long <- pivot_longer(genednds_merged_hits_cancer_driver, 
                                                       cols = c("wmis_cv", "wnon_cv", "wspl_cv"),
                                                       names_to = "variable", 
                                                       values_to = "value")

pdf("cancer_driver_gene_dNdScv_condition_violin.pdf",
    width = 6,
    height = 5)
 ggplot(genednds_merged_hits_cancer_driver_long, aes(x = list_name, y = value, fill = list_name)) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_violin() +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(y = "Gene dN/dS ratio", fill = "Condition") + 
    scale_fill_Publication() +
    theme_Publication(base_size = 16) +
    facet_wrap(.~variable, labeller = labeller(variable = wrap_labels(new_labels, 12))) +
   scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()

write.csv(genednds_merged_hits_cancer_driver,
          "cancer_driver_dnds.csv",
          row.names = FALSE)


```


## Expressed genes 

### By condition

#### Run dndscv
- not filtering out cancer genes from indel background model
- using all mutations per gene instead of limiting to 3
- using all coding mutations per sample instead of limiting to 3000
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed/expressed")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed/expressed")

# split into groups
control <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "control")$sample,]
enu <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "enu")$sample,]
mutations_list <- list(control = control, enu = enu)

observed.res.list <- lapply(mutations_list, function(x) 
              dndscv(mutations = x,
                     gene_list = expressed.genes,
                     max_muts_per_gene_per_sample = Inf,
                     max_coding_muts_per_sample = Inf,
                     refdb = "hg19",
                     kc = ""))

saveRDS(observed.res.list, "observed_global_dndscv.RDS")
```

#### Global dN/dS estimates
- wmis = missense
- wnon = nonsense
- wspl = essential splice site mutations
- wtru = truncating substitutions
- wall = all non-synonymous substitutions

- Global dN/dS ratios in somatic evolution of cancer, and seemingly of healthy somatic tissues, appear to show a near-universal pattern of dN/dS~1, with exome-wide dN/dS ratios typically slightly higher than 1
```{r}
# merge conditions
globaldnds_list <- lapply(observed.res.list, function(x) x$globaldnds)

globaldnds_merged <- do.call(rbind, lapply(names(globaldnds_list), function(df_name) {
  tmp_df <- globaldnds_list[[df_name]]
  tmp_df$list_name <- df_name
  return(tmp_df)}))

# remove Inf
globaldnds_merged <- data.frame(lapply(globaldnds_merged, function(x) replace(x, is.infinite(x), NA)))

# stats
# Adding standard errors to the data frame
group_comparisons <- compare_control_enu(globaldnds_merged)
write.csv(group_comparisons, "group_comparison_stats.csv")

# plot by group
globaldnds_merged$condition <- factor(globaldnds_merged$list_name, levels = c("control", "enu"))

new_labels <- c(
  wmis = "Missense",
  wnon = "Nonsense",
  wspl = "Essential splice site",
  wtru = "Truncating",
  wall = "All mutations"
)

pdf("global_dnds_condition_bar.pdf",
    width = 6,
    height = 4)
 ggplot(globaldnds_merged, aes(x = condition, y = mle, fill = condition)) +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +  # Average bar plot
    geom_errorbar(aes(ymin = cilow, ymax = cihigh),
                    width = 0.5, position = position_dodge(0.9)) +
       geom_hline(yintercept = 1, linetype = "dashed") +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(y = "Observed\nGlobal dN/dS Ratio", fill = "Condition") + 
    scale_fill_Publication(labels = c("Control", "ENU")) +
    theme_Publication(base_size = 16) +
   scale_y_continuous(trans=scales::pseudo_log_trans(base = 10),
                   breaks = c(0, 1, 10, 100),
                   limits = c(0, 100)) +
    facet_wrap(.~name, ncol = 5, labeller = labeller(name = wrap_labels(new_labels, 12))) +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()
```

##### Missense only
```{r}
pdf("global_dnds_condition_missense_bar.pdf",
    width = 3,
    height = 4)
 ggplot(subset(globaldnds_merged, name == "wmis"), aes(x = condition, y = mle, fill = condition)) +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +  # Average bar plot
    geom_errorbar(aes(ymin = cilow, ymax = cihigh),
                    width = 0.5, position = position_dodge(0.9)) +
       geom_hline(yintercept = 1, linetype = "dashed") +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(y = "Observed\nGlobal dN/dS Ratio", fill = "Condition") + 
    scale_fill_Publication(labels = c("Control", "ENU")) +
    scale_x_discrete(labels=c("Control", "ENU")) +
   scale_y_continuous(breaks = c(1:6))
    theme_Publication(base_size = 16) +
    facet_wrap(.~name, ncol = 5, labeller = labeller(name = wrap_labels(new_labels, 12))) +
    theme(axis.title.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "none")
dev.off()
```


#### Gene-wise dN/dS estimates

##### Using dNdScv
- Plotting distribution of dn/ds values for all genes
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed/expressed")

# merge conditions
genednds_list <- lapply(observed.res.list, function(x) x$sel_cv)

genednds_merged <- do.call(rbind, lapply(names(genednds_list), function(df_name) {
  tmp_df <- genednds_list[[df_name]]
  tmp_df$list_name <- df_name
  return(tmp_df)}))



# stats
# Adding standard errors to the data frame
#group_comparisons <- compare_control_enu(globaldnds_merged)
#write.csv(group_comparisons, "group_comparison_stats.csv")

# make long for plotting
genednds_merged_long <- pivot_longer(genednds_merged, 
                                     cols = c("wmis_cv", "wnon_cv", "wspl_cv"),
                                     names_to = "variable", 
                                     values_to = "value")

# plot by group
genednds_merged_long$list_name <- factor(genednds_merged_long$list_name, levels = c("control", "enu"))

new_labels <- c(
  wmis_cv = "missense",
  wnon_cv = "nonsense",
  wspl_cv = "splice"
)

pdf("gene_dNdScv_condition_violin.pdf",
    width = 6,
    height = 5)
 ggplot(genednds_merged_long, aes(x = list_name, y = value, fill = list_name)) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_violin() +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(y = "Gene dN/dS ratio", fill = "Condition") + 
    scale_fill_Publication() +
    theme_Publication(base_size = 16) +
    facet_wrap(.~variable, labeller = labeller(variable = wrap_labels(new_labels, 12))) +
   scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()

# subset significant
genednds_merged_sig <- genednds_merged %>%
  filter(pallsubs_cv < 0.1)

# make long for plotting
genednds_merged_long <- pivot_longer(genednds_merged_sig, 
                                     cols = c("wmis_cv", "wnon_cv"),
                                     names_to = "variable", 
                                     values_to = "value")

# plot by group
genednds_merged_long$list_name <- factor(genednds_merged_long$list_name, levels = c("control", "enu"))

new_labels <- c(
  wmis_cv = "missense",
  wnon_cv = "nonsense"
)

pdf("gene_dNdScv_sig_condition_bar.pdf",
    width = 4,
    height = 4)
 ggplot(genednds_merged_long, aes(x = list_name, y = value, fill = list_name)) +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +
    stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), geom="errorbar", color="black", width=0.5, position = position_dodge(0.9)) +
    stat_compare_means(comparisons = list(c("control", "enu")), method = "wilcox.test", tip.length = 0, label = "p.signif", vjust = 0.5, hide.ns = TRUE)+
    geom_hline(yintercept = 1, linetype = "dashed") +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(y = "Gene dN/dS ratio", fill = "Condition") + 
    scale_fill_Publication() +
    theme_Publication(base_size = 16) +
    facet_wrap(.~variable, labeller = labeller(variable = wrap_labels(new_labels, 12))) +
    scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()
```

###### Overlap with cancer driver genes
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/11-cancer-mutations")

cancer.genes <- read.csv("Census_allMon Mar 24 23_07_21 2025.csv",
                         header = TRUE)

# filter for genes with hits in cancer driver that have significant dnds
genednds_merged_hits_cancer_driver <- genednds_merged %>% 
  filter(n_syn > 0 | n_mis > 0 | n_non > 0 | n_spl > 0) %>%
  filter(gene_name %in% cancer.genes$Gene.Symbol)

genednds_merged_hits_cancer_driver_long <- pivot_longer(genednds_merged_hits_cancer_driver, 
                                                       cols = c("wmis_cv", "wnon_cv", "wspl_cv"),
                                                       names_to = "variable", 
                                                       values_to = "value")

pdf("cancer_driver_gene_dNdScv_condition_violin.pdf",
    width = 6,
    height = 5)
 ggplot(genednds_merged_hits_cancer_driver_long, aes(x = list_name, y = value, fill = list_name)) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_violin() +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(y = "Gene dN/dS ratio", fill = "Condition") + 
    scale_fill_Publication() +
    theme_Publication(base_size = 16) +
    facet_wrap(.~variable, labeller = labeller(variable = wrap_labels(new_labels, 12))) +
   scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()

write.csv(genednds_merged_hits_cancer_driver,
          "cancer_driver_dnds.csv",
          row.names = FALSE)
```


##### Using dNdSloc
- Plotting distribution of dn/ds values for all genes
```{r}
# merge conditions
genednds_list <- lapply(observed.res.list, function(x) x$sel_loc)

genednds_merged <- do.call(rbind, lapply(names(genednds_list), function(df_name) {
  tmp_df <- genednds_list[[df_name]]
  tmp_df$list_name <- df_name
  return(tmp_df)}))

# stats
# Adding standard errors to the data frame
#group_comparisons <- compare_control_enu(globaldnds_merged)
#write.csv(group_comparisons, "group_comparison_stats.csv")

# make long for plotting
genednds_merged_long <- pivot_longer(genednds_merged, 
                                     cols = c("wnon_loc", "wspl_loc"),
                                     names_to = "variable", 
                                     values_to = "value")

# plot by group
genednds_merged_long$list_name <- factor(genednds_merged_long$list_name, levels = c("control", "enu"))

new_labels <- c(
  wnon_loc = "missense",
  wspl_loc = "nonsense"
)

pdf("gene_dNdSloc_group_violin.pdf",
    width = 6,
    height = 5)
 ggplot(genednds_merged_long, aes(x = list_name, y = value, fill = list_name)) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_violin() +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Group", y = "dN/dS rati0", fill = "Condition") + 
    scale_fill_Publication() +
    theme_Publication(base_size = 16) +
    facet_wrap(.~variable, labeller = labeller(variable = wrap_labels(new_labels, 12))) +
   scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()
```

#### Manually calculate global dN/dS estimates
- Note that this does not account for things like the substitution model
```{r}
observed.res.manual <- dndscv(mutations = obs_all,
             max_muts_per_gene_per_sample = Inf,
             max_coding_muts_per_sample = Inf,
             refdb = "hg19",
             kc = "")

# get ratio of SNV dN/dS per cell
# this is simply the number of non-synonymous (Essential_Splice, Missense, Synonymous, Nonsense)
# Comapred to those annotated as synonymous
impacts <- observed.res.manual$annotmuts %>%
  filter(impact != "no-SNV") %>%
  mutate(grouped_impact = ifelse(impact == "Synonymous", "Synonymous", "Non-synonymous")) %>%
  group_by(sampleID, grouped_impact) %>%
  dplyr::count() %>%
  spread(key = grouped_impact, value = n) %>%
  replace(is.na(.), 1) %>% # replace NA with 0
  mutate(ratio = (`Non-synonymous` + 1) / (Synonymous + 1))

# add meta data
impacts <- merge(impacts, sample_table, by.x = "sampleID", by.y = "sample", all = TRUE)
impacts$condition <- factor(impacts$condition, levels = c("control", "enu"))
impacts$cycle <- factor(impacts$cycle, levels = c(1,3,6,9))

# by group per cell
stats_summary <- impacts %>%
  group_by(condition) %>%
  summarise(mean = mean(ratio, na.rm = TRUE),
            sd = sd(ratio, na.rm = TRUE))

pdf("global_dnds_manual_group.pdf",
    width = 4,
    height = 5)
ggplot(impacts, aes(x = condition, y = ratio, fill = condition)) +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +  # Average bar plot
    geom_errorbar(data = stats_summary, aes(x = condition, y = mean, ymin = mean - sd, ymax = mean + sd),
                    width = 0.5, position = position_dodge(0.9)) +
    geom_point(data = impacts, aes(x = condition, y = ratio), position = position_jitterdodge(jitter.width = 0.2), size = 2, alpha = 0.5) +
    stat_compare_means(method = "t.test", label = "p.signif", label.y.npc = "top", label.x.npc = "center", size = 8, color = "black", vjust = 0.5, hjust = 0.5, hide.ns = TRUE) +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Group", y = "dN/dS ratio", title = "All mutations") + 
    scale_fill_Publication() +
    theme_Publication(base_size = 16)
dev.off()
```

#### Expected vs observed
```{r}
gene_list <- lapply(observed.res.list, function(x) x$genemuts)

gene_list_merged <- do.call(rbind, lapply(names(gene_list), function(df_name) {
  tmp_df <- gene_list[[df_name]]
  tmp_df$list_name <- df_name
  return(tmp_df)}))

# stats
# Adding standard errors to the data frame
#group_comparisons <- compare_control_enu(globaldnds_merged)
#write.csv(group_comparisons, "group_comparison_stats.csv")

# calculate observed vs expected ratio
gene_list_merged <- gene_list_merged %>%
  mutate(synonymous.oe = log2((n_syn + 1) / (exp_syn + 1)),
         missense.oe = log2((n_mis + 1) / (exp_mis + 1)),
         nonsynonymous.oe = log2((n_non + 1) / (exp_non + 1)),
         splicesite.oe = log2((n_spl + 1) / (exp_spl + 1)))

# summary
gene_list_merged %>%
  group_by(list_name) %>%
  summarize(synonymous.oe = log2((sum(n_syn) + 1) / (sum(exp_syn) + 1)),
         missense.oe = log2((sum(n_mis) + 1) / (sum(exp_mis) + 1)),
         nonsynonymous.oe = log2((sum(n_non) + 1) / (sum(exp_non) + 1)),
         splicesite.oe = log2((sum(n_spl) + 1) / (sum(exp_spl) + 1)))

# make long for plotting
gene_list_merged_long <- pivot_longer(gene_list_merged, 
                                     cols = c("synonymous.oe", "missense.oe", "nonsynonymous.oe", "splicesite.oe"),
                                     names_to = "variable", 
                                     values_to = "value")

# plot by group
gene_list_merged_long$list_name <- factor(gene_list_merged_long$list_name, levels = c("control", "enu"))

new_labels <- c(
  synonymous.oe = "synonymous",
  missense.oe = "nonsense",
  nonsynonymous.oe = "nonsynonymous",
  splicesite.oe = "splicesite"
)

pdf("gene_expected_observed_group_jitter.pdf",
    width = 6,
    height = 5) 
 ggplot(gene_list_merged_long, aes(x = list_name, y = value, fill = list_name)) +
    geom_violin() +
    geom_jitter(alpha = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Group", y = "Log2(Observed / Expected)", fill = "Condition") + 
    scale_fill_Publication() +
    theme_Publication(base_size = 16) +
    facet_wrap(.~variable, scale = "free_y", labeller = labeller(variable = wrap_labels(new_labels, 12))) +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()

pdf("gene_expected_observed_group_bar.pdf",
    width = 6,
    height = 5) 
 ggplot(gene_list_merged_long, aes(x = list_name, y = value, fill = list_name)) +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +  # Average bar plot
    stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
        geom="errorbar", color="black", width=0.5, position = position_dodge(0.9)) +
    # geom_errorbar(aes(ymin = cilow, ymax = cihigh),
    #                 width = 0.5, position = position_dodge(0.9)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(x = "Group", y = "Log2(Observed / Expected)", fill = "Condition") + 
    scale_fill_Publication() +
    theme_Publication(base_size = 16) +
    facet_wrap(.~variable, scale = "free_y", labeller = labeller(variable = wrap_labels(new_labels, 12))) +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()
```


#### Specific genes under positive or negative selection
- none for either group
```{r}
observed.res.list.sig.genes <- lapply(observed.res.list, function(x) x$sel_cv[x$sel_cv$qglobal_cv<0.1, c("gene_name","qallsubs_cv")])
```

## By cycle
- Can only run on ENU and not control because too few mutations

### Run dndscv
```{r}
# split into groups
control_1 <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "control" & cycle == 1)$sample,]
enu_3 <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "enu" & cycle == 3)$sample,]
enu_6 <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "enu" & cycle == 6)$sample,]
control_9 <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "control" & cycle == 9)$sample,]
enu_9 <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "enu" & cycle == 9)$sample,]

mutations_list <- list(enu_3 = enu_3,
                       enu_6 = enu_6,
                       enu_9 = enu_9)

res.cycle.list <- lapply(mutations_list, function(x) 
              dndscv(mutations = x,
                     gene_list = expressed.genes,
                     max_muts_per_gene_per_sample = Inf,
                     max_coding_muts_per_sample = Inf,
                     refdb = "hg19"))
```

### Table of significant genes for positive selection
- none for either group
```{r}
res.cycle.list.sig.genes <- lapply(res.cycle.list, function(x) x$sel_cv[x$sel_cv$qglobal_cv<0.1, c("gene_name","qallsubs_cv")])
```

### Global dN/dS estimates
- wmis = missense
- wnon = nonsense
- wspl = essential splice site mutations
- wtru = truncating substitutions
- wall = all non-synonymous substitutions

- Global dN/dS ratios in somatic evolution of cancer, and seemingly of healthy somatic tissues, appear to show a near-universal pattern of dN/dS~1, with exome-wide dN/dS ratios typically slightly higher than 1
```{r}
# merge conditions
globaldnds_list <- lapply(res.cycle.list, function(x) x$globaldnds)

globaldnds_merged <- do.call(rbind, lapply(names(globaldnds_list), function(df_name) {
  tmp_df <- globaldnds_list[[df_name]]
  tmp_df$list_name <- df_name
  return(tmp_df)}))

data_split <- strsplit(as.character(globaldnds_merged$list_name), "_")
globaldnds_merged$cycle <- factor(sapply(data_split, function(x) x[2]), levels = c(3,6,9))

# stats
# cycle_comparisons <- compare_condition_cycles(globaldnds_merged)
# write.csv(cycle_comparisons, "cycle_comparisons_stats.csv")

# plot
new_labels <- c(
  wmis = "missense",
  wnon = "nonsense",
  wspl = "essential splice site",
  wtru = "truncating",
  wall = "all mutations"
)

pdf("global_dnds_cycle_enu_bar.pdf",
    width = 6,
    height = 4)
 ggplot(globaldnds_merged, aes(x = cycle, y = mle, fill = cycle)) +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +  # Average bar plot
    geom_errorbar(aes(ymin = cilow, ymax = cihigh),
                    width = 0.5, position = position_dodge(0.9)) +
       geom_hline(yintercept = 1, linetype = "dashed") +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(y = "Global dN/dS ratio", x = "Cycle", title = "ENU condition") + 
    theme_Publication(base_size = 16) +
    scale_fill_viridis(discrete = TRUE, end = 0.8) +
    facet_wrap(.~name, ncol = 5, labeller = labeller(name = wrap_labels(new_labels, 12))) +
    theme(strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "none")
dev.off()
```

## Essential genes 

### By condition
- Can only run on ENU and not control because too few mutations

#### Run dndscv
- not filtering out cancer genes from indel background model
- using all mutations per gene instead of limiting to 3
- using all coding mutations per sample instead of limiting to 3000
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed/essential")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/1-observed/essential")

# split into groups
control <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "control")$sample,]
enu <- obs_all[obs_all$sampleID %in% subset(sample_table, condition == "enu")$sample,]
mutations_list <- list(enu = enu)

observed.res.list <- lapply(mutations_list, function(x) 
              dndscv(mutations = x,
                     gene_list = essential.expressed$Gene,
                     max_muts_per_gene_per_sample = Inf,
                     max_coding_muts_per_sample = Inf,
                     refdb = "hg19",
                     kc = ""))

saveRDS(observed.res.list, "observed_global_dndscv.RDS")
```

#### Global dN/dS estimates
- wmis = missense
- wnon = nonsense
- wspl = essential splice site mutations
- wtru = truncating substitutions
- wall = all non-synonymous substitutions

- Global dN/dS ratios in somatic evolution of cancer, and seemingly of healthy somatic tissues, appear to show a near-universal pattern of dN/dS~1, with exome-wide dN/dS ratios typically slightly higher than 1
```{r}
# merge conditions
globaldnds_list <- lapply(observed.res.list, function(x) x$globaldnds)

globaldnds_merged <- do.call(rbind, lapply(names(globaldnds_list), function(df_name) {
  tmp_df <- globaldnds_list[[df_name]]
  tmp_df$list_name <- df_name
  return(tmp_df)}))

# stats
# Adding standard errors to the data frame
# group_comparisons <- compare_control_enu(globaldnds_merged)
# write.csv(group_comparisons, "group_comparison_stats.csv")

new_labels <- c(
  wmis = "missense",
  wnon = "nonsense",
  wspl = "essential splice site",
  wtru = "truncating",
  wall = "all mutations"
)

pdf("global_dnds_condition_bar.pdf",
    width = 6,
    height = 4)
 ggplot(globaldnds_merged, aes(x = list_name, y = mle)) +
    geom_bar(stat = "summary", fun.y = "mean", position = "dodge") +  # Average bar plot
    geom_errorbar(aes(ymin = cilow, ymax = cihigh),
                    width = 0.5, position = position_dodge(0.9)) +
       geom_hline(yintercept = 1, linetype = "dashed") +
    guides(fill = guide_legend(override.aes = list(shape = NA))) +
    labs(y = "Global dN/dS ratio", title = "ENU condition") + 
    #scale_fill_Publication(labels = c("Control", "ENU")) +
    theme_Publication(base_size = 16) +
   # scale_y_continuous(trans=scales::pseudo_log_trans(base = 10),
   #                 breaks = c(0, 1, 10, 100),
   #                 limits = c(0, 100)) +
    facet_wrap(.~name, ncol = 5, labeller = labeller(name = wrap_labels(new_labels, 12))) +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size = 12),
          legend.justification = c("center","top"),
          legend.position = "top")
dev.off()
```

# Compare calculation of expected to results from SigProfilerSimulation

## Run dndscv
- not filtering out cancer genes from indel background model
- using all mutations per gene instead of limiting to 3
- using all coding mutations per sample instead of limiting to 3000
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/4-compare-vep/")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/10-dnds-ratio/4-compare-vep/")

# split into cells
cell_list <- split(obs_all, obs_all$sampleID)

# subset just for ENU
cell_list_subset <- cell_list[sample_table$condition == "enu"]

cell_list_subset_dnds <- lapply(cell_list_subset, function(x) 
              dndscv(mutations = x,
                     gene_list = expressed.genes,
                     max_muts_per_gene_per_sample = Inf,
                     max_coding_muts_per_sample = Inf,
                     refdb = "hg19",
                     kc = ""))

# saveRDS(cell_list_subset_dnds, "global_cell_dndscv.RDS")
```

## synonymous

### Get expected number per gene and sum per cell
```{r}
cell_list_subset_dnds_expected <- lapply(cell_list_subset_dnds, function(x) x$genemuts)

saveRDS(cell_list_subset_dnds_expected, "cell_list_subset_dnds_expected.RDS")

# sum for each cell
cell_list_subset_dnds_expected_sum <- unlist(lapply(cell_list_subset_dnds_expected, function(x) sum(x$exp_syn, na.rm = TRUE)))

cell_list_subset_dnds_observed_sum <- unlist(lapply(cell_list_subset_dnds_expected, function(x) sum(x$n_syn, na.rm = TRUE)))

# make df
df_compare <- data.frame(dnds_expected = cell_list_subset_dnds_expected_sum,
                         dnds_observed = cell_list_subset_dnds_observed_sum,
                         dnds_delta = cell_list_subset_dnds_expected_sum - cell_list_subset_dnds_observed_sum,
                         row.names = names(cell_list_subset_dnds_expected_sum))
```

### Get expected from SigProfilerSimulation for variant effect prediction
- This gives the average from the 1000 simulations per cell
```{r}
vep_sim <- read.csv("/Users/ronaldcutler/Dropbox (EinsteinMed)/Vijg-lab/Projects/mutation accumulation/231009 multiple ENU analysis/5-variant-effect-prediction/1-calculate-ratio/observed_consequence_main_ratio_mean.csv", header = TRUE)

vep_sim_subset <- subset(vep_sim, variant_type == "synonymous variant" & condition == "enu")

vep_sim_subset$vep_delta <- vep_sim_subset$sim.mut - vep_sim_subset$obs.mut

rownames(vep_sim_subset) <- vep_sim_subset$sample
```

### Compare dndscv and variant effect prediction
```{r}
df_compare <- merge(df_compare, vep_sim_subset, by = "row.names")
rownames(df_compare) <- df_compare$Row.names

pdf("compare_synonymous_expected.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = dnds_expected, y = sim.mut)) +
    geom_point(aes(color = as.factor(cycle)), size = 2) +  # Apply fill only to points
    geom_smooth(method = lm) +  # Fitted line without fill
    stat_cor() + # Correlation coefficient
    labs(x = "dndscv Expected\nVariants Per Cell",
         y = "Average Simulated\nVariants Per Cell", 
         color = "Cycle") + 
    theme_Publication(base_size = 16) +
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()

pdf("compare_synonymous_observed.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = dnds_observed, y = obs.mut)) +
    geom_point(aes(color = as.factor(cycle)), size = 2) +  # Apply fill only to points
    geom_smooth(method = lm) +  # Fitted line without fill
    stat_cor() + # Correlation coefficient
    labs(x = "dndscv Observed\nVariants Per Cell",
         y = "Observed\nVariants Per Cell", 
         color = "Cycle") + 
    theme_Publication(base_size = 16) +
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()

pdf("compare_synonymous_delta.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = dnds_delta, y = vep_delta)) +
    geom_point(aes(color = as.factor(cycle)), size = 2) +  # Apply fill only to points
    geom_smooth(method = lm) +  # Fitted line without fill
    stat_cor() + # Correlation coefficient
    labs(x = "dndscv Expected - Observed\nVariants Per Cell",
         y = "Observed - Average Simulated\nVariants Per Cell", 
         color = "Cycle") + 
    theme_Publication(base_size = 16) +
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()
```

## Missense

### Get expected number per gene and sum per cell
```{r}
cell_list_subset_dnds_expected <- lapply(cell_list_subset_dnds, function(x) x$genemuts)

saveRDS(cell_list_subset_dnds_expected, "cell_list_subset_dnds_expected.RDS")

# sum for each cell
cell_list_subset_dnds_expected_sum <- unlist(lapply(cell_list_subset_dnds_expected, function(x) sum(x$exp_mis, na.rm = TRUE)))

cell_list_subset_dnds_observed_sum <- unlist(lapply(cell_list_subset_dnds_expected, function(x) sum(x$n_mis, na.rm = TRUE)))

# make df
df_compare <- data.frame(dnds_expected = cell_list_subset_dnds_expected_sum,
                         dnds_observed = cell_list_subset_dnds_observed_sum,
                         dnds_delta = cell_list_subset_dnds_expected_sum - cell_list_subset_dnds_observed_sum,
                         row.names = names(cell_list_subset_dnds_expected_sum))
```

### Get expected from SigProfilerSimulation for variant effect prediction
- This gives the average from the 1000 simulations per cell
```{r}
vep_sim <- read.csv("/Users/ronaldcutler/Dropbox (EinsteinMed)/Vijg-lab/Projects/mutation accumulation/231009 multiple ENU analysis/5-variant-effect-prediction/1-calculate-ratio/observed_consequence_main_ratio_mean.csv", header = TRUE)

vep_sim_subset <- subset(vep_sim, variant_type == "missense variant" & condition == "enu")

vep_sim_subset$vep_delta <- vep_sim_subset$sim.mut - vep_sim_subset$obs.mut

rownames(vep_sim_subset) <- vep_sim_subset$sample
```

### Compare dndscv and variant effect prediction
```{r}
df_compare <- merge(df_compare, vep_sim_subset, by = "row.names")
rownames(df_compare) <- df_compare$Row.names

pdf("compare_missense_expected.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = dnds_expected, y = sim.mut)) +
    geom_point(aes(color = as.factor(cycle)), size = 2) +  # Apply fill only to points
    geom_smooth(method = lm) +  # Fitted line without fill
    stat_cor() + # Correlation coefficient
    labs(x = "dndscv Expected\nVariants Per Cell",
         y = "Average Simulated\nVariants Per Cell", 
         color = "Cycle") + 
    theme_Publication(base_size = 16) +
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()

pdf("compare_missense_observed.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = dnds_observed, y = obs.mut)) +
    geom_point(aes(color = as.factor(cycle)), size = 2) +  # Apply fill only to points
    geom_smooth(method = lm) +  # Fitted line without fill
    stat_cor() + # Correlation coefficient
    labs(x = "dndscv Observed\nVariants Per Cell",
         y = "Observed\nVariants Per Cell", 
         color = "Cycle") + 
    theme_Publication(base_size = 16) +
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()

pdf("compare_missense_delta.pdf",
    width = 5,
    height = 5)
ggplot(df_compare, aes(x = dnds_delta, y = vep_delta)) +
    geom_point(aes(color = as.factor(cycle)), size = 2) +  # Apply fill only to points
    geom_smooth(method = lm) +  # Fitted line without fill
    stat_cor() + # Correlation coefficient
    labs(x = "dndscv Expected - Observed\nVariants Per Cell",
         y = "Observed - Average Simulated\nVariants Per Cell", 
         color = "Cycle") + 
    theme_Publication(base_size = 16) +
    theme(legend.justification = c("center","top"),
          legend.position = "top") +
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.8)
dev.off()
```
