---
title: "Mutational Signatures analysis"
output: html_document
date: "2023-09-05"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

## Load libraries
```{r}
library(VariantAnnotation)
library(GenomicRanges)
library(GenomicFeatures)
library(rtracklayer)
library(MutationalPatterns)
library(ggplot2)
library(zFPKM)
library(biomaRt)

library(BSgenome.Hsapiens.UCSC.hg19)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
TxDb.Hsapiens.UCSC.hg19.knownGene <- keepStandardChromosomes(TxDb.Hsapiens.UCSC.hg19.knownGene)
TxDb.Hsapiens.UCSC.hg19.knownGene <- dropSeqlevels(TxDb.Hsapiens.UCSC.hg19.knownGene, c("chrY", "chrM"), pruning.mode="coarse")
hg19_size <- sum(seqlengths(TxDb.Hsapiens.UCSC.hg19.knownGene))
```

## Functions
```{r}
theme_Publication <- function(base_size=16, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family = "")
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               #legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

group_GRanges <- function(variants_list, sample_table, grouping_column = "") {
  
  # 1. Get unique groupings
  unique_groupings <- unique(sample_table[[grouping_column]])
  
  # 2. Iterate over each grouping
  combined_variants_list <- lapply(unique_groupings, function(group) {
    
    # Indices of variants belonging to this group
    indices <- which(sample_table[[grouping_column]] == group)
    
    # Combine variants
    combined_variants <- unlist(variants_list[indices])
    return(combined_variants)
  })
  
  names(combined_variants_list) <- unique_groupings
  return(GRangesList(combined_variants_list))
}
```

## Load mutations

### Our study
- Remove control cells from cycle 1
- See with-control-cells folder for analysis with these
- Removed because C>T technical artifacts
```{r}
vcf_files <- list.files(path = "~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/1-data/final/filtered",
 pattern = "\\.vcf$", full.names = TRUE)

sample_table <- read.table("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/1-data/sample_table_filtered.txt", header = TRUE, sep = "\t")
sample_table$condition_cycle <- paste(sample_table$condition, sample_table$cycle, sep = "_")

# Remove control cells from cycle 1
sample_table <- sample_table[sample_table$condition_cycle != "control_1",]

# reorder vcf_files to match sample_table
vcf_files <- vcf_files[match(sample_table$sample,tools::file_path_sans_ext(basename(vcf_files)))]

# mutations
mutations <- read_vcfs_as_granges(vcf_files, sample_table$sample, BSgenome.Hsapiens.UCSC.hg19, type = "all")

mutations_snv <- read_vcfs_as_granges(vcf_files, sample_table$sample, BSgenome.Hsapiens.UCSC.hg19, type = "snv")
mutations_snv_condition <- group_GRanges(mutations_snv, sample_table, "condition")
mutations_snv_condition_cycle <- group_GRanges(mutations_snv, sample_table, "condition_cycle")

mutations_indel <- read_vcfs_as_granges(vcf_files, sample_table$sample, BSgenome.Hsapiens.UCSC.hg19, type = "indel")
mutations_indel_condition <- group_GRanges(mutations_indel, sample_table, "condition")
mutations_indel_condition_cycle <- group_GRanges(mutations_indel, sample_table, "condition_cycle")
```

### fibroblast
- https://www.nature.com/articles/nmeth.4227
```{r}
vcf_files <- list.files(path = "~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/1-data/somamutdb/fibroblast/single-cells",
 pattern = "\\.vcf$", full.names = TRUE)

# mutations
mutations.fibroblasts <- read_vcfs_as_granges(vcf_files, tools::file_path_sans_ext(basename(vcf_files)), BSgenome.Hsapiens.UCSC.hg19, type = "snv")
```

### lung
- https://www.nature.com/articles/s41588-022-01035-w
```{r}
vcf_files <- list.files(path = "~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/1-data/somamutdb/lung/single-cells",
 pattern = "\\.vcf$", full.names = TRUE)

# mutations
mutations.lung <- read_vcfs_as_granges(vcf_files, tools::file_path_sans_ext(basename(vcf_files)), BSgenome.Hsapiens.UCSC.hg19, type = "snv")
```

### brain
- https://www.nature.com/articles/s41586-022-04640-1
```{r}
vcf_files <- list.files(path = "~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/1-data/somamutdb/brain",
 pattern = "\\.vcf$", full.names = TRUE)

# mutations
mutations.brain <- read_vcfs_as_granges(vcf_files, tools::file_path_sans_ext(basename(vcf_files)), BSgenome.Hsapiens.UCSC.hg19, type = "snv")
```

## Save objects
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells-with-background")

saveRDS(mutations.fibroblasts, "mutations.fibroblasts.RDS")

saveRDS(mutations.lung, "mutations.lung.RDS")

saveRDS(mutations.brain, "mutations.brain.RDS")
```

## Combine objects
```{r}
mutations.all <- c(mutations_snv,
                   mutations.fibroblasts,
                   mutations.lung,
                   mutations.brain)

mutations.all.annot <- data.frame(sample = names(mutations.all),
                                  group = c(sample_table$condition,
                                            rep("fibroblasts", length(mutations.fibroblasts)),
                                            rep("lung", length(mutations.lung)),
                                            rep("brain", length(mutations.brain))))
```


# Mutation specturm, profiile, and contexts

## SNV

### SNV mutation spectrum
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells-with-background")

type_occurrences <- mut_type_occurrences(mutations.all, BSgenome.Hsapiens.UCSC.hg19)

pdf("SNV_spectrum_group.pdf",
    width = 10,
    height = 10)
plot_spectrum(type_occurrences, 
              CT = TRUE, 
              indv_points = TRUE, 
              legend = TRUE,
              error_bars = 'stdev',
              by = mutations.all.annot$group,
              condensed = T) +
  theme_Publication(base_size = 16) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "bottom",
        legend.direction = "horizontal",
        strip.text = element_text(size = 12),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12))
dev.off()

type_occurrences_ref <- type_occurrences[-c(1:18),]

pdf("SNV_spectrum_reference.pdf",
    width = 7,
    height = 4)
plot_spectrum(type_occurrences_ref, 
              CT = TRUE, 
              indv_points = TRUE, 
              legend = TRUE,
              error_bars = 'stdev',
              by = mutations.all.annot[-c(1:18),]$group,
              condensed = T) +
  theme_Publication(base_size = 16) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "bottom",
        legend.direction = "horizontal",
        strip.text = element_text(size = 12),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12))
dev.off()
```

### SNV mutational profile
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells-with-background")

# by cell
mut_mat_cell <- mut_matrix(mutations.all, BSgenome.Hsapiens.UCSC.hg19)

pdf("SNV_96_profile_sample.pdf",
    width = 10,
    height = 20)
plot_96_profile(mut_mat_cell,
                ymax = 0.1,
                condensed = TRUE)
dev.off()
```

# Mutation signature analysis

## Calculate factorization rank using NMF
- How to choose the rank: Choose the smallest rank for which cophenetic correlation coefficient starts decreasing. Another approach is to choose the rank for which the plot of the residual sum of squares (RSS) between the input matrix and its estimate shows an inflection point
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells-with-background")

# generate an estimate rank plot
estimate <- nmf(mut_mat_cell, rank = 2:5, method = "brunet", 
                nrun = 10, seed = 123456, .opt = "v-p")

# run same estimation on randomized data
rV <- randomize(mut_mat_cell)
rand <- nmf(rV, rank = 2:5, method = "brunet", 
                nrun = 10, seed = 123456, .opt = "v-p")

# plot
pdf("SNV_NMF_rank_survey_estimate.pdf",
    width = 5,
    height = 5)
plot(estimate)
dev.off()

pdf("SNV_NMF_rank_survey_random.pdf",
    width = 5,
    height = 5)
plot(rand)
dev.off()
```

#### De novo signature extraction
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells-with-background")

# get similarity between samples
# cos_sim_samples <- cos_sim_matrix(mut_mat_condition, mut_mat_condition)
# 
# pdf("SNV_sample_cosine_heatmap.pdf",
#     width = 5,
#     height = 5)
# plot_cosine_heatmap(cos_sim_samples, cluster_rows = TRUE, cluster_cols = TRUE)
# dev.off()

# extract signatures
nmf_res_snv <- extract_signatures(mut_mat_cell, 
                              rank = 2, 
                              nrun = 1000, 
                              nmf_type = "regular",
                              single_core = FALSE)

# name signatures - reverse order to match last aanalysis
colnames(nmf_res_snv$signatures) <- c("SBSB", "SBSA")
#nmf_res_snv$signatures <- nmf_res_snv$signatures[,c(2,1)]

rownames(nmf_res_snv$contribution) <- c("SBSB", "SBSA")
#nmf_res_snv$contribution <- nmf_res_snv$contribution[c(1,2),]



write.csv(nmf_res_snv, "nmf_res_snv.csv", quote = FALSE)

# plot profile
pdf("SNV_de_novo_signature_profile.pdf",
    width = 7,
    height = 4)
plot_96_profile(nmf_res_snv$signatures,
                ymax= 0.06,
                condensed = TRUE) +
  scale_y_continuous(breaks = seq(0, 0.1, by = 0.01))
dev.off()

# compare original vs reconstructed
pdf("SNV_de_novo_signature_profile_reconstruction.pdf",
    width = 5,
    height = 5)
plot_original_vs_reconstructed(mut_mat_cell, nmf_res_snv$reconstructed, 
                               y_intercept = 0.95)
dev.off()

# plot contribution
pdf("SNV_de_novo_profile_contribution_relative.pdf",
    width = 10,
    height = 20)
plot_contribution(nmf_res_snv$contribution, 
                  nmf_res_snv$signature,
                  mode = "relative",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()

# plot contribution of only enu and control
nmf_res_snv_contrib.sub <- nmf_res_snv$contribution[,c(1:18)]
nmf_res_snv_contrib.sub <- nmf_res_snv_contrib.sub[,c("MMEA9_C1", "MMEA9_C2", "MMEA9_C3", "MMEB9_C1", "MMEB9_C2", "MMEB9_C3",
                                                      "E1_06_14", "E2_06_14", "E1_08_16",
                                                      "E3_06_14", "E2_08_16", "E3_08_16",
                                                      "MMEA9_E1", "MMEA9_E2", "MMEA9_E3", "MMEB9_E1", "MMEB9_E2", "MMEB9_E3")]

pdf("SNV_de_novo_profile_contribution_relative_enu_control.pdf",
    width = 5,
    height = 7)
plot_contribution(nmf_res_snv_contrib.sub, 
                  nmf_res_snv$signature,
                  mode = "relative",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()


sample_counts <- colSums(mut_mat_cell)
sample_counts.sub <- sample_counts[colnames(nmf_res_snv_contrib.sub)]
absolute_contribution.sub <- nmf_res_snv_contrib.sub * rep(sample_counts.sub, each = nrow(nmf_res_snv_contrib.sub))

pdf("SNV_de_novo_profile_contribution_absolute.pdf",
    width = 5,
    height = 7)
plot_contribution(absolute_contribution.sub, 
                  nmf_res_snv$signature,
                  mode = "absolute",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14) +
  scale_y_log10(limits = c(1, NA)) +
  theme(legend.position = "top") +
  labs(y = "Observed mutations")
dev.off()

pdf("SNV_de_novo_profile_contribution_heatmap.pdf",
    width = 5,
    height = 5)
plot_contribution_heatmap(nmf_res_snv$contribution, 
                          cluster_samples = TRUE, 
                          cluster_sigs = TRUE)
dev.off()

# similarity between samples at signature level
cos_sim_samples_signatures <- cos_sim_matrix(mut_mat_cell, nmf_res_snv$signature)

pdf("SNV_de_novo_signature_cosine_heatmap.pdf",
    width = 5,
    height = 5)
plot_cosine_heatmap(cos_sim_samples_signatures, 
                    cluster_rows = TRUE, cluster_cols = TRUE)
dev.off()
```

#### COSMIC Signature refitting
```{r}
# compare to known signatures
cosmic_signatures = get_known_signatures(muttype = "snv",
                                         source = "COSMIC",
                                         tissue_type = NULL,
                                         incl_poss_artifacts = TRUE)

# get cosine similarity of all COSMIC signatures to the denovo 

# SBSA
sbsa.cos_sim_all <- apply(cosmic_signatures, 2, function(x) cos_sim(x, nmf_res_snv$signatures[,"SBSA"]))
sbsa.cos_sim_all <- sbsa.cos_sim_all[order(sbsa.cos_sim_all, decreasing = TRUE)]

pdf("SNV_signature_refit_COSMIC_SBSA_vs_SBS5.pdf",
    width = 7,
    height = 4)
plot_compare_profiles(nmf_res_snv$signatures[,"SBSA"],
  cosmic_signatures[, "SBS5"],
  profile_names = c("SBSA", "SBS5"),
  profile_ymax = 0.06,
  condensed = TRUE)
dev.off()


# SBSB
sbsb.cos_sim_all <- apply(cosmic_signatures, 2, function(x) cos_sim(x, nmf_res_snv$signatures[,"SBSB"]))
sbsb.cos_sim_all <- sbsb.cos_sim_all[order(sbsb.cos_sim_all, decreasing = TRUE)]

pdf("SNV_signature_refit_COSMIC_SBSB_vs_SBS5.pdf",
    width = 7,
    height = 4)
plot_compare_profiles(nmf_res_snv$signatures[,"SBSB"],
  cosmic_signatures[, "SBS5"],
  profile_names = c("SBSB", "SBS5"),
  profile_ymax = 0.06,
  condensed = TRUE
)
dev.off()

pdf("SNV_signature_refit_COSMIC_SBSB_vs_SBS25.pdf",
    width = 7,
    height = 4)
plot_compare_profiles(nmf_res_snv$signatures[,"SBSB"],
  cosmic_signatures[, "SBS25"],
  profile_names = c("SBSB", "SBS25"),
  profile_ymax = 0.06,
  condensed = TRUE
)
dev.off()

# bootstrapped refitting to ensure stability
# contri_boots <- fit_to_signatures_bootstrapped(mut_mat_condition,
#   cosmic_signatures,
#   n_boots = 1000,
#   method = "strict"
# )
# 
# pdf("SNV_refit_signature_bootstrap.pdf",
#     width = 10,
#     height = 10)
# plot_bootstrapped_contribution(contri_boots,
#                                mode = "relative")
# dev.off()

# get top 10 COSMIC signatures that correalte the bset to each de novo signature
combined_top <- unique(c(names(sbsa.cos_sim_all[1:10]), names(sbsb.cos_sim_all[1:10])))

fit_res_snv <- fit_to_signatures_strict(mut_mat_cell, cosmic_signatures[,colnames(cosmic_signatures) %in% combined_top], max_delta = 0.002, method = "best_subset")

# compare original vs reconstructed
pdf("SNV_signature_refit_COSMIC_profile_reconstruction.pdf",
    width = 5,
    height = 5)
plot_original_vs_reconstructed(mut_mat_condition, fit_res_snv$fit_res$reconstructed, 
                               y_intercept = 0.95)
dev.off()

# plot contribution
pdf("SNV_signature_refit_COSMIC_profile_contribution_relative.pdf",
    width = 5,
    height = 5)
plot_contribution(fit_res_snv$fit_res$contribution, 
                  mode = "relative",
                  coord_flip = FALSE) +
  theme_Publication(base_size = 14)
dev.off()


pdf("SNV_signature_refit_COSMIC_profile_contribution_absolute.pdf",
    width = 8,
    height = 8)
plot_contribution(fit_res_snv$fit_res$contribution, 
                  mode = "absolute",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()

pdf("SNV_signature_refit_COSMIC_contribution_heatmap.pdf",
    width = 10,
    height = 5)
plot_contribution_heatmap(fit_res_snv$fit_res$contribution, 
                          cluster_samples = TRUE, 
                          cluster_sigs = TRUE)
dev.off()

# similarity between samples at signature level
cos_sim_samples_signatures <- cos_sim_matrix(mut_mat_condition, COSMIC_signatures)

pdf("SNV_signature_refit_COSMIC_cosine_heatmap.pdf",
    width = 10,
    height = 5)
plot_cosine_heatmap(cos_sim_samples_signatures, 
                    cluster_rows = TRUE, cluster_cols = TRUE)
dev.off()
```

#### SIGNAL exposure signature refitting
- Purpose is to see what mutagen the ENU signature most resembles
```{r}
# compare to known signatures
signal_signatures = get_known_signatures(muttype = "snv",
                                         source = "SIGNAL",
                                         tissue_type = NULL,
                                         sig_type = "exposure",
                                         incl_poss_artifacts = TRUE)

# get cosine similarity of all SIGNAL signatures to the denovo 

# SBSA
sbsa.cos_sim_all <- apply(signal_signatures, 2, function(x) cos_sim(x, nmf_res_snv$signatures[,"SBSA"]))
sbsa.cos_sim_all <- sbsa.cos_sim_all[order(sbsa.cos_sim_all, decreasing = TRUE)]

pdf("SNV_signature_refit_SIGNAL_SBSA_vs_ENU.pdf",
    width = 7,
    height = 4)
plot_compare_profiles(nmf_res_snv$signatures[,"SBSA"],
  signal_signatures[, "ENU..400.uM."],
  profile_names = c("SBSA", "ENU"),
  profile_ymax = 0.06,
  condensed = TRUE
)
dev.off()

# SBSB
sbsb.cos_sim_all <- apply(signal_signatures, 2, function(x) cos_sim(x, nmf_res_snv$signatures[,"SBSB"]))
sbsb.cos_sim_all <- sbsb.cos_sim_all[order(sbsb.cos_sim_all, decreasing = TRUE)]

pdf("SNV_signature_refit_SIGNAL_SBSB_vs_DES.pdf",
    width = 7,
    height = 4)
plot_compare_profiles(nmf_res_snv$signatures[,"SBSB"],
  signal_signatures[, "DES..0.938.mM."],
  profile_names = c("SBSB", "DES..0.938.mM."),
  profile_ymax = 0.06,
  condensed = TRUE
)
dev.off()

# bootstrapped refitting to ensure stability
# contri_boots <- fit_to_signatures_bootstrapped(mut_mat_condition,
#   signal_signatures,
#   n_boots = 1000,
#   method = "strict"
# )
# 
# pdf("SNV_refit_signature_bootstrap.pdf",
#     width = 10,
#     height = 10)
# plot_bootstrapped_contribution(contri_boots,
#                                mode = "relative")
# dev.off()

# get top 10 signal signatures that correalte the bset to each de novo signature
combined_top <- unique(c(names(sbs40.cos_sim_all[1:10]), names(sbsa.cos_sim_all[1:10])))

fit_res_snv <- fit_to_signatures_strict(mut_mat_condition, signal_signatures[,colnames(signal_signatures) %in% combined_top], max_delta = 0.002, method = "best_subset")

# compare original vs reconstructed
pdf("SNV_signature_refit_SIGNAL_profile_reconstruction.pdf",
    width = 5,
    height = 5)
plot_original_vs_reconstructed(mut_mat_condition, fit_res_snv$fit_res$reconstructed, 
                               y_intercept = 0.95)
dev.off()

# plot contribution
pdf("SNV_signature_refit_SIGNAL_profile_contribution_relative.pdf",
    width = 5,
    height = 5)
plot_contribution(fit_res_snv$fit_res$contribution, 
                  mode = "relative",
                  coord_flip = FALSE) +
  theme_Publication(base_size = 14)
dev.off()


pdf("SNV_signature_refit_SIGNAL_profile_contribution_absolute.pdf",
    width = 8,
    height = 8)
plot_contribution(fit_res_snv$fit_res$contribution, 
                  mode = "absolute",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()

pdf("SNV_signature_refit_SIGNAL_contribution_heatmap.pdf",
    width = 10,
    height = 5)
plot_contribution_heatmap(fit_res_snv$fit_res$contribution, 
                          cluster_samples = TRUE, 
                          cluster_sigs = TRUE)
dev.off()

# similarity between samples at signature level
cos_sim_samples_signatures <- cos_sim_matrix(mut_mat_condition, signal_signatures)

pdf("SNV_signature_refit_SIGNAL_cosine_heatmap.pdf",
    width = 10,
    height = 5)
plot_cosine_heatmap(cos_sim_samples_signatures, 
                    cluster_rows = TRUE, cluster_cols = TRUE)
dev.off()
```

#### Signature potential damage analysis
- Can use different subset of genes (e.g. growth, nonessential) - Looking at genes determined to be expressed in IMR90
- Using best fit signatures: SBS5, SBS40, ENU
- ratio_by_background is normalized ratio. Noramlized ratio of 2 means that a signature is twice as likely to cause that type of mutation, compared to a completely random “flat” signature
```{r, eval = FALSE}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/annotation/rna")

# rna expression table
rna <- read.table("ENCFF245ZOB-IMR90-total-RNA-seq-hg19.tsv", header = TRUE)

# remove version from ensemble gene id
rna$gene_id <- gsub("\\..*", "", rna$gene_id)

# plot TPM distribution
nrow(rna)
hist(log2(rna$TPM))

# zFPKM calculation
zfpkm <- zFPKM(rna[,"TPM", drop = FALSE])
zFPKMPlot(rna[,"TPM", drop = FALSE])

# plot zfpkm  distribution
hist(zfpkm$TPM)

# keep genes with > -3 zfpkm
rna.filter <- rna[zfpkm$TPM > -3,]

# plot filtered TPM distribution
nrow(rna.filter)
hist(log2(rna.filter$TPM))

# get entrz id
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
gene_names <- getBM(attributes = c('ensembl_gene_id', 'external_gene_name', 'entrezgene_id'),
                filters = 'ensembl_gene_id',
                values = rna$gene_id, 
                mart = ensembl)

rna.filter <- merge(rna.filter, gene_names, by.x = "gene_id", by.y = "ensembl_gene_id")
sum(duplicated(rna.filter$entrezgene_id))
rna.filter <- rna.filter[!duplicated(rna.filter$entrezgene_id),] # remove duplicates

setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells/signature/condition")

# get the ratio of “stop gain”, “mismatch”, “synonymous” and “splice site” mutations can be determined per genomic context
contexts <- rownames(mut_mat_condition)
context_mismatches <- context_potential_damage_analysis(contexts, TxDb.Hsapiens.UCSC.hg19.knownGene, BSgenome.Hsapiens.UCSC.hg19, rna.filter$entrezgene_id, verbose = TRUE)

write.csv(context_mismatches, "context_mismatches_potential_damage.csv", row.names = FALSE)

# get the ratios per signature
signatures <- cbind(cosmic_signatures[,c("SBS5", "SBS40")], signal_signatures[,"ENU..400.uM.", drop = FALSE])
sig_damage <- signature_potential_damage_analysis(signatures, contexts, context_mismatches)

write.csv(sig_damage, "fitted_sig_potential_damage.csv", row.names = FALSE)
```


### INDEL

#### Calculate factorization rank using NMF
- How to choose the rank: Choose the smallest rank for which cophenetic correlation coefficient starts decreasing. Another approach is to choose the rank for which the plot of the residual sum of squares (RSS) between the input matrix and its estimate shows an inflection point
- Choosing rank = 2
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells/signature/condition")

# add small pseudocount
indel_counts_condition <- indel_counts_condition + 0.0001

# generate an estimate rank plot
estimate <- nmf(indel_counts_condition, rank = 2:5, method = "brunet", 
                nrun = 10, seed = 123456, .opt = "v-p")

# run same estimation on randomized data
rV <- randomize(indel_counts_condition)
rand <- nmf(rV, rank = 2:5, method = "brunet", 
                nrun = 10, seed = 123456, .opt = "v-p")

# plot
pdf("INDEL_NMF_rank_survey_estimate.pdf",
    width = 5,
    height = 5)
plot(estimate)
dev.off()

pdf("INDEL_NMF_rank_survey_random.pdf",
    width = 5,
    height = 5)
plot(rand)
dev.off()
```

#### De novo signature extraction
- Using COSMIC database for comparison
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells/signature/condition")

# get similarity between samples
cos_sim_samples <- cos_sim_matrix(indel_counts, indel_counts)

pdf("INDEL_sample_cosine_heatmap.pdf",
    width = 5,
    height = 5)
plot_cosine_heatmap(cos_sim_samples, cluster_rows = TRUE, cluster_cols = TRUE)
dev.off()

# extract signatures
nmf_res_indel <- extract_signatures(indel_counts_condition, 
                              rank = 1, 
                              nrun = 1000, 
                              nmf_type = "regular",
                              single_core = FALSE)

# name signatures
colnames(nmf_res_indel$signatures) <- c("IDA")
rownames(nmf_res_indel$contribution) <- c("IDA")

# bootstrapped refitting to ensure stability
# contri_boots <- fit_to_signatures_bootstrapped(indel_counts,
#   nmf_res_indel$signatures,
#   n_boots = 1000,
#   method = "strict"
# )
# 
# pdf("INDEL_de_novo_signature_context.pdf",
#     width = 10,
#     height = 5)
# plot_indel_contexts(nmf_res_indel$signatures,
#                     same_y = TRUE,
#                     extra_labels = FALSE,
#                     condensed = TRUE) +
#   theme_Publication(base_size = 12) +
#   theme(axis.text.x = element_text(size = 5),
#         legend.position = "right",
#         legend.direction = "vertical",
#         strip.text = element_text(size = 10)) +
#   labs(y = "Relative contribution", x = "No. repeat units")
# dev.off()

pdf("INDEL_de_novo_signature_main_context.pdf",
    width = 4,
    height = 4)
plot_main_indel_contexts(nmf_res_indel$signatures,
                    same_y = TRUE) +
    theme_Publication(base_size = 12) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "right",
        legend.direction = "vertical",
        strip.text = element_text(size = 10)) +
  labs(y = "Relative contribution")
dev.off()

# compare original vs reconstructed
pdf("INDEL_de_novo_signature_profile_reconstruction.pdf",
    width = 5,
    height = 5)
plot_original_vs_reconstructed(indel_counts_condition, nmf_res_indel$reconstructed, 
                               y_intercept = 0.95)
dev.off()

# plot contribution
pdf("INDEL_de_novo_profile_contribution_relative.pdf",
    width = 4,
    height = 5)
plot_contribution(nmf_res_indel$contribution, 
                  nmf_res_indel$signature,
                  mode = "relative",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()


pdf("INDEL_de_novo_profile_contribution_absolute.pdf",
    width = 4,
    height = 5)
plot_contribution(nmf_res_indel$contribution, 
                  nmf_res_indel$signature,
                  mode = "absolute",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()

# pdf("INDEL_de_novo_profile_contribution_heatmap.pdf",
#     width = 5,
#     height = 5)
# plot_contribution_heatmap(nmf_res_indel$contribution, 
#                           cluster_samples = TRUE, 
#                           cluster_sigs = TRUE)
# dev.off()

# # similarity between samples at signature level
# cos_sim_samples_signatures <- cos_sim_matrix(indel_counts_condition, nmf_res_indel$signature)
# 
# pdf("INDEL_de_novo_signature_cosine_heatmap.pdf",
#     width = 5,
#     height = 5)
# plot_cosine_heatmap(cos_sim_samples_signatures, 
#                     cluster_rows = TRUE, cluster_cols = TRUE)
# dev.off()
```

#### COSMIC Signature refitting
```{r}
# compare to known signatures
indel_signatures <- get_known_signatures(muttype = "indel", incl_poss_artifacts = TRUE)

# get similarity
indel.cos_sim_all <- apply(indel_signatures, 2, function(x) cos_sim(x, nmf_res_indel$signatures[,1]))
indel.cos_sim_all <- indel.cos_sim_all[order(indel.cos_sim_all, decreasing = TRUE)]

# pdf("INDEL_signature_refit_COSMIC_IDA_vs_ID1.pdf",
#     width = 7,
#     height = 4)
# plot_compare_profiles(nmf_res_indel$signatures[,"IDA"],
#   indel_signatures[, "ID1"],
#   profile_names = c("IDA", "ID1"),
#   profile_ymax = 0.06,
#   condensed = TRUE,
#   colors = NA)
# dev.off()

fit_res_indel <- fit_to_signatures_strict(indel_counts_condition, indel_signatures[,1:10], max_delta = 0.002, method = "best_subset")

# # bootstrapped refitting to ensure stability
# contri_boots <- fit_to_signatures_bootstrapped(indel_counts,
#   cosmic_signatures,
#   n_boots = 1000,
#   method = "strict"
# )
# 
# pdf("INDEL_refit_signature_bootstrap.pdf",
#     width = 10,
#     height = 4)
# plot_bootstrapped_contribution(contri_boots,
#                                mode = "relative")
# dev.off()

# compare original vs reconstructed
pdf("INDEL_refit_signature_profile_reconstruction.pdf",
    width = 5,
    height = 5)
plot_original_vs_reconstructed(indel_counts_condition, fit_res_indel$fit_res$reconstructed, 
                               y_intercept = 0.95)
dev.off()

# plot contribution
pdf("INDEL_refit_signature_profile_contribution_relative.pdf",
    width = 8,
    height = 8)
plot_contribution(fit_res_indel$fit_res$contribution, 
                  mode = "relative",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()


pdf("INDEL_refit_signature_profile_contribution_absolute.pdf",
    width = 8,
    height = 8)
plot_contribution(fit_res_indel$fit_res$contribution, 
                  mode = "absolute",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()

pdf("INDEL_refit_profile_contribution_heatmap.pdf",
    width = 10,
    height = 5)
plot_contribution_heatmap(fit_res_indel$fit_res$contribution, 
                          cluster_samples = TRUE, 
                          cluster_sigs = TRUE)
dev.off()

# similarity between samples at signature level
cos_sim_samples_signatures <- cos_sim_matrix(indel_counts, indel_signatures)

pdf("INDEL_refit_signature_cosine_heatmap.pdf",
    width = 10,
    height = 5)
plot_cosine_heatmap(cos_sim_samples_signatures, 
                    cluster_rows = TRUE, cluster_cols = TRUE)
dev.off()
```

#### Signature potential damage analysis
- Can use different subset of genes (e.g. growth, nonessential) - Looking at genes determined to be expressed in IMR90
- Using de novo signatures
- ratio_by_background is normalized ratio. Noramlized ratio of 2 means that a signature is twice as likely to cause that type of mutation, compared to a completely random “flat” signature
```{r, eval = FALSE}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/annotation/rna")

# rna expression table
rna <- read.table("ENCFF245ZOB-IMR90-total-RNA-seq-hg19.tsv", header = TRUE)

# remove version from ensemble gene id
rna$gene_id <- gsub("\\..*", "", rna$gene_id)

# plot TPM distribution
nrow(rna)
hist(log2(rna$TPM))

# zFPKM calculation
zfpkm <- zFPKM(rna[,"TPM", drop = FALSE])
zFPKMPlot(rna[,"TPM", drop = FALSE])

# plot zfpkm  distribution
hist(zfpkm$TPM)

# keep genes with > -3 zfpkm
rna.filter <- rna[zfpkm$TPM > -3,]

# plot filtered TPM distribution
nrow(rna.filter)
hist(log2(rna.filter$TPM))

# get entrz id
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
gene_names <- getBM(attributes = c('ensembl_gene_id', 'external_gene_name', 'entrezgene_id'),
                filters = 'ensembl_gene_id',
                values = rna$gene_id, 
                mart = ensembl)

rna.filter <- merge(rna.filter, gene_names, by.x = "gene_id", by.y = "ensembl_gene_id")
sum(duplicated(rna.filter$entrezgene_id))
rna.filter <- rna.filter[!duplicated(rna.filter$entrezgene_id),] # remove duplicates

# get the ratio of “stop gain”, “mismatch”, “synonymous” and “splice site” mutations can be determined per genomic context
contexts <- rownames(indel_counts)
context_mismatches <- context_potential_damage_analysis(contexts, TxDb.Hsapiens.UCSC.hg19.knownGene, BSgenome.Hsapiens.UCSC.hg19, rna.filter$entrezgene_id)

# get the ratios per signature
sig_damage_indel <- signature_potential_damage_analysis(nmf_res_indel$signatures, contexts, context_mismatches)
sig_damage_indel <- sig_damage_indel[order(sig_damage_indel$ratio_by_background, decreasing = TRUE),]
```

## By Cell

### SNV

#### Calculate factorization rank using NMF
- How to choose the rank: Choose the smallest rank for which cophenetic correlation coefficient starts decreasing. Another approach is to choose the rank for which the plot of the residual sum of squares (RSS) between the input matrix and its estimate shows an inflection point
- This doesn't work right becasue there are too few conditions
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells/signature/cell")

# generate an estimate rank plot
estimate <- nmf(mut_mat_cell, rank = 2:5, method = "brunet", 
                nrun = 10, seed = 123456, .opt = "v-p")

# run same estimation on randomized data
rV <- randomize(mut_mat_cell)
rand <- nmf(rV, rank = 2:5, method = "brunet", 
                nrun = 10, seed = 123456, .opt = "v-p")

# plot
pdf("SNV_NMF_rank_survey_estimate.pdf",
    width = 5,
    height = 5)
plot(estimate)
dev.off()

pdf("SNV_NMF_rank_survey_random.pdf",
    width = 5,
    height = 5)
plot(rand)
dev.off()
```

#### De novo signature extraction
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells/signature/cell")

# get similarity between samples
# cos_sim_samples <- cos_sim_matrix(mut_mat_cell, mut_mat_cell)
# 
# pdf("SNV_sample_cosine_heatmap.pdf",
#     width = 5,
#     height = 5)
# plot_cosine_heatmap(cos_sim_samples, cluster_rows = TRUE, cluster_cols = TRUE)
# dev.off()

# extract signatures
nmf_res_snv <- extract_signatures(mut_mat_cell, 
                              rank = 2, 
                              nrun = 1000, 
                              nmf_type = "regular",
                              single_core = FALSE)

# name signatures
colnames(nmf_res_snv$signatures) <- c("SBSB", "SBSA")
rownames(nmf_res_snv$contribution) <- c("SBSB", "SBSA")

write.csv(nmf_res_snv, "nmf_res_snv.csv", quote = FALSE)

# plot profile
pdf("SNV_de_novo_signature_profile.pdf",
    width = 7,
    height = 4)
plot_96_profile(nmf_res_snv$signatures,
                ymax= 0.06,
                condensed = TRUE) +
  scale_y_continuous(breaks = seq(0, 0.1, by = 0.01))
dev.off()

# compare original vs reconstructed
pdf("SNV_de_novo_signature_profile_reconstruction.pdf",
    width = 5,
    height = 5)
plot_original_vs_reconstructed(mut_mat_cell, nmf_res_snv$reconstructed, 
                               y_intercept = 0.95)
dev.off()

# plot contribution
pdf("SNV_de_novo_profile_contribution_relative.pdf",
    width = 4,
    height = 5)
plot_contribution(nmf_res_snv$contribution, 
                  nmf_res_snv$signature,
                  mode = "relative",
                  coord_flip = TRUE,
                  index = c(7:9, 13:15, 1:6, 10:12, 16:18)) +
  theme_Publication(base_size = 14)
dev.off()


pdf("SNV_de_novo_profile_contribution_absolute.pdf",
    width = 4,
    height = 5)
plot_contribution(nmf_res_snv$contribution, 
                  nmf_res_snv$signature,
                  mode = "absolute",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()

pdf("SNV_de_novo_profile_contribution_heatmap.pdf",
    width = 5,
    height = 5)
plot_contribution_heatmap(nmf_res_snv$contribution, 
                          cluster_samples = TRUE, 
                          cluster_sigs = TRUE)
dev.off()

# similarity between samples at signature level
cos_sim_samples_signatures <- cos_sim_matrix(mut_mat_cell, nmf_res_snv$signature)

pdf("SNV_de_novo_signature_cosine_heatmap.pdf",
    width = 5,
    height = 5)
plot_cosine_heatmap(cos_sim_samples_signatures, 
                    cluster_rows = TRUE, cluster_cols = TRUE)
dev.off()
```

#### COSMIC Signature refitting
```{r}
# compare to known signatures
cosmic_signatures = get_known_signatures(muttype = "snv",
                                         source = "COSMIC",
                                         tissue_type = NULL,
                                         incl_poss_artifacts = TRUE)

# get cosine similarity of all COSMIC signatures to the denovo 

# SBSA
sbsa.cos_sim_all <- apply(cosmic_signatures, 2, function(x) cos_sim(x, nmf_res_snv$signatures[,"SBSA"]))
sbsa.cos_sim_all <- sbsa.cos_sim_all[order(sbsa.cos_sim_all, decreasing = TRUE)]

pdf("SNV_signature_refit_COSMIC_SBSA_vs_SBS40.pdf",
    width = 7,
    height = 4)
plot_compare_profiles(nmf_res_snv$signatures[,"SBSA"],
  cosmic_signatures[, "SBS40"],
  profile_names = c("SBSA", "SBS40"),
  profile_ymax = 0.06,
  condensed = TRUE)
dev.off()


# SBSB
sbsb.cos_sim_all <- apply(cosmic_signatures, 2, function(x) cos_sim(x, nmf_res_snv$signatures[,"SBSB"]))
sbsb.cos_sim_all <- sbsb.cos_sim_all[order(sbsb.cos_sim_all, decreasing = TRUE)]

pdf("SNV_signature_refit_COSMIC_SBSB_vs_SBS5.pdf",
    width = 7,
    height = 4)
plot_compare_profiles(nmf_res_snv$signatures[,"SBSB"],
  cosmic_signatures[, "SBS5"],
  profile_names = c("SBSB", "SBS5"),
  profile_ymax = 0.06,
  condensed = TRUE
)
dev.off()

pdf("SNV_signature_refit_COSMIC_SBSB_vs_SBS25.pdf",
    width = 7,
    height = 4)
plot_compare_profiles(nmf_res_snv$signatures[,"SBSB"],
  cosmic_signatures[, "SBS25"],
  profile_names = c("SBSB", "SBS25"),
  profile_ymax = 0.06,
  condensed = TRUE
)
dev.off()

# bootstrapped refitting to ensure stability
# contri_boots <- fit_to_signatures_bootstrapped(mut_mat_cell,
#   cosmic_signatures,
#   n_boots = 1000,
#   method = "strict"
# )
# 
# pdf("SNV_refit_signature_bootstrap.pdf",
#     width = 10,
#     height = 10)
# plot_bootstrapped_contribution(contri_boots,
#                                mode = "relative")
# dev.off()

# get top 10 COSMIC signatures that correalte the bset to each de novo signature
combined_top <- unique(c(names(sbsa.cos_sim_all[1:10]), names(sbsb.cos_sim_all[1:10])))

fit_res_snv <- fit_to_signatures_strict(mut_mat_cell, cosmic_signatures[,colnames(cosmic_signatures) %in% combined_top], max_delta = 0.002, method = "best_subset")

# compare original vs reconstructed
pdf("SNV_signature_refit_COSMIC_profile_reconstruction.pdf",
    width = 5,
    height = 5)
plot_original_vs_reconstructed(mut_mat_cell, fit_res_snv$fit_res$reconstructed, 
                               y_intercept = 0.95)
dev.off()

# plot contribution
pdf("SNV_signature_refit_COSMIC_profile_contribution_relative.pdf",
    width = 5,
    height = 5)
plot_contribution(fit_res_snv$fit_res$contribution, 
                  mode = "relative",
                  coord_flip = FALSE) +
  theme_Publication(base_size = 14)
dev.off()


pdf("SNV_signature_refit_COSMIC_profile_contribution_absolute.pdf",
    width = 8,
    height = 8)
plot_contribution(fit_res_snv$fit_res$contribution, 
                  mode = "absolute",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()

pdf("SNV_signature_refit_COSMIC_contribution_heatmap.pdf",
    width = 10,
    height = 5)
plot_contribution_heatmap(fit_res_snv$fit_res$contribution, 
                          cluster_samples = TRUE, 
                          cluster_sigs = TRUE)
dev.off()

# similarity between samples at signature level
cos_sim_samples_signatures <- cos_sim_matrix(mut_mat_cell, COSMIC_signatures)

pdf("SNV_signature_refit_COSMIC_cosine_heatmap.pdf",
    width = 10,
    height = 5)
plot_cosine_heatmap(cos_sim_samples_signatures, 
                    cluster_rows = TRUE, cluster_cols = TRUE)
dev.off()
```

#### SIGNAL exposure signature refitting
- Purpose is to see what mutagen the ENU signature most resembles
```{r}
# compare to known signatures
signal_signatures = get_known_signatures(muttype = "snv",
                                         source = "SIGNAL",
                                         tissue_type = NULL,
                                         sig_type = "exposure",
                                         incl_poss_artifacts = TRUE)

# get cosine similarity of all SIGNAL signatures to the denovo 

# SBSA
sbsa.cos_sim_all <- apply(signal_signatures, 2, function(x) cos_sim(x, nmf_res_snv$signatures[,"SBSA"]))
sbsa.cos_sim_all <- sbsa.cos_sim_all[order(sbsa.cos_sim_all, decreasing = TRUE)]

pdf("SNV_signature_refit_SIGNAL_SBSA_vs_Control.pdf",
    width = 7,
    height = 4)
plot_compare_profiles(nmf_res_snv$signatures[,"SBSA"],
  signal_signatures[, "Control"],
  profile_names = c("SBSA", "Control"),
  profile_ymax = 0.06,
  condensed = TRUE
)
dev.off()

# SBSB
sbsb.cos_sim_all <- apply(signal_signatures, 2, function(x) cos_sim(x, nmf_res_snv$signatures[,"SBSB"]))
sbsb.cos_sim_all <- sbsb.cos_sim_all[order(sbsb.cos_sim_all, decreasing = TRUE)]

pdf("SNV_signature_refit_SIGNAL_SBSB_vs_ENU.pdf",
    width = 7,
    height = 4)
plot_compare_profiles(nmf_res_snv$signatures[,"SBSB"],
  signal_signatures[, "ENU..400.uM."],
  profile_names = c("SBSB", "ENU"),
  profile_ymax = 0.06,
  condensed = TRUE
)
dev.off()

# bootstrapped refitting to ensure stability
# contri_boots <- fit_to_signatures_bootstrapped(mut_mat_cell,
#   signal_signatures,
#   n_boots = 1000,
#   method = "strict"
# )
# 
# pdf("SNV_refit_signature_bootstrap.pdf",
#     width = 10,
#     height = 10)
# plot_bootstrapped_contribution(contri_boots,
#                                mode = "relative")
# dev.off()

# get top 10 signal signatures that correalte the bset to each de novo signature
combined_top <- unique(c(names(sbs40.cos_sim_all[1:10]), names(sbsa.cos_sim_all[1:10])))

fit_res_snv <- fit_to_signatures_strict(mut_mat_cell, signal_signatures[,colnames(signal_signatures) %in% combined_top], max_delta = 0.002, method = "best_subset")

# compare original vs reconstructed
pdf("SNV_signature_refit_SIGNAL_profile_reconstruction.pdf",
    width = 5,
    height = 5)
plot_original_vs_reconstructed(mut_mat_cell, fit_res_snv$fit_res$reconstructed, 
                               y_intercept = 0.95)
dev.off()

# plot contribution
pdf("SNV_signature_refit_SIGNAL_profile_contribution_relative.pdf",
    width = 5,
    height = 5)
plot_contribution(fit_res_snv$fit_res$contribution, 
                  mode = "relative",
                  coord_flip = FALSE) +
  theme_Publication(base_size = 14)
dev.off()


pdf("SNV_signature_refit_SIGNAL_profile_contribution_absolute.pdf",
    width = 8,
    height = 8)
plot_contribution(fit_res_snv$fit_res$contribution, 
                  mode = "absolute",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()

pdf("SNV_signature_refit_SIGNAL_contribution_heatmap.pdf",
    width = 10,
    height = 5)
plot_contribution_heatmap(fit_res_snv$fit_res$contribution, 
                          cluster_samples = TRUE, 
                          cluster_sigs = TRUE)
dev.off()

# similarity between samples at signature level
cos_sim_samples_signatures <- cos_sim_matrix(mut_mat_cell, signal_signatures)

pdf("SNV_signature_refit_SIGNAL_cosine_heatmap.pdf",
    width = 10,
    height = 5)
plot_cosine_heatmap(cos_sim_samples_signatures, 
                    cluster_rows = TRUE, cluster_cols = TRUE)
dev.off()
```

### INDEL

#### Calculate factorization rank using NMF
- How to choose the rank: Choose the smallest rank for which cophenetic correlation coefficient starts decreasing. Another approach is to choose the rank for which the plot of the residual sum of squares (RSS) between the input matrix and its estimate shows an inflection point
- Choosing rank = 2
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells/signature/condition")

# add small pseudocount
indel_counts_condition <- indel_counts_condition + 0.0001

# generate an estimate rank plot
estimate <- nmf(indel_counts_condition, rank = 2:5, method = "brunet", 
                nrun = 10, seed = 123456, .opt = "v-p")

# run same estimation on randomized data
rV <- randomize(indel_counts_condition)
rand <- nmf(rV, rank = 2:5, method = "brunet", 
                nrun = 10, seed = 123456, .opt = "v-p")

# plot
pdf("INDEL_NMF_rank_survey_estimate.pdf",
    width = 5,
    height = 5)
plot(estimate)
dev.off()

pdf("INDEL_NMF_rank_survey_random.pdf",
    width = 5,
    height = 5)
plot(rand)
dev.off()
```

#### De novo signature extraction
- Using COSMIC database for comparison
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/4-mutational-signatures/without-control-cells/signature/condition")

# get similarity between samples
cos_sim_samples <- cos_sim_matrix(indel_counts, indel_counts)

pdf("INDEL_sample_cosine_heatmap.pdf",
    width = 5,
    height = 5)
plot_cosine_heatmap(cos_sim_samples, cluster_rows = TRUE, cluster_cols = TRUE)
dev.off()

# extract signatures
nmf_res_indel <- extract_signatures(indel_counts_condition, 
                              rank = 1, 
                              nrun = 1000, 
                              nmf_type = "regular",
                              single_core = FALSE)

# name signatures
colnames(nmf_res_indel$signatures) <- c("IDA")
rownames(nmf_res_indel$contribution) <- c("IDA")

# bootstrapped refitting to ensure stability
# contri_boots <- fit_to_signatures_bootstrapped(indel_counts,
#   nmf_res_indel$signatures,
#   n_boots = 1000,
#   method = "strict"
# )
# 
# pdf("INDEL_de_novo_signature_context.pdf",
#     width = 10,
#     height = 5)
# plot_indel_contexts(nmf_res_indel$signatures,
#                     same_y = TRUE,
#                     extra_labels = FALSE,
#                     condensed = TRUE) +
#   theme_Publication(base_size = 12) +
#   theme(axis.text.x = element_text(size = 5),
#         legend.position = "right",
#         legend.direction = "vertical",
#         strip.text = element_text(size = 10)) +
#   labs(y = "Relative contribution", x = "No. repeat units")
# dev.off()

pdf("INDEL_de_novo_signature_main_context.pdf",
    width = 4,
    height = 4)
plot_main_indel_contexts(nmf_res_indel$signatures,
                    same_y = TRUE) +
    theme_Publication(base_size = 12) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "right",
        legend.direction = "vertical",
        strip.text = element_text(size = 10)) +
  labs(y = "Relative contribution")
dev.off()

# compare original vs reconstructed
pdf("INDEL_de_novo_signature_profile_reconstruction.pdf",
    width = 5,
    height = 5)
plot_original_vs_reconstructed(indel_counts_condition, nmf_res_indel$reconstructed, 
                               y_intercept = 0.95)
dev.off()

# plot contribution
pdf("INDEL_de_novo_profile_contribution_relative.pdf",
    width = 4,
    height = 5)
plot_contribution(nmf_res_indel$contribution, 
                  nmf_res_indel$signature,
                  mode = "relative",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()


pdf("INDEL_de_novo_profile_contribution_absolute.pdf",
    width = 4,
    height = 5)
plot_contribution(nmf_res_indel$contribution, 
                  nmf_res_indel$signature,
                  mode = "absolute",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()

# pdf("INDEL_de_novo_profile_contribution_heatmap.pdf",
#     width = 5,
#     height = 5)
# plot_contribution_heatmap(nmf_res_indel$contribution, 
#                           cluster_samples = TRUE, 
#                           cluster_sigs = TRUE)
# dev.off()

# # similarity between samples at signature level
# cos_sim_samples_signatures <- cos_sim_matrix(indel_counts_condition, nmf_res_indel$signature)
# 
# pdf("INDEL_de_novo_signature_cosine_heatmap.pdf",
#     width = 5,
#     height = 5)
# plot_cosine_heatmap(cos_sim_samples_signatures, 
#                     cluster_rows = TRUE, cluster_cols = TRUE)
# dev.off()
```

#### COSMIC Signature refitting
```{r}
# compare to known signatures
indel_signatures <- get_known_signatures(muttype = "indel", incl_poss_artifacts = TRUE)

# get similarity
indel.cos_sim_all <- apply(indel_signatures, 2, function(x) cos_sim(x, nmf_res_indel$signatures[,1]))
indel.cos_sim_all <- indel.cos_sim_all[order(indel.cos_sim_all, decreasing = TRUE)]

pdf("INDEL_signature_refit_COSMIC_IDA_vs_ID1.pdf",
    width = 7,
    height = 4)
plot_compare_profiles(nmf_res_indel$signatures[,"IDA"],
  indel_signatures[, "ID1"],
  profile_names = c("IDA", "ID1"),
  profile_ymax = 0.06,
  condensed = TRUE,
  colors = NA)
dev.off()

fit_res_indel <- fit_to_signatures_strict(indel_counts_condition, indel_signatures[,1:10], max_delta = 0.002, method = "best_subset")

# # bootstrapped refitting to ensure stability
# contri_boots <- fit_to_signatures_bootstrapped(indel_counts,
#   cosmic_signatures,
#   n_boots = 1000,
#   method = "strict"
# )
# 
# pdf("INDEL_refit_signature_bootstrap.pdf",
#     width = 10,
#     height = 4)
# plot_bootstrapped_contribution(contri_boots,
#                                mode = "relative")
# dev.off()

# compare original vs reconstructed
pdf("INDEL_refit_signature_profile_reconstruction.pdf",
    width = 5,
    height = 5)
plot_original_vs_reconstructed(indel_counts_condition, fit_res_indel$fit_res$reconstructed, 
                               y_intercept = 0.95)
dev.off()

# plot contribution
pdf("INDEL_refit_signature_profile_contribution_relative.pdf",
    width = 8,
    height = 8)
plot_contribution(fit_res_indel$fit_res$contribution, 
                  mode = "relative",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()


pdf("INDEL_refit_signature_profile_contribution_absolute.pdf",
    width = 8,
    height = 8)
plot_contribution(fit_res_indel$fit_res$contribution, 
                  mode = "absolute",
                  coord_flip = TRUE) +
  theme_Publication(base_size = 14)
dev.off()

pdf("INDEL_refit_profile_contribution_heatmap.pdf",
    width = 10,
    height = 5)
plot_contribution_heatmap(fit_res_indel$fit_res$contribution, 
                          cluster_samples = TRUE, 
                          cluster_sigs = TRUE)
dev.off()

# similarity between samples at signature level
cos_sim_samples_signatures <- cos_sim_matrix(indel_counts, indel_signatures)

pdf("INDEL_refit_signature_cosine_heatmap.pdf",
    width = 10,
    height = 5)
plot_cosine_heatmap(cos_sim_samples_signatures, 
                    cluster_rows = TRUE, cluster_cols = TRUE)
dev.off()
```

#### Signature potential damage analysis
- Can use different subset of genes (e.g. growth, nonessential) - Looking at genes determined to be expressed in IMR90
- Using de novo signatures
- ratio_by_background is normalized ratio. Noramlized ratio of 2 means that a signature is twice as likely to cause that type of mutation, compared to a completely random “flat” signature
```{r}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/annotation/rna")

# rna expression table
rna <- read.table("ENCFF245ZOB-IMR90-total-RNA-seq-hg19.tsv", header = TRUE)

# remove version from ensemble gene id
rna$gene_id <- gsub("\\..*", "", rna$gene_id)

# plot TPM distribution
nrow(rna)
hist(log2(rna$TPM))

# zFPKM calculation
zfpkm <- zFPKM(rna[,"TPM", drop = FALSE])
zFPKMPlot(rna[,"TPM", drop = FALSE])

# plot zfpkm  distribution
hist(zfpkm$TPM)

# keep genes with > -3 zfpkm
rna.filter <- rna[zfpkm$TPM > -3,]

# plot filtered TPM distribution
nrow(rna.filter)
hist(log2(rna.filter$TPM))

# get entrz id
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
gene_names <- getBM(attributes = c('ensembl_gene_id', 'external_gene_name', 'entrezgene_id'),
                filters = 'ensembl_gene_id',
                values = rna$gene_id, 
                mart = ensembl)

rna.filter <- merge(rna.filter, gene_names, by.x = "gene_id", by.y = "ensembl_gene_id")
sum(duplicated(rna.filter$entrezgene_id))
rna.filter <- rna.filter[!duplicated(rna.filter$entrezgene_id),] # remove duplicates

# get the ratio of “stop gain”, “mismatch”, “synonymous” and “splice site” mutations can be determined per genomic context
contexts <- rownames(indel_counts)
context_mismatches <- context_potential_damage_analysis(contexts, TxDb.Hsapiens.UCSC.hg19.knownGene, BSgenome.Hsapiens.UCSC.hg19, rna.filter$entrezgene_id)

# get the ratios per signature
sig_damage_indel <- signature_potential_damage_analysis(nmf_res_indel$signatures, contexts, context_mismatches)
sig_damage_indel <- sig_damage_indel[order(sig_damage_indel$ratio_by_background, decreasing = TRUE),]
```

# Strand Bias analysis
## By condition
### Transcriptional strand bias
```{r}

```

### Replicative strand bias

```{r}

```


# Mutational patterns of genomic regions
- only do interesting regions - methylation

## Load genomic regions
```{r}
# generic


```

- By condition
```{r}
# signature contribution
plot_contribution_heatmap(nmf_res_region$contribution, 
                          cluster_samples = TRUE, 
                          cluster_sigs = TRUE)

# mutation specturm
type_occurrences_region <- mut_type_occurrences(grl_region, ref_genome)
plot_spectrum_region(type_occurrences_region)
```


# Lesion Segregation