---
title: "Load regions of interest for pathway oe ratio calculation"
output: html_document
date: "2023-10-10"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

## Load libraries
```{r}
library(VariantAnnotation)
library(GenomicRanges)
library(GenomicFeatures)
library(rtracklayer)
library(MutationalPatterns)
library(ggplot2)
library(AnnotationHub)
library(tximport)
library(zFPKM)
library(dplyr)
library(tidyr)
library(SummarizedExperiment)
library(biomaRt)
library(clusterProfiler)
library(ggbio)
library(tibble)
library(foreach)
library(doParallel)

library(org.Hs.eg.db)
library(BSgenome.Hsapiens.UCSC.hg19)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
TxDb.Hsapiens.UCSC.hg19.knownGene <- keepStandardChromosomes(TxDb.Hsapiens.UCSC.hg19.knownGene)
TxDb.Hsapiens.UCSC.hg19.knownGene <- dropSeqlevels(TxDb.Hsapiens.UCSC.hg19.knownGene, c("chrY", "chrM"), pruning.mode="coarse")
hg19_size <- sum(seqlengths(TxDb.Hsapiens.UCSC.hg19.knownGene))
```

## Functions
```{r}
# function to create bed file from granges
export.granges.bed <- function(gr, outputDir, fileName) {
  if (!dir.exists(outputDir)) {
    dir.create(outputDir, recursive = TRUE)
  }

  sanitizeFileName <- function(name) {
    # Replace problematic characters with an underscore
    name <- gsub("[/\\\\:*?\"<>|,()']", "_", name)
    # Replace multiple underscores with a single one
    gsub("_+", "_", name)
  }

  # Sanitize file name
  name <- sanitizeFileName(fileName)

  bed <- data.frame(
    seqnames = as.character(seqnames(gr)),
    start = start(gr) - 1, # bed uses 0 coordinate system
    end = end(gr),
    name = seq(1:length(gr)),
    score = rep(1, length(gr)),
    stringsAsFactors = FALSE
  )

  filePath <- paste0(outputDir, "/", name)
  write.table(bed, file = filePath, quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)
}

# function to create bed files from a list of granges
export.grangesList.bed <- function(grList, outputDir) {
  if (!dir.exists(outputDir)) {
    dir.create(outputDir, recursive = TRUE)
  }

  sanitizeFileName <- function(name) {
    # Replace problematic characters with an underscore
    name <- gsub("[/\\\\:*?\"<>|,()']", "_", name)
    # Replace multiple underscores with a single one
    gsub("_+", "_", name)
  }

  for (i in seq_along(grList)) {
    name <- sanitizeFileName(names(grList)[i])
    gr <- grList[[i]]

    bed <- data.frame(
      seqnames = as.character(seqnames(gr)),
      start = start(gr) - 1, # bed uses 0 coordinate system
      end = end(gr),
      name = seq(1:length(gr)),
      score = rep(1, length(gr)),
      stringsAsFactors = FALSE
    )

    fileName <- paste0(outputDir, "/", name, ".bed")
    write.table(bed, file = fileName, quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)
  }
}

# function to subset granges based on genes from a pathway enrichment
subset_regions_pathways <- function(expressed_regions, enrich_result, intron = FALSE){
  
  # split 
  exon <- expressed_regions[["exons"]]
  if(intron){introns <- expressed_regions[["introns"]]}
  promoter <- expressed_regions[["promoters"]]
  enhancer <- expressed_regions[["enhancers"]]

  pathways <- GRangesList()
  for(i in 1:nrow(enrich_result@result)){
    # get genes
    genes <- unlist(strsplit(enrich_result@result$geneID[i], "/"))
    
    # susbet lists
    exons.subset <- exon[exon$entrezgene_id %in% genes]
    if(intron){introns.subset <- introns[introns$entrezgene_id %in% genes]}
    promoter.subset <- promoter[promoter$entrezgene_id %in% genes]
    enhancer.subset <- enhancer[enhancer$entrezgene_id %in% genes]
    
    # merge 
    merged.subset <- c(exons.subset, 
                       promoter.subset,
                       enhancer.subset)
    
    if(intron){
      merged.subset <- c(exons.subset,
                         introns.subset,
                       promoter.subset,
                       enhancer.subset)
    }
    
    # add to pathway list
    pathways <- append(pathways, GRangesList(merged.subset))
    
    # name
    names(pathways)[i] <- enrich_result@result$Description[i]
  }
  
  return(pathways)
}

# function to create pathways with random genes
# for each pathway get the number of genes
# randomly sample genes from all the genes that were found in enrichment
# create list list of 100000 different shuffled objects
subset_regions_pathways_shuffle <- function(expressed_regions, enrich_result, intron = TRUE, n_iterations = 10000) {
  # Split expressed_regions into types
  exon <- expressed_regions[["exons"]]
  if(intron){introns <- expressed_regions[["introns"]]}
  promoter <- expressed_regions[["promoters"]]
  enhancer <- expressed_regions[["enhancers"]]
  
  # Get all unique genes found in all pathways
  all.genes <- unique(unlist(sapply(enrich_result@result$geneID, function(x) strsplit(x, "/"))))
  
  # Initialize a list to store each iteration's results
  all_pathways_list <- vector("list", n_iterations)
  
  # Perform shuffling n_iterations times
  for (n in 1:n_iterations) {
    pathways <- GRangesList()
    
    for (i in 1:nrow(enrich_result@result)) {
      # Get number of genes in each pathway
      genes.length <- length(unlist(strsplit(enrich_result@result$geneID[i], "/")))
      
      # Randomly sample genes from all the genes that were found in enrichment
      genes <- sample(all.genes, genes.length)
      
      # Subset lists
      exons.subset <- exon[exon$entrezgene_id %in% genes]
      if(intron){introns.subset <- introns[introns$entrezgene_id %in% genes]}
      promoter.subset <- promoter[promoter$entrezgene_id %in% genes]
      enhancer.subset <- enhancer[enhancer$entrezgene_id %in% genes]
      
      # merge 
      merged.subset <- c(exons.subset,
                         promoter.subset,
                         enhancer.subset)
  
      if(intron){
        merged.subset <- c(exons.subset,
                           introns.subset,
                         promoter.subset,
                         enhancer.subset)
      }

      # Add to pathway list
      pathways <- append(pathways, GRangesList(merged.subset))
      
      # Name the current pathway according to the description in enrich_result
      names(pathways)[i] <- enrich_result@result$Description[i]
    }
    
    # Store the shuffled pathways of the current iteration
    all_pathways_list[[n]] <- pathways
  }
  
  return(all_pathways_list)
}


# function to subset granges based on genes from a pathway enrichment
# subset_regions_gene_sets <- function(expressed_regions, gene_sets){
#   
#   # split 
#   exon <- expressed_regions[["exons"]]
#   promoter <- expressed_regions[["promoters"]]
#   enhancer <- expressed_regions[["enhancers"]]
# 
#   pathways <- GRangesList()
#   for(i in 1:length(gene_sets)){
#     # get genes
#     genes <- gene_sets[[i]][,"entrezgene_id"]
#     
#     
#     # susbet lists
#     exons.subset <- exon[exon$entrezgene_id %in% genes]
#     promoter.subset <- promoter[promoter$entrezgene_id %in% genes]
#     enhancer.subset <- enhancer[enhancer$entrezgene_id %in% genes]
#     
#     # merge 
#     merged.subset <- c(exons.subset, 
#                        promoter.subset,
#                        enhancer.subset)
#     
#     # add to pathway list
#     pathways <- append(pathways, GRangesList(merged.subset))
#     
#     # name
#     names(pathways)[i] <- names(gene_sets)[i]
#   }
#   
#   return(pathways)
# }
# 
# subset_regions_genes <- function(expressed_regions, genes){
#   
#   # split 
#   exon <- expressed_regions[["exons"]]
#   promoter <- expressed_regions[["promoters"]]
#   enhancer <- expressed_regions[["enhancers"]]
#     
#   # susbet lists
#   exons.subset <- exon[exon$entrezgene_id %in% genes]
#   promoter.subset <- promoter[promoter$entrezgene_id %in% genes]
#   enhancer.subset <- enhancer[enhancer$entrezgene_id %in% genes]
#   
#   # merge 
#   pathways <- c(exons.subset, 
#                      promoter.subset,
#                      enhancer.subset)
#     
#   return(pathways)
# }

load_bed <- function(dir, file, hg19_size, includeMetadata = FALSE, reduce = TRUE) {
  setwd(dir)
  
  # Read the BED file
  if (includeMetadata) {
    bed <- read.table(file, header = FALSE, stringsAsFactors = FALSE)
    # Assuming the first three columns are always chr, start, end
    colnames(bed)[1:3] <- c("chr", "start", "end")
  } else {
    bed <- read.table(file, header = FALSE, stringsAsFactors = FALSE)
    colnames(bed) <- c("chr", "start", "end")
  }

  # Create GRanges object
  bed <- makeGRangesFromDataFrame(bed, start.field = "start", end.field = "end", keep.extra.columns = includeMetadata)
  bed <- keepStandardChromosomes(bed, pruning.mode = "coarse")
  bed <- dropSeqlevels(bed, c("chrY", "chrM"), pruning.mode = "coarse")
  
  # merge overlapping regions
  if(reduce){
    bed <- reduce(bed)
  }
  
  # Calculate and print coverage of genome
  print(sum(width(bed)) / hg19_size)

  return(bed)
}

export_GRanges_to_VCF <- function(gr_obj, sample_name) {
  # Step 1: Convert to data frame
  gr_df <- as.data.frame(gr_obj)
  
  # Step 2: Convert columns to appropriate types
  gr_df$REF <- as.character(gr_df$REF)
  gr_df$ALT <- sapply(gr_df$ALT, function(alt) {
    as.character(alt)[1]
  })

  # Step 3: Generate paramRangeID if necessary
  gr_df$paramRangeID <- paste0(gr_df$seqnames, ":", gr_df$start, "_", gr_df$REF, "/", gr_df$ALT)
  
  # Step 4: Create VCF header
  vcf_header <- paste0('##fileformat=VCFv4.3\n',
                       '#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO\tFORMAT\t', 
                       sample_name)
  
  # Step 5: Generate VCF body
  vcf_body <- apply(gr_df, 1, function(row) {
    paste(row['seqnames'], row['start'], row['paramRangeID'], row['REF'], row['ALT'], row['QUAL'], row['FILTER'], '.', '.', '.', sep = '\t')
  })
  vcf_body <- paste(vcf_body, collapse = '\n')
  
  # Step 6: Combine header and body and write to file
  vcf_text <- paste(vcf_header, vcf_body, sep = '\n')
  writeLines(vcf_text, paste0(sample_name, ".vcf"))
}

theme_Publication <- function(base_size=16, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family = "")
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               #legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

wrap_labels <- function(labels, width = 5) {
  wrapped_labels <- str_wrap(labels, width)
  names(wrapped_labels) <- names(labels)
  return(wrapped_labels)
}
```

# Get expressed regions

## Load RNA-Seq data
- Loading control and ENU samples
- IMPROVEMENT: account for differences between control and ENU groups
```{r}
activeGenes <- readRDS("activeGenes.RDS")
nonactiveGenes <- readRDS("nonactiveGenes.RDS")
```

## Get exons

### Expressed
```{r}
# get all exons
exons <- exonsBy(TxDb.Hsapiens.UCSC.hg19.knownGene, by="gene")
# exons <- unlist(exons)
# exons$entrezgene_id <- names(exons)

# subset for expressed genes
exons.expressed <- exons[names(exons) %in% activeGenes]
length(exons.expressed)/length(exons)  # proportion of total
sum(unlist(exons.expressed)@ranges@width)/hg19_size # coverage of genome

# unlist them
exons.expressed <- unlist(exons.expressed)
exons.expressed$entrezgene_id <- names(exons.expressed)
```

### Non-expressed
```{r}
# subset for nonexpressed genes
exons.nonexpressed <- exons[names(exons) %in% nonactiveGenes]
length(exons.nonexpressed)/length(exons)  # proportion of total
sum(unlist(exons.nonexpressed)@ranges@width)/hg19_size # coverage of genome

# unlist them
exons.nonexpressed <- unlist(exons.nonexpressed)
exons.nonexpressed$entrezgene_id <- names(exons.nonexpressed)
```

## Get introns

### Expressed
```{r}
# Get introns by transcript
introns <- intronsByTranscript(TxDb.Hsapiens.UCSC.hg19.knownGene)

# subset for expressed genes
introns.expressed <- introns[names(introns) %in% activeGenes]
length(introns.expressed)/length(introns)  # proportion of total
sum(unlist(introns.expressed)@ranges@width)/hg19_size # coverage of genome

# unlist them
introns.expressed <- unlist(introns.expressed)
introns.expressed$entrezgene_id <- names(introns.expressed)
```

### Non-expressed
```{r}
# subset for nonexpressed genes
introns.nonexpressed <- introns[names(introns) %in% nonactiveGenes]
length(introns.nonexpressed)/length(introns)  # proportion of total
sum(unlist(introns.nonexpressed)@ranges@width)/hg19_size # coverage of genome

# unlist them
introns.nonexpressed <- unlist(introns.nonexpressed)
introns.nonexpressed$entrezgene_id <- names(introns.nonexpressed)
```

## Get promoters (+5KB)

### Expressed
```{r}
# get all promoters
promoters <- genes(TxDb.Hsapiens.UCSC.hg19.knownGene)
promoters <- resize(promoters, width = 5000, fix = "start")
promoters$entrezgene_id <- promoters$gene_id

# subset for expressed genes
promoters.expressed <- promoters[promoters$entrezgene_id %in% activeGenes]
length(promoters.expressed)/length(promoters) # proportion of total
sum(promoters.expressed@ranges@width)/hg19_size # coverage of genome
```
### Non-expressed
```{r}
# subset for nonexpressed genes
promoters.nonexpressed <- promoters[promoters$entrezgene_id %in% nonactiveGenes]
length(promoters.nonexpressed)/length(promoters) # proportion of total
sum(promoters.nonexpressed@ranges@width)/hg19_size # coverage of genome
```

## Get connected enhancer regions

### Expressed
```{r}
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")

enhancers <- read.table("BSS00720_EPIMAP-IMR90-links_corr_only-hg19.tsv", header = TRUE)
enhancers <- makeGRangesFromDataFrame(enhancers, start.field = "enhStart", end.field = "enhEnd", keep.extra.columns=TRUE)
enhancers <- keepStandardChromosomes(enhancers)
enhancers <- dropSeqlevels(enhancers, c("chrY", "chrM"), pruning.mode="coarse")
sum(enhancers@ranges@width)/hg19_size # coverage of genome

# convert gene names
convert <- bitr(enhancers$gene, "ENSEMBL", "ENTREZID", org.Hs.eg.db)
sum(is.na(convert$ENTREZID[match(enhancers$gene, convert$ENSEMBL)]))/length(enhancers$gene) # proportion of missing

# map
entrez_vector <- setNames(convert$ENTREZID, convert$ENSEMBL)
entrez_ids <- entrez_vector[enhancers$gene]
mcols(enhancers)$entrezgene_id <- entrez_ids
sum(is.na(enhancers$entrezgene_id)/length(enhancers$entrezgene_id)) # proportion of missing

# subset for expressed genes
enhancers.expressed <- enhancers[enhancers$entrezgene_id %in% activeGenes]
length(enhancers.expressed)/length(enhancers)
sum(enhancers.expressed@ranges@width)/hg19_size # coverage of genome
```

### Non-Expressed
```{r}
# subset for nonexpressed genes
enhancers.nonexpressed <- enhancers[enhancers$entrezgene_id %in% nonactiveGenes]
length(enhancers.nonexpressed)/length(enhancers)
sum(enhancers.nonexpressed@ranges@width)/hg19_size # coverage of genome
```

## Combine exons, introns, pomoters, enhancers

### Expressed
```{r}
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")

expressed_regions_exons_introns_promoters_enhancers <- GRangesList(exons = exons.expressed,
                                 introns = introns.expressed,
                                 promoters = promoters.expressed,
                                 enhancers = enhancers.expressed,
                                 compress=FALSE)

# save
export.grangesList.bed(expressed_regions_exons_introns_promoters_enhancers,
                       "/Users/ronaldcutler/Dropbox (EinsteinMed)/Vijg-lab/Projects/mutation accumulation/231009 multiple ENU analysis/6-pathway/1-load-data")
saveRDS(expressed_regions, "expressed_regions_exons_introns_promoters_enhancers.RDS")

# plot
vec <- unlist(lapply(expressed_regions_exons_introns_promoters_enhancers, function(x) sum(x@ranges@width)/hg19_size))
df <- tibble(Category = names(vec), Value = as.numeric(vec))
df$Category <- factor(df$Category, levels = df$Category)

pdf("expressed_regions_exons_introns_promoters_enhancers.pdf",
    width = 5,
    height = 5)
ggplot(df, aes(x = Category, y = Value)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Expressed regions", y = "Proportion of genome") +
  theme_Publication() 
dev.off()
```

### Non-expressed
```{r}
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")

nonexpressed_regions_exons_introns_promoters_enhancers <- GRangesList(exons = exons.nonexpressed,
                                    introns = introns.nonexpressed,
                                    promoters = promoters.nonexpressed,
                                    enhancers = enhancers.nonexpressed,
                                    compress=FALSE)

# save
export.grangesList.bed(nonexpressed_regions_exons_introns_promoters_enhancers,
                       "/Users/ronaldcutler/Dropbox (EinsteinMed)/Vijg-lab/Projects/mutation accumulation/231009 multiple ENU analysis/6-pathway/1-load-data")
saveRDS(expressed_regions, "nonexpressed_regions_exons_introns_promoters_enhancers.RDS")

# plot
vec <- unlist(lapply(nonexpressed_regions_exons_introns_promoters_enhancers, function(x) sum(x@ranges@width)/hg19_size))
df <- tibble(Category = names(vec), Value = as.numeric(vec))
df$Category <- factor(df$Category, levels = df$Category)

pdf("nonexpressed_regions_exons_introns_promoters_enhancers.pdf",
    width = 5,
    height = 5)
ggplot(df, aes(x = Category, y = Value)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Expressed regions", y = "Proportion of genome") +
  theme_Publication() 
dev.off()
```

## Combine
```{r}
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")

expressed_nonexpressed_regions_exons_introns_promoters_enhancers <- list(expressed = unlist(expressed_regions_exons_introns_promoters_enhancers),
                                       nonexpressed = unlist(nonexpressed_regions_exons_introns_promoters_enhancers))
```

## Combine exons, pomoters, enhancers

### Expressed
```{r}
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")

expressed_regions_exons_promoters_enhancers <- GRangesList(exons = exons.expressed,
                                 promoters = promoters.expressed,
                                 enhancers = enhancers.expressed,
                                 compress=FALSE)

# save
export.grangesList.bed(expressed_regions_exons_promoters_enhancers,
                       "/Users/ronaldcutler/Dropbox (EinsteinMed)/Vijg-lab/Projects/mutation accumulation/231009 multiple ENU analysis/6-pathway/1-load-data")
saveRDS(expressed_regions, "expressed_regions_exons_promoters_enhancers.RDS")

# plot
vec <- unlist(lapply(expressed_regions_exons_promoters_enhancers, function(x) sum(x@ranges@width)/hg19_size))
df <- tibble(Category = names(vec), Value = as.numeric(vec))
df$Category <- factor(df$Category, levels = df$Category)

pdf("expressed_regions_exons_promoters_enhancers.pdf",
    width = 5,
    height = 5)
ggplot(df, aes(x = Category, y = Value)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Expressed regions", y = "Proportion of genome") +
  theme_Publication() 
dev.off()
```

### Non-expressed
```{r}
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")

nonexpressed_regions_exons_promoters_enhancers <- GRangesList(exons = exons.nonexpressed,
                                    promoters = promoters.nonexpressed,
                                    enhancers = enhancers.nonexpressed,
                                    compress=FALSE)

# save
export.grangesList.bed(nonexpressed_regions_exons_promoters_enhancers,
                       "/Users/ronaldcutler/Dropbox (EinsteinMed)/Vijg-lab/Projects/mutation accumulation/231009 multiple ENU analysis/6-pathway/1-load-data")
saveRDS(expressed_regions, "nonexpressed_regions_exons_promoters_enhancers.RDS")

# plot
vec <- unlist(lapply(nonexpressed_regions_exons_promoters_enhancers, function(x) sum(x@ranges@width)/hg19_size))
df <- tibble(Category = names(vec), Value = as.numeric(vec))
df$Category <- factor(df$Category, levels = df$Category)

pdf("nonexpressed_regions_exons_promoters_enhancers.pdf",
    width = 5,
    height = 5)
ggplot(df, aes(x = Category, y = Value)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Expressed regions", y = "Proportion of genome") +
  theme_Publication() 
dev.off()
```

## Combine
```{r}
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")

expressed_nonexpressed_regions_exons_promoters_enhancers <- list(expressed = unlist(expressed_regions_exons_promoters_enhancers),
                                       nonexpressed = unlist(nonexpressed_regions_exons_promoters_enhancers))
```

# Pathway enrichment

## Expressed

### Gene Ontology Biological Process
```{r}
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")

# get enrichment of genes expressed
enrich.go.expressed <- enrichGO(as.character(activeGenes), 
                   OrgDb = org.Hs.eg.db, 
                   keyType = 'ENTREZID', 
                   ont = 'BP',
                   universe = as.character(rowData(se)$entrezgene_id),
                   readable = FALSE)
enrich.go.expressed <- simplify(enrich.go.expressed)
enrich.go.expressed <- mutate(enrich.go.expressed, qscore = -log(pvalue, base=10))

# ouput
write.csv(enrich.go.expressed@result, "enrich.go.expressed.csv", row.names = FALSE)
saveRDS(enrich.go.expressed, "enrich.go.expressed.RDS")

# plot
pdf("enrich.go.expressed.bar.pdf",
    width = 7,
    height = 7)
barplot(enrich.go.expressed,
        x = "qscore",
        showCategory = 20)
dev.off()

pdf("enrich.go.expressed.dot.pdf",
    width = 7,
    height = 7)
dotplot(enrich.go.expressed,
        x = "Count",
        showCategory = 20)
dev.off()
```

#### Divide into pathways

##### exons, introns, promoters, enhancers
```{r}
go_expressed_regions_exons_introns_promoters_enhancers <- subset_regions_pathways(expressed_regions_exons_introns_promoters_enhancers,
                                        enrich.go.expressed,
                                        intron = TRUE)

go_expressed_shuffled_regions_exons_introns_promoters_enhancers <- subset_regions_pathways_shuffle(expressed_regions_exons_introns_promoters_enhancers,
                                                         enrich.go.expressed,
                                                         intron = TRUE,
                                                         n_iterations = 10000)

# save
saveRDS(go_expressed_regions_exons_introns_promoters_enhancers, "go_expressed_regions_exons_introns_promoters_enhancers.RDS")
saveRDS(go_expressed_shuffled_regions_exons_introns_promoters_enhancers, "go_expressed_shuffled_regions_exons_introns_promoters_enhancers.RDS")
```

##### exons, promoters, enhancers
```{r}
go_expressed_regions_exons_promoters_enhancers <- subset_regions_pathways(expressed_regions_exons_promoters_enhancers,
                                        enrich.go.expressed,
                                        intron = TRUE)

go_expressed_shuffled_regions_exons_promoters_enhancers <- subset_regions_pathways_shuffle(expressed_regions_exons_promoters_enhancers,
                                                         enrich.go.expressed,
                                                         intron = TRUE,
                                                         n_iterations = 10000)

# save
saveRDS(go_expressed_regions_exons_promoters_enhancers, "go_expressed_regions_exons_promoters_enhancers.RDS")
saveRDS(go_expressed_shuffled_regions_exons_promoters_enhancers, "go_expressed_shuffled_regions_exons_promoters_enhancers.RDS")
```

### Essential genes
```{r, eval = FALSE}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/annotation/gene-sets")

essential <- read.table("DEG_database_eukaryote_essential_genes.csv")

# load
essential <- read.csv("AchillesCommonEssentialControls.csv", header = TRUE)
essential$Gene <- sub(" \\(.*", "", essential$Gene)

nonessential <- read.csv("AchillesNonessentialControls.csv", header = TRUE)
nonessential$Gene <- sub(" \\(.*", "", nonessential$Gene)

# convert
essential <- merge(essential, gene_names, by.x = "Gene", by.y = "external_gene_name")
sum(duplicated(essential$entrezgene_id))
essential <- essential[!duplicated(essential$entrezgene_id),] # remove duplicates

nonessential <- merge(nonessential, gene_names, by.x = "Gene", by.y = "external_gene_name")
sum(duplicated(nonessential$entrezgene_id))
nonessential <- nonessential[!duplicated(nonessential$entrezgene_id),] # remove duplicates

# combine
essential_list <- list(essential = essential,
                       nonessential = nonessential)

# divide regions by genes
essential_genes <- subset_regions_gene_sets(expressed_regions,
                                        essential_list)

# save 
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")
essential_genes <- readRDS("essential_genes.RDS")

saveRDS(essential_genes, "essential_genes.RDS")
```

## Non-Expressed

### Gene Ontology Biological Process
```{r, eval = FALSE}
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")

# get enrichment of genes expressed
enrich.go.nonexpressed <- enrichGO(as.character(nonactiveGenes), 
                   OrgDb = org.Hs.eg.db, 
                   keyType = 'ENTREZID', 
                   ont = 'BP',
                   universe = as.character(rowData(se)$entrezgene_id),
                   readable = FALSE)
enrich.go.nonexpressed <- simplify(enrich.go.nonexpressed)
enrich.go.nonexpressed <- mutate(enrich.go.nonexpressed, qscore = -log(pvalue, base=10))

# ouput
write.csv(enrich.go.nonexpressed@result, "enrich.go.nonexpressed.csv", row.names = FALSE)
saveRDS(enrich.go.nonexpressed, "enrich.go.nonexpressed.RDS")

# plot
pdf("enrich.go.nonexpressed.bar.pdf",
    width = 7,
    height = 7)
barplot(enrich.go.nonexpressed,
        x = "qscore",
        showCategory = 20)
dev.off()

pdf("enrich.go.nonexpressed.dot.pdf",
    width = 7,
    height = 7)
dotplot(enrich.go.nonexpressed,
        x = "Count",
        showCategory = 20)
dev.off()
```

#### Divide into pathways

##### exons, introns, promoters, enhancers
```{r, eval = FALSE}
go_nonexpressed_regions_exons_introns_promoters_enhancers <- subset_regions_pathways(nonexpressed_regions_exons_introns_promoters_enhancers,
                                        enrich.go.nonexpressed,
                                        intron = TRUE)

go_nonexpressed_shuffled_regions_exons_introns_promoters_enhancers <- subset_regions_pathways_shuffle(nonexpressed_regions_exons_introns_promoters_enhancers,
                                                         enrich.go.nonexpressed,
                                                         intron = TRUE,
                                                         n_iterations = 10000)

# save
saveRDS(go_nonexpressed_regions_exons_introns_promoters_enhancers, "go_nonexpressed_regions_exons_introns_promoters_enhancers.RDS")
saveRDS(go_nonexpressed_shuffled_regions_exons_introns_promoters_enhancers, "go_nonexpressed_shuffled_regions_exons_introns_promoters_enhancers.RDS")
```

##### exons, promoters, enhancers
```{r, eval = FALSE}
go_nonexpressed_regions_exons_promoters_enhancers <- subset_regions_pathways(nonexpressed_regions_exons_promoters_enhancers,
                                        enrich.go.nonexpressed,
                                        intron = TRUE)

go_nonexpressed_shuffled_regions_exons_promoters_enhancers <- subset_regions_pathways_shuffle(nonexpressed_regions_exons_promoters_enhancers,
                                                         enrich.go.nonexpressed,
                                                         intron = TRUE,
                                                         n_iterations = 10000)

# save
saveRDS(go_nonexpressed_regions_exons_promoters_enhancers, "go_nonexpressed_regions_exons_promoters_enhancers.RDS")
saveRDS(go_nonexpressed_shuffled_regions_exons_promoters_enhancers, "go_nonexpressed_shuffled_regions_exons_promoters_enhancers.RDS")
```

## KEGG
```{r, eval = FALSE}
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")

# get enrichment of genes expressed
enrich.kegg.expressed <- enrichKEGG(as.character(activeGenes), 
                   organism = "hsa",
                   keyType = 'kegg', 
                   universe = as.character(rna$entrezgene_id))
enrich.kegg.expressed <- mutate(enrich.kegg.expressed, qscore = -log(pvalue, base=10))

# ouput
write.csv(enrich.kegg.expressed@result, "expressed_genes_kegg.csv", row.names = FALSE)
saveRDS(enrich.kegg.expressed, "enrich.kegg.expressed.RDS")

# plot
pdf("enrich.kegg.expressed.bar.pdf",
    width = 7,
    height = 7)
barplot(enrich.kegg.expressed,
        x = "qscore",
        showCategory = 20)
dev.off()

pdf("enrich.kegg.expressed.dot.pdf",
    width = 7,
    height = 7)
dotplot(enrich.kegg.expressed,
        x = "Count",
        showCategory = 20)
dev.off()

# divide regions by pathways
kegg.expressed <- subset_regions_pathways(expressed_regions,
                                              enrich.kegg.expressed)

kegg.expressed.shuffled <- subset_regions_pathways_shuffle(expressed_regions,
                                                               enrich.kegg.expressed)

# save
saveRDS(kegg.expressed, "kegg.expressed.RDS")
saveRDS(kegg.expressed.shuffled, "kegg.expressed.shuffled.RDS")
```

## Marker genes
```{r, eval = FALSE}
setwd("/Users/ronaldcutler/Dropbox\ (EinsteinMed)/Vijg-lab/Projects/mutation\ accumulation/annotation/gene-sets")

markers <- read.table("fibroblast_markers.txt")

# Map gene ID to gene name
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
gene_names <- getBM(attributes = c('external_gene_name', 'entrezgene_id'),
                filters = 'external_gene_name',
                values = markers$V1, 
                mart = ensembl)
markers <- merge(markers, gene_names, by.x = "V1", by.y = "external_gene_name")

# duplicate removal
sum(duplicated(markers$entrezgene_id))
markers <- markers[!duplicated(markers$entrezgene_id),] # remove duplicates

# divide regions by genes
marker_genes <- subset_regions_genes(expressed_regions,
                                        markers$entrezgene_id)

# save 
setwd("/gs/gsfs0/usersb/vijg-lab/2023-Ronnie/231009_multiple_ENU_analysis/pathway/1-load-data")

saveRDS(marker_genes, "marker_genes.RDS")
```

## Cytopus