---
title: "Multiple ENU treatment bulk mRNA analysis"
output: html_document
date: "2023-09-27"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Notes

# Preparation

## Working directory
```{r}
working.dir <- "~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA"
```


## Load libraries
```{r}
library(biomaRt)
#library(msigdbr)
library(KEGGREST)

library(tximport)
library(SummarizedExperiment)
library(DESeq2)
library(PCAtools)
#library(genefilter)
library(zFPKM)
#library(sva)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(limma)
library(GSVA)

library(ggplot2)
library(cowplot)
library(PerformanceAnalytics)
library(ggrepel)
library(EnhancedVolcano)
library(ComplexHeatmap)
library(UpSetR)
library(ggpubr)
library(ggcorrplot)
library(ggvenn)

library(data.table)
library(tibble)
library(purrr)
library(dplyr)
library(tidyr)
library(reshape2)
library(stringr)

library(viridis)
library(RColorBrewer)

library(BiocParallel)
library(parallel)
num_cores <- detectCores() # Detect the number of cores
```

## Functions
```{r}
theme_Publication <- function(base_size=16, base_family="helvetica") {
  library(grid)
  library(ggthemes)
  (theme_foundation(base_size=base_size, base_family = "")
    + theme(plot.title = element_text(face = "bold",
                                      size = rel(1.2), hjust = 0.5),
            text = element_text(),
            panel.background = element_rect(colour = NA),
            plot.background = element_rect(colour = NA),
            panel.border = element_rect(colour = NA),
            axis.title = element_text(face = "bold",size = rel(1)),
            axis.title.y = element_text(angle=90,vjust =2),
            axis.title.x = element_text(vjust = -0.2),
            axis.text = element_text(), 
            axis.line = element_line(colour="black"),
            axis.ticks = element_line(),
            panel.grid.major = element_line(colour="#f0f0f0"),
            panel.grid.minor = element_blank(),
            legend.key = element_rect(colour = NA),
            legend.position = "top",
            legend.direction = "horizontal",
            #legend.key.size= unit(0.2, "cm"),
            legend.margin = unit(0, "cm"),
            legend.title = element_text(face="italic"),
            plot.margin=unit(c(10,5,5,5),"mm"),
            strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
            strip.text = element_text(face="bold")
    ))
  
}

scale_fill_Publication <- function(...){
  library(scales)
  discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)
  
}

scale_colour_Publication <- function(...){
  library(scales)
  discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)
  
}

# Function to clean row names
remove_version <- function(df) {
  rownames(df) <- sub("\\..*", "", rownames(df))
  return(df)
}
```

## Load data
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/1-data")

# sample table
sample.table <- read.csv('sample_table.csv', header = TRUE)
rownames(sample.table) <- sample.table$sample

# files
files <- paste0(sample.table$sample, ".genes.results")
names(files) <- sample.table$sample
all(file.exists(files))

# import gene level counts
txi <- tximport(files, 
                type = "rsem",
                txIn = FALSE,
                txOut = FALSE)

# remove ensemble id version
# there are duplicate genes when you do this so store in row data
row.data <- data.frame(ensemble_id.clean = rownames(remove_version(txi[[1]])),
                       row.names = rownames(txi[[1]]))
sum(duplicated(row.data$ensemble_id.clean))

# create summarize experiment object
se <- SummarizedExperiment(assays=SimpleList(counts = as.matrix(round(txi$counts)),
                                             tpm = as.matrix(txi$abundance),
                                             effective.length = as.matrix(txi$length)),
                           colData = sample.table,
                           rowData = row.data)
```

### Save
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA")

saveRDS(se, "se.RDS")
```

# Quality Control

## Raw genes expressed per sample
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")

df <- data.frame(Sample = se$sample, 
                 Frequency = colSums(assay(se, "tpm") > 1)) 

pdf("raw_genes_expressed.pdf",
    width = 5,
    height = 5)
ggplot(df, aes(x = Sample, y = Frequency)) +
    geom_col() + 
    theme_Publication() +
    coord_flip() +
    labs(subtitle = "TPM > 1")
dev.off()
```

## Distribution of raw counts
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")

df <- as.data.frame(log2(assay(se) + 1))
colnames(df) <- se$sample
df <- gather(df, key = "Sample", value = "Value")

pdf("raw_count_distribution.pdf",
    width = 5,
    height = 5)
ggplot(df, aes(x = Value, fill = Sample)) +  # Use Value for density and Sample for color fill
    geom_density(alpha = 0.7) +  # Adjust alpha for transparency
    theme_Publication() +
    labs(x = "log2(Counts + 1)",  # Adjust the x-axis label for the density plot
         y = "Density") +  # Density plots typically have "Density" as the y-axis label
    theme(legend.position = "right",
          legend.direction = "vertical")
dev.off()
```

## Filter for expressed genes
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")

# filtering non-expressed genes
assay(se, "zfpkm") <- zFPKM(se, assayName = "tpm")
assay(se, "zfpkm")[assay(se, "zfpkm") == -Inf] <- NA

pdf("zFPKM_distribution.pdf",
    width = 5,
    height = 5)
zFPKMPlot(se, assayName = "tpm", FacetTitles = TRUE)
dev.off()

# To determine which genes are active, we compute the median expression within each group
# detach(package:plyr)
activeGenes <- assay(se, "zfpkm") %>%
  mutate(gene=rownames(assay(se, "zfpkm"))) %>%
  gather(sample, zfpkm, -gene) %>%
  left_join(dplyr::select(as.data.frame(colData(se)), sample, condition, cycle), by="sample") %>%
  group_by(gene, condition) %>%
  summarize(median_zfpkm=median(zfpkm)) %>%
  ungroup() %>%
  mutate(active=(median_zfpkm > -3)) %>%
  filter(active) %>%
  dplyr::select(gene) %>%
  distinct()

# filter for active genes
se.filter <- se[rownames(se) %in% activeGenes$gene]
```

### Save
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA")

saveRDS(se.filter, "se.filter.RDS")
```

## Filtered genes expressed per sample
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")

df <- data.frame(Sample = se$sample, 
                 Frequency = colSums(assay(se.filter, "tpm") > 1)) 

pdf("filtered_genes_expressed.pdf",
    width = 5,
    height = 5)
ggplot(df, aes(x = Sample, y = Frequency)) +
    geom_col() + 
    theme_Publication() +
    coord_flip() +
    labs(subtitle = "TPM > 1")
dev.off()
```

## Distribution of filtered counts
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")

df <- as.data.frame(log2(assay(se.filter) + 1))
colnames(df) <- se$sample
df <- gather(df, key = "Sample", value = "Value")

pdf("filtered_count_distribution.pdf",
    width = 5,
    height = 5)
ggplot(df, aes(x = Value, fill = Sample)) +  # Use Value for density and Sample for color fill
    geom_density(alpha = 0.25) +  # Adjust alpha for transparency
    theme_Publication() +
    labs(x = "log2(Counts + 1)",  # Adjust the x-axis label for the density plot
         y = "Density") +  # Density plots typically have "Density" as the y-axis label
    theme(legend.position = "right",
          legend.direction = "vertical")
dev.off()
```

## Sample Correlation

### All samples
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")

# tpm
df <- log2(assay(se.filter, "tpm") + 1)
colnames(df) <- se.filter$sample

pdf("sample_correlation.pdf",
    width = 10,
    height = 10)
chart.Correlation(df, histogram=TRUE, pch=19)
dev.off()
```

### Control cycle 1
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")

# tpm
df <- log2(assay(se.filter, "tpm") + 1)
colnames(df) <- se.filter$sample
df <- df %>%
  as.data.frame() %>%
  select(contains("C1E"))

pdf("control_cycle_1_correlation.pdf",
    width = 5,
    height = 5)
chart.Correlation(df, histogram=TRUE, pch=19)
dev.off()
```

### Control cycle 9
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")

# tpm
df <- log2(assay(se.filter, "tpm") + 1)
colnames(df) <- se.filter$sample
df <- df %>%
  as.data.frame() %>%
  select(contains("CTL"))

pdf("control_cycle_9_correlation.pdf",
    width = 5,
    height = 5)
chart.Correlation(df, histogram=TRUE, pch=19)
dev.off()
```

### ENU cycle 9
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")

# tpm
df <- log2(assay(se.filter, "tpm") + 1)
colnames(df) <- se.filter$sample
df <- df %>%
  as.data.frame() %>%
  select(contains("ENU"))

pdf("enu_cycle_9_correlation.pdf",
    width = 5,
    height = 5)
chart.Correlation(df, histogram=TRUE, pch=19)
dev.off()
```

## Overlap analysis
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")

binary_expr <- as.data.frame(t(apply(assay(se.filter, "tpm"), 1, function(x) as.integer(x > 0))))
colnames(binary_expr) <- se.filter$sample

pdf("overlap_expressed_genes_all.pdf",
    width = 10,
    height = 7)
upset(binary_expr, 
      sets = colnames(binary_expr),
      text.scale = 1.25,
      #nintersects = 10,
      order.by = "freq",
      show.numbers = FALSE)
dev.off()

pdf("overlap_expressed_genes_top_10.pdf",
    width = 10,
    height = 7)
upset(binary_expr, 
      sets = colnames(binary_expr),
      text.scale = 1.25,
      nintersects = 10,
      order.by = "freq",
      show.numbers = FALSE)
dev.off()
```

## Map gene IDs
```{r}
# get database
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
                dataset = "hsapiens_gene_ensembl")

# mapping
names <- getBM(attributes = c('ensembl_gene_id', 
                              'external_gene_name', 
                              'entrezgene_id',
                              'gene_biotype',
                              'chromosome_name',
                              'start_position',
                              'end_position',
                              'percentage_gene_gc_content'),
                filters = 'ensembl_gene_id',
                values = rowData(se.filter)$ensemble_id.clean,
                mart = mart)

# remove duplicates - 1739
sum(duplicated(names$external_gene_name))
names <- names %>%
  distinct(external_gene_name, .keep_all = TRUE)

# match to genes in dataset
names.match <- names[match(rowData(se.filter)$ensemble_id.clean, names$ensembl_gene_id),]
nrow(names.match) == nrow(rowData(se.filter))
all(rowData(se.filter)$ensemble_id.clean == names.match$ensembl_gene_id, na.rm = TRUE)

# add to se object row data
rowData(se.filter) <- names.match
```

### Save
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA")

saveRDS(se.filter, "se.filter.RDS")
#se.filter <- readRDS("se.filter.RDS")
```

## Output tables
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/2-quality-control")

write.csv(merge(rowData(se.filter), assay(se.filter, "counts"), by = 'row.names'), "counts.filtered",
          row.names = FALSE)

write.csv(merge(rowData(se.filter), assay(se.filter, "tpm"), by = 'row.names'), "tpm.filtered",
          row.names = FALSE)
```

# Normalization

## DeSeq2 normalization and modeling
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/3-normalization")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/3-normalization")

# combine condition and cycle variables
se.filter$condition.cycle <- paste(se.filter$condition,
                                   se.filter$cycle,
                                   sep = "_")
# set factors
se.filter$condition <- factor(se.filter$condition,
                              levels = c("control", "enu"))

se.filter$cycle <- factor(se.filter$cycle,
                              levels = c("1", "9"))

se.filter$batch <- factor(se.filter$batch,
                              levels = c("1", "2", "3"))

se.filter$condition.cycle <- factor(se.filter$condition.cycle,
                                    levels = c("control_1", "control_9", "enu_9"))


# run deseq
dds <- DESeqDataSet(se.filter, 
                    design = ~condition.cycle + batch)

dds <- DESeq(dds)

resultsNames(dds)
```

### Save and output normalized counts
```{r}
# save object
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA")

saveRDS(dds, "dds.RDS")
dds <- readRDS("dds.RDS")

# save norm counts
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/3-normalization")

write.csv(merge(rowData(se.filter), counts(dds, normalized = TRUE), by = 'row.names'), "counts.filtered.normalized.csv",
          row.names = FALSE)
```

## DESeq2 model estimates
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/3-normalization")

# Mean and variance relationship
pdf("mean-variance.pdf",
    width = 5,
    height = 5)
norm.counts <- counts(dds, normalized=TRUE) 
mean.counts <- rowMeans(norm.counts)
variance.counts <- apply(norm.counts, MARGIN = 1, var)
mean.var.col <- densCols(x=log2(mean.counts), y=log2(variance.counts)) 
plot(x=log2(mean.counts), y=log2(variance.counts), pch=16, cex=0.5,
col=mean.var.col, main="Mean-variance relationship", xlab="Mean log2(normalized counts) per gene", ylab="Variance of log2(normalized counts)", panel.first = grid())
abline(a=1, b=1, col="red") # a linear line to compare against 
dev.off()

# dispersion
pdf("dispersion.pdf",
    width = 5,
    height = 5)
plotDispEsts(dds)
dev.off()
```

## Distribution of normalized counts
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/3-normalization")

df <- as.data.frame(log2(counts(dds, normalized=TRUE) + 1))
colnames(df) <- dds$sample
df <- gather(df, key = "Sample", value = "Value")

pdf("normalized_count_distribution_box.pdf",
    width = 5,
    height = 5)
ggplot(df, aes(x = Sample, y = Value)) +
    geom_boxplot() +
    theme_Publication() +
    coord_flip() +
    labs(y = "Log2(counts + 1)",
         subtitle = "Normalized counts")
dev.off()

pdf("normalized_count_distribution_density.pdf",
    width = 6,
    height = 5)
ggplot(df, aes(x = Value, fill = Sample)) +
    geom_density(alpha = 0.25) +
    theme_Publication() +
    labs(x = "Log2(counts + 1)",
         subtitle = "Normalized counts") +
    theme(legend.position = "right",
          legend.direction = "vertical")
dev.off()
```

## Sample-sample correlations
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/3-normalization")

df <- as.data.frame(log2(counts(dds, normalized=TRUE) + 1))
colnames(df) <- dds$sample
df.cor <- cor(df)
df.cor <- df.cor[order(rownames(df.cor)), order(colnames(df.cor))]

pdf("normalized_count_sample_correlations.pdf",
    width = 8,
    height = 8)
ggcorrplot(df.cor, 
           hc.order = FALSE, 
           type = "lower",
           lab = TRUE)
dev.off()

# Transform counts to stabalize mean-variance
# blind = TRUE to compare samples in a manner unbiased by prior information on samples
df <- assay(vst(dds, blind = TRUE))
colnames(df) <- dds$sample
df.cor <- cor(df)
df.cor <- df.cor[order(rownames(df.cor)), order(colnames(df.cor))]

pdf("vst_count_sample_correlations.pdf",
    width = 8,
    height = 8)
ggcorrplot(df.cor, 
           hc.order = FALSE, 
           type = "lower",
           lab = TRUE)
dev.off()
```

## Principle component analysis
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/3-normalization")

# Transform counts to stabalize mean-variance
# blind = FALSE for downstream analysis, where the full use of the design information should be made
dds_vst <- vst(dds, blind = FALSE)

# PCA
p <- pca(assay(dds_vst), 
         metadata = colData(dds), 
         removeVar = 0.1)

pdf("screeplot.pdf",
    width = 5,
    height = 5)
screeplot(p)
dev.off()

pdf("eingencor.pdf",
    width = 5,
    height = 5)
eigencorplot(p,
             components = getComponents(p, seq_len(9)),
             main = "Principle component correlations",
             cexMain = 1.5,
             metavars = c("condition", "cycle", "batch"),
             col = viridis(100),
             colCorval = 'firebrick',
             fontCorval = 2,
             cexCorval = 0.5,
             rotLabX = 45,
             posColKey = 'top')
dev.off()

pdf("pca.pdf",
    width = 5.5,
    height = 5)
biplot(p,
       colby = "condition.cycle",
       shape = "batch",
       hline = 0,
       vline = 0,
       gridlines.major = FALSE, 
       gridlines.minor = FALSE,
       max.overlaps = Inf,
       lab = dds$sample.description) +
       theme_Publication() +
       theme(legend.position = "right",
             legend.box="vertical",
             legend.direction = "vertical",
             legend.margin=margin())
dev.off()
```

# Batch effect correction
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/3-normalization")
# mod <- model.matrix(~condition.cycle, data=colData(dds))
# mod
# 
# counts <- counts(dds, normalized = FALSE)
# # counts <- na.omit(counts)
# # zero_var_genes <- apply(counts, 1, function(x) var(x) == 0)
# # counts <- counts[!zero_var_genes,]
# # head(counts)
# 
# dds$batch
# 
# param <- MulticoreParam(workers = num_cores - 1)
# 
# ComBat(dat = counts,
#        batch = dds$batch,
#        ref.batch = "1",
#        mod = mod,
#        par.prior = FALSE,
#        mean.only = TRUE,
#        BPPARAM = param)
# 
# assays(dds, withDimnames=FALSE)[["counts.batchCorrect"]] <- ComBat(dat = na.omit(counts(dds, normalized = FALSE)),
#                                                                                  batch = dds$batch,
#                                                                                  ref.batch = "2",
#                                                                                  mod = mod)

assays(dds, withDimnames=FALSE)[["counts.normalized.batchCorrect"]] <- removeBatchEffect(log2(counts(dds, normalized = TRUE) + 1),
                                                                              batch = dds$batch,
                                                                              group = dds$condition.cycle)
```

## Principle component analysis
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/3-normalization")

# Transform counts to stabalize mean-variance
# blind = FALSE for downstream analysis, where the full use of the design information should be made
# dds_vst <- vst(assay(dds, "counts.batchCorrect"), blind = FALSE)

# PCA
p <- pca(assay(dds, "counts.normalized.batchCorrect"), 
         metadata = colData(dds), 
         removeVar = 0.1)

pdf("batchCorrect_screeplot.pdf",
    width = 5,
    height = 5)
screeplot(p)
dev.off()

pdf("batchCorrect_eingencor.pdf",
    width = 5,
    height = 5)
eigencorplot(p,
             components = getComponents(p, seq_len(9)),
             main = "Principle component correlations",
             cexMain = 1.5,
             metavars = c("condition", "cycle", "batch"),
             col = viridis(100),
             colCorval = 'firebrick',
             fontCorval = 2,
             cexCorval = 0.5,
             rotLabX = 45,
             posColKey = 'top')
dev.off()

pdf("batchCorrect_pca.pdf",
    width = 5.5,
    height = 5)
biplot(p,
       colby = "condition.cycle",
       shape = "batch",
       hline = 0,
       vline = 0,
       gridlines.major = FALSE, 
       gridlines.minor = FALSE,
       max.overlaps = Inf,
       lab = dds$sample.description) +
       theme_Publication() +
       theme(legend.position = "right",
             legend.box="vertical",
             legend.direction = "vertical",
             legend.margin=margin())
dev.off()
```


# Differential Expression

## Control 9 vs Control 1
- Effect of passaging
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/4-differential-expression")

# get results
control9_vs_control1 <- results(dds, 
                  alpha = 0.05, 
                  contrast = c("condition.cycle", "control_9", "control_1"))

# get a summary of the amount of significantly DEG 
summary(control9_vs_control1, 
        alpha = 0.05)

# pdf("control9_vs_control1_ma.pdf",
#     width = 5,
#     height = 5)
# DESeq2::plotMA(control9_vs_control1)
# dev.off()

# get shrunken fold change
# control9_vs_control1 <- lfcShrink(dds = dds, 
#                               res = control9_vs_control1, 
#                               contrast = c("condition", "no infect", "background"),
#                               type = "ashr")

cols <- c("ensembl_gene_id",
          "external_gene_name",
          "entrezgene_id",
          "gene_biotype",
          "chromosome_name",
          "start_position",
          "end_position",
          "percentage_gene_gc_content")

control9_vs_control1 <- merge(as.data.frame(rowData(dds)[,cols]),
                         as.data.frame(control9_vs_control1),
                         by = "row.names")

counts <- counts(dds, normalized = TRUE)
colnames(counts) <- dds$sample.description
control9_vs_control1 <- merge(control9_vs_control1, 
                         counts,
                         by.x = "Row.names",
                         by = "row.names")

control9_vs_control1$ensembl_gene_id <- NULL
colnames(control9_vs_control1)[1] <- "ensembl_gene_id"

# write results
write.csv(control9_vs_control1, "control9_vs_control1.csv",
          row.names = FALSE)

# p-value distribution histogram
pdf("control9_vs_control1_pvalue_distribution.pdf",
    width = 5,
    height = 5)
hist(control9_vs_control1$pvalue, 
     main = "P-value distribution", 
     xlab = "P-value", 
     ylab = "Frequency")
dev.off()


# ma plot
control9_vs_control1 <- control9_vs_control1 %>%
  as.data.frame() %>%
  mutate(sig = case_when(log2FoldChange > 0 & control9_vs_control1$padj < 0.05 ~ "sig.up",
                         log2FoldChange < 0 & control9_vs_control1$padj < 0.05 ~ "sig.down",
                         .default = "")
         )


pdf("control9_vs_control1_ma.pdf",
    width = 5,
    height = 5)
ggplot(control9_vs_control1, aes(baseMean, log2FoldChange, color = sig)) +
  geom_point(alpha = 0.5, size = 4) +
  theme_Publication() +
  scale_colour_manual(values = c("grey", "blue", "red")) +
  geom_hline(yintercept= 0, 
             size=1,
             linetype = "dashed") +
  labs(title = "Control 9 vs Control 1",
       x = "Average Normalized counts",
       y = "Log2(fold change)")
dev.off()

pdf("control9_vs_control1_ma_log2.pdf",
    width = 5,
    height = 5)
ggplot(control9_vs_control1, aes(log2(baseMean + 1), log2FoldChange, color = sig)) +
  geom_point(alpha = 0.5, size = 4) +
  theme_Publication() +
  scale_colour_manual(values = c("grey", "blue", "red")) +
  geom_hline(yintercept= 0, 
             size=1,
             linetype = "dashed") +
  labs(title = "Control 9 vs Control 1",
       x = "Average Normalized Log2(counts) + 1",
       y = "Log2(fold change)")
dev.off()

# volcano plot
# control9_vs_control1$volcano.label <- ""
# control9_vs_control1$volcano.score <- abs(control9_vs_control1$log2FoldChange) * -log10(control9_vs_control1$pvalue)
# control9_vs_control1 <- control9_vs_control1[order(control9_vs_control1$volcano.score, decreasing = TRUE),]
# control9_vs_control1$volcano.label[1:10] <- control9_vs_control1$external_gene_name[1:10]

pdf("control9_vs_control1_volcano.pdf",
    width = 5,
    height = 5)
EnhancedVolcano(control9_vs_control1,
                lab = control9_vs_control1$external_gene_name,
                #selectLab = control9_vs_control1$volcano.label,
                title = "Control Cycle 9 vs Control Cycle 1",
                col = c("grey30", "grey30", "grey30", "red2"),
                subtitle = NULL,
                caption = "FC > 2; padj < 0.05",
                x = 'log2FoldChange',
                y = 'pvalue',
                pointSize = 2,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 2,
                pCutoff = 0.05,
                pCutoffCol = 'padj',
                labSize = 3,
                labCol = 'black',
                labFace = 'bold',
                boxedLabels = TRUE,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black',
                max.overlaps = 10,
                gridlines.major = FALSE,
                gridlines.minor = FALSE,
                ylim = c(0, 40),
                xlim = c(-10, 10))
dev.off()

# heatmap of significant changes
ha <- HeatmapAnnotation(Condition = subset(colData(dds), condition.cycle %in% c("control_1", "control_9"))$condition,
                        Cycle = subset(colData(dds), condition.cycle %in% c("control_1", "control_9"))$cycle,
                        Batch = subset(colData(dds), condition.cycle %in% c("control_1", "control_9"))$batch
)

# get normalized counts
mat <- as.matrix(counts(dds, normalized = TRUE)[,colData(dds)$condition.cycle %in% c("control_1", "control_9")])

# order by results
mat.order <- mat[match(control9_vs_control1$ensembl_gene_id, rownames(mat)),]

# get significant genes
mat.sig <- mat.order[control9_vs_control1$padj < 0.05,]
mat.sig <- mat.sig[!is.na(row.names(mat.sig)),]

# scale and center normalized counts
mat.sig.scaled <- t(scale(t(mat.sig)))

pdf("control9_vs_control1_sig_heatmap.pdf",
    width = 4,
    height = 5)
Heatmap(mat.sig.scaled, 
        name = "Z-score",
        top_annotation = ha,
        column_split = subset(colData(dds), condition.cycle %in% c("control_1", "control_9"))$condition.cycle,
        border = TRUE,
        cluster_rows = TRUE,
        cluster_columns = TRUE,
        show_row_names = FALSE,
        show_column_names = FALSE,
        row_dend_width = unit(10, "mm"),
        column_dend_height = unit(10, "mm"))
dev.off()
```

### Comparison to previous data
- https://onlinelibrary.wiley.com/doi/10.1111/acel.12234
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/4-differential-expression")

previous_data <- read.csv("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/E-MTAB-2086/acel.12234&file=acel12234-sup-0006-TableS1.csv")

previous_data_matched <- left_join(previous_data, 
                                   control9_vs_control1,
                                   by = c("Gene.Symbol" = "external_gene_name"))

# remove rows with NA values
previous_data_matched <- previous_data_matched %>%
  filter(!is.na(ensembl_gene_id))

# Group 1: PD 50 vs PD 30
df_pd50 <- previous_data_matched %>%
  filter(P.value..Middle.Young. < 0.05 & padj < 0.05) %>%
  select(log2FoldChange, 
         Ratio = Ratio..Middle.Young.) %>%
  mutate(Group = "PD 50 vs PD 30")

# Group 2: PD 70 vs PD 30
df_pd70 <- previous_data_matched %>%
  filter(P.value..Old.Young. < 0.05 & padj < 0.05) %>%
  select(log2FoldChange, 
         Ratio = Ratio..Old.Young.) %>%
  mutate(Group = "PD 70 vs PD 30")

# Group 3: PD 80 vs PD 30
df_pd80 <- previous_data_matched %>%
  filter(P.value..Senescent.Young. < 0.05 & padj < 0.05) %>%
  select(log2FoldChange, 
         Ratio = Ratio..Senescent.Young.) %>%
  mutate(Group = "PD 80 vs PD 30")

# 2. Combine them into one large dataframe
plot_data_combined <- bind_rows(df_pd50, df_pd70, df_pd80)

# 3. Plot everything on one graph using the 'color' aesthetic
pdf("control9_vs_control1_lackner2014.pdf", width = 6, height = 5)

ggplot(plot_data_combined, aes(x = log2FoldChange, y = log2(Ratio), color = Group)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_Publication() +
  labs(x = "Log2FC Control 9 vs Control 1",
       y = "Log2FC (Lackner et al. 2014)",
       color = "Lackner et al. 2014\nComparisons") +
  stat_cor(method = "pearson",
           label.x.npc = "left",
           label.y.npc = "top", 
           show.legend = FALSE) +
  theme(legend.position = "right",
        legend.direction = "vertical")

dev.off()
```


## ENU 9 vs Control 9
- Effect of ENu-treatment
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/4-differential-expression")

# get results
enu9_vs_control9 <- results(dds, 
                  alpha = 0.05, 
                  contrast = c("condition.cycle", "enu_9", "control_9"))

# get a summary of the amount of significantly DEG 
summary(enu9_vs_control9, 
        alpha = 0.05)

# pdf("enu9_vs_control9_ma.pdf",
#     width = 5,
#     height = 5)
# DESeq2::plotMA(enu9_vs_control9)
# dev.off()

# get shrunken fold change
# enu9_vs_control9 <- lfcShrink(dds = dds, 
#                               res = enu9_vs_control9, 
#                               contrast = c("condition", "no infect", "background"),
#                               type = "ashr")

cols <- c("ensembl_gene_id",
          "external_gene_name",
          "entrezgene_id",
          "gene_biotype",
          "chromosome_name",
          "start_position",
          "end_position",
          "percentage_gene_gc_content")

enu9_vs_control9 <- merge(as.data.frame(rowData(dds)[,cols]),
                         as.data.frame(enu9_vs_control9),
                         by = "row.names")

counts <- counts(dds, normalized = TRUE)
colnames(counts) <- dds$sample.description
enu9_vs_control9 <- merge(enu9_vs_control9, 
                         counts,
                         by.x = "Row.names",
                         by = "row.names")

enu9_vs_control9$ensembl_gene_id <- NULL
colnames(enu9_vs_control9)[1] <- "ensembl_gene_id"

# write results
write.csv(enu9_vs_control9, "enu9_vs_control9.csv",
          row.names = FALSE)

# p-value distribution histogram
pdf("enu9_vs_control9_pvalue_distribution.pdf",
    width = 5,
    height = 5)
hist(enu9_vs_control9$pvalue, 
     main = "P-value distribution", 
     xlab = "P-value", 
     ylab = "Frequency")
dev.off()


# ma plot
enu9_vs_control9 <- enu9_vs_control9 %>%
  as.data.frame() %>%
  mutate(sig = case_when(log2FoldChange > 0 & enu9_vs_control9$padj < 0.05 ~ "sig.up",
                         log2FoldChange < 0 & enu9_vs_control9$padj < 0.05 ~ "sig.down",
                         .default = "")
         )


pdf("enu9_vs_control9_ma.pdf",
    width = 5,
    height = 5)
ggplot(enu9_vs_control9, aes(baseMean, log2FoldChange, color = sig)) +
  geom_point(alpha = 0.5, size = 4) +
  theme_Publication() +
  scale_colour_manual(values = c("grey", "blue", "red")) +
  geom_hline(yintercept= 0, 
             size=1,
             linetype = "dashed") +
  labs(title = "ENU 9 vs Control 9",
       x = "Average Normalized counts",
       y = "Log2(fold change)")
dev.off()

pdf("enu9_vs_control9_ma_log2.pdf",
    width = 5,
    height = 5)
ggplot(enu9_vs_control9, aes(log2(baseMean + 1), log2FoldChange, color = sig)) +
  geom_point(alpha = 0.5, size = 4) +
  theme_Publication() +
  scale_colour_manual(values = c("grey", "blue", "red")) +
  geom_hline(yintercept= 0, 
             size=1,
             linetype = "dashed") +
  labs(title = "ENU 9 vs Control 9",
       x = "Average Normalized Log2(counts) + 1",
       y = "Log2(fold change)")
dev.off()

# volcano plot
enu9_vs_control9$volcano.label <- ""
# enu9_vs_control9$volcano.score <- abs(enu9_vs_control9$log2FoldChange) * -log10(enu9_vs_control9$pvalue)
enu9_vs_control9$volcano.score <- -log10(enu9_vs_control9$pvalue)
enu9_vs_control9 <- enu9_vs_control9[order(enu9_vs_control9$volcano.score, decreasing = TRUE),]
enu9_vs_control9$volcano.label[1:10] <- enu9_vs_control9$external_gene_name[1:10]

pdf("enu9_vs_control9_volcano.pdf",
    width = 5,
    height = 5)
EnhancedVolcano(enu9_vs_control9,
                lab = enu9_vs_control9$external_gene_name,
                selectLab = enu9_vs_control9$volcano.label,
                col = c("grey30", "grey30", "grey30", "red2"),
                title = "ENU Cycle 9 vs Control Cycle 9",
                subtitle = NULL,
                caption = "FC > 2; padj < 0.05",
                x = 'log2FoldChange',
                y = 'pvalue',
                pointSize = 2,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 2,
                pCutoff = 0.05,
                pCutoffCol = 'padj',
                labSize = 3,
                labCol = 'black',
                labFace = 'bold',
                boxedLabels = TRUE,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black',
                max.overlaps = 10,
                gridlines.major = FALSE,
                gridlines.minor = FALSE,
                ylim = c(0,6))
dev.off()

# heatmap of significant changes
ha <- HeatmapAnnotation(Condition = subset(colData(dds), condition.cycle %in% c("enu_9", "control_9"))$condition,
                        Cycle = subset(colData(dds), condition.cycle %in% c("enu_9", "control_9"))$cycle,
                        Batch = subset(colData(dds), condition.cycle %in% c("enu_9", "control_9"))$batch
)

# get normalized counts
mat <- as.matrix(counts(dds, normalized = TRUE)[,colData(dds)$condition.cycle %in% c("enu_9", "control_9")])

# order by results
mat.order <- mat[match(enu9_vs_control9$ensembl_gene_id, rownames(mat)),]

# get significant genes
mat.sig <- mat.order[enu9_vs_control9$padj < 0.05,]
mat.sig <- mat.sig[!is.na(row.names(mat.sig)),]

# scale and center normalized counts
mat.sig.scaled <- t(scale(t(mat.sig)))

pdf("enu9_vs_control9_sig_heatmap.pdf",
    width = 4,
    height = 5)
Heatmap(mat.sig.scaled, 
        name = "Z-score",
        top_annotation = ha,
        column_split = subset(colData(dds), condition.cycle %in% c("enu_9", "control_9"))$condition.cycle,
        border = TRUE,
        cluster_rows = TRUE,
        cluster_columns = TRUE,
        show_row_names = FALSE,
        show_column_names = FALSE,
        row_dend_width = unit(10, "mm"),
        column_dend_height = unit(10, "mm"))
dev.off()
```

# Functional enrichment

## Control 9 vs Control 1
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get score to use in enrichment
control9_vs_control1$score <- sign(control9_vs_control1$log2FoldChange) * -log10(control9_vs_control1$pvalue)
res.sorted <- control9_vs_control1[order(control9_vs_control1$score, decreasing = TRUE),]
res.list <- res.sorted$score
names(res.list) <- res.sorted$entrezgene_id
res.list <- res.list[!is.na(names(res.list))] # remove genes witn no ID
res.list <- res.list[!is.na(res.list)] # remove genes with no score
```

### Biological process
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

bp <- gseGO(geneList     = res.list,
            OrgDb        = org.Hs.eg.db,
            ont          = "BP",
            keyType = "ENTREZID",
            pvalueCutoff = 0.1,
            verbose      = FALSE)
bp.simple <- clusterProfiler::simplify(bp)

write.csv(bp.simple@result, "control9_vs_control1_go_bp.csv")

pdf("control9_vs_control1_go_bp.pdf",
    width = 6,
    height = 5)
dotplot(bp.simple, 
        showCategory=10, 
        title = paste("GSE GO BP", "Control 9 vs Control 1", sep = " "), 
        split=".sign",
        x = "NES", 
        font.size = 10,
        label_format = 50) +
        geom_vline(xintercept = 0, linetype="dotted", color = "#444444", size=1) +
        scale_color_viridis(option = "viridis", direction = 1) +
        labs(fill = "test")
dev.off()
```

### Moelcular function
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

mf <- gseGO(geneList     = res.list,
            OrgDb        = org.Hs.eg.db,
            ont          = "MF",
            keyType = "ENTREZID",
            pvalueCutoff = 0.1,
            verbose      = FALSE)
mf.simple <- clusterProfiler::simplify(mf)

write.csv(mf.simple@result, "control9_vs_control1_go_mf.csv")

pdf("control9_vs_control1_go_mf.pdf",
    width = 6,
    height = 5)
dotplot(mf.simple, 
        showCategory=10, 
        title = paste("GSE GO MF", "Control 9 vs Control 1", sep = " "), 
        split=".sign",
        x = "NES", 
        font.size = 10,
        label_format = 50) +
        geom_vline(xintercept = 0, linetype="dotted", color = "#444444", size=1) +
        scale_color_viridis(option = "viridis", direction = 1) +
        labs(fill = "test")
dev.off()
```

### Cellular component
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

cc <- gseGO(geneList     = res.list,
            OrgDb        = org.Hs.eg.db,
            ont          = "CC",
            keyType = "ENTREZID",
            pvalueCutoff = 0.1,
            verbose      = FALSE)
cc.simple <- clusterProfiler::simplify(cc)

write.csv(cc.simple@result, "control9_vs_control1_go_cc.csv")

pdf("control9_vs_control1_go_cc.pdf",
    width = 6,
    height = 5)
dotplot(cc.simple, 
        showCategory=10, 
        title = paste("GSE GO CC", "Control 9 vs Control 1", sep = " "), 
        split=".sign",
        x = "NES", 
        font.size = 10,
        label_format = 50) +
        geom_vline(xintercept = 0, linetype="dotted", color = "#444444", size=1) +
        scale_color_viridis(option = "viridis", direction = 1) +
        labs(fill = "test")
dev.off()
```

### Hallmark DNA repair
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get gene set
msigdb_dna_repair <- msigdbr(species = "human") %>%
    filter(str_detect(gs_name, "HALLMARK_DNA_REPAIR")) %>%
    dplyr::select(gs_name, entrez_gene) %>%
    dplyr::rename(term = gs_name, entrez_gene = entrez_gene)

# gsea
hallmark_dna_repair <- GSEA(res.list, 
                   TERM2GENE = msigdb_dna_repair,
                   pvalueCutoff = 0.1)

# plot
anno <- hallmark_dna_repair[,c("NES", "pvalue", "p.adjust")]
lab <- paste0(names(anno), "=",  round(anno, 3), collapse="\n")

pdf("control9_vs_control1_hallmark_dna_repair.pdf",
    width = 5,
    height = 5)
gsearank(hallmark_dna_repair, 1, title = hallmark_dna_repair[1, "Description"]) +
  annotate("text", 10000, hallmark_dna_repair[, "enrichmentScore"] * .75, label = lab, hjust=0, vjust=0)
dev.off()
```

### KEGG DNA repair
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get gene sets
dna_repair_pathways <- c("Base excision repair",
                         "Nucleotide excision repair",
                         "Mismatch repair",
                         "Homologous recombination",
                         "Non-homologous end-joining",
                         "Fanconi anemia pathway")

pathways_list <- keggList("pathway", "hsa")

# Loop through each DNA repair pathway and filter the pathways_list
filtered_pathways <- c()
for (pathway in dna_repair_pathways) {
    matching_pathways <- grep(pathway, pathways_list, value = TRUE)
    filtered_pathways <- c(filtered_pathways, matching_pathways)
}

# Loop through each DNA repair pathway and retrieve the associated genes
filtered_pathways_genes <- list()
for (pathway_id in names(filtered_pathways)) {
    # Get pathway information
    pathway_info <- keggGet(pathway_id)
    
    # Extract gene information
    if (!is.null(pathway_info[[1]]$GENE)) {
        genes <- pathway_info[[1]]$GENE
        gene_ids <- genes[seq(1, length(genes), 2)]
        filtered_pathways_genes[[pathway_id]] <- gene_ids
    }
}
names(filtered_pathways_genes) <- filtered_pathways

# Convert the list to a tibble
kegg_dna_repair <- enframe(filtered_pathways_genes, name = "pathway", value = "entrez_id")

# Unnest the tibble to create a long-format table
kegg_dna_repair <- kegg_dna_repair %>%
  unnest(entrez_id)

# trim names
kegg_dna_repair$pathway <-  sub(" - Homo sapiens \\(human\\)", "", kegg_dna_repair$pathway)

# gsea
kegg_dna_repair_gsea <- GSEA(res.list, 
                   TERM2GENE = kegg_dna_repair,
                   pvalueCutoff = 0.1)

# plot
pp <- lapply(1:nrow(kegg_dna_repair_gsea), function(i) {
    anno <- kegg_dna_repair_gsea[i, c("NES", "pvalue", "p.adjust")]
    lab <- paste0(names(anno), "=",  round(anno, 5), collapse="\n")

    gsearank(kegg_dna_repair_gsea, i, kegg_dna_repair_gsea[i, 2]) + xlab(NULL) +ylab(NULL) +
        annotate("text", 7000, kegg_dna_repair_gsea[i, "enrichmentScore"] * .75, label = lab, hjust=0, vjust=0, size = 2.25)
})

pdf("control9_vs_control1_kegg_dna_repair.pdf",
    width = 6,
    height = 6)
plot_grid(plotlist=pp, ncol=2)
dev.off()
```

### KEGG Cell Cycle
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get gene sets
pathways <- c("Cell cycle")

pathways_list <- keggList("pathway", "hsa")

# Loop through each DNA repair pathway and filter the pathways_list
filtered_pathways <- c()
for (pathway in pathways) {
    matching_pathways <- grep(pathway, pathways_list, value = TRUE)
    filtered_pathways <- c(filtered_pathways, matching_pathways)
}

# Loop through each DNA repair pathway and retrieve the associated genes
filtered_pathways_genes <- list()
for (pathway_id in names(filtered_pathways)) {
    # Get pathway information
    pathway_info <- keggGet(pathway_id)
    
    # Extract gene information
    if (!is.null(pathway_info[[1]]$GENE)) {
        genes <- pathway_info[[1]]$GENE
        gene_ids <- genes[seq(1, length(genes), 2)]
        filtered_pathways_genes[[pathway_id]] <- gene_ids
    }
}
names(filtered_pathways_genes) <- filtered_pathways

# Convert the list to a tibble
kegg_cell_cycle <- enframe(filtered_pathways_genes, name = "pathway", value = "entrez_id")

# Unnest the tibble to create a long-format table
kegg_cell_cycle <- kegg_cell_cycle %>%
  unnest(entrez_id)

# trim names
kegg_cell_cycle$pathway <-  sub(" - Homo sapiens \\(human\\)", "", kegg_cell_cycle$pathway)

# gsea
kegg_cell_cycle_gsea <- GSEA(res.list, 
                   TERM2GENE = kegg_cell_cycle,
                   pvalueCutoff = 0.1)

# plot
anno <- kegg_cell_cycle_gsea[,c("NES", "pvalue")]
anno$NES <- round(anno$NES, 2)
lab <- paste0(names(anno), "=",  anno, collapse="\n")

pdf("control9_vs_control1_kegg_cell_cycle.pdf",
    width = 5,
    height = 5)
gsearank(kegg_cell_cycle_gsea, 1, title = "KEGG Cell Cycle Gene Set") +
  annotate("text", 1000, -0.5, label = lab, hjust=0, vjust=0, size = 5) +
  theme(text=element_text(size=15))
dev.off()
```

### KEGG Apoptosis
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get gene sets
pathways <- c("Apoptosis - Homo sapiens")

pathways_list <- keggList("pathway", "hsa")

# Loop through each DNA repair pathway and filter the pathways_list
filtered_pathways <- c()
for (pathway in pathways) {
    matching_pathways <- grep(pathway, pathways_list, value = TRUE)
    filtered_pathways <- c(filtered_pathways, matching_pathways)
}

# Loop through each DNA repair pathway and retrieve the associated genes
filtered_pathways_genes <- list()
for (pathway_id in names(filtered_pathways)) {
    # Get pathway information
    pathway_info <- keggGet(pathway_id)
    
    # Extract gene information
    if (!is.null(pathway_info[[1]]$GENE)) {
        genes <- pathway_info[[1]]$GENE
        gene_ids <- genes[seq(1, length(genes), 2)]
        filtered_pathways_genes[[pathway_id]] <- gene_ids
    }
}
names(filtered_pathways_genes) <- filtered_pathways

# Convert the list to a tibble
kegg_apoptosis <- enframe(filtered_pathways_genes, name = "pathway", value = "entrez_id")

# Unnest the tibble to create a long-format table
kegg_apoptosis <- kegg_apoptosis %>%
  unnest(entrez_id)

# trim names
kegg_apoptosis$pathway <-  sub(" - Homo sapiens \\(human\\)", "", kegg_apoptosis$pathway)

# gsea
kegg_apoptosis_gsea <- GSEA(res.list, 
                   TERM2GENE = kegg_apoptosis,
                   pvalueCutoff = 1)

# plot
anno <- kegg_apoptosis_gsea[,c("NES", "pvalue")]
anno$NES <- round(anno$NES, 2)
anno$pvalue <- round(anno$pvalue, 2)
lab <- paste0(names(anno), "=",  anno, collapse="\n")

pdf("control9_vs_control1_kegg_apoptosis.pdf",
    width = 5,
    height = 5)
gsearank(kegg_apoptosis_gsea, 1, title = "KEGG Apoptosis Gene Set") +
  annotate("text", 10000, 0.2, label = lab, hjust=0, vjust=0, size = 5) +
  theme(text=element_text(size=15))
dev.off()
```

### SenMayo
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get gene sets
senmayo <- read.csv("senmayo.csv",
                    header = TRUE)

# convert symbol to entrez
senmayo$entrez_id <- bitr(senmayo$symbol, fromType="SYMBOL", toType=c("ENTREZID"), OrgDb="org.Hs.eg.db")$ENTREZID
senmayo <- data.frame(set = senmayo$set,
                      entrez_id = senmayo$entrez_id)

# gsea
senmayo_gsea <- GSEA(res.list, 
                   TERM2GENE = senmayo,
                   pvalueCutoff = 0.1)

# plot
anno <- senmayo_gsea[,c("NES", "pvalue")]
lab <- paste0(names(anno), "=",  round(anno, 3), collapse="\n")

pdf("control9_vs_control1_senmayo.pdf",
    width = 5,
    height = 5)
gsearank(senmayo_gsea, 1, title = "SenMayo Senescence Gene Set") +
  annotate("text", 1000, -0.3, label = lab, hjust=0, vjust=0, size = 5) +
  theme(text=element_text(size=15))
dev.off()
```



## ENU 9 vs Control 9
```{r}
dir.create("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get score to use in enrichment
enu9_vs_control9$score <- sign(enu9_vs_control9$log2FoldChange) * -log10(enu9_vs_control9$pvalue)
res.sorted <- enu9_vs_control9[order(enu9_vs_control9$score, decreasing = TRUE),]
res.list <- res.sorted$score
names(res.list) <- res.sorted$entrezgene_id
res.list <- res.list[!is.na(names(res.list))] # remove genes witn no ID
res.list <- res.list[!is.na(res.list)] # remove genes with no score
```

### Biological process
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

bp <- gseGO(geneList     = res.list,
            OrgDb        = org.Hs.eg.db,
            ont          = "BP",
            keyType = "ENTREZID",
            pvalueCutoff = 0.1,
            verbose      = FALSE)
bp.simple <- clusterProfiler::simplify(bp)

write.csv(bp.simple@result, "enu9_vs_control9_go_bp.csv")

pdf("enu9_vs_control9_go_bp.pdf",
    width = 6,
    height = 5)
dotplot(bp.simple, 
        showCategory=10, 
        title = paste("GSE GO BP", "ENU 9 vs Control 9", sep = " "), 
        split=".sign",
        x = "NES", 
        font.size = 10,
        label_format = 50) +
        geom_vline(xintercept = 0, linetype="dotted", color = "#444444", size=1) +
        scale_color_viridis(option = "viridis", direction = 1) +
        labs(fill = "test")
dev.off()
```

### Moelcular function
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

mf <- gseGO(geneList     = res.list,
            OrgDb        = org.Hs.eg.db,
            ont          = "MF",
            keyType = "ENTREZID",
            pvalueCutoff = 0.1,
            verbose      = FALSE)
mf.simple <- clusterProfiler::simplify(mf)

write.csv(mf.simple@result, "enu9_vs_control9_go_mf.csv")

pdf("enu9_vs_control9_go_mf.pdf",
    width = 6,
    height = 5)
dotplot(mf.simple, 
        showCategory=10, 
        title = paste("GSE GO MF", "ENU 9 vs Control 9", sep = " "), 
        split=".sign",
        x = "NES", 
        font.size = 10,
        label_format = 50) +
        geom_vline(xintercept = 0, linetype="dotted", color = "#444444", size=1) +
        scale_color_viridis(option = "viridis", direction = 1) +
        labs(fill = "test")
dev.off()
```

### Cellular component
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

cc <- gseGO(geneList     = res.list,
            OrgDb        = org.Hs.eg.db,
            ont          = "CC",
            keyType = "ENTREZID",
            pvalueCutoff = 0.1,
            verbose      = FALSE)
cc.simple <- clusterProfiler::simplify(cc)

write.csv(cc.simple@result, "enu9_vs_control9_go_cc.csv")

pdf("enu9_vs_control9_go_cc.pdf",
    width = 6,
    height = 5)
dotplot(cc.simple, 
        showCategory=10, 
        title = paste("GSE GO CC", "ENU 9 vs Control 9", sep = " "), 
        split=".sign",
        x = "NES", 
        font.size = 10,
        label_format = 50) +
        geom_vline(xintercept = 0, linetype="dotted", color = "#444444", size=1) +
        scale_color_viridis(option = "viridis", direction = 1) +
        labs(fill = "test")
dev.off()
```

### Hallmark DNA repair
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get gene set
msigdb_dna_repair <- msigdbr(species = "human") %>%
    filter(str_detect(gs_name, "HALLMARK_DNA_REPAIR")) %>%
    dplyr::select(gs_name, entrez_gene) %>%
    dplyr::rename(term = gs_name, entrez_gene = entrez_gene)

# gsea
hallmark_dna_repair <- GSEA(res.list, 
                   TERM2GENE = msigdb_dna_repair,
                   pvalueCutoff = 0.1)

# plot
anno <- hallmark_dna_repair[,c("NES", "pvalue", "p.adjust")]
lab <- paste0(names(anno), "=",  round(anno, 3), collapse="\n")

pdf("enu9_vs_control9_hallmark_dna_repair.pdf",
    width = 5,
    height = 5)
gsearank(hallmark_dna_repair, 1, title = hallmark_dna_repair[1, "Description"]) +
  annotate("text", 10000, hallmark_dna_repair[, "enrichmentScore"] * .75, label = lab, hjust=0, vjust=0)
dev.off()
```

### KEGG DNA repair
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get gene sets
dna_repair_pathways <- c("Base excision repair",
                         "Nucleotide excision repair",
                         "Mismatch repair",
                         "Homologous recombination",
                         "Non-homologous end-joining",
                         "Fanconi anemia pathway")

pathways_list <- keggList("pathway", "hsa")

# Loop through each DNA repair pathway and filter the pathways_list
filtered_pathways <- c()
for (pathway in dna_repair_pathways) {
    matching_pathways <- grep(pathway, pathways_list, value = TRUE)
    filtered_pathways <- c(filtered_pathways, matching_pathways)
}

# Loop through each DNA repair pathway and retrieve the associated genes
filtered_pathways_genes <- list()
for (pathway_id in names(filtered_pathways)) {
    # Get pathway information
    pathway_info <- keggGet(pathway_id)
    
    # Extract gene information
    if (!is.null(pathway_info[[1]]$GENE)) {
        genes <- pathway_info[[1]]$GENE
        gene_ids <- genes[seq(1, length(genes), 2)]
        filtered_pathways_genes[[pathway_id]] <- gene_ids
    }
}
names(filtered_pathways_genes) <- filtered_pathways

# Convert the list to a tibble
kegg_dna_repair <- enframe(filtered_pathways_genes, name = "pathway", value = "entrez_id")

# Unnest the tibble to create a long-format table
kegg_dna_repair <- kegg_dna_repair %>%
  unnest(entrez_id)

# trim names
kegg_dna_repair$pathway <-  sub(" - Homo sapiens \\(human\\)", "", kegg_dna_repair$pathway)

# gsea
kegg_dna_repair_gsea <- GSEA(res.list, 
                   TERM2GENE = kegg_dna_repair,
                   pvalueCutoff = 0.1)

# plot
pp <- lapply(1:nrow(kegg_dna_repair_gsea), function(i) {
    anno <- kegg_dna_repair_gsea[i, c("NES", "pvalue", "p.adjust")]
    lab <- paste0(names(anno), "=",  round(anno, 5), collapse="\n")

    gsearank(kegg_dna_repair_gsea, i, kegg_dna_repair_gsea[i, 2]) + xlab(NULL) +ylab(NULL) +
        annotate("text", 7000, kegg_dna_repair_gsea[i, "enrichmentScore"] * .75, label = lab, hjust=0, vjust=0, size = 2.25)
})

pdf("enu9_vs_control9_kegg_dna_repair.pdf",
    width = 6,
    height = 6)
plot_grid(plotlist=pp, ncol=2)
dev.off()
```

### KEGG Cell Cycle
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get gene sets
pathways <- c("Cell cycle")

pathways_list <- keggList("pathway", "hsa")

# Loop through each DNA repair pathway and filter the pathways_list
filtered_pathways <- c()
for (pathway in pathways) {
    matching_pathways <- grep(pathway, pathways_list, value = TRUE)
    filtered_pathways <- c(filtered_pathways, matching_pathways)
}

# Loop through each DNA repair pathway and retrieve the associated genes
filtered_pathways_genes <- list()
for (pathway_id in names(filtered_pathways)) {
    # Get pathway information
    pathway_info <- keggGet(pathway_id)
    
    # Extract gene information
    if (!is.null(pathway_info[[1]]$GENE)) {
        genes <- pathway_info[[1]]$GENE
        gene_ids <- genes[seq(1, length(genes), 2)]
        filtered_pathways_genes[[pathway_id]] <- gene_ids
    }
}
names(filtered_pathways_genes) <- filtered_pathways

# Convert the list to a tibble
kegg_cell_cycle <- enframe(filtered_pathways_genes, name = "pathway", value = "entrez_id")

# Unnest the tibble to create a long-format table
kegg_cell_cycle <- kegg_cell_cycle %>%
  unnest(entrez_id)

# trim names
kegg_cell_cycle$pathway <-  sub(" - Homo sapiens \\(human\\)", "", kegg_cell_cycle$pathway)

# gsea
kegg_cell_cycle_gsea <- GSEA(res.list, 
                   TERM2GENE = kegg_cell_cycle,
                   pvalueCutoff = 0.1)

# plot
anno <- kegg_cell_cycle_gsea[,c("NES", "pvalue")]
anno$NES <- round(anno$NES, 2)
anno$pvalue <- round(anno$pvalue, 8)
lab <- paste0(names(anno), "=",  anno, collapse="\n")

pdf("enu9_vs_control9_kegg_cell_cycle.pdf",
    width = 5,
    height = 5)
gsearank(kegg_cell_cycle_gsea, 1, title = "KEGG Cell Cycle Gene Set") +
  annotate("text", 1000, -0.3, label = lab, hjust=0, vjust=0, size = 5) +
  theme(text=element_text(size=15))
dev.off()
```

### KEGG Apoptosis
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get gene sets
pathways <- c("Apoptosis - Homo sapiens")

pathways_list <- keggList("pathway", "hsa")

# Loop through each DNA repair pathway and filter the pathways_list
filtered_pathways <- c()
for (pathway in pathways) {
    matching_pathways <- grep(pathway, pathways_list, value = TRUE)
    filtered_pathways <- c(filtered_pathways, matching_pathways)
}

# Loop through each DNA repair pathway and retrieve the associated genes
filtered_pathways_genes <- list()
for (pathway_id in names(filtered_pathways)) {
    # Get pathway information
    pathway_info <- keggGet(pathway_id)
    
    # Extract gene information
    if (!is.null(pathway_info[[1]]$GENE)) {
        genes <- pathway_info[[1]]$GENE
        gene_ids <- genes[seq(1, length(genes), 2)]
        filtered_pathways_genes[[pathway_id]] <- gene_ids
    }
}
names(filtered_pathways_genes) <- filtered_pathways

# Convert the list to a tibble
kegg_apoptosis <- enframe(filtered_pathways_genes, name = "pathway", value = "entrez_id")

# Unnest the tibble to create a long-format table
kegg_apoptosis <- kegg_apoptosis %>%
  unnest(entrez_id)

# trim names
kegg_apoptosis$pathway <-  sub(" - Homo sapiens \\(human\\)", "", kegg_apoptosis$pathway)

# gsea
kegg_apoptosis_gsea <- GSEA(sort(-res.list, decreasing = TRUE), 
                   TERM2GENE = kegg_apoptosis,
                   pvalueCutoff = 1)

# plot
anno <- kegg_apoptosis_gsea[,c("NES", "pvalue")]
anno$NES <- round(anno$NES, 2)
anno$pvalue <- round(anno$pvalue, 2)
lab <- paste0(names(anno), "=",  anno, collapse="\n")

pdf("enu9_vs_control9_kegg_apoptosis.pdf",
    width = 5,
    height = 5)
gsearank(kegg_apoptosis_gsea, 1, title = "KEGG Apoptosis Gene Set") +
  annotate("text", 10000, 0.2, label = lab, hjust=0, vjust=0, size = 5) +
  theme(text=element_text(size=15))
dev.off()
```

### SenMayo
```{r}
setwd("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/bulk RNA/5-functional-enrichment")

# get gene sets
senmayo <- read.csv("senmayo.csv",
                    header = TRUE)

# convert symbol to entrez
senmayo$entrez_id <- bitr(senmayo$symbol, fromType="SYMBOL", toType=c("ENTREZID"), OrgDb="org.Hs.eg.db")$ENTREZID
senmayo <- data.frame(set = senmayo$set,
                      entrez_id = senmayo$entrez_id)

# gsea
senmayo_gsea <- GSEA(res.list, 
                   TERM2GENE = senmayo,
                   pvalueCutoff = 0.1)

# plot
anno <- senmayo_gsea[,c("NES", "pvalue")]
lab <- paste0(names(anno), "=",  round(anno, 3), collapse="\n")

pdf("enu9_vs_control9_senmayo.pdf",
    width = 5,
    height = 5)
gsearank(senmayo_gsea, 1, title = "SenMayo Senescence Gene Set") +
  annotate("text", 1000, -0.3, label = lab, hjust=0, vjust=0, size = 5) +
  theme(text=element_text(size=15))
dev.off()
```

# Gene set variation analysis

## Get gene sets
- Using GO gene sets found to be enriched
```{r}
dir.create(paste0(working.dir, "7-gene-set-variation"))
setwd(paste0(working.dir, "7-gene-set-variation"))

# Load GO gene sets
enrich.ge.expressed <- read.csv("~/EinsteinMed Dropbox/Ronald Cutler/Vijg-lab/Projects/mutation accumulation/9 cycles ENU/single-cell DNA/6-pathway/1-load-data/enrich.go.expressed.csv")

# Convert to a list that matches with row names of expression matrix
gene_sets <- split(enrich.ge.expressed$geneID,
                      enrich.ge.expressed$ID)
```

## Calcaulte GSVA enrichment score
```{r}
setwd(paste0(working.dir, "7-gene-set-variation"))

exprData <- assay(dds, "counts.normalized.batchCorrect") 
rownames(exprData) <- rowData(dds)$entrezgene_id
exprData <- exprData[!is.na(rownames(exprData)), ]

se.filter.param <- gsvaParam(exprData = exprData,
                         geneSets = m2_gene_sets, 
                         minSize=5, 
                         maxSize=500,
                         kcdf="Gaussian")

se.filter.gsva <- gsva(se.filter.param,
                           verbose = TRUE)

write.csv(se.filter.gsva, "se.filter.gsva.csv")
```

## Differential abundance

### Fit linear model
```{r}
setwd(paste0(working.dir, "7-gene-set-variation"))

# Create the design matrix
design <- model.matrix(~ sex * diet, data = colData(dds))

# make names in correct format=
colnames(design) <- make.names(colnames(design))

# Fit the linear model
fit <- lmFit(se.filter.gsva, design)

# Compute empirical Bayes statistics
fit.eb <- eBayes(fit, robust=TRUE)

gssizes <- sapply(se.filter.param@geneSets, length)
gssizes <- gssizes[rownames(se.filter.gsva)]
all(names(gssizes) == names(fit.eb$sigma))

pdf("gene_set_size_vs_precision.pdf",
    width = 5,
    height = 5)
plot(sqrt(gssizes), sqrt(fit.eb$sigma), xlab="Sqrt(gene sets sizes)",
          ylab="Sqrt(standard deviation)", las=1, pch=".", cex=4)
lines(lowess(sqrt(gssizes), sqrt(fit.eb$sigma)), col="red", lwd=2)
dev.off()

# account for increased precision of larger gene sets using limma trend
fit.eb.trend <- eBayes(fit, robust=TRUE, trend=gssizes)

# set up contrasts
contrasts <- makeContrasts(
  # Diet effect in males: (MHF - MNC) = diethigh.fat coefficient
  Diet_Male       = diethigh.fat,

  # Diet effect in females: (FHF - FNC) = diethigh.fat + interaction
  Diet_Female     = diethigh.fat + sexfemale.diethigh.fat,

  # Sex effect in normal chow: (FNC - MNC) = sexfemale coefficient
  Sex_NC          = sexfemale,

  # Sex effect in high fat: (FHF - MHF) = sexfemale + interaction
  Sex_HF          = sexfemale + sexfemale.diethigh.fat,

  # Interaction: (Diet_Female - Diet_Male) = interaction coefficient
  Interaction     = sexfemale.diethigh.fat,

  levels = colnames(design)
)

# Apply the contrasts to the fit object
fit.eb.trend <- contrasts.fit(fit.eb.trend, contrasts)

# Re-compute the empirical Bayes statistics for this specific contrast
fit.eb.trend <- eBayes(fit.eb.trend, robust=TRUE, trend=gssizes)

results.adjusted <- summary(decideTests(fit.eb.trend,
                    adjust.method = 'BH',
                    p.value = 0.05,
                    lfc = 0.5))
write.csv(results.adjusted, "summary_expression_abundance_pathway_lfc1_pvalue0.05.csv")
```

### Male High-Fat vs Male Normal
```{r}
setwd(paste0(working.dir, "7-gene-set-variation"))

effector_male_142oe_vs_control_gsva <- topTable(fit.eb.trend, coef="Diet_Male", number = Inf)

# make pathways easier to read
effector_male_142oe_vs_control_gsva$pathway <- tools::toTitleCase(tolower(str_wrap(gsub("_", " ", sub(".*?_", "", rownames(effector_male_142oe_vs_control_gsva))), width = 50)))
effector_male_142oe_vs_control_gsva$database <- sub("_.*", "", rownames(effector_male_142oe_vs_control_gsva))

# save
write.csv(effector_male_142oe_vs_control_gsva, "effector_male_142oe_vs_control_gsva.csv")

# p-value distribution histogram
pdf("effector_male_142oe_vs_control_gsva_pvalue_distribution.pdf",
    width = 5,
    height = 5)
hist(effector_male_142oe_vs_control_gsva$P.Value, 
     main = "Male High-Fat vs Male Normal", 
     xlab = "P-value", 
     ylab = "Frequency")
dev.off()


# ma plot
effector_male_142oe_vs_control_gsva <- effector_male_142oe_vs_control_gsva %>%
  as.data.frame() %>%
  mutate(sig = case_when(logFC > 0 & effector_male_142oe_vs_control_gsva$adj.P.Val < 0.05 ~ "Upregulated",
                         logFC < 0 & effector_male_142oe_vs_control_gsva$adj.P.Val < 0.05 ~ "Downregulated",
                         .default = "")
         )

pdf("effector_male_142oe_vs_control_gsva_ma_log2.pdf",
    width = 5,
    height = 5)
ggplot(effector_male_142oe_vs_control_gsva, aes(log2(AveExpr + 1), logFC, color = sig)) +
  geom_point(alpha = 0.5, size = 4) +
  theme_Publication() +
  scale_colour_manual(values = c("grey", "blue", "red")) +
  geom_hline(yintercept= 0, 
             size=1,
             linetype = "dashed") +
  labs(title = "Male High-Fat vs Male Normal",
       subtitle = "Signficant Pathways: padj < 0.05",
       x = "Average Normalized Log2(pathway score) + 1",
       y = "Log2(fold change)")
dev.off()

# volcano plot

# pick the top 20 most significant gene names
res_df <- as.data.frame(effector_male_142oe_vs_control_gsva) %>%
  arrange(P.Value)
top20 <- res_df$pathway[1:20]

# color volcano points
thresh <- 0.5 # match with what is in plot below
keyvals <- rep('grey', nrow(effector_male_142oe_vs_control_gsva))
names(keyvals) <- rownames(effector_male_142oe_vs_control_gsva)
keyvals[effector_male_142oe_vs_control_gsva$logFC < -thresh & effector_male_142oe_vs_control_gsva$adj.P.Val < 0.05] <- 'blue'
keyvals[effector_male_142oe_vs_control_gsva$logFC > thresh & effector_male_142oe_vs_control_gsva$adj.P.Val < 0.05] <- 'red'

# volcano 
pdf("effector_male_142oe_vs_control_gsva_volcano.pdf",
    width = 5, 
    height = 5)
EnhancedVolcano(effector_male_142oe_vs_control_gsva,
                lab = effector_male_142oe_vs_control_gsva$pathway,
                selectLab = top20,
                title = "Male High-Fat vs Male Normal",
                subtitle = "Log2FC > 0.5; padj < 0.05",
                caption = NULL,
                x = 'logFC',
                y = 'P.Value',
                pCutoffCol = "adj.P.Val",
                pointSize = 3,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 2,
                labCol = 'black',
                labFace = 'bold',
                drawConnectors = TRUE,
                colConnectors = 'black',
                maxoverlapsConnectors = Inf,
                gridlines.major = FALSE,
                gridlines.minor = FALSE,
                #ylim = c(0, 8),
                colCustom = keyvals)
dev.off()

# heatmap of significant changes
ha <- HeatmapAnnotation(Diet = dds$diet,
                        Sex = dds$sex)

# get normalized counts
mat <- as.matrix(se.filter.gsva)

# order by results
mat.order <- mat[match(rownames(effector_male_142oe_vs_control_gsva), rownames(mat)),]

# get significant genes
mat.sig <- mat.order[effector_male_142oe_vs_control_gsva$adj.P.Val < 0.05 & abs(effector_male_142oe_vs_control_gsva$logFC) > 0.5,]
mat.sig <- mat.sig[!is.na(row.names(mat.sig)),]

# scale and center normalized counts
mat.sig.scaled <- t(scale(t(mat.sig)))

# replace underscores with spaces & wrap at 30 characters
new.rn <- rownames(mat.sig.scaled) %>%
  str_replace_all("_", " ")
  #str_wrap(width = 20)

rownames(mat.sig.scaled) <- new.rn

ht <- Heatmap(
  mat.sig.scaled,
  name = "Z-score",
  column_title = "CIMSS Infected vs CIMSS Mock\nSignficant pathways: log2FC > 0.5 & padj < 0.05",
  top_annotation = ha,
  border = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 12),
  show_column_names = FALSE,
  row_dend_width = unit(20, "mm")
)


pdf("effector_male_142oe_vs_control_gsva_sig_heatmap.pdf",
    width = 20, height = 50)
draw(ht,
     heatmap_legend_side      = "top",
     annotation_legend_side   = "top",
     padding               = unit(c(2, 2, 2, 200), "mm")  # top, right, bottom, left
)
dev.off()
```

### Female High-Fat vs Female Normal
```{r}
setwd(paste0(working.dir, "7-gene-set-variation"))

neg_guide2_vs_scramble_gsva <- topTable(fit.eb.trend, coef="Diet_Female", number = Inf)

# make pathways easier to read
neg_guide2_vs_scramble_gsva$pathway <- tools::toTitleCase(tolower(str_wrap(gsub("_", " ", sub(".*?_", "", rownames(neg_guide2_vs_scramble_gsva))), width = 50)))
neg_guide2_vs_scramble_gsva$database <- sub("_.*", "", rownames(neg_guide2_vs_scramble_gsva))

# save
write.csv(neg_guide2_vs_scramble_gsva, "neg_guide2_vs_scramble_gsva.csv")

# p-value distribution histogram
pdf("neg_guide2_vs_scramble_gsva_pvalue_distribution.pdf",
    width = 5,
    height = 5)
hist(neg_guide2_vs_scramble_gsva$P.Value, 
     main = "Female High-Fat vs Female Normal", 
     xlab = "P-value", 
     ylab = "Frequency")
dev.off()


# ma plot
neg_guide2_vs_scramble_gsva <- neg_guide2_vs_scramble_gsva %>%
  as.data.frame() %>%
  mutate(sig = case_when(logFC > 0 & neg_guide2_vs_scramble_gsva$adj.P.Val < 0.05 ~ "Upregulated",
                         logFC < 0 & neg_guide2_vs_scramble_gsva$adj.P.Val < 0.05 ~ "Downregulated",
                         .default = "")
         )

pdf("neg_guide2_vs_scramble_gsva_ma_log2.pdf",
    width = 5,
    height = 5)
ggplot(neg_guide2_vs_scramble_gsva, aes(log2(AveExpr + 1), logFC, color = sig)) +
  geom_point(alpha = 0.5, size = 4) +
  theme_Publication() +
  scale_colour_manual(values = c("grey", "blue", "red")) +
  geom_hline(yintercept= 0, 
             size=1,
             linetype = "dashed") +
  labs(title = "Female High-Fat vs Female Normal",
       subtitle = "Signficant Pathways: padj < 0.05",
       x = "Average Normalized Log2(pathway score) + 1",
       y = "Log2(fold change)")
dev.off()

# volcano plot

# pick the top 20 most significant gene names
res_df <- as.data.frame(neg_guide2_vs_scramble_gsva) %>%
  arrange(P.Value)
top20 <- res_df$pathway[1:20]

# color volcano points
thresh <- 0.5 # match with what is in plot below
keyvals <- rep('grey', nrow(neg_guide2_vs_scramble_gsva))
names(keyvals) <- rownames(neg_guide2_vs_scramble_gsva)
keyvals[neg_guide2_vs_scramble_gsva$logFC < -thresh & neg_guide2_vs_scramble_gsva$adj.P.Val < 0.05] <- 'blue'
keyvals[neg_guide2_vs_scramble_gsva$logFC > thresh & neg_guide2_vs_scramble_gsva$adj.P.Val < 0.05] <- 'red'

# volcano 
pdf("neg_guide2_vs_scramble_gsva_volcano.pdf",
    width = 5, 
    height = 5)
EnhancedVolcano(neg_guide2_vs_scramble_gsva,
                lab = neg_guide2_vs_scramble_gsva$pathway,
                selectLab = top20,
                title = "Female High-Fat vs Female Normal",
                subtitle = "Log2FC > 0.5; padj < 0.05",
                caption = NULL,
                x = 'logFC',
                y = 'P.Value',
                pCutoffCol = "adj.P.Val",
                pointSize = 3,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 2,
                labCol = 'black',
                labFace = 'bold',
                drawConnectors = TRUE,
                colConnectors = 'black',
                maxoverlapsConnectors = Inf,
                gridlines.major = FALSE,
                gridlines.minor = FALSE,
                #ylim = c(0, 8),
                colCustom = keyvals)
dev.off()

# heatmap of significant changes
ha <- HeatmapAnnotation(Diet = dds$diet,
                        Sex = dds$sex)

# get normalized counts
mat <- as.matrix(se.filter.gsva)

# order by results
mat.order <- mat[match(rownames(neg_guide2_vs_scramble_gsva), rownames(mat)),]

# get significant genes
mat.sig <- mat.order[neg_guide2_vs_scramble_gsva$adj.P.Val < 0.05 & abs(neg_guide2_vs_scramble_gsva$logFC) > 0.5,]
mat.sig <- mat.sig[!is.na(row.names(mat.sig)),]

# scale and center normalized counts
mat.sig.scaled <- t(scale(t(mat.sig)))

# replace underscores with spaces & wrap at 30 characters
new.rn <- rownames(mat.sig.scaled) %>%
  str_replace_all("_", " ")
  #str_wrap(width = 20)

rownames(mat.sig.scaled) <- new.rn

ht <- Heatmap(
  mat.sig.scaled,
  name = "Z-score",
  column_title = "CIMSS Infected vs CIMSS Mock\nSignficant pathways: log2FC > 0.5 & padj < 0.05",
  top_annotation = ha,
  border = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 12),
  show_column_names = FALSE,
  row_dend_width = unit(20, "mm")
)


pdf("neg_guide2_vs_scramble_gsva_sig_heatmap.pdf",
    width = 20, height = 50)
draw(ht,
     heatmap_legend_side      = "top",
     annotation_legend_side   = "top",
     padding               = unit(c(2, 2, 2, 200), "mm")  # top, right, bottom, left
)
dev.off()
```

### Interaction between diet and sex
```{r}
setwd(paste0(working.dir, "7-gene-set-variation"))

diet_sex_interaction_gsva <- topTable(fit.eb.trend, coef="Interaction", number = Inf)

# make pathways easier to read
diet_sex_interaction_gsva$pathway <- tools::toTitleCase(tolower(str_wrap(gsub("_", " ", sub(".*?_", "", rownames(diet_sex_interaction_gsva))), width = 30)))
diet_sex_interaction_gsva$database <- sub("_.*", "", rownames(diet_sex_interaction_gsva))

# save
write.csv(diet_sex_interaction_gsva, "diet_sex_interaction_gsva.csv")

# p-value distribution histogram
pdf("diet_sex_interaction_gsva_pvalue_distribution.pdf",
    width = 5,
    height = 5)
hist(diet_sex_interaction_gsva$P.Value, 
     main = "Interaction between drug and infection", 
     xlab = "P-value", 
     ylab = "Frequency")
dev.off()


# ma plot
diet_sex_interaction_gsva <- diet_sex_interaction_gsva %>%
  as.data.frame() %>%
  mutate(sig = case_when(logFC > 0 & diet_sex_interaction_gsva$adj.P.Val < 0.05 ~ "Upregulated",
                         logFC < 0 & diet_sex_interaction_gsva$adj.P.Val < 0.05 ~ "Downregulated",
                         .default = "")
         )

pdf("diet_sex_interaction_gsva_ma_log2.pdf",
    width = 5,
    height = 5)
ggplot(diet_sex_interaction_gsva, aes(log2(AveExpr + 1), logFC, color = sig)) +
  geom_point(alpha = 0.5, size = 4) +
  theme_Publication() +
  scale_colour_manual(values = c("grey", "blue", "red")) +
  geom_hline(yintercept= 0, 
             size=1,
             linetype = "dashed") +
  labs(title = "Interaction between drug and infection",
       subtitle = "Signficant Pathways: padj < 0.05",
       x = "Average Normalized Log2(pathway score) + 1",
       y = "Log2(fold change)")
dev.off()

# volcano plot

# pick the top 20 most significant gene names
res_df <- as.data.frame(diet_sex_interaction_gsva) %>%
  arrange(P.Value)
top20 <- res_df$pathway[1:10]

# color volcano points
thresh <- 0.5 # match with what is in plot below
keyvals <- rep('grey', nrow(diet_sex_interaction_gsva))
names(keyvals) <- rownames(diet_sex_interaction_gsva)
keyvals[diet_sex_interaction_gsva$logFC < -thresh & diet_sex_interaction_gsva$adj.P.Val < 0.05] <- 'blue'
keyvals[diet_sex_interaction_gsva$logFC > thresh & diet_sex_interaction_gsva$adj.P.Val < 0.05] <- 'red'

# volcano 
pdf("diet_sex_interaction_gsva_volcano.pdf",
    width = 5, 
    height = 5)
EnhancedVolcano(diet_sex_interaction_gsva,
                lab = diet_sex_interaction_gsva$pathway,
                selectLab = top20,
                title = "Interaction between drug and infection",
                subtitle = "Log2FC > 0.5; padj < 0.05",
                caption = NULL,
                x = 'logFC',
                y = 'P.Value',
                pCutoffCol = "adj.P.Val",
                pointSize = 3,
                colAlpha = 0.5,
                legendPosition = "none",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                labSize = 2,
                labCol = 'black',
                labFace = 'bold',
                drawConnectors = TRUE,
                colConnectors = 'black',
                maxoverlapsConnectors = Inf,
                gridlines.major = FALSE,
                gridlines.minor = FALSE,
                #ylim = c(0, 5),
                colCustom = keyvals)
dev.off()

# heatmap of significant changes
ha <- HeatmapAnnotation(Diet = dds$diet,
                        Sex = dds$sex)

# get normalized counts
mat <- as.matrix(se.filter.gsva)

# order by results
mat.order <- mat[match(rownames(diet_sex_interaction_gsva), rownames(mat)),]

# get significant genes
mat.sig <- mat.order[diet_sex_interaction_gsva$adj.P.Val < 0.05 & abs(diet_sex_interaction_gsva$logFC) > 0.5,]
mat.sig <- mat.sig[!is.na(row.names(mat.sig)),]

# scale and center normalized counts
mat.sig.scaled <- t(scale(t(mat.sig)))

# replace underscores with spaces & wrap at 30 characters
new.rn <- rownames(mat.sig.scaled) %>%
  str_replace_all("_", " ") %>%
  str_wrap(width = 20)

rownames(mat.sig.scaled) <- new.rn

ht <- Heatmap(
  mat.sig.scaled,
  name = "Z-score",
  column_title = "Interaction between drug and infection\nSignficant pathways: log2FC > 0.5 & padj < 0.05",
  top_annotation = ha,
  border = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 12),
  show_column_names = FALSE,
  row_dend_width = unit(20, "mm")
)


pdf("diet_sex_interaction_gsva_sig_heatmap.pdf",
    width = 10, height = 10)
draw(ht,
     heatmap_legend_side      = "top",
     annotation_legend_side   = "top",
     padding               = unit(c(2, 2, 2, 100), "mm")  # top, right, bottom, left
)
dev.off()
```

#### Scatter plot
```{r}
# merge tables
diet_sex_interaction_gsva_merge <- merge(effector_male_142oe_vs_control_gsva,
                        neg_guide2_vs_scramble_gsva[,c("logFC", "P.Value", "adj.P.Val")],
                        by = "row.names")
rownames(diet_sex_interaction_gsva_merge) <- diet_sex_interaction_gsva_merge$Row.names

diet_sex_interaction_gsva_merge <- merge(diet_sex_interaction_gsva_merge, 
                        diet_sex_interaction_gsva[,c("logFC", "P.Value", "adj.P.Val")],
                        by = "row.names")
rownames(diet_sex_interaction_gsva_merge) <- diet_sex_interaction_gsva_merge$Row.names
diet_sex_interaction_gsva_merge$Row.names <- NULL

# clean up columns
diet_sex_interaction_gsva_merge <- diet_sex_interaction_gsva_merge[c("pathway", "database",
                                   "logFC.x", "P.Value.x", "adj.P.Val.x",
                                   "logFC.y", "P.Value.y", "adj.P.Val.y",
                                   "logFC", "P.Value", "adj.P.Val")]

# rename columns
colnames(diet_sex_interaction_gsva_merge)[3:11] <- c("effector_male_142oe_vs_control_gsva.logFC",
                                    "effector_male_142oe_vs_control_gsva.P.Value",
                                    "effector_male_142oe_vs_control_gsva.adj.P.Val",
                                    "neg_guide2_vs_scramble_gsva.logFC",
                                    "neg_guide2_vs_scramble_gsva.P.Value",
                                    "neg_guide2_vs_scramble_gsva.adj.P.Val",
                                    "diet_sex_interaction_gsva_merge.logFC",
                                    "diet_sex_interaction_gsva_merge.P.Value",
                                    "diet_sex_interaction_gsva_merge.adj.P.Val")

fc_thr    <- 0.5
p_thr     <- 0.05
delta_thr <- 0.1

diet_sex_interaction_gsva_merge <- diet_sex_interaction_gsva_merge %>%
  mutate(
    x = effector_male_142oe_vs_control_gsva.logFC,
    y = neg_guide2_vs_scramble_gsva.logFC,
    delta = diet_sex_interaction_gsva_merge.logFC,
    quadrant = case_when(
      x > 0 & y > 0 ~ "Concordant Up",
      x < 0 & y < 0 ~ "Concordant Down",
      x > 0 & y < 0 ~ "Negative Reversal: Up -> Down",
      x < 0 & y > 0 ~ "Positive Reversal: Down -> Up",
      TRUE ~ "Near Zero / Mixed"
    ),
    modulation = case_when(
      quadrant == "Concordant Up"   & delta >  delta_thr ~ "Amplified Up",
      quadrant == "Concordant Up"   & delta < -delta_thr ~ "Attenuated Up",
      quadrant == "Concordant Down" & delta < -delta_thr ~ "Amplified Down",
      quadrant == "Concordant Down" & delta >  delta_thr ~ "Attenuated Down",
      TRUE ~ NA_character_
    ),
    sig_int = !is.na(diet_sex_interaction_gsva_merge.adj.P.Val) &
              diet_sex_interaction_gsva_merge.adj.P.Val < p_thr,
    sig_simple = ((abs(x) > fc_thr) & !is.na(effector_male_142oe_vs_control_gsva.adj.P.Val) &
                   effector_male_142oe_vs_control_gsva.adj.P.Val < p_thr) |
                 ((abs(y) > fc_thr) & !is.na(neg_guide2_vs_scramble_gsva.adj.P.Val) &
                   neg_guide2_vs_scramble_gsva.adj.P.Val < p_thr),
    Label = case_when(
      sig_int    ~ if_else(!is.na(modulation), modulation, quadrant),
      sig_simple ~ if_else(!is.na(modulation), modulation, quadrant),
      TRUE       ~ "Not Significant"
    )
  )

table(diet_sex_interaction_gsva_merge$modulation)

# output
write.csv(diet_sex_interaction_gsva_merge, "diet_sex_interaction_gsva_merge.csv", row.names = FALSE)

## ---- plot ----

# only significant-interaction genes
out <- diet_sex_interaction_gsva_merge %>%
  mutate(sig_int = !is.na(diet_sex_interaction_gsva_merge.adj.P.Val) &
                   diet_sex_interaction_gsva_merge.adj.P.Val < p_thr)
sig <- out %>% filter(sig_int)  

# label top genes by interaction padj
highlight <- out %>%
  filter(sig_int) %>%
  arrange(diet_sex_interaction_gsva_merge.adj.P.Val) %>%
  slice_head(n = 20)

pdf("diet_sex_interaction_gsva_merge_log2fc_corr.pdf", height = 4, width = 6)
ggplot() +
  # background: nonsignificant interaction genes
  geom_point(data = out, aes(x = x, y = y),
             color = "grey40", alpha = 0.1, size = 1, show.legend = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  # overlay: only significant interaction genes are colored
 geom_point(data = sig, aes(x = x, y = y, color = Label),
             size = 2, alpha = 0.75) +
 geom_text_repel(
    data = highlight,
    aes(x = x, y = y, label = pathway),
    size = 2, fontface = "bold", colour = "black",
    segment.color = "black", segment.size = 0.3,
    box.padding = 0.4, point.padding = 0.3,
    max.overlaps = Inf, min.segment.length = 0
  ) +
  lims(x = c(-2, 2), y = c(-2, 2)) +
  scale_color_manual(
    name = "Interaction class",
    values = c(
    "Amplified Up"       = "firebrick",
    "Attenuated Up"      = "orange3",
    "Amplified Down"     = "navy",
    "Attenuated Down"    = "skyblue4",
    "Negative Reversal: Up -> Down"   = "purple",
    "Positive Reversal: Down -> Up"   = "darkgreen"
  ) # , drop = FALSE  # uncomment to keep all legend entries even if absent
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  theme_bw() + theme(panel.grid = element_blank()) +
  labs(x = "Effect of infection with DMSO (log2FC)",
       y = "Effect of infection with CIMSS (log2FC)",
       title = "Drug × Infection: Significant Interactions",
       subtitle = "Log2FC > 0.5; padj < 0.5",)
dev.off()
```

